### 2020-10-16 papers/ijcar2020/response-to-rebuttal.txt
Move the svn papers to git.  I tried to move the histories also but failed with that.  In the end, I just did a unix cp and git add .

>>  ----------------------- REVIEW 1 ---------------------

>>  ... My understanding is that they have a Coq tactic that takes
>>  a proof goal, exports to to an external mini-prover, which generates
>>  an FPC certificate. Then the FPC certificate is imported back to Coq
>>  via a deep embedding, and Coq uses a reflection theorem to turn it
>>  into an ordinary proof. But I am only guessing: this is not made
>>  clear.

A more accurate outline is: an external proof certificate is checked
via lambda Prolog (ELPI running within Coq).  If that check is
successful, lambda Prolog will build a proof structure that the Coq
kernel should be able to check (only the Coq kernel needs to be
trusted).  Part of the novelty here is the use of FPCs and logic
programming, which makes defining external proof structures, checking
such proofs, and generating Coq proofs all part of the same, modular
process.  Another novelty here is that Coq proofs are built entirely
within the ELPI plugin without using, say, Ltac or reflection.

>>  The example, a proof of "(A -> B) -> A -> B", is singularly
>>  unimpressive. Can the tool do something more complex? How long does it
>>  take?

Since this is an ongoing project, there are now more examples in the
GitHub repository.  Many other kinds of FPCs are described in 
papers [5,9,11].

>>  I don't understand what is in the repository. The READMEs don't
>>  provide any guidance.

The repository has been reorganized to make it much more clear.

>>  ----------------------- REVIEW 2 ---------------------

>>  - The presented example (example 1) that uses the system is
>>    trivial. It leaves me wondering how easy it is to apply the system
>>    to realistic examples.

There are more examples in the GitHub repository.

>>  * I am uncertain of the accuracy of the second sentence in the related
>>    work section. It states that this version of FPC-Coq only works for
>>    proving theorems in first order intuitionistic logic. Then continues
>>    to state that FPC-Coq can be used to prove the double negation of a
>>    formula for which there is a classical proof. Double negating a
>>    formula and finding an intuitionistic proof is only true for
>>    propositional logic and not first-order logic. First order negative
>>    translations are more involved (e.g. also double negating the body
>>    of each forall quantifier).

We meant to say "double negation translation" and not "the double
negation."  As this reviewer points out, many more negations need to
be inserted into first-order formulas for this scheme to work.  The
use of such double-negation translations in this focused setting has
been studied more carefully in reference [11] and in the technical
report https://hal.inria.fr/hal-01354298 (by Chihani, Ilik, Miller).

>>  - Section 2.1, could you please provide a short explanation why there
>>    are 101 introduction rules (and not 100)?

Since there are 100 formulas on the left and one on the right, there
are possibly 101 places to use an introduction rule.

>>  - Section 2.2: it may be helpful to explicitly state that decide and
>>    store are examples of experts and not necessarily part of every
>>    system.

We can be more clear about this: Decide and store are part of the
focusing mechanism and, hence, part of every FPC.  They are not,
however, part of the usual presentation of the sequent calculus.

### 2020-10-16 papers/ijcar2020/reviews-from-rebuttal.txt
Move the svn papers to git.  I tried to move the histories also but failed with that.  In the end, I just did a unix cp and git add .

IJCAR 2020 review response (submission [*NUMBER*])

Dear Dale Miller,

Thank you for your submission to IJCAR 2020. The IJCAR 2020
rebuttal period will be between March 9 and March 12 (AoE).

During this time, you will have access to the current state of
your reviews and have the opportunity to submit a response.
Please keep in mind the following during this
process:

* The response must focus on any factual errors in the reviews
and any questions posed by the reviewers. It must not provide
new research results or reformulate the presentation. Try to be
as concise and to the point as possible.

* The rebuttal period is an opportunity to react to the
reviews, but not a requirement to do so. Thus, if you feel the
reviews are accurate and the reviewers have not asked any
questions, then you do not have to respond.

* The reviews are as submitted by the PC members, without any
coordination between them. Thus, there may be inconsistencies.
Furthermore, these are not the final versions of the reviews.
The reviews can later be updated to take into account the
discussions at the program committee meeting, and we may find
it necessary to solicit other outside reviews after the
rebuttal period.

* The program committee will read your responses carefully and
take this information into account during the discussions. On
the other hand, the program committee will not directly respond
to your responses, either before the program committee meeting
or in the final versions of the reviews.

* Your response will be seen by all PC members who have access
to the discussion of your paper, so please try to be polite and
constructive.

The reviews on your paper are attached to this letter. To
submit your response you should log on the EasyChair Web page
for IJCAR 2020 and select your submission on the menu.

----------------------- REVIEW 1 ---------------------
SUBMISSION: 122
TITLE: FPC-Coq: Using ELPI to elaborate external proof evidence into Coq proofs (System Description)
AUTHORS: Roberto Blanco, Matteo Manighetti and Dale Miller

----------- Overall evaluation -----------
SCORE: -2 (reject)
----- TEXT:

This paper describes a connection between Coq and a lambda-Prolog that
allows one to generate proof certificates externally and verify them
in Coq. Even as a short paper, it is not suitable for publication in
its present form.

Four out of the six pages are devoted to describing the FPC proof
format, which has been described in detail in other publications. So
the "new" part is only two pages.

The paper is virtually unreadable. It is hard to understand what has
been done. My understanding is that they have a Coq tactic that takes
a proof goal, exports to to an external mini-prover, which generates
an FPC certificate. Then the FPC certificate is imported back to Coq
via a deep embedding, and Coq uses a reflection theorem to turn it
into an ordinary proof. But I am only guessing: this is not made
clear.

The example, a proof of "(A -> B) -> A -> B", is singularly
unimpressive. Can the tool do something more complex? How long does it
take?

The last page and a half drops down to a smaller font. Why? To fit in
extra text? That is cheating.

The paper is very sloppily written. There are very many typographical
errors, spelling errors, grammatical errors, sentences that don't make
sense, and a missing citation. I wonder whether the authors even
proofread the paper.

I don't understand what is in the repository. The READMEs don't
provide any guidance. There is no information as to how to install the
system and its dependencies, or how to test anything. There are only a
couple of hundred lines of Coq code, and it is not clear what they
do. There are number of commits made after the submission deadline.

----------------------- REVIEW 2 ---------------------
SUBMISSION: 122
TITLE: FPC-Coq: Using ELPI to elaborate external proof evidence into Coq proofs (System Description)
AUTHORS: Roberto Blanco, Matteo Manighetti and Dale Miller

----------- Overall evaluation -----------
SCORE: 2 (accept)
----- TEXT:
Summary:

This paper describes system FPC-Coq that combines two systems: the
foundational proof certificate framework and the Coq proof
assistant. This allows for using Coq to check partial proofs that are
missing some detailed information and not formed by explicit
lambda-terms. This paper extends the FPC framework to allow for proof
focussing and for augmenting proofs with certificates (that may be
used to fill in the missing information in the proof term).

Feedback and Questions:

+ The paper is well-written and easy to follow

+ The tool developed is exciting and seems potentially useful in proof
exchange between theorem provers.

- The presented example (example 1) that uses the system is
  trivial. It leaves me wondering how easy it is to apply the system
  to realistic examples.

* I am uncertain of the accuracy of the second sentence in the related
  work section. It states that this version of FPC-Coq only works for
  proving theorems in first order intuitionistic logic. Then continues
  to state that FPC-Coq can be used to prove the double negation of a
  formula for which there is a classical proof. Double negating a
  formula and finding an intuitionistic proof is only true for
  propositional logic and not first-order logic. First order negative
  translations are more involved (e.g. also double negating the body
  of each forall quantifier). Am I misunderstanding this paragraph, is
  this an error, or is there a justification for this that I am
  missing?

Further clarifications and typos:

- Section 2.1, could you please provide a short explanation why there
  are 101 introduction rules (and not 100)? 
- Section 2.1, it would be nice at the end of the section to go back
  to the example at the beginning and give the number of choices for
  this example using the focused proof system (with a short
  explanation). 
- Section 2.2: it may be helpful to explicitly state that decide and
  store are examples of experts and not necessarily part of every
  system. 
- There is a missing reference in the related work section [1,2, ?].

----------------------- REVIEW 3 ---------------------
SUBMISSION: 122
TITLE: FPC-Coq: Using ELPI to elaborate external proof evidence into Coq proofs (System Description)
AUTHORS: Roberto Blanco, Matteo Manighetti and Dale Miller

----------- Overall evaluation -----------
SCORE: 2 (accept)
----- TEXT:

The FPC-Coq system is built using Coq-ELPI that is a new Coq plugin
based on the ELPI implementation of λProlog. The foundational proof
certificate (FPC) framework is a system capable of checking and
elaborating a wide range of proof evidence and in FPC-Coq, it is
implemented using the logic programming language λProlog. The authors
propose the recently released Coq-ELPI plugin or Coq in which the ELPI
implementation of λProlog can send information to the Coq kernel.

Then it is possible to provide an additional procedure (by going
around Coq’s tactic language) for building and submitting formal
proofs to be checked by the Coq kernel. Thus, proof certificates
coming from external provers or from users can be converted to proofs
appropriate for the Coq kernel. Then, in this context, it is an
significant contribution to the FPC framework.

Section 2 presents the FPC framework with a focus on two problems with
attempting to use sequent proofs for theorem proving and proof
checking. The first is the lack of focus that is addressed in the
paper with a subsection dedicated to focused proof systems. The second
one is the way to determine from where missing information comes and
then another subsection considers how to augment the proof system with
certificates.  This section is well-written with enough details to
understand the questions and the approach followed by the authors and
its specificity. They are illustrated through some examples of FPC
specifications

Section 3 presents FPC-Coq and the way proposed to make Coq interact
with proof certificates. Moreover the Elpi extension for Coq allows
some facilities for building tactics written in λProlog.

This paper is a system description that is well written with enough
details about the design and the implementation and also with a
logical account of some steps.  It is a part of a more global project
that consists of using λProlog within a type theory setting, in
general, and within Coq, in particular.  The conversion of external
proof certificates to proofs appropriate for the Coq kernel is an
interesting and significant result and its implementation through
FPC-Coq will have an impact on some further formal proof developments.
Then I recommend to accept this system description.

Other comments:

p5. analisis -> analysis

P7. In related work paragraph a reference is missing [1,2,?]

------------------------------------------------------

Best wishes,

Nicolas Peltier
Viorica Sofronie-Stokkermans

### 2020-10-16 papers/coq-workshop-2020/main.tex
Move the svn papers to git.  I tried to move the histories also but failed with that.  In the end, I just did a unix cp and git add .

\documentclass[a4paper]{easychair}
\usepackage{xspace}
\usepackage{fullpage}

\newcommand{\lP}{$\lambda$Prolog\xspace}
\newcommand{\LKF}{\hbox{\sl LKF}\xspace}
\newcommand{\LJF}{\hbox{\sl LJF}\xspace}
\newcommand{\fpccoq}{FPC-Coq\xspace}

\title{\fpccoq: Using ELPI to elaborate external proof evidence into Coq proofs}
\author{Roberto Blanco\inst{1}\and Matteo Manighetti\inst{2}\and Dale Miller\inst{2}}
\institute{Inria Paris\and Inria Saclay}
% RB: I wonder whether/when my change of affiliation will be relevant.
\authorrunning{Blanco, Miller, and Manighetti}
\titlerunning{\fpccoq: Using ELPI in Coq}

\begin{document}
\maketitle
\begin{abstract}\noindent
Logic programming implementations of the \emph{Foundational Proof
  Certificate} (FPC) framework are capable of checking a wide range of
proof evidence.  Proof checkers based on logic programming can make
use of both unification and backtracking search to allow varying
degrees of proof reconstruction to take place during proof checking.
Such proof checkers are also able to elaborate proofs lacking full
details into proofs containing much more detail.  We outline here a
proof-of-concept system in which the Coq-Elpi plugin, which embeds an
implementation of \lP into Coq, is used to check proof certificates
supplied by external (to Coq) provers and to elaborate them into the
fully detailed proof terms that can be checked by the Coq kernel.
\end{abstract}

The trusted base of Coq is its kernel, which is a type-checking
program that certifies that a dependently typed $\lambda$-term has a
given type.  If type checking succeeds, we accept that the formula
corresponding to that type is, in fact, a theorem of intuitionistic
logic.  The rest of the Coq system, especially its tactic language, is
designed to help a human user build proofs-cum-$\lambda$-terms that
can be checked by the kernel.  Of course, there are many theorem
provers for intuitionistic logic for which a completed proof is not
the kind of detailed $\lambda$-term required by the Coq kernel but is,
instead, a trace of some key aspects of a proof: some proof details
might not be captured in such a trace.  For example:
\begin{enumerate}

\item Substitution instances of quantifiers might not be recorded in a
  proof since such instances can, in principle, be reconstructed using
  unification.

\item Detailed typing information might not need to be stored within a
  proof certificate since types can often be reconstructed during
  proof checking.

\item Some simplifications steps might be applied within a proof
  without recording which rewrites were used.  A simple 
  non-deterministic proof-search engine might be expected to
  reconstruct an equivalent simplification.
\end{enumerate}
%RB: Not intuitionistic, but SAT/SMT certificates also are like this.

The Foundational Proof Certificate (FPC) framework \cite{chihani17jar}
has been designed for formally defining a rich collection of proof
certificates that do not necessarily contain full details.  Proof
checkers for FPC certificates can be written using simple logic
programs: the standard implementation mechanisms of unification and
backtracking search can then be used to reconstruct missing proof
details \cite{miller17fac}.  Such proof checking kernels can also be
used to elaborate such certificates into fully detailed proofs
\cite{blanco17phd,blanco17cade}.

We built the \fpccoq system to demonstrate the feasibility of using
logic programming to check proof certificates produced by external
provers and to elaborate such certificates into proofs with sufficient
details that they can be given directly to Coq's kernel to
check.  Note that one does not need to trust the external prover nor
the \fpccoq tool-chain: one only needs to trust the Coq kernel.

Because of the need to represent and compute with quantificational
formulas and dependently-typed $\lambda$-terms, we chose \lP
\cite{miller12proghol} as our implementation language since it has a
direct and elegant treatment of binding in data structures.  The ELPI
implementation \cite{dunchev15lpar} of \lP has been embedded recently
into the Coq-Elpi plugin \cite{coq-elpi20web}, and we have used this
plugin to implement a system that checks proof certificates in the FPC
framework and elaborates them into proof terms that can be submitted
to the Coq kernel.

The \fpccoq system can be seen as part of the larger effort to use \lP
within a type theory setting \cite{coen19mscs}, in general, and within
Coq \cite{tassi19itp}, in particular. In doing so, the first step is to
define how to use Coq terms as proof evidence for a formula, i.e.,
to program a type checker in the FPC framework. Initially, the
presentation of connectives as inductive types in Coq is translated to
the usual connectives of intuitionistic first-order logic, the only ones
known to the FPC checker.

The current version of \fpccoq only works for proving theorems in
first-order intuitionistic logic.  Given that the \lP proof checker,
based on \LJF (a focused version of intuitionistic sequent calculus),
can also serve as a proof checker for \LKF (its classical
counterpart), \fpccoq can be used to prove double-negation
translations of a formula for which there is a proof certificate in
classical logic \cite{chihani17jar}.  Extending the FPC framework to
involve inductive and coinductive reasoning has also been considered
\cite{heath15pxtp,heath19jar}, and a future version of \fpccoq could
well include proof certificates that are output from inductive theorem
provers as well as model checkers.

Given its origins in the theory of focused sequent calculus, our first
FPC proof checker only implements a minimalistic sequent calculus in a
purely declarative fashion.  We have also developed a second approach
to building a proof checker based on dependent types in the style of
\cite{lengrand10lmcs}.  This second approach allows for a much more
concise implementation, where there is no translation needed from Coq
formulas to intuitionistic formulas since the kernel directly operates
with dependent types. The usual definitions of proof certificates
remain directly usable also in this context.

In particular, when the user loads the \lP code for an FPC definition
named, say \texttt{fpc}, into the system, that FPC definition is made
available as a Coq-Elpi tactic that takes a proof certificate, say
\texttt{cert}, as an argument.  Inside Coq's proof mode, that tactic
can be invoked by \texttt{elpi fpc cert}.

The code for both of these implementations is available at
\url{https://github.com/proofcert/fpc-elpi}, along with some simple
examples to illustrate the steps taken to elaborate and check proof
certificates.  We have not yet explored the cost of performing proof
checking in this setting: it is likely to vary greatly among different
certificate formats.

Although \fpccoq currently targets the construction of proofs for the
Coq kernel, it should be straightforward to use the same setup
described here to build Dedukti proofs as well \cite{assaf16types}.
Our development could also provide the facilities for cleanly and
easily building Coq tactics that are written in \lP and are controlled
by (partial) proof certificates.

%\bibliography{../../references/master}
\bibliographystyle{plain}
\begin{thebibliography}{10}

\bibitem{assaf16types}
Ali Assaf, Guillaume Burel, Raphal Cauderlier, David Delahaye, Gilles Dowek,
  Catherine Dubois, Fr{\'e}d{\'e}ric Gilbert, Pierre Halmagrand, Olivier
  Hermant, and Ronan Saillard.
\newblock Expressing theories in the $\lambda$${\Pi}$-calculus modulo theory
  and in the {Dedukti} system.
\newblock In {\em TYPES: Types for Proofs and Programs}, Novi Sad, Serbia,
  2016.

\bibitem{blanco17phd}
Roberto Blanco.
\newblock {\em Applications for Foundational Proof Certificates in theorem
  proving}.
\newblock PhD thesis, Universit{\'e} Paris-Saclay, December 2017.

\bibitem{blanco17cade}
Roberto Blanco, Zakaria Chihani, and Dale Miller.
\newblock Translating between implicit and explicit versions of proof.
\newblock In Leonardo de~Moura, editor, {\em Automated Deduction - {CADE} 26 -
  26th International Conference on Automated Deduction, Gothenburg, Sweden,
  August 6-11, 2017, Proceedings}, volume 10395 of {\em Lecture Notes in
  Computer Science}, pages 255--273. Springer, 2017.

\bibitem{chihani17jar}
Zakaria Chihani, Dale Miller, and Fabien Renaud.
\newblock A semantic framework for proof evidence.
\newblock {\em J. of Automated Reasoning}, 59(3):287--330, 2017.

\bibitem{dunchev15lpar}
Cvetan Dunchev, Ferruccio Guidi, Claudio~Sacerdoti Coen, and Enrico Tassi.
\newblock {ELPI:} fast, embeddable, $\lambda${Prolog} interpreter.
\newblock In M. Davis, A. Fehnker, A. McIver, and A. Voronkov, editors, 
{\em Logic for Programming, Artificial Intelligence, and
  Reasoning, LPAR-20}, LNCS 9450, pages 460--468. Springer, 2015.

\bibitem{coen19mscs}
Ferruccio Guidi, Claudio~Sacerdoti Coen, and Enrico Tassi.
\newblock Implementing type theory in higher order constraint logic
  programming.
\newblock {\em Mathematical Structures in Computer Science}, 29(8):1125--1150,
  2019.

\bibitem{heath15pxtp}
Quentin Heath and Dale Miller.
\newblock A framework for proof certificates in finite state exploration.
\newblock In Cezary Kaliszyk and Andrei Paskevich, editors, {\em Proceedings of
  the Fourth Workshop on Proof eXchange for Theorem Proving}, number 186 in
  Electronic Proceedings in Theoretical Computer Science, pages 11--26. Open
  Publishing Association, August 2015.

\bibitem{heath19jar}
Quentin Heath and Dale Miller.
\newblock A proof theory for model checking.
\newblock {\em J. of Automated Reasoning}, 63(4):857--885, 2019.

\bibitem{lengrand10lmcs}
{St\'ephane} Lengrand, Roy Dyckhoff, and James McKinna.
\newblock A focused sequent calculus framework for proof search in pure type
  systems.
\newblock {\em Logical Methods in Computer Science}, 7(1), 2011.

\bibitem{miller17fac}
Dale Miller.
\newblock Proof checking and logic programming.
\newblock {\em Formal Aspects of Computing}, 29(3):383--399, 2017.

\bibitem{miller12proghol}
Dale Miller and Gopalan Nadathur.
\newblock {\em Programming with Higher-Order Logic}.
\newblock Cambridge Univ. Press, 2012.

\bibitem{tassi19itp}
Enrico Tassi.
\newblock {Deriving Proved Equality Tests in Coq-Elpi: Stronger Induction
  Principles for Containers in Coq}.
\newblock In John Harrison, John O'Leary, and Andrew Tolmach, editors, {\em
  10th International Conference on Interactive Theorem Proving (ITP 2019)},
  volume 141 of {\em Leibniz International Proceedings in Informatics
  (LIPIcs)}, pages 29:1--29:18, Dagstuhl, Germany, 2019. Schloss
  Dagstuhl--Leibniz-Zentrum fuer Informatik.

\bibitem{coq-elpi20web}
Enrico Tassi.
\newblock Coq plugin embedding {ELPI}.
\newblock \url{https://github.com/LPCIC/coq-elpi}, 2020.

\end{thebibliography}

\end{document}

%%  LocalWords:  ELPI Coq's minimalistic Dedukti Elpi fpc elpi

### 2020-10-16 papers/ijcar2020/outtakes.tex
Move the svn papers to git.  I tried to move the histories also but failed with that.  In the end, I just did a unix cp and git add .

which is capable to bridging the
gap between such high-level and partial proof evidence supplied by
some external prover and the specific needs of the Coq kernel.
%
%
The Foundational Proof Certificate (FPC) framework can use
the unification and backtracking mechanisms of \lP to perform proof
reconstruction in order to provide a fully detailed natural deduction
proof that can be checked by the Coq kernel.
%
As a result, we are able to have Coq check proof evidence supplied by
various external provers in such a way that we do not need to trust
either the external prover or the Coq-ELPI system.

### 2020-10-16 papers/ijcar2020/paper.tex
Move the svn papers to git.  I tried to move the histories also but failed with that.  In the end, I just did a unix cp and git add .

%% IJCAR 2020 CFP: System Descriptions

%% Submissions, not exceeding seven (7) pages excluding bibliography,
%% should describe the implemented tool and its novel features. One
%% author is expected to be able to perform a demonstration on demand to
%% accompany a tool presentation. Papers describing tools that have
%% already been presented in other conferences before will be accepted
%% only if significant and clear enhancements to the tool are reported
%% and implemented.

%\documentclass[runningheads,11pt]{llncs}
\documentclass{llncs}

\usepackage{proof}
%\usepackage{fullpage}
\usepackage{xspace}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{letltxmacro}
\usepackage{wrapfig}

%\usepackage{hyperref}
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
%\renewcommand\UrlFont{\color{blue}\rmfamily}

\input macros

\begin{document}
\title{\fpccoq: Using ELPI to elaborate external proof evidence into Coq proofs}
%\titlerunning{\fpccoq: Using ELPI in Coq}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here

\author{Roberto Blanco, Matteo Manighetti, and Dale Miller}
\institute{INRIA}
\date{Draft: \today}

\maketitle

\begin{abstract}
The \fpccoq system mixes two systems.
%
The \emph{foundational proof certificate} (FPC) framework is capable
of checking and elaborating a wide range of proof evidence and in
\fpccoq, it is implemented using the logic programming language \lP.
%
The checking a proof certificate in this higher-order logic
programming language can yield a fully detailed, dependently
typed $\lambda$-terms that can be checked by the Coq kernel.
%
We use the recently released Coq-ELPI plugin for Coq in which the ELPI
implementation of \lP can send information to the Coq kernel.
%
As a result of using this plugin, it is possible to provide an
additional procedure (by going around Coq's tactic language) for
building and submitting formal proofs to be checked by the Coq
kernel.
%
Thus, proof certificates come from external provers or from
users can be converted to proofs appropriate for the Coq kernel.
%
When using \fpccoq, as with other Coq tools, 
one only needs to trust the Coq kernel. 
\end{abstract}

% DM Care using proof, proof certificate, proof evidence

\section{Introduction}

The trusted base of Coq is its kernel, which is used to certify that a
given formula has been proved.
%
That kernel checks natural deduction proofs encoded as a certain kind
of dependently typed $\lambda$-term.
%
The rest of the Coq system, especially its tactic language, is
designed to help a human user build proofs-cum-$\lambda$-terms that
can be checked by the kernel.

% RB In standard Coq parlance, the kernel of Coq is the type checker
% that verifies whether the type of the goal (the ``formula'' to be
% proved) can be inhabited by a witnessing term of that type.

Of course, there are many theorem provers for intuitionistic logic for
which a completed proof is not the kind of detailed $\lambda$-term
required by the Coq kernel but is, instead, a trace of some key
aspects of a proof: some proof details might not be captured in such a
trace.  For example:
\begin{enumerate}
\item  Substitution instances of quantifiers might not be
recorded in a proof since such instances can, in principle, be
reconstructed using unification.
\item  Detailed typing information might not need to be stored
within a proof certificate since types can often be reconstructed
during proof checking.
%
% Thus, an untyped $\lambda$-term might provide the key proof-building
% information.  Such terms can be more compact than the fully typed
% term.
%
\item Some simplifications steps might be applied within a proof
  without recording which rewrites were used.  A simple and shallow
  non-deterministic proof-search engine might be expected to
  reconstruct an equivalent simplification.
\end{enumerate}

% RB The connection may not be very useful right now, but usually
% tactics fulfill this role in Coq proof scripts. Pursuing the
% analogy later, FPCs can assume the role of tactical languages with
% clean and proper foundations.

The \emph{foundational proof certificate} (FPC) framework
\cite{chihani17jar} was designed to allow for the formal specification
of such partial proofs.
%
The FPC framework exploits the availability of unification and
backtracking search within the logic programming paradigm to blur the
distinction between proof checking and proof reconstruction. 
%
In particular, an FPC is a logic programming specification of the
\emph{clerk} and \emph{expert} predicates which mediates between the
given (partial) proof evidence and the needs of the logic programming
kernel that both checks and reconstructs all the details needed to
build an internal and formal proof.

The \fpccoq system is built using E. Tassi's Coq-ELPI
\cite{tassi18coqpl,coq-elpi20web}, a new Coq plugin based on the
ELPI~\cite{dunchev15lpar} implementation of \lP
\cite{miller12proghol}.

\section{The FPC framework}

There are at least two significant problems with attempting to use
Gentzen-style sequent calculus proofs for theorem proving and proof
checking: its lack of ``focus'' and determining from where missing
information comes.  We address these two in the next two subsections.

\subsection{Focused proof systems}

\begin{wrapfigure}{r}{45mm}
\vspace{-.9cm}
\[
\infer{\seq{\Gamma}{B_1\lor B_2}}{\seq{\Gamma}{B_i}} 
\quad
\infer{\seq{\Gamma}{\exists x.B}}{\seq{\Gamma}{B[t/x]}} 
\]
\vspace{-.9cm}
\caption{From the LJ calculus}
\label{fig:two unfocused}
\vspace{-.3cm}
\end{wrapfigure}

Consider attempting to prove the LJ sequent
\(\seq{\Gamma}{\exists x\exists y [(p~x~y)\lor ((q~x~y)\lor (r~x~y))]},\)
where $\Gamma$
contains, say, 100 formulas and where two of the introduction
rules are displayed in Fig.~\ref{fig:two unfocused}.
%
The search for a (cut-free) proof
of this sequent confronts the need to choose from among 101
introduction rules.  If the right-side introduction
rule is chosen, the premise of that rule will again present
101 possible introduction rules to consider.
%
Thus, reducing this sequent to, say, $\seq{\Gamma}{(q~t~s)}$ requires
picking one path of choices in a space of $101^4$ choices.

\begin{wrapfigure}{l}{46mm}
\vspace{-.9cm}
\[
\infer{\jRf{\Gamma}{B_1\lor B_2}}{\jRf{\Gamma}{B_i}} 
\quad
\infer{\jRf{\Gamma}{\exists x.B}}{\jRf{\Gamma}{B[t/x]}} 
\]
\vspace{-.8cm}
\caption{Focusing annotations}
\vspace{-.6cm}
\label{fig:two up}
\end{wrapfigure}

Focused proof systems (such as \LJF \cite{liang09tcs}) address this
explosion by organizing rules into two phases in which the sequents
are marked with either $\Downarrow$ or $\Uparrow$.
%
For example, the rules in Fig.~\ref{fig:two unfocused} are written
instead as in Fig.~\ref{fig:two up}.
%
Here, the formula between the $\Downarrow$ and the $\vdash$ is
distinguished: it is the only formula on which an introduction can
take place: as a result, reducing the sequent
\(\jRf{\Gamma}{\exists x\exists y [(p~x~y)\lor ((q~x~y)\lor (r~x~y))]}\)
to $\jRf{\Gamma}{(q~t~s)}$ involves only those choices related to the
formula marked for focus: no interleaving of other choices need to be
considered.

While the $\Downarrow$ phase involves rules that may not be
invertible, the $\Uparrow$ phase involves rules that must be
invertible.  For example, the left rules for $\lor$
and $\exists$ are invertible and their introduction rule is listed as
\[
  \infer{\jUnf{\Gamma}{B_1\veep B_2,\Theta}{\Delta}{E}} %[\veep_l]
        {\jUnf{\Gamma}{B_1,         \Theta}{\Delta}{E}\quad
         \jUnf{\Gamma}{B_2,         \Theta}{\Delta}{E}}
\qquad
  \infer{\jUnf{\Gamma}{\exists x.B, \Theta}{\Delta}{E}}
        {\jUnf{\Gamma}{[y/x]B,      \Theta}{\Delta}{E}} %[\exists_l]
\]
These rules need no external information.  In these
two rules, the zone between $\Uparrow$ and $\vdash$ contains a
\emph{list} of formulas.  When there are no more invertible rules that
can be applied to that first formula, that formula is moved to (i.e.,
stored in) the zone written as $\Gamma$, using the following
\emph{store-left} rule 
\[
  \infer[S_l.]{\jUnf{  \Gamma}{C,\Theta}{\Delta}{E}}
              {\jUnf{C,\Gamma}{  \Theta}{\Delta}{E}}
\]
Finally, when the zone between the $\Uparrow$ and the $\vdash$ is
empty (i.e., all invertible inference rules have been completed),
it is time to select a (possibly non-invertible) introduction rule to
attempt.  For that, we have the two \emph{decide} rules
\[
  \vcenter{\infer[D_l]{\jUnf{\Gamma,N}{\cdot}{\cdot}{E}}{\jLf{\Gamma,N}{N}{E}}}
  \qquad\hbox{and}\qquad
  \vcenter{\infer[D_r.]{\jUnf{\Gamma}{\cdot}{\cdot}{P}}{\jRf{\Gamma}{P}}}
\]
Such focused proof systems can remove a great deal of non-determinism
from proof systems while still being complete \cite{liang09tcs}.

\subsection{Augmenting the proof system with certificates}

The main idea behind the FPC framework \cite{chihani13cade} is that
proof evidence discovered by some theorem prover is packaged up as
some term structure (not necessarily a $\lambda$-term) and 
that term can be given to an independent and trustable proof checker.
%
In order for such a scheme to work, the semantics of the output proof
evidence must be formally defined.
%
The FPC framework provides such formal definitions that are executable
when interpreted using a suitable logic programming
engine~\cite{miller17fac}.

Consider the following \emph{augmented} versions of inference
rules given in Fig.~\ref{fig:two unfocused}.
\[
\infer{\bXi0 \jRf{\Gamma}{B_1\lor B_2}}
      {\bXi1 \jRf{\Gamma}{B_i}\quad \blue{\orExpert{\Xi_0}{\Xi_1}{i}}}
\qquad
\infer{\bXi0 \jRf{\Gamma}{\exists x.B}}
      {\bXi1 \jRf{\Gamma}{B[t/x]}\quad\blue{\someExpert{\Xi_0}{\Xi_1}{t}}}
\]
In these augmented rules the schema variable $\Xi$ ranges over terms
that denote proof evidence (certificates) and the additional premises
involve \emph{experts} which are predicates that relate the concluding
certificate $\Xi_0$ to a continuation certificate $\Xi_1$ and some
additional information.
%
The expert predicate for the disjunction can provide an indication of
which disjunct to pick and the expert for the existential quantifier
can provide an indication of which instance of the quantifier to use.
%
Presumably, these expert predicates are capable of digging into a
certificate and extracting such information.
%
It is not required, however, for an expert to be able to extract
such information.
%
For example, the disjunction expert might guess both 1 and 2 for $i$
and the proof checker will need to handle such non-determinism during
the checking of certificates.

When formulas are stored (removed from the $\Downarrow$ zone) and
later recalled (using a decide rule), indexes are used to address such
formulas. 
\[
  \infer[S_l]{\bXi0 \jUnfamb{\Gamma}{C,\Theta}{\null}}
             {\bXi1 \jUnfamb{\tupp{l}{C},\Gamma}{\Theta}{\null}\quad
              \blue{\storeClerk{\Xi_0}{\Xi_1}{l}}}
\]
\[
  \infer[D_l]{\bXi0 \jUnf{\Gamma,\tupp{l}{N}}{\cdot}{\cdot}{E}}
             {\bXi1 \jLf{\Gamma,\tupp{l}{N}}{N}{E}\quad
              \blue{\decideExpert{\Xi_0}{\Xi_1}{l}}}
\]
The augmented store-left ($S_l$) rule uses a \emph{clerk} predicate
to determine the index $l$ that is associated to the stored formula
(here, $C$).  The augmented decide-left ($D_l$) rule is given an extra
premise that uses an expert predicate: that premise can be used to
compute the index of the formula that is to be selected for focus.
%
The indexing mechanism does not need to give unique indexes to
different formulas: thus, the decide rule may be seen as
non-deterministically selecting a formula from storage.
%
With FPCs, indexes have been identified with
structures as diverse as formula occurrences and de Bruijn numerals
\cite{chihani17jar}.

\subsection{Examples of FPC specifications}

A simple proof certificate is essentially a natural number that denotes
the maximum number of decide rules that can occur along a path from
the root sequent.
%
Thus, checking that such a number is a valid proof certificate for a
formula essentially causes the proof check to enumerate all proofs
that have such a bounded number of decide rules.
%
The definition of some key experts and clerks for the definition of
this certificate are produced below: here, \lsti{s} is the non-zero
constructor for natural numbers and \lsti{dd} maps natural numbers
into the type of certificate terms (the syntax \lsti{x\} denotes
$\lambda x$ in \lP---here, this binding is vacuous).
{\small
\begin{lstlisting}
some_je    (dd D) (dd D) Term.       % _j refers to LJ 
or_jc      (dd D) (dd D) (dd D).     % e denotes experts
or_je      (dd D) (dd D) I.          % c denotes clerks
storeL_jc  (dd D) (x\ dd D) (x\ indx).
decideL_je (dd (s D)) (dd D) indx.
\end{lstlisting}
}
Here, we note that all stored formulas get the same index \lsti{indx}
and that the \lsti{decideL_je} expert always provides that index as
the source of the formulas to select.  Note also that both the expert
\lsti{some_je} allows any term to instantiate the existential
quantifier and that \lsti{or_je} allows both disjunctions to be
selected.
%
Many other (more sophisticated) FPCs appear in 
\cite{blanco15wof,blanco17lfmtp,chihani15tableaux,chihani16lfsa,chihani17jar,heath15pxtp}.

The \emph{pairing} FPC was presented in \cite{blanco17cade} and it was
used to show that two FPCs can be checked to hold simultaneously for a
given sequent.
%
Since the logic programming setting allows us to see proof checking
and proof reconstruction as two aspects of the \emph{same} execution,
it is possible to use pairing to check one certificate following one
FPC definition and to reconstruct a second certificate following a
different FPC definition.
%
In \cite{blanco17cade}, that pairing FPC was used to show how a given
proof certificate can be \emph{elaborated} (converted to a format with
more details) or \emph{distilled} (converted to a format with fewer
details).

\section{\fpccoq: making Coq interact with proof certificates}

\subsection{The FPC for Coq kernel terms}

%\lstinputlisting[linerange={7-10}]{termrep/repterm.sig}

%  - Illustrate both interactive and batch uses.
The first step towards making the framework of Foundational Proof
Certificates and the Coq proof assistant cooperate is the definition of a
proof certificate that explains how to consume Coq terms as proof evidence
for a formula. We can restate the problem with the terminology of the
\emph{formulas-as-types} tradition Coq sits in: here, for a given term, being
a proof of some formula amounts to having a certain type; the problem of
devising a definiton of a proof certificate that can consume Coq terms is
then equivalent to the problem of programming a type checker in terms of the
clerks and experts.

Coq lambda terms are directly isomorphic to proofs in a flavour of natural
deduction. In contrast, our approach is based on the sequent calculus: this
requires our approach to type checking to be slightly finer, and akin to
other analysis of the interpretations of focusing in the lambda calculus
\cite{brocknannestad15mfps} %% TODO: At which moment should I mention that we don't do inductive types?
Moreover, we should stress once again that we are only
targeting the first order fragment of the Calculus of Constructions, without
inductively defined types. The usual presentation of connectives as inductive
types is directly translated to the usual connectives of intuitionistic first
order logic, that are the only ones to be known to the FPC checker.

Let's start looking into the definition of the clerks and experts for type
checking by examining the handling of the implication. An implication
occurring in the type must correspond to a lambda abstraction occurring in
the term, and is handled by the \lsti{arr_jc} clerk that will simply remove
the binder, and return a term with an open variable.

{\small
\begin{lstlisting}
arr_jc (coqcert (fun _name _type F)) (coqabs F).
\end{lstlisting} }

Note that we need two certificate constructors: \lsti{coqcert} indicates that
the certificate being handled is a closed term; \lsti{coqabs} indicates that
it is a term with an open variable. The open variable is then bound to the
proof tree by the \lsti{storeL_jc} that takes an open term and returns to the
kernel an open certificate and an open index: the kernel will then bind both the
certificate and the term to the same variable.

{\small
\begin{lstlisting}
storeL_jc (coqabs T) (x\ coqcert (T x)) (x\ idxoftm x).
\end{lstlisting}
}

A logical account of the process is the following:

\[
  \infer[I_r]{\Xi_0 : (fun\; x : A \Rightarrow t), \jUnfG{\null}{A\to B }}
             {\infer[S_l]{\Xi_1 : (t), \jUnfG{A}{B}\quad
              \blue{\arrClerk{\Xi_0}{\Xi_1}}}{
                \Xi_2 : (t x), \jUnf{\Gamma, x : A}{\null}{B}{\null} \quad \blue{\storeClerk{\Xi_1}{\Xi_2(x)}{x}}
              }
             }
\]
The certificate now contains the variable $x$ that is bound in the kernel, and allows
a reference to the formula $A$. The immediate store of the formula happens only if it is
atomic; otherwise, a phase similar to a pattern matching phase happens.
For example, if the formula to the left of the implication sign is of the form
$A \lor B$, then the \lsti{or_jc} clerk is used. Its specification is
{\small
\begin{lstlisting}
or_jc (coqabs (x\ app {{or_ind F1 F2 lp:x}})) (coqabs T1) (coqabs T1) :-
  F1 = fun _ _ T1,
  F2 = fun _ _ T2.
\end{lstlisting}
}

Thus, the term with an open variable that would accept a disjunction type is
immediately decomposed into two terms with open variables for each of the
disjuncts. The initial proof term must contain a disjunction elimination
applied to the open variable: note that thanks to Coq-Elpi we can use a quoting operator and include
a term in the syntax of Coq inside a \lP clause. The new proof terms are just
the cases for the two disjuncts that we extract.

% % Not used at the moment: arr_je
% In the case the formula to the left of an
% implication is another implication, this phase is handled by the
% \lsti{arr_je} expert:

% {\small
% \begin{lstlisting}
% arr_je (coqabs (x\ app [x,T])) (coqcert T) (coqabs (x\ x)).
% \end{lstlisting} }

% This specifies that if the free variable of an open term (of some type
% $A$) has the type of an implication $B \to C$, then the term must be of the
% form $xt$, and one can use the implication left rule and continue checking
% that $x$ is an open term of type $A$ and $t$ is a closed term of type

\subsection{Tactics for proof certificates}
As we mentioned at the end of the previous section and at the beginning of 
this one, thanks to the purely declarative spirit of the implementation
of the checkers for FPCs the difference between proof checking and proof
reconstruction is blurred; the \emph{pairing} FPC helps in this direction
and allows for a declarative way to extract synthesised proofs in a format
starting from proofs in another format. Employing it together with the FPC
specification we have just given for Coq proof terms, we obtain a system
that can build proofs that can be checked by the Coq kernel starting from
external proofs given in any format for which an FPC definition is available.

The Elpi extension for Coq provides the facilities for building tactics
written in \lP. The code for the kernel of our proof checker, as well
as the FPC definitions that have already been produced could thus be
used right away to build our tactic. Some preprocessing code transforms Coq
types into first order formulas that our kernel can recognize,
handles the external proof evidence and pairs it (in the sense of the
\emph{pairing} FPC) with an empty proof term,
so that a run of the proof checker reconstructs it.

Let's consider a definition of a Proof Certificate, such as the one described above
\small{
  \begin{lstlisting}
ljf_entry ((dd N) <c> (coqcert Term)) Form.
  \end{lstlisting}
}
Where \lsti{<c>} denotes the pairing between the two certificate formats.
We then have a tactic that will have a natural as parameter, and that we can call in 
the usual way in a Coq proof:
\small{
  \begin{lstlisting}
Lemma example1 : forall A B: Prop, (A -> B) -> A -> B.
Proof.
elpi dd_fpc 2.
Qed.
  \end{lstlisting}
}

\subsection{Handling dependent types in the kernel}
The proof checking kernel in the Foundational Proof Certificates program is
meant to be a small trusted base. For this reason, given its origins in the
theory of focused sequent calculus, it only implements a minimalistic
sequent calculus in a purely decladative fashion.
In this context, however, we could also decide to keep Coq kernel as a trusted checker,
and to allow for extensions in the FPC checker kernel.

With this in mind, we developed a second solution where we extend the kernel 
of our proof checker with dependent types and reconstruction of Coq proof terms, 
in the style of \cite{lengrand10lmcs}.

This allows for a much more concise implementation, where there is no translation 
needed from Coq formulas to intuitionistic formulas since the kernel directly operates
with dependent types. The usual definitions of proof certificates remain directly usable
also in this context.

The code for both implementations is available at \url{https://gitlab.inria.fr/rblancom/fpc-elpi}

\section{Related work}

This paper can be seen as part of the larger effort to use \lP within
a type theory setting \cite{coen19mscs}, in general, and within Coq
\cite{tassi19itp}, in particular.

This particular version of \fpccoq only works on proving theorems in
first-order intuitionistic logic.
%
Given that the \lP proof checker based on \LJF can also serve as a
proof checker for \LKF (a focused version of classical sequent
calculus), \fpccoq can be used to prove double negation translations 
of a formula for which there is a proof certificate in classical
logic \cite{chihani17jar}.

The goals of the Dedukti framework overlap some of the goals of the
\fpccoq system. 
%
In particular, a great deal of work has gone into providing back-ends
to numerous proof assistant so that their proofs can be output in the
Dedukti format \cite{assaf16unp,assaf16types,cauderlier16phd}. 
%
The Dedukti interpreter can then provide an independent certification
of the correctness of the proof.
%
Although \fpccoq targets the construction of proofs for the Coq kernel
only, it should be straightforward to use the same setup described
here to build Dedukti proofs as well.

% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystyle{splncs04}
\bibliography{../../references/master}
%\bibliography{export}  % Use the Makefile to great this smaller bib file
\end{document}

%%  LocalWords:  Matteo Coq's Tassi's trustable basicstyle keepspaces
%%  LocalWords:  storeL jc indx decideL je storeR

### 2020-10-16 papers/lfmtp2020/abstract/paper.tex
Move the svn papers to git.  I tried to move the histories also but failed with that.  In the end, I just did a unix cp and git add .

\RequirePackage[utf8]{inputenc}
\RequirePackage[T1]{fontenc}
\documentclass[submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{LFMTP 2020}
\usepackage{url,xcolor}
\usepackage{xspace}
\newcommand{\lP}{$\lambda$Prolog\xspace}
\newcommand{\LKF}{\hbox{\sl LKF}\xspace}
\newcommand{\LJF}{\hbox{\sl LJF}\xspace}
\newcommand{\fpccoq}{FPC-Coq\xspace}

\title{Linking a \lP proof checker to the Coq kernel\\
       An extended abstract}
\author{Roberto Blanco
\institute{INRIA Paris, France}
\and Matteo Manighetti and 
\institute{INRIA Saclay \\ LIX, \'Ecole Polytechnique, France}
\and Dale Miller
\institute{INRIA Saclay \\ LIX, \'Ecole Polytechnique, France}}
\def\authorrunning{Blanco, Manighetti, and Miller}
\def\titlerunning{Linking a \lP proof checker to the Coq kernel}

\begin{document}
\maketitle

\noindent\textbf{Synopsis.}  The Coq kernel works with a specific form
of proof structure, and that proof structure is fully detailed.  In
particular, the Coq kernel performs type-checking on a dependently
typed $\lambda$-term.  If type checking succeeds, the formula
corresponding to its type is, in fact, a theorem of intuitionistic
logic.  Most external-to-Coq theorem provers do not generally build
such detailed, dependently typed $\lambda$-terms for proofs.  To the
extent that theorem provers output proof objects (called \emph{proof
  certificates} here), their structure can vary a great deal.  Also,
since some details can be reconstructed, they are seldom traced and
inserted into proof certificates.  In this extended abstract, we
describe the design of the \fpccoq system that can take externally
generated proof certificates and elaborate them into the kind of proof
structures required by the Coq kernel.  This \emph{elaboration} system
is built using three technologies: (1) the foundational proof
certificate (FPC) framework \cite{chihani17jar} that allows for the
specification of a wide range of proof certificates, (2) the \lP
programming language \cite{miller12proghol} that can interpret a given
FPC definition and then perform both proof checking and proof
reconstruction \cite{blanco17phd,blanco17cade}, and (3) the Coq-Elpi
plugin \cite{coq-elpi20web}, which embeds the Elpi implementation
\cite{dunchev15lpar} of \lP into Coq.

\bigskip

Consider an external (to Coq) prover capable of exporting its proofs
as certificates in some particular format.  Such a theorem prover (for
intuitionistic logic) is unlikely to build the detailed $\lambda$-term
required by the Coq kernel.  More likely, its notion of a proof
certificate is, instead, a trace of some key aspects of a proof: some
proof details might not be captured in such a trace.  For example:
\begin{enumerate}
\item Substitution instances of quantifiers might not be recorded in a
  proof since such instances can, in principle, be reconstructed using
  unification.

\item Detailed typing information might not need to be stored within a
  proof certificate since types can often be reconstructed during
  proof checking.

\item Some simplifications steps might be applied within a proof
  without recording which rewrites were used.  A simple
  non-deterministic proof-search engine might be expected to
  reconstruct an equivalent simplification.
\end{enumerate}
The FPC framework can be used to formally define the proof evidence of
many different formats \cite{chihani17jar} including those with such
details missing.  Combining such a definition, which is a simple logic
program, with an FPC proof checking kernel yields a complete logic
programming proof kernel.  During proof checking, unification and
backtracking search can infer missing proof details.  The FPC
framework also allows for the convenient capture and reorganization of
such details and, as a result, it can be used to build proof
structures appropriate for the Coq kernel.

Given its origins in the theory of focused sequent calculus, our first
FPC proof checker implements a purely declarative sequent calculus
proof checker, along the lines described in \cite{chihani17jar}.  We
have also developed a second approach to building a proof checker
based on dependent types in the style of \cite{lengrand10lmcs}.  This
second approach allows for a more concise implementation, where there
is no translation needed from Coq formulas to intuitionistic formulas
since the kernel directly operates with dependent types. The usual FPC
definitions of proof certificates remain directly usable also in this
context.

The code for both of these implementations is available at
\url{https://github.com/proofcert/fpc-elpi}.  It is important to
recognize that both \lP and Coq-Elpi can be seen as parts of a tool
chain, connecting an external prover to the Coq kernel and that these
two pieces of technology do not need to be trusted by the Coq kernel.

The current version of \fpccoq only works on proving theorems in
first-order intuitionistic logic.  We also assume that external proof
certificates are in a format that is formally defined by some given
FPC: examples of such formats which have already been developed are
resolution refutations, Hilbert proofs, de Bruijn notation, and
natural deduction \cite{chihani17jar}.  Given that the \lP proof
checker internally implements \LJF (a focused version of
intuitionistic sequent calculus \cite{liang09tcs}), it can also serve
as a proof checker for \LKF (its classical counterpart).  As a result,
\fpccoq can be used to prove double-negation translations of a formula
for which there is a proof certificate in classical logic
\cite{chihani17jar}.  Extending the FPC framework to involve inductive
and coinductive reasoning has also been considered
\cite{heath15pxtp,heath19jar}, and a future version of \fpccoq could
include proof certificates that are output from inductive theorem
provers as well as model checkers.

\bibliographystyle{eptcs}
%\bibliography{../../references/master}
\begin{thebibliography}{10}
\providecommand{\bibitemdeclare}[2]{}
\providecommand{\surnamestart}{}
\providecommand{\surnameend}{}
\providecommand{\urlprefix}{Available at }
\providecommand{\url}[1]{\texttt{#1}}
\providecommand{\href}[2]{\texttt{#2}}
\providecommand{\urlalt}[2]{\href{#1}{#2}}
\providecommand{\doi}[1]{doi:\urlalt{http://dx.doi.org/#1}{#1}}
\providecommand{\bibinfo}[2]{#2}

\bibitemdeclare{phdthesis}{blanco17phd}
\bibitem{blanco17phd}
\bibinfo{author}{Roberto \surnamestart Blanco\surnameend}
  (\bibinfo{year}{2017}): \emph{\bibinfo{title}{Applications for Foundational
  Proof Certificates in theorem proving}}.
\newblock Ph.D. thesis, \bibinfo{school}{Universit{\'e} Paris-Saclay}.
\newblock \urlprefix\url{https://tel.archives-ouvertes.fr/tel-01743857}.

\bibitemdeclare{inproceedings}{blanco17cade}
\bibitem{blanco17cade}
\bibinfo{author}{Roberto \surnamestart Blanco\surnameend},
  \bibinfo{author}{Zakaria \surnamestart Chihani\surnameend} \&
  \bibinfo{author}{Dale \surnamestart Miller\surnameend}
  (\bibinfo{year}{2017}): \emph{\bibinfo{title}{Translating Between Implicit
  and Explicit Versions of Proof}}.
\newblock In \bibinfo{editor}{Leonardo \surnamestart de~Moura\surnameend},
  editor: {\sl \bibinfo{booktitle}{Automated Deduction - {CADE} 26 - 26th
  International Conference on Automated Deduction, Gothenburg, Sweden, August
  6-11, 2017, Proceedings}}, {\sl \bibinfo{series}{Lecture Notes in Computer
  Science}} \bibinfo{volume}{10395}, \bibinfo{publisher}{Springer}, pp.
  \bibinfo{pages}{255--273}, \doi{10.1007/978-3-319-63046-5\_16}.

\bibitemdeclare{article}{chihani17jar}
\bibitem{chihani17jar}
\bibinfo{author}{Zakaria \surnamestart Chihani\surnameend},
  \bibinfo{author}{Dale \surnamestart Miller\surnameend} \&
  \bibinfo{author}{Fabien \surnamestart Renaud\surnameend}
  (\bibinfo{year}{2017}): \emph{\bibinfo{title}{A semantic framework for proof
  evidence}}.
\newblock {\sl \bibinfo{journal}{J. of Automated Reasoning}}
  \bibinfo{volume}{59}(\bibinfo{number}{3}), pp. \bibinfo{pages}{287--330},
  \doi{10.1007/s10817-016-9380-6}.
\newblock \urlprefix\url{https://doi.org/10.1007/s10817-016-9380-6}.

\bibitemdeclare{inproceedings}{dunchev15lpar}
\bibitem{dunchev15lpar}
\bibinfo{author}{Cvetan \surnamestart Dunchev\surnameend},
  \bibinfo{author}{Ferruccio \surnamestart Guidi\surnameend},
  \bibinfo{author}{Claudio~Sacerdoti \surnamestart Coen\surnameend} \&
  \bibinfo{author}{Enrico \surnamestart Tassi\surnameend}
  (\bibinfo{year}{2015}): \emph{\bibinfo{title}{{ELPI:} Fast, Embeddable,
  $\lambda${Prolog} Interpreter}}.
\newblock In \bibinfo{editor}{Martin \surnamestart Davis\surnameend},
  \bibinfo{editor}{Ansgar \surnamestart Fehnker\surnameend},
  \bibinfo{editor}{Annabelle \surnamestart McIver\surnameend} \&
  \bibinfo{editor}{Andrei \surnamestart Voronkov\surnameend}, editors: {\sl
  \bibinfo{booktitle}{Logic for Programming, Artificial Intelligence, and
  Reasoning - 20th International Conference, {LPAR-20} 2015}}, {\sl \bibinfo{series}{LNCS}}
  \bibinfo{volume}{9450}, \bibinfo{publisher}{Springer}, pp.
  \bibinfo{pages}{460--468}, \doi{10.1007/978-3-662-48899-7\_32}.

\bibitemdeclare{inproceedings}{heath15pxtp}
\bibitem{heath15pxtp}
\bibinfo{author}{Quentin \surnamestart Heath\surnameend} \&
  \bibinfo{author}{Dale \surnamestart Miller\surnameend}
  (\bibinfo{year}{2015}): \emph{\bibinfo{title}{A framework for proof
  certificates in finite state exploration}}.
\newblock In \bibinfo{editor}{Cezary \surnamestart Kaliszyk\surnameend} \&
  \bibinfo{editor}{Andrei \surnamestart Paskevich\surnameend}, editors: {\sl
  \bibinfo{booktitle}{Proceedings of the Fourth Workshop on Proof eXchange for
  Theorem Proving}}, {\sl \bibinfo{series}{Electronic Proceedings in
  Theoretical Computer Science}} \bibinfo{volume}{186}, pp.
  \bibinfo{pages}{11--26}, \doi{10.4204/EPTCS.186.4}.

\bibitemdeclare{article}{heath19jar}
\bibitem{heath19jar}
\bibinfo{author}{Quentin \surnamestart Heath\surnameend} \&
  \bibinfo{author}{Dale \surnamestart Miller\surnameend}
  (\bibinfo{year}{2019}): \emph{\bibinfo{title}{A proof theory for model
  checking}}.
\newblock {\sl \bibinfo{journal}{J. of Automated Reasoning}}
  \bibinfo{volume}{63}(\bibinfo{number}{4}), pp. \bibinfo{pages}{857--885},
  \doi{10.1007/s10817-018-9475-3}.

\bibitemdeclare{article}{lengrand10lmcs}
\bibitem{lengrand10lmcs}
\bibinfo{author}{{St\'ephane} \surnamestart Lengrand\surnameend},
  \bibinfo{author}{Roy \surnamestart Dyckhoff\surnameend} \&
  \bibinfo{author}{James \surnamestart McKinna\surnameend}
  (\bibinfo{year}{2011}): \emph{\bibinfo{title}{A Focused Sequent Calculus
  Framework for Proof Search in Pure Type Systems}}.
\newblock {\sl \bibinfo{journal}{Logical Methods in Computer Science}}
  \bibinfo{volume}{7}(\bibinfo{number}{1}).
\newblock
  \urlprefix\url{http://www.lix.polytechnique.fr/~lengrand/Work/Reports/TTSC09.pdf}.

\bibitemdeclare{article}{liang09tcs}
\bibitem{liang09tcs}
\bibinfo{author}{Chuck \surnamestart Liang\surnameend} \& \bibinfo{author}{Dale
  \surnamestart Miller\surnameend} (\bibinfo{year}{2009}):
  \emph{\bibinfo{title}{Focusing and Polarization in Linear, Intuitionistic,
  and Classical Logics}}.
\newblock {\sl \bibinfo{journal}{Theoretical Computer Science}}
  \bibinfo{volume}{410}(\bibinfo{number}{46}), pp. \bibinfo{pages}{4747--4768},
  \doi{10.1016/j.tcs.2009.07.041}.

\bibitemdeclare{book}{miller12proghol}
\bibitem{miller12proghol}
\bibinfo{author}{Dale \surnamestart Miller\surnameend} \&
  \bibinfo{author}{Gopalan \surnamestart Nadathur\surnameend}
  (\bibinfo{year}{2012}): \emph{\bibinfo{title}{Programming with Higher-Order
  Logic}}.
\newblock \bibinfo{publisher}{Cambridge University Press},
  \doi{10.1017/CBO9781139021326}.

\bibitemdeclare{misc}{coq-elpi20web}
\bibitem{coq-elpi20web}
\bibinfo{author}{Enrico \surnamestart Tassi\surnameend} (\bibinfo{year}{2020}):
  \emph{\bibinfo{title}{Coq plugin embedding {ELPI}}}.
\newblock \bibinfo{howpublished}{\url{https://github.com/LPCIC/coq-elpi}}.

\end{thebibliography}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

%%  LocalWords:  inputenc fontenc Elpi minimalistic fpc Coq's
%%  LocalWords:  elpi

### 2020-10-16 papers/coLP/lfmtp20/lfm20.tex
Added another paper.

\RequirePackage[utf8]{inputenc}
\RequirePackage[T1]{fontenc}
\documentclass[submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{LFMTP 2020}
\usepackage{url,xcolor}
\usepackage{listings}
\usepackage{letltxmacro}
\input{../types20/listing-macros}  % Use to get better highlighing of code
\lstset{language=lprolog}
\lstset{language=abella}
\textheight             632.01pt         % 46 lines exactly = 21.98cm  %
\title{On the Proof Theory of Property-Based Testing of Coinductive Specifications, or:\\
 PBT to Infinity and beyond}
% \author{Rob van Glabbeek
% \institute{NICTA\\ Sydney, Australia}
% \institute{School of Computer Science and Engineering\\
% University of New South Wales\thanks{A fine university.}\\
% Sydney, Australia}
% \email{rvg@cs.stanford.edu}
% \and
% Co Author \qquad\qquad Yet S. Else
% \institute{Stanford Univeristy\\
% California, USA}
% \email{\quad is@gmail.com \quad\qquad somebody@else.org}
% }
\author{Roberto Blanco
\institute{INRIA Paris, France}
\and Dale Miller
\institute{INRIA Saclay \\ LIX, \'Ecole Polytechnique, France}
\and Alberto Momigliano
\institute{DI, Universit\`a degli Studi di Milano, Italy}
}

\def\authorrunning{Blanco, Miller and Momigliano}
\def\titlerunning{PBT to Infinity and   beyond}
\begin{document}

\maketitle

Reasoning about infinite computations via coinduction and corecursion
has an ever-increasing relevance in formal methods and, in particular,
in the semantics of programming languages, starting
from~\cite{milner91tcs} (see~\cite{2007-Leroy-Grall} for a
compelling example) and, of course, coinduction underlies (the
meta-theory of) process calculi. This importance was acknowledged by researchers
in proof assistants, who promptly provided support for coinduction and
corecursion from the early '90s on: see~\cite{Paulson97,Gim95types}
for the beginning of the story concerning two popular frameworks.

It also became apparent that tools that search for
refutations/counter-examples of conjectures before attempting a
formal proof are invaluable: this is particularly true in PL theory,
where proofs tend to be shallow but may have hundreds of cases.  One
such approach is \emph{property-based testing} (PBT), which employs
automatic test data generation to try and refute executable
specifications.  Pioneered by \emph{QuickCheck} for functional
programming~\cite{claessen00icfp}, it has now spread to most major
proof assistants~\cite{BlanchetteBN11,QChick}.

In general, PBT does not extend immediately to testing coinductive specifications (an
exception being Isabelle's \emph{Nitpick}, which is, however, a
counter-model generator).  Extending Coq's
\emph{QuickChick}~\cite{QChick} to deal with Coq's notion of
coinduction via \emph{guarded} recursion (which is generally seen to
be a less than satisfactory approach to coinduction) is particularly
challenging.  We are not aware of applications of PBT to other forms
of coinduction, such as \emph{co-patterns}~\cite{AbelPTS13}.

While PBT originated in the functional programming community, we have
given in a previous paper (\cite{Blanco0M19}) a reconstruction of some
of its features (operational semantics, different flavors of
generation, shrinking) in proof-theoretic terms using \emph{focused
  proof systems}~\cite{andreoli92jlc,liang09tcs} and
\emph{Foundational Proof Certificates} (FPC)~\cite{chihani17jar}.  An
FPC can be used to define a range of proof structures, such as
resolution refutations, Herbrand disjuncts, tableaux, etc.
%
In the context of PBT, the proof theory setup is rather simple.
Consider an attempt to find counter-examples to a conjecture of the
form \(\forall x [(\tau(x)\wedge P(x)) \supset Q(x)]\) where $\tau$ is
a typing predicate and $P$ and $Q$ are two other predicates defined
using Horn clause specifications.
%
The negation of this conjecture is
\(\exists x [(\tau(x)\land P(x)) \land \neg Q(x)]\).
%
In searching for a focused proof of this negation, the \emph{positive
  phase} (which corresponds to the generation of possible
counter-examples) is represented by \(\exists x\) and \((\tau(x)\land
P(x))\).
%
That phase is followed by the \emph{negative phase} (which correspond
to counter-example testing) and is represented by \(\neg Q(x)\).
%
FPCs are simple logic programs that guide the search for potential
counter-examples using different generation strategies; they further
capture diverse features such as $\delta$-debugging, fault isolation,
explanation, etc.  Such a range of features can be programmed as the
\emph{clerks and experts} predicates~\cite{chihani17jar} that decorate
the sequent rules used in an FPC proof checking kernel: the kernel is
also able to do a limited amount of proof reconstruction.

% As explained in~\cite{Blanco0M19}, the standard PBT setup needs little
% more than Horn logic. % specifications.
% %
% However, when addressing potentially infinite computations, we need
% richer specifications.  While coinductive logic programming (see for
% example~\cite{Luke07} and~\cite{BasoldKL19}) may at first seem to fit
% the bill, the need to model infinite \emph{behavior} rather than
% infinite objects (i.e., (ir)rational terms on the domain of discourse)
% has lead us to adopt a much stronger logic with explicit rules for
% induction and coinduction.
% As explained in~\cite{Blanco0M19}, the standard PBT setup needs little
% more than Horn logic. % specifications.
%
%% new paragraph for R2
There are at least two ways to address potentially infinite
computations in logical terms. We can introduce infinite terms,
rational or even irrational, in our semantics, as already accounted
for in Lloyd's textbook and in~\cite{maher88lics}: % same as Maher88 ?
this has recently
been revisited in \emph{coinductive logic programming}, see, for
example,~\cite{Luke07} and the definitive~\cite{BasoldKL19}. Or we can
concentrate on modeling infinite \emph{behavior} of finite terms, for
example, divergence of a given (finite) program. We choose the latter and this requires a
logic stronger than a logic programming interpreter, namely one with explicit rules for induction and coinduction.

A natural choice for such a logic is the fixed point logic $\cal
G$~\cite{gacek11ic} % Gacek2012 is not best reference for G
and its linear logic cousin
$\mu\mathrm{MALL}$~\cite{Baelde12}, which are associated to the
\emph{Abella} proof assistant~\cite{baelde14jfr} and the \emph{Bedwyr}
model-checker.  In fact, the latter has already been used for related
aims~\cite{HeathM15}.

For a concrete example, consider a \emph{coinductive} definition
for CBV evaluation in the $\lambda$-calculus with constants
(following~\cite{2007-Leroy-Grall}). Using Bedwyr's
concrete syntax, this is written as: 
\begin{verbatim}
Define coinductive coeval: tm -> tm -> prop by
 coeval (con C) (con C);
 coeval (fun R) (fun R);
 coeval (app M N) V := 
   exists R W, coeval M (fun R) /\ coeval N W /\ coeval (R W) V.
\end{verbatim}
As is well-known, co-evaluation fails to be deterministic, since a divergent
term such as $\Omega$ co-evaluates to anything.  We can confirm this by
searching for a proof of the following formula:
\begin{verbatim}
exists E V1 V2, coeval E V1 /\ coeval E V2 /\ (V1 = V2 -> false)
\end{verbatim}
Proving this query entails a way to generate such (finite) terms and then checking,
with a crucial appeal to the coinduction, that they are in the required relation.

Other applications of PBT include separating the various
notions of equivalences in the lambda-calculus and various process
calculi: for example,  applicative and ground similarity in
PCFL~\cite{PITobtpe}, or analogous standard results in the
$\pi$-calculus.
%
While similar goals have been achieved in the literature for labeled transition
systems (using, for example, the \emph{Concurrency
  Workbench}), it is a remarkable feature of the proof-theoretic
setting that we can generalize PBT from a system without bindings
(say, CCS) to a system with bindings (say, the $\pi$-calculus).
%
Such ease is possible since proof theory accommodates the
\emph{$\lambda$-tree syntax} approach to treating
bindings~\cite{miller18jar}: in particular, both Abella and Bedwyr
include the $\nabla$ quantifier~\cite{miller05tocl}.

In our current setup, we attempt to find counter-examples using
Bedwyr to execute both the generation of test cases (controlled by
using specific FPCs~\cite{Blanco0M19}) and the testing phase.
%
Such an implementation of PBT allows us to
piggyback on Bedwyr's facilities for efficient proof 
search via tabling for (co)inductive predicates.
%
The treatment of the negation in the testing phase is, as usual, a
sticky point~\cite{Momigliano00}.
%
%One approach to eliminating negation from intuitionistic specification
%can be based on the techniques in.
%
However, if we identify, as we do, the proof theory behind model
checking as based on the linear logic $\mu$MALL~\cite{HeathM19}, in
that setting, occurrences of negations can be eliminated by using De
Morgan duality and inequality.

\bibliographystyle{eptcs}
\bibliography{../colp,../../ppdp19/l}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

%  LocalWords:  Coinductive QuickCheck utf inputenc fontenc Herbrand
%  LocalWords:  corecursion coinductive corecursive QuickChick CBV ir
%  LocalWords:  disjuncts PCFL LTS Harrop PBT Coq's coinductively CCS
%  LocalWords:  intuitionistic FPCs Bedwyr Bedwyr's

### 2020-10-16 papers/coLP/lfmtp20/talk-nopause.tex
Added another paper.

\documentclass{beamer}
\mode<presentation>
%\usepackage{tikz}
%\usetheme{Frankfurt}
\usecolortheme{seahorse}
%\usepackage{xcolor}
\usepackage{graphics}
\usepackage{amsthm,verbatim,booktabs}
\usepackage{epsfig}
%\input{nominal}
\usepackage{boxedminipage}     % for surrounding boxes in figures
\usepackage{proof}
\usepackage{amsmath} 
\usepackage{xspace}
\newcommand{\red}[1]{{\color{red} #1}}
\newcommand{\NF}{\emph{NAF}\xspace}
\newcommand{\trueExpert }[1]{{\true_e}(#1)}
\newcommand{\eqExpert }[1]{{=_e}(#1)}
\newcommand{\unfoldExpert}[2]{{\hbox{\sl unfold}_e}(#1,#2)}
\newcommand{\andExpert}[3]{{\wedge_e}(#1,#2,#3)}
\newcommand{\andExpertLJF}[6]{{\wedge_e}(#1,#2,#3,#4,#5,#6)}
\newcommand{\orExpert  }[3]{{\vee_e}(#1,#2,#3)}
\newcommand{\someExpert}[3]{\exists_e(#1,#2,#3)}
\newcommand{\initExpert}[2]{\hbox{\sl init}_e(#1,#2)}
\newcommand{\lP}{$\lambda$Prolog\xspace}
\newcommand{\true}{tt}
  \newcommand{\XXi}{{\color{blue}{\Xi}}}
\newcommand{\ok}{\checkmark}
  \def\lam{\lambda}
\def\Lam{\Lambda}
\def\arrow{\rightarrow}
\def\oftp{\mathord{:}}
\def\hastype{\mathrel{:}}
\def\bnfas{\mathrel{::=}}
\def\bnfalt{\mid}

\newenvironment{myslide}[1]
    {\begin{frame}\frametitle{#1}}
    {\end{frame}}

\newcommand{\new}[1][]{\reflectbox{\sf{#1{}N}}}
% \newcommand{\aprolog}{{\ensuremath \alpha}{Pro\-log}\xspace}
% \newcommand{\acheck}{{\ensuremath \alpha}{Check}\xspace}
\newcommand{\lprolog}{{\ensuremath \lam}{Pro\-log}\xspace}
\newcommand{\plus}{\hbox{\sl app}}

%\newtheorem{theorem}{Theorem}
%\newtheorem{lemma}{Lemma}
%\newtheorem{conj}{Conjecture}
\usepackage{listings}
\input{listing-macros}
\lstset{language=lprolog}
\lstset{language=abella}
\begin{document}

\title{
Property-Based Testing to Infinity and beyond!}

\author{Alberto Momigliano\\ joint work with {Roberto Blanco} and Dale Miller}
\institute{LFMTP 2020}
\date{June 29, 2020}

\begin{frame}
\titlepage
\end{frame}
 %---------------------------------------------------------
\begin{frame}
  \frametitle{Take away from the talk}
  \begin{itemize}
  \item Standard PBT  is just search in a focused sequent calculus akin to uniform proofs
  \item PBT's data generation strategies (exhaustive, random) and
    other features (shrinking, bug provenance) can be programmed as
    instantiations of the Foundational Proof Certificates paradigm.
  \item Co-PBT requires a stronger logic with fixed points, preferably linear.
  \end{itemize}
    % \pause

   Now you can go (back) to sleep.

\end{frame}

%\section{Part 1: Motivation}
%----------------------------------------------------------------------------
%\begin{frame}[fragile]\frametitle{Off the record}
%-------------------------------------------------
 %  \begin{itemize}
%   \item After almost 20 years of formal verification with Twelf,
%     Isabelle/HOL, Coq, Abella, I'm a bit worn out
%   \item I still find it  a very demanding, often frustrating, day
%     job.
% % \pause
%   \item Especially when the theorem I'm trying to prove is, ehm,
%     wrong. I mean, \emph{almost right}:
% % \pause
%     \begin{itemize}
%     \item statement is too strong/weak
%     \item there are minor mistakes in the spec I'm reasoning about
%     \end{itemize}
%   \item A failed proof attempt not the best way to debug those kind of
%     mistakes
%   \item That's why I'm inclined to give \emph{testing} a try (and I'm
%     in good company!)  % \pause
% \item Not any testing: \textbf{property-based testing}
%   \end{itemize}

% \end{frame}
%----------------------------------------------------------------------------
\begin{frame}[fragile]\frametitle{Property-based Testing}
  \begin{itemize}
  \item A light-weight validation approach to software correctness
    % \begin{enumerate}
    % \item automatic generation of test data, against
    % \item executable program specifications.
    % \end{enumerate}
    % % \pause
  \item The programmer specifies executable properties  the code should satisfy
  \item PBT tries to refute them  by trying a  large number of
    automatically generated cases.
    % \pause
     \item Brought together in \emph{QuickCheck} (Claessen \& Hughes '00) for Haskell \dots

     \item \dots and now available for pretty much every PL (and
       commercialized for Erlang).
  \end{itemize}

\end{frame}

% %----------------------------------------------------------------------------
\begin{frame}[fragile]
  \frametitle{FsCheck example: 1/2}
  Does insertion in an ordered list preserve order?
  \begin{small}
    \begin{lstlisting}
let rec ordered = function                      
 |[] -> true
 |[x] -> true
 |x::y::ys -> x <= y && ordered ys

let rec insert x  = function 
 |[] -> [x]
 |c::cs -> 
   if x <= c then x::c::cs else c::insert x cs

let prop_insert (x:int,  xs) =  
  ordered xs ==> ordered (insert  x xs)

do Check.Quick prop_insert
    \end{lstlisting}
  \end{small}

% \pause
  \begin{small}
    \begin{lstlisting}
Falsifiable, after 16 tests ... (0, [0; 0; -1])
    \end{lstlisting}
  \end{small}
\end{frame}

% %----------------------------------------------------------------------------
\begin{frame}[fragile]
  \frametitle{FsCheck example: 2/2}
Well, it would, had I the correct encoding of \texttt{ordered}
  \begin{small}
    \begin{lstlisting}
let rec ordered = function                      
   |[] -> true
   |[x] -> true
   |x::y::ys -> x <= y && ordered (y :: ys) 
                    // was ordered ys

// rest stays the same

do Check.Quick prop_insert
 \end{lstlisting}
  \end{small}
  % \pause
    \begin{small}
    \begin{lstlisting}
OK. Arguments exhausted after 37 tests.
    \end{lstlisting}
  \end{small}
  Since this is a conditional property with a  sparse pre-condition, it
  warns me of bad coverage, but this is another story\dots
\end{frame}

%\begin{frame}[fragile]\frametitle{QuickCheck's Hello World!\hfill
%     (FsCheck, actually)}
%   \begin{small}
% \begin{verbatim}
% let rec rev ls =  
%     match ls with 
%     | [] -> [] 
%     | x :: xs -> append (rev xs, [x])

% // a true spec
% let prop_revRevIsOrig (xs:int list) =  rev (rev xs) = xs;;

% do Check.Quick prop_revRevIsOrig ;;
% >> Ok, passed 100 tests.

% // a false spec
% let prop_revIsOrig (xs:int list) = rev xs = xs

% do Check.Quick prop_revIsOrig ;;

% >>  Falsifiable, after 3 tests (5 shrinks) (StdGen (518275965,...)):
% [1; 0]
%   \end{verbatim}
%   \end{small}
%   % \begin{itemize}
%   % \item 
%   % \end{itemize}

% \end{frame}

\begin{frame}
  \frametitle{PBT and theorem proving}
  \begin{itemize}
  \item One of PBT's success stories is the integration with proof assistants
  % \pause
  \item Isabelle/HOL broke the ice adopting
    \emph{random} testing some 15 years ago %and many     followed suit:
    \begin{itemize}
    \item  a la QC: Agda ['04], PVS ['06], Coq
    with QuickChick ['15]
  \item exhaustive/smart generators (Isabelle/HOL ['12])
  \item model finders (Nitpick, again in Isabelle/HOL ['11])
  \end{itemize}
% \item In fact, Pierce and co.'s \emph{Software Foundations} has a volume dedicated to PBT.
%   % \pause
\item But, wait! Isn't testing the very thing theorem proving wants to replace?
  % \pause 
\item Oh, no: test a conjecture before attempting to prove it and/or
  test a subgoal (a lemma) inside a proof
\item The beauty (wrt general testing) is: you don't
  have to invent the specs, they're exactly what you want to prove
  anyway.
  \end{itemize}
\end{frame}
% ---------
\begin{frame}\frametitle{My pet interest: \emph{mechanized
    meta-theory verification}}
\begin{itemize}
\item Consider the verification of the meta-theory of programming languages and related calculi with a proof assistant.
\item  Type soundness, proofs by logical relations, non-interference\dots
\item Proofs are mostly standard, but
  \begin{itemize}
  \item lots of work, mostly wasted in the
  \red{design} phase
\item only worthwhile
  if \red{we already ``know'' the system is correct}
\end{itemize}

% \pause
\item PBT to the rescue:
\begin{itemize}
  \item produces helpful counterexamples for incorrect systems
%  \item unhelpfully diverges for correct systems
  \item little expertise required, fully (well, sort of) automatic
\end{itemize}
  \end{itemize}
\end{frame}

% ----------------------------------------------------------------------------
\begin{frame}
  \frametitle{What about infinite computations?}
  \begin{itemize}
  \item I don't have to argue (but I will) about the relevance of
    coinduction/corecursion in PL theory:
    \begin{itemize}
    \item divergence/co-evaluation
    \item recursive (sub)typing
    \item program equivalence
    \item pretty much the whole meta-theory of process calculi
    \item \dots [add your own]
    \end{itemize}
    % \pause
  \item Lots of work on the \textbf{proving} side in most proof assistants
    \begin{itemize}
    \item guarded recursion, greatest fixed points, co-patterns, co-datatypes \dots
    \end{itemize}
    % \pause
  \item Much less on the \textbf{testing} side:
    \begin{itemize}
    \item a little with Haskell's QC (but basically using the \emph{take} lemma)
    \item a little with \emph{Nitpick}, possibly with (co)datatypes, I have to check
    \item seems hard for Coq's QuickChick.
    \end{itemize}

  \end{itemize}
\end{frame}

% ----------------------
\begin{frame}
  \frametitle{Motivating examples: co-evaluation}
  \newcommand{\step}{\Downarrow}
  \begin{itemize}
  \item Consider the CBV big step semantics of the lambda calculus,
    but \red{coinductively} [Leroy \& Grall '07]:
    \begin{small}
  \[
  \begin{array}{c}
    % \infer=[\texttt{E-V}]{V \step V}{\mathrm{value\ } V}
        \infer=[\texttt{E-L}]{\lambda x.\, M \step \lambda x.\, M}{}
  \qquad
    \infer=[\texttt{E-A}]{M_1\cdot M_2 \step V}{M_1 \step \lambda x.\ M\quad M_2\step V_2\quad  M\{V_2/x\}\step V}
    \end{array}
  \]
\end{small}
\item Is this deterministic? Is it type preserving?
  % \pause

\item No, on both counts:
  \begin{itemize}
  \item a divergent term such as $\Omega$ co-evaluates to anything.
  \item a variant of the Y-combinator falsifies preservation [Filinski].
  \end{itemize}

\item In fact, one can argue that this notion of co-evaluation makes
  little sense [Ancona et al. '15] and wouldn't it be nice if your PBT
  tool would help you with that?
\end{itemize}
\end{frame}
% ----------------------
\begin{frame}
  \frametitle{Motivating examples: separating equivalences}
  \begin{itemize}
  \item Consider  \textbf{PCF} with lazy lists and
    observational equivalence via \red{bisimilarity}
    [Pitts '98]
  \item There are several notions of equivalence, e.g., depending if you observe
    convergence at any type (\red{applicative}) or at base type
    (\red{ground}). Can we separate them?
    \begin{itemize}
    \item $\lambda x.\, \perp$ is \red{ground} but not \red{applicative} bisimilar to $\perp$
    \item Same for the eta and surjective pairing laws. 
      \end{itemize}

  \item Similar results in the $\pi$ calculus:
    \begin{itemize}
    \item Ground bisimilarity not a congruence
      \item  Early and late bisimilarity not preserved by inputs
    \end{itemize}
  \end{itemize}
\end{frame}
% ----------------------
\begin{frame}
  \frametitle{\textbf{Non}-examples}
  \begin{itemize}
  \item What about \red{streams}? All those nice Haskell-like equations, as in Louise Dennis' thesis?
    % \pause
  \item Thanks, but no thanks
    % \pause
  \item Here we concentrate on \red{infinite behavior} (e.g.,
    is a finite program diverging) rather than \red{infinite objects}
    (e.g., is $2$ the last element of the infinite stream $1::1::1\dots$?).
  \item In a logical view of coPBT, we would need to
    \textbf{construct} such infinite terms as cex, and the literature is not satisfactory:
    % \pause
    \begin{itemize}
    \item coinductive LP works with rational terms: problematic and hopelessly incomplete
% \pause
    \item ``Coinduction in uniform'' uses a fixed point term
      constructor: does not sit well within a logical framework
      (adequacy of encodings, canonical forms etc.).
    \end{itemize}
  \end{itemize}
\end{frame}

% \begin{frame}[fragile]\frametitle{Not so fast/quick\dots}

%     \begin{small}
% \begin{verbatim}
% ordered xs ==> ordered (insert x xs)
% \end{verbatim}
%     \end{small}

% \end{frame}
%----------------------------------------------------------------------------

%---------------------------\-------------
\begin{myslide}{PBT: from FP to LP}
  \begin{itemize}
%  \item Lots of activities, mainly in the FP community --- lack of logical common thread
  \item PBT  was born and raised functional, but is rediscovering logic programming:
    \begin{itemize}
    \item mode analysis in Isabelle/HOL and QuickChick's automatic derivation of 
        generators
    \item (Randomized) backchaining in PLT-Redex
    \item Narrowing in LazySmallCheck \dots

    \end{itemize}
  % \pause
  \item What the last 30 years have taught us is that if we take a
    \red{proof-theoretic} view of LP, good things start to happen
  \item In particular: \red{focusing} and a \red{treatment of
      bindings}
  % \item While standard PBT can be accounted for by basically
  %   \red{focused proofs} on a bit more than Horn logic, for coPBT we
  %   need to resort to stronger (fixed points) logic, but still keep a LP vibe.

  \end{itemize}
\end{myslide}
% ----------------------
\begin{frame}
  \frametitle{PBT: the logical view}
  \begin{itemize}
  \item Specifications and code (think \texttt{ordered} or more interestingly the operational semantics of a PL) are logical theories 
  % \pause
  \item Trying to refute a property of the form
    \[\forall x\colon\tau\ [P(x) \supset Q(x)]\] means searching for a (focused) proof of
    \[\exists x [(\tau(x)\land P(x)) \land \neg Q(x)]\] yielding a
    a $t$ of type $\tau$ s.t.\ $P(t)$ holds and $Q(t)$ does not % \pause
  \item The generate-and-test approach of PBT can be seen in terms of
    focused sequent calculus proof where the \red{positive} phase
    corresponds to generation and a single \red{negative} one to
    testing.
  \item Intuition: generating is hard (lots of backtracking), testing is easy (deterministic computation).
  \end{itemize}
\end{frame}

% ----------------------
\begin{myslide}{Going deeper: PBT via FPC}
  \begin{itemize}
  \item A flexible  way to look at those proofs is as a
    \red{proof reconstruction} problem in the \red{Foundational Proof
      Certificate} framework [Chihani, Miller \& Renaud 2017]
\item FPC proposed as a means of defining proof
structures used in a range of different theorem provers
% \pause
\item Think of a focused sequent calculus
  augmented with predicates (\textbf{clerks} for the negative phase and \textbf{experts}
  for the positive one) that produce and process information to drive
  the checking/reconstruction of a proof.
% \pause
\item For PBT, use of FPC as a way to
describe \red{generation strategies} and as a way to combine them.
  \end{itemize}

\end{myslide}
%----------------
\newcommand{\andd}{\wedge}
\newcommand{\impp}{\supset}

\newcommand{\instan}[1]{\hbox{\sl grnd}~(#1)}
\newcommand{\Pscr}{{\mathcal P}}
%
% -----------------------------------------------
\begin{frame}
  \frametitle{Proof system for Horn logic with certificates and experts}

\[
\infer{\XXi\vdash G_1\wedge G_2}
      {\XXi_1\vdash G_1\qquad \XXi_2\vdash G_2\qquad \andExpert{\XXi}{\XXi_1}{\XXi_2}}
\qquad
\infer{\XXi\vdash \true}
      {\trueExpert{\XXi}}
\]
\vskip -6pt
\[
\infer{\XXi\vdash G_1\vee G_2}
      {\XXi'\vdash G_i\qquad \orExpert{\XXi}{\XXi'}{i}}
\qquad
\infer{\XXi\vdash \exists x. G}
      {\XXi'\vdash G[t/x]\qquad \someExpert{\XXi}{\XXi'}{t}}
\]
\vskip -6pt
\[
\infer{\XXi\vdash t = t}
      {\eqExpert{\XXi}}
\qquad
\infer{\XXi\vdash A}
      {\XXi'\vdash G \quad (A~\hbox{\tt :-}~G)\in\instan\Pscr
                     \quad \unfoldExpert{\XXi}{\XXi'}}
\]
% \pause
\begin{itemize}
% \item Certificates can be seen as proof terms or more in general any
%   evidence/oracle that can help, via the specification of the \textbf{expert}
%   predicates, to reconstruct a proof
\item An FPC is an instantiation of $\XXi$ and experts predicates
\end{itemize}
\end{frame}
% -----------------------
% \begin{frame}
%   \frametitle{Sample FPC 1: height bound}
%   \begin{itemize}
%   \item To begin with, we encode exhaustive generation by building proofs bounded by their
%     height (the gen strategy of $\alpha$Check)
%   % \pause
%   \item Certificates are just integers and the only active expert is a
%     simple non-zero check while backchaining: \fbox{$n\vdash G$}
%     \begin{itemize}
%     \item Note that the bound is \red{additive} wrt conjunction
%     \end{itemize}
%  \end{itemize}
%  % \pause
%  \[
% \infer{n\vdash G_1\wedge G_2}
% {n\vdash G_1\qquad n\vdash G_2 %\qquad \andExpert{\XXi}{\XXi_1}{\XXi_2}
% }
% \qquad
% \infer{n\vdash \true}
% {
%   % \trueExpert{\XXi}
% }
% \]
% \vskip -6pt
% \[
% \infer{n\vdash G_1\vee G_2}
% {n\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{n\vdash \exists x. G}
% {n\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \]
% \vskip -6pt
% \[
% \infer{n\vdash t = t}
% {
%   % \eqExpert{\XXi}
% }
% \qquad
% \infer{n + 1\vdash A}
%       {n\vdash G \quad (A~\hbox{\tt :-}~G)\in\instan\Pscr
%                      \quad n \geq 0}
% \]

% \end{frame}
% -----------------------------
\begin{frame}
  \frametitle{FPC for exhaustive generation}
  \begin{itemize}
  \item We capture exhaustive generation by building proofs bounded by their 
    size -- many others in [PPDP '19] paper
  % \pause
  \item Certificates are pairs of integers and the only active expert
    is the non-zero check while backchaining: \fbox{$n,m\vdash G$}
    %     \begin{itemize}
    % \item Note that the certificate behaves \red{multiplicatively} wrt conjunction
    % \end{itemize}

 \end{itemize}
 % \pause
 \[
\infer{n,m\vdash G_1\wedge G_2}
{n,n_1\vdash G_1\qquad n_1,m\vdash G_2 %\qquad \andExpert{\XXi}{\XXi_1}{\XXi_2}
}
\qquad
\infer{n,n\vdash \true}
{
  % \trueExpert{\XXi}
}
\]
\vskip -6pt
\[
\infer{n,m\vdash G_1\vee G_2}
{n,m\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
}
\qquad
\infer{n,m\vdash \exists x. G}
{n,m\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
}
\]
\vskip -6pt
\[
\infer{n,n\vdash t = t}
{
  % \eqExpert{\XXi}
}
\qquad
\infer{n + 1,m\vdash A}
      {n,m\vdash G \quad (A~\hbox{\tt :-}~G)\in\instan\Pscr
                     \quad n \geq 0}
\]

\end{frame}

% -----------------------------
% \begin{frame}
%   \frametitle{Sample FPC 3: maximal info}
%   \begin{itemize}
%   \item What about recording explicitly all information within a
% proof:
% \begin{enumerate}
% \item all disjunctive choices
% \item all substitution
% instances for existential quantifiers  
% \end{enumerate}
% % \pause
%   \item Certificates are basically proof terms for Horn logic
%      \fbox{$M\vdash G$}
%     \end{itemize}

% % \pause
%  \[
% \infer{\langle M_i,M_2\rangle\vdash G_1\wedge G_2}
% {M_1\vdash G_1\qquad M_2\vdash G_2}
% \qquad
% \infer{\langle\ \rangle\vdash \true}
% {}
% \]
% \vskip -6pt
% \[
% \infer{inj_i(M)\vdash G_1\vee G_2}
% {M\vdash G_i}
% \qquad
% \infer{\langle t,M\rangle\vdash \exists x. G}
% {M\vdash G[t/x]}
% \]
% \vskip -6pt
% \[
% \infer{\mathit{refl}\vdash t = t}
% {}
% \qquad
% \infer{M\vdash A}
%       {M\vdash G \quad  (A~\hbox{\tt :-}~G)\in\instan\Pscr}                    
% \]

% \end{frame}

% % ----------------
% \begin{frame}
%   \frametitle{FPC for random generation}
%   \begin{itemize}
%   \item We support random generation by implementing \red{randomized}
%     backtracking, as opposed to the usual chronological one.
%   % \pause
%   \item There are many possible variations: the simplest one  permutes the program at each backchaining --- certificates are empty
%     \[\infer{\vdash A}
%       {\vdash G \quad  (A~\hbox{\tt :-}~G)\in\instan{\pi\Pscr}}                 
% \]
% % \pause
% \item Or we can use non-uniform distributions, by viewing certificates as assigning \textbf{weights} to clause ids
%   $$\XXi \quad\bnfas\quad\cdot\bnfalt\XXi, (id, n) $$
%   and have the backchaining expert randomically select a clause according to the weight. In QuickCheck:
%   \begin{enumerate}
%   \item the \texttt{oneof} combinator
%   \item the \texttt{frequency} combinator
%   \end{enumerate}

%   \end{itemize}
% \end{frame}

% ----------------

% \begin{frame}[fragile]
%   \frametitle{More FPC's for PBT}
%   \begin{itemize}
%   \item 
%   In previous paper we showed how FPC's reconstruct most features of PBT:
%   \begin{itemize}
%   \item random generation
%   \item \red{Pairing}  \emph{restricts} proofs
%     satisfying two FPCs simultaneously and
%  accounts for \textbf{shrinking} of (random)
%   cex:  
% \item for other proof management duties, such as bug provenance and
%   explanation
% % % \pause
% % \item The FPC way:
% %   \begin{enumerate}
% %   \item Collect all substitution terms in an existing proof (e.g., with the \textbf{Max} FPC
% %   \item Search again restricting the existential quantifiers to use
% %     (strict) subterms of terms collected in the first pass.
% %   \end{enumerate}
% \end{itemize}
% % \pause
%  \item Moves seamlessly from
%     standard first-order terms to one containing \textbf{binders}
%     (via $\lambda$-tree syntax)  with
%   $\nabla$, a formula level quantifier encoding binders mobility.
% %% \pause
% \end{itemize}
% \end{frame}
% % ----------------

% \begin{frame}[fragile]
%   \frametitle{More FPC's}
%   \begin{itemize}
%   \item FPC collecting maximal info (basically proof terms)
%   \item FPC for random generation
%   \item It is possible to build an FPC that \emph{restricts} proofs
%     satisfying two FPCs simultaneously:
%     \begin{itemize}
%     \item \red{Pairing} $\XXi_1 \oplus \XXi_2$ is  such an
%       operations on certificates% , whereby we require the corresponding
%       % experts also succeed for both $\XXi_i$ and for \lsti{orE} and
%       % \lsti{someE}, also return the same choice and substitution term
%     \end{itemize}
% % \pause
% \item For example, \fbox{$\langle (5,0), M\rangle\vdash G$} collects  the
% maximal   information of a proof $M$ of $G$ of size at most $5$.
%   % \pause
% \item Pairing also accounts for \textbf{shrinking} of (random)
%   cex:  smaller variants are generated and
%   tested to see if they are still cex.
%   \item We can use certificates also for other proof management duties, such
%   as bug provenance and explanation
% % % \pause
% % \item The FPC way:
% %   \begin{enumerate}
% %   \item Collect all substitution terms in an existing proof (e.g., with the \textbf{Max} FPC
% %   \item Search again restricting the existential quantifiers to use
% %     (strict) subterms of terms collected in the first pass.
% %   \end{enumerate}
%   \end{itemize}
% \end{frame}

% ----------------

% \begin{frame}[fragile]
%   \frametitle{From algebraic to binding signatures}

%   \begin{itemize}
%   \item The proof-theoretic view allows us to move seamlessly from
%     standard first-order terms to one containing \textbf{binders}
%     ($\lambda$-tree syntax)
%   \item Note: no current reasoning tool supports (I think) both proofs \textbf{and}
%     disproofs with native binders, unless via DB encodings 
% %% \pause
% \item In meta-logics based fixed points, this is accomplished with
%   $\nabla$, a formula level quantifier encoding binders mobility.
% \item $\nabla$ needs no experts

%     \end{itemize}
% \end{frame}

% --------------------------
\begin{frame}
  \frametitle{From PBT to coPBT}
  \begin{itemize}
  \item Standard PBT requires only finite computations and can be accounted for with logic programming (Horn) queries
  \item Recall failure of determinism of \emph{co-evaluation}:
    $\exists M\, V_1\, V_2\ [istm(M) \land istm( V_1) \land istm(V_2) \land M\Downarrow V_1\land M\Downarrow V_2 \land V_1 \not = V_2 ]$
    % \pause
    \item Generation (pred \texttt{istm}) is still Horn and finitary: it will be driven by FPC

    \item If you want to capture its
      \red{infinite} behavior of  $ M\Downarrow V_i$, you need a proof-theory with rules for
      fixed points, such as the ones underlying \emph{Abella} and
      \emph{Bedwyr}% \\
      % (there are alternatives: co-patterns, guarded recursive types, I just don't know enough)
    \item Other specs such as bisimilarity are Harrop and will also need proofs by \red{case analysis}, which is readily available with fixed points.

  \end{itemize}
\end{frame}
% --------------------------
\begin{frame}
  \frametitle{$\mu$MALL 1/2}

  \begin{itemize}
  \item Extends multiplicative additive linear logic with fixed points and free equality [Baelde \& Miller '07]:

    $\Rightarrow$ co-evaluation is encoded (using $\lambda$-tree syntax) as this greatest fixed point 
    (see linear version of Clark's completion):
    $$
\begin{small}
\begin{array}{l}
    \nu (\lambda {CE}.\lambda m.\lambda m'. (\exists M.\, m = (fun\  M) \otimes  m' = (fun\  M)  \oplus\mbox{}\\
  \quad(\exists M_1\, M_2\, M\,  V_2\, V.\ m = (app\ M_1\ M_2)  \otimes m' =  V \otimes
     ({CE} \ M_1 \ (fun\  M)) \\
  \quad \mbox{} \otimes ({CE\ } M_2\ V_2)   \otimes ({CE} \ (M \ V_2) \ V))
 \end{array}
\end{small}
 $$
\item Left sequent rule of $\nu$ is unfolding (case analysis), right is coinduction via simulation:
  \begin{small}
    \[
      \begin{array}{l}
        \infer[\nu L]{\Gamma, \nu B\vec t\vdash G}{\Gamma, B(\nu B)\vec t\vdash G}\qquad
        \infer[\nu R]{\Gamma\vdash \nu B\vec t}{\Gamma\vdash S\vec t\quad S\vec x\vdash B S\vec x}
      \end{array}
      \]
  \end{small}
  \end{itemize}
\end{frame}
% --------------------------
\begin{frame}
  \frametitle{$\mu$MALL 2/2}
  \begin{itemize}
  \item $\mu$MALL provides a proof-search interpretation for some aspects of model checking, see [Heath \& Miller '19 ] for the finite case.
%      \item $\mu$MALL has a significant and rigid focusing theorem and is sound wrt  Peano  arithmetic (which lacks focusing).
  \item Polarization: Prolog programs = least fixed points = purely positive. Dually, greatest fixed points:
    \begin{itemize}
    \item Often in model checking problems we just need arbitrary fixed points to be unrolled and polarization is neutral
    \end{itemize}
  \item Fixed points build in contraction, but for coPBT that's the only place we need it.
  \end{itemize}
\end{frame}
% ------------------------------------------------
% skip this
\begin{frame}
  \frametitle{$\mu$MALL and negation}
  \begin{itemize}
  \item Recall the PBT query: $\exists x [(\tau(x)\land P(x)) \land \neg Q(x)$
  \item With \red{negation} and logic (programming), things get hairy:
  \item In our account of PBT, we managed to see $\neg$ as negation-as-failure, no certificate needed % (just exploration of the search tree)
  % \pause
  \item $\mu$MALL builds in  for Horn spec the \red{CWA}:  $\neg Q$ as $Q\rightarrow\perp$
  \item Otherwise, we can use de Morgan and inequalities to \red{eliminate} negation. E.g., non applicative simulation $m \not\leq_a n$:
  \end{itemize}
  \begin{small}
   $$
\begin{array}{l}
 \mu(\lambda NAS.\lambda m.\lambda n.\exists M'.\,  m\Downarrow (fun\ M')
  \otimes
                         \forall N'.\, n\Downarrow (fun\ N') \multimap\mbox{}\\
  \qquad\exists R.\ ({NAS\ } (M' \ R) \  (N'\ R)))
\end{array}
$$
\end{small}
\end{frame}

\begin{frame}[fragile]
  \frametitle{A Proof-of-concept implementation: \textbf{Bedwyr}}
  We piggy-back our prototype on top of the \emph{Bedwyr} model-checker, which:
  \begin{itemize}
  \item implements a depth-first fully automatic search for focused proofs of a fragment of $\mu$MALL
  \item supports $\lambda$-tree syntax: binding in terms, pattern unification, binder mobility, $\nabla$-quantification
  \item implements (co)induction as loop detection via \textbf{tables} (modulo equivariance).
  \end{itemize}
\end{frame}

% --------------------------
\begin{frame}[fragile]
  \frametitle{A Proof-of-concept implementation: architecture}
  \begin{itemize}
  \item We use a ``multi-level'' approach where FPC drive generation and Bedwyr does the rest
  % \pause
    \item generators are reified in \lsti{prog} clauses over object
      level formulae of type \lsti{oo};
      \begin{small}
\begin{lstlisting}
Define prog : oo -> oo -> prop by
 prog (istm Ctx (app Exp1 Exp2))
    (istm Ctx Exp1) & (istm Ctx Exp2) ; ...
\end{lstlisting}
      \end{small}
      % \pause
  \item For each certificate format \lsti{cert},
    generation is driven by a meta-interpreter \lsti{check}
    parameterized by a \lsti{prog}
    \begin{small}
\begin{lstlisting}
Define check: cert -> oo -> (oo -> oo -> prop) -> prop by
 check Cert A Prog := unfoldE Cert Cert' /\ Prog A G /\ check Cert' G Prog.
\end{lstlisting}
  \end{small}

  %    % \pause

%    \end{itemize}

  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Our fav example}
  Plain Bedwyr is in charge of precondition and testing phase:
      \begin{small}
\begin{lstlisting}
Define coinductive coeval : i -> i -> prop by
 coeval (fun R) (fun R) ;
 coeval (app M N) V := coeval M (fun R) /\ coeval N W /\ coeval (R W) V.

% the query: gen by height
?= check (qheight 4) (istm nl M) & (istm nl M1) & (istm nl M2) /\ coeval M M1 /\ coeval M M2 /\ (M1 = M2 -> false).
\end{lstlisting}
      \end{small}
      % \pause
      \begin{small}
\begin{lstlisting}
Found a solution  + 45ms:
 M2 = app (fun (x\ x)) (fun (x\ x))
 M1 = fun (x\ x)
 M  = app (fun (x\ app x x)) (fun (x\ app x x))
\end{lstlisting}
      \end{small}
\end{frame}

%-----------------------
  \begin{myslide}{Conclusions}
  \begin{itemize}
  \item PBT successfully
    complements theorem proving with a preliminary phase of conjecture
    testing, but its support for checking coinductive spec is unsatisfactory

\item We have shown as the FPC-based proof-theoretic reconstruction of
  PBT extends to such specs by relying on stronger logics.

   \item We have presented a proof-of-concept implementation in Bedwyr
  \end{itemize}

\end{myslide}

%------------------------------------------------------------
\begin{myslide}{Current and Future Work: implementation}
  \begin{itemize}
  \item PBT uses \textbf{mode} information to restrict and delay term generation:
    \begin{itemize}
    \item w.r.t.\ standard evaluation $M\Downarrow V$,  need to generate only $M$ and often can use the judgment  as a generator.
    \end{itemize}
  \item Not the case for coinductive judgments that typically do not compute, only check (see also bisimilarity etc.)
    \begin{itemize}
    \item refuting non-det of \texttt{coeval} requires the unbounded, orthogonal generation of 3 terms. Does not scale.
    \end{itemize}
    % \pause
  \item deal with the combinatorial explosion via \red{fuzzying}:
    \begin{itemize}
    \item Generate one term and obtain the others by (random)
      mutations, possibly preserving global invariants (viz.,
      typing) %(see statesin non-interefernce)
    \end{itemize}
  \item Pair it with the idea of \emph{pre-computing} equivalence
    classes of terms of given depth 'a la Tarau ['18]
  \end{itemize}
  \end{myslide}
%------------------------------------------------------------
\begin{myslide}{Current and Future Work: extensions}
  \begin{itemize}
  \item Integrate with Abella's workflow, both at the top-level
    (disproving conjectures) and inside a proof attempt (disproving
    subgoals).    
\item  Investigate an \emph{explanation} tool for attributing ``blame'' for
the origin of a counterexample
% \begin{enumerate}
% \item certificate distillation for wrong answers
% \item abduction for missing answers
% \end{enumerate}
  \item Look into sub-structural object logics for PBT-ing specs about state and concurrency
  \end{itemize}
\end{myslide}
\begin{frame}
  \begin{center}
    \begin{LARGE}
      Thank you!
    \end{LARGE}
  \end{center}

\end{frame}

\end{document}
% --------------------------------------------------------------

% JUNK

  \newcommand{\wedgep}{\wedge^{\!+}}
\begin{myslide}{$\mu$MALL}
  \begin{itemize}
  \item As the plan is to have a PBT tool for \textbf{Abella}, we
    have in mind specs and checks in \red{multiplicative additive linear
      logic} with (for the time being) \red{least fixed points}
    (Baelde \& Miller)
  \item E.g.~, the \texttt{append} predicate is:
\begin{align*}
\plus \equiv &\mu\lambda A\lambda xs\lambda ys\lambda zs~(xs={\bf nl}\wedgep ys=zs)~\vee\\
&\exists x'\exists xs'\exists zs'(xs=\mathbf{cns}~x'~xs'\wedgep zs=\mathbf{cns}~x'~zs'\wedgep A~xs'~ys~zs')
\end{align*}
\item Usual \red{polarization} for LP: everything is \red{positive}
  --- note, no atoms.
\item Searching for a cex is searching for a proof of a formula like
  \(\exists x\colon\tau\ [P(x) \wedgep \neg Q(x)]\)
  is a single \red{bipole} --- a positive phase followed by a negative
  one.
\item Correspond to the intuition that generation is hard, testing a deterministic computation
  \end{itemize}
\end{myslide}

%--------------------------------------
%--------------------------------------
\begin{frame}[fragile]
  \frametitle{FPC for the common man}
  \begin{itemize}
  \item We defined certificates for families of proofs (the generation
    phase) limited either by the number of inference rules that they
    contain, by their size, or by both.
\item They essentially translate into meta-interpreters that perform bounded generation, not only of terms but of derivations.
\item As a proof of concept, we implement this in $\lambda$Prolog and
  we use \NF to implement negation --- it's a shortcut, but theoretically, think fixed point and negation as $A\rightarrow\perp$.

\item We use the two-level approach: OL specs are encoded as
  \texttt{prog} clauses and a \texttt{check} predicates will
  meta-interpret them using the size/height certificates to guide the
  generation.
\item Checking $\forall x \oftp elt, \forall xs, ys\oftp eltlist\ [\mathit{rev\ xs\ ys \rightarrow xs = ys }] $ is
  \begin{small}
\begin{verbatim}
cexrev Xs Ys :- 
   check (qgen (qheight 3)) (is_eltlist Xs), % generate
   solve (rev Xs Ys), not (Xs = Ys).         % test

\end{verbatim}
  \end{small}
  \end{itemize}
\end{frame}
%--------------------------------------

%-----------------------------------

%--------------------------------------
\begin{myslide}{Future Work: more case studies}
  \begin{itemize}
  \item Search for deeper known bugs
    \begin{itemize}
    \item ``value'' restriction in ML with references and let-polymorphous
    \item intersection types with computational effects
    \end{itemize}
  \item Search for unknown bugs in ($\lambda$)Prolog code ``in the
    wild'' (e.g.\ Hannan's ``Extended natural semantics'' or even old
    CENTAUR stuff)
  \item Tackle \red{coinductive} specs, to look for
    \begin{itemize}
    \item Two process that are similar but \textbf{not} bisimilar
    \item $\lambda$-terms that are ground- but \textbf{not} applicative-bisimilar\dots
    \end{itemize}
   \red{Tabling}  could prove handy.
  \item Implement \red{random} generators e.g.\ with an
    \texttt{unfold} expert that may flip a coin when selecting a
    clause to backchain on.
  \end{itemize}
\end{myslide}

%--------------------------------------
\begin{myslide}{Future Work: architecture}
  \begin{itemize}
  \item Integrate with Abella's workflow, both at the top-level
    (disproving conjectures) and inside a proof attempt (disproving
    subgoals).

  \item Long-ish time view: a mini Sledgehammer protocol for Abella,
    by which conjectures are under the hood PB-tested: if no cex
    reported \red{proof outlines} are used to try and conclude the
    proof.

    \item Keeping in mind that Abella's implementation
      \textbf{not} immediately meant for search
    \item 
  Previous attempts with FPC
      kernels with primitive $\nabla$ written as inductive definition in
      Abella proper seems too slow for generation
  \end{itemize}
\end{myslide}

%--------------------------------------
\begin{frame}
  \frametitle{The blame game}
  \begin{itemize}
  \item Suppose your PBT tool reports a cex. Now what? You're not
    getting payed just for finding faults\dots
  \item Staring at a potentially huge spec even with a cex in hand not
    the best way to go. Two issues:
    \begin{enumerate}
    \item Soundness: your spec is plain wrong and returns an answer that should not hold
    \item Completeness: you've forgotten to encode some info and some answers are not produced.
    \end{enumerate}
  \item FPC to the rescue (possibly):
    \begin{enumerate}
    \item Use certificate \red{distillation} to restrict to a more manageable set of suspects
    \item Use \red{abductive} experts to collect sets of assumptions that should hold but don't
    \end{enumerate}
  \end{itemize}
\end{frame}
%--------------------------------------
\begin{frame}
  \begin{center}
    \begin{LARGE}
      Thanks!
    \end{LARGE}
  \end{center}

\end{frame}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

%  LocalWords:  Blanco LFMTP Twelf Coq Abella ehm PBT QuickCheck ICFP Momigliano PBT's wrt coinductively combinator al QuickChick's LazySmallCheck Renaud SmallCheck pred DM Heyting reified Ctx qheight det combinatorial fuzzying subgoals OL
%  LocalWords:  Claessen Haskell QuickCheck's FsCheck pre QC's Quis
%  LocalWords:  custodiet ipsos custodes SUT Coq's cex shrinkers Agda
%  LocalWords:  subgoal PVS QuickChick PLT Redex Spoofax backchaining
%  LocalWords:  llcl ilist hd tl instantiations intensional MMT xs ys
%  LocalWords:  Baelde zs nl cns bipole FPC Foundational clllll arg
%  LocalWords:  pos eval lookup reachability bisimulation NEs neq bg
%  LocalWords:  bisimilarity disunification clllllll prog TAPL LF et
%  LocalWords:  documemtation POPLMark Findler atomatically DFS nabla
%  LocalWords:  datatypes Subsumption equivariant Prolog Hannah's ish
%  LocalWords:  coinductive bisimilar applicative Bedwyr backchain FP
%  LocalWords:  abductive elt eltlist Hannan's PPDP focusable grnd oo
%  LocalWords:  ing Erlang corecursion Clarks multiplicatively interp
%  LocalWords:  coPBT CBV Grall Filinski Ancona PCF surjective Tarau
%  LocalWords:  FPC's istm Harrop NAS fav Chihani Peano equivariance
%  LocalWords:  CWA

### 2020-10-16 papers/coLP/long-related.tex
Added another paper.

\paragraph{Haskell}

QuickCheck can partially deal with infinite objects, provided that the properties themselves are computable. Lazy evaluation is no magic bullets: stream equality loops, but we can use the \emph{take} lemma to observe finite portions.

\paragraph{coLP}

Gupta and his students have been advocating (very, very incrementally)
for extending LP first only coinduction, then allowing both least and
greatest semantics~\cite{Luke07}. The approach is semantically based:
you start with the greatest complete Herbrand model, containing finite
and infinite (both rational and irrational) ground terms. Then they
modify the operational semantics dropping the occur check and relying
on cycle (same atom) and loop detection (same atom modulo
unification). They show soundness of the procedure wrt the greatest
complete Herbrand model. Completeness seems hopeless for
recursion-theoretic reasons as shown in~\cite{AnconaD15}.

\paragraph{Katya}
This has been much improved by Komendantskaya and friends (see~\cite{BasoldKL19}) \dots

Rather than relying on infinite terms via rational unification, they
introduce in the language of the logic a fixed point constructor
(which is then restricted to occur guarded, so as to be normalizing)
and its equational theory. While this may make sense for programming,
it seems problematic for a logical framework. Do we still have
adequacy of the encodings and canonical forms? Do \texttt{fix z.~cons
  0 z} and \texttt{fix z.~cons 0 (cons 0 z)} denote the same infinite
term or now the encodings refers to equivalence classes (perhaps wrt
bi-similarity) ? It would be hard to give a Curry-Howard reading of
this, since fixed points terms do not have a type classifying them and
the later modality does not have a proof term. Still, it could be
straighten up following~\cite{abelVezzosi:aplas14}. However, the
approach seems limited to guarded coinduction and may suffer from the
very problems raised by Ancona et al., namely  that a greatest fixed point semantics plus infinite (although) guarded terms gives unwanted results. For example, I think that  this is
still provable: \texttt{last (fix ones.~cons 1 ones) 2}. Finally, it's hard to implement in our setup, since we have to change the unification algorithm to take into account unfolding of fixed points (it's equi-recursive, in Peirce's terminology).

\paragraph{Terms Generation}

There's been considerable work on generation (mostly random) of
lambda-terms, with an emphasis on uniform
distributions~\cite{BendkowskiGT17}. Tarau generalizes it to arbitrary
first-order signatures~\cite{Tarau18} extending Remy's algorithm for
generating random binary trees \dots

\paragraph{Coaxioms}

In a lengthy list of papers such as ~\cite{AnconaDZ17, Dagnino19},
Ancona at al.\ argue against the use of gfixpoints as the semantics of
Aczel's style inference system. Starting from the desire to improve on
Leroy's coinductive operational semantics given its shortcomings, as
well as those of co-logic programming, they try to carve out undesired
results: take for example the logic program for last element of a list
and an infinite list of ones \texttt{Os = [1|Os]}. Then by a circular
proof you can prove \texttt{last Os 2}. We would argue that (1) we do
not care about infinite objects (2) in a guarded sense, the definition
of \texttt{last} is \emph{not productive} and it is rejected by Coq
(as a \textbf{Fixpoint} (but that proof holds with a
\textbf{Coinductive}) and by Agda's copatterns. They instead use
coaxioms (and then corules) to rule out those derivations. Those are
standard rules, but with a different interpretation: an atom is
provable if it has an infinite (even irrational ) proof, but whose
nodes have all a finite proof from the coaxioms only. Semantically
this corresponds to a fixed point that lies somewhere between least
and greatest. In the \texttt{last} example, the coaxiom would state
\texttt{last [X|XS] X}, which should force the element to be in the
list and rule out the offending derivation.  To me this notion of a
proof does not seem effective (how do I check the coaxiom condition
finitely?) and the coaxioms do not seem to have a declarative reading,
or, less charitably, that reading makes no sense. Still, they may have
more details in later papers. Essentially, it's an answer to a problem
I do not much care for.

\paragraph{Abella's current setup}

I think we all agree that infinite terms are an hassle, better to stay
clear from. Still, Abella's approach makes me worry: first, there is
no formal connection between the guarded version implemented and the
fixed point rules of the meta-logic, in particular compared to the strict correspondence exhibited by Beluga's script language
\emph{Harpoon}.  This can be fixed, I assume. While we cannot have a
``paradox'' such as the \texttt{last} example just because we have no
way to construct such a term, there are others mentioned by Ancona
(graph reachability) that are raised by finite terms. Moreover,
examples in the library such as \texttt{colist.thm} are bizarre,
since we prove properties of objects that we cannot construct. So, we fail to prove that
stream equality entails free equality,  but we cannot exhibit two streams that
refute the property.
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "outline"
%%% End:

%  LocalWords:  Tarau Remy's Coaxioms Ancona gfixpoints Aczel's coLP
%  LocalWords:  coinductive Fixpoint Agda's coaxioms corules coaxiom
%  LocalWords:  Herbrand Katya Komendantskaya QuickCheck

### 2020-10-16 papers/coLP/outline.tex
Added another paper.

\RequirePackage[utf8]{inputenc}
\RequirePackage[T1]{fontenc}
\documentclass[a4paper]{easychair}
\usepackage{url,xcolor}
\usepackage{proof}
\usepackage{listings}
\usepackage{letltxmacro}

\newcommand{\instan}[1]{\hbox{\sl grnd}~(#1)}
\newcommand{\Pscr}{{\mathcal P}}

\newcommand{\ok}{\checkmark}
% DM The use of pifont and txfonts broke typesetting for me: equality
% signs never appearred and some parentheses did not appear either.
%\newcommand{\noc}{\ding{56}}
\newcommand{\noc}{\ddag}
\newcommand{\lP}{$\lambda$Prolog\xspace}
\newcommand{\nat }{\hbox{\sl nat}\xspace}
\newcommand{\plus}{\hbox{\sl app}\xspace}
\newcommand{\lst}{\hbox{\sl lst}\xspace}
\newcommand{\sort}{\hbox{\sl sort}}
\newcommand{\rev}{\hbox{\sl rev}}
\newcommand{\ordered}{\hbox{\sl ordered}}

%\newcommand{\atac}{A2Tac\xspace}
\newcommand{\atac}{ACheck\xspace}
%\newcommand{\lra}{\mathrel{\longrightarrow}}
\newcommand{\lra}{\mathrel{\vdash}}
\newcommand{\seq}[2]{#1\lra #2}

\newcommand{\true}{tt}

%%%%%%%%%%%%%%%% LJF
\newcommand{\truen}{t^-}
\newcommand{\truep}{t^+}
\newcommand{\falsen}{f^-}
\newcommand{\falsep}{f^+}
\newcommand{\wedgep}{\wedge^{\!+}}
\newcommand{\wedgen}{\wedge^{\!-}}
\newcommand{\veep}{\vee^{\!+}}
\newcommand{\veen}{\vee^{\!-}}
\newcommand{\with}{\&}

\newcommand{\Nscr}{{\cal N}}
\newcommand{\Rscr}{{\cal R}}                                   % Used for an ambiguous rhs
%\newcommand{\Rscr}{\Delta_1\Downarrow\Delta_2}                % Used for an ambiguous rhs
\newcommand{\jUnf    }[4]{#1\mathbin\Uparrow#2\vdash#3\mathbin\Uparrow #4} % unfocused sequent
\newcommand{\jUnfG   }[2]{\jUnf{\Gamma}{#1}{#2}{{}}}           % unf sequ with \Gamma
\newcommand{\jUnfamb }[3]{#1\mathbin\Uparrow#2\vdash#3 \Rscr}  % unfocused sequent
\newcommand{\jUnfGamb}[1]{\Gamma\mathbin\Uparrow#1\vdash \Rscr}% unf sequ with \Gamma
\newcommand{\jLf     }[3]{#1\Downarrow#2\vdash#3}              % left focused sequent
\newcommand{\jLfG    }[1]{\jLf{\Gamma}{#1}{E}}                 % left foc seq with \Gamma 
\newcommand{\jRf     }[2]{#1\vdash #2\Downarrow}               % right focused sequent
\newcommand{\jRfG    }[1]{\jRf{\Gamma}{#1}}                    % right foc seq with \Gamma
%%%%%%%%%%%%%%%% 
\newcommand{\bxi}[1]{\blue{\Xi_{#1}}}
\newcommand{\bXi}[1]{\blue{\Xi_{#1} :\null}}

\newcommand{\andClerk}[3]{{\wedge_c}(#1,#2,#3)}
\newcommand{\falseClerk}[2]{f_c(#1,#2)}
\newcommand{\orClerk}[2]{{\vee_c}(#1,#2)}
\newcommand{\allClerk}[2]{\forall_c(#1,#2)}
\newcommand{\storeClerk}[3]{\hbox{\sl store}_c(#1,#2,#3)}

\newcommand{\trueExpert }[1]{{\true_e}(#1)}
\newcommand{\eqExpert }[1]{{=_e}(#1)}
\newcommand{\unfoldExpert}[2]{{\hbox{\sl unfold}_e}(#1,#2)}
\newcommand{\andExpert}[3]{{\wedge_e}(#1,#2,#3)}
\newcommand{\andExpertLJF}[6]{{\wedge_e}(#1,#2,#3,#4,#5,#6)}
\newcommand{\orExpert  }[3]{{\vee_e}(#1,#2,#3)}
\newcommand{\someExpert}[3]{\exists_e(#1,#2,#3)}
\newcommand{\initExpert}[2]{\hbox{\sl init}_e(#1,#2)}
\newcommand{\cutExpert}[4]{\hbox{\sl cut}_e(#1,#2,#3,#4)}
\newcommand{\decideExpert}[3]{\hbox{\sl decide}_e(#1,#2,#3)}
\newcommand{\releaseExpert}[2]{\hbox{\sl release}_e(#1,#2)}
%%%%%%%%%%%%%%%%

%
\newcommand{\step}{\longrightarrow}
\newcommand{\sstlc}{\textit{STLC}\xspace}
\newcommand{\vds}{\vdash_\Sigma}

\def\bnfas{\mathrel{::=}}
\def\bnfalt{\mid}

\def\lam{\lambda}
\def\Lam{\Lambda}
\def\arrow{\rightarrow}
\def\oftp{\mathord{:}}
\def\hastype{\mathrel{:}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Editorials
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\long\def\ednote#1{\footnote{[{\it #1\/}]}\message{ednote!}}
%\long\def\ednote#1{\begin{quote}[{\it #1\/}]\end{quote}\message{note!}}
\newenvironment{metanote}{\begin{quote}\message{note!}[\begingroup\it}%
                         {\endgroup]\end{quote}}
\long\def\ignore#1{}

\newcommand{\todo}[1]{\begin{metanote}TODO: #1\end{metanote}}
%
%\setlength{\textwidth}{13.2cm}
%\setlength{\textheight}{19.9cm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{types20/listing-macros}  % Use to get better highlighing of code
\lstset{language=lprolog}
\lstset{language=abella}
\title{On the Proof Theory of Property-Based Testing of Coinductive Specifications, or:\\
 PBT to Infinity and beyond}
\authorrunning{Blanco, Miller and Momigliano}
\titlerunning{PBT to Infinity and
   beyond}
\author{Roberto Blanco\inst{1} \and Dale Miller\inst{2} \and Alberto Momigliano\inst{3}}

\institute{
  INRIA Paris, France
  \and
  INRIA Saclay \& LIX, \'Ecole Polytechnique, France\\
\and
DI, Universit\`a degli Studi di Milano, Italy
}
\begin{document}

\maketitle

\section{Introduction}
\label{sec:intro}

\begin{center}
  [For the moment taken from the TYPES abstract, to be revised]
\end{center}

Reasoning about infinite computations via coinduction and corecursion
has an ever increasing relevance in formal methods and, in particular,
in the semantics of programming languages, starting
from~\cite{milner91tcs}; see also~\cite{2007-Leroy-Grall} for a
compelling example --- and, of course, coinduction underlies (the
meta-theory of) process calculi. This was acknowledged by researchers
in proof assistants, who promptly provided support for coinduction and
corecursion from the early 90's on, see~\cite{Paulson97,Gim95types}
for the beginning of the story concerning the most popular frameworks.

It also became apparent that tools that search for
refutations/counter-examples of conjectures prior to attempting a
formal proof are invaluable: this is particularly true in PL theory,
where proofs tend to be shallow but may have hundreds of cases.  One
such approach is \emph{property-based testing} (PBT), which employs
automatic test data generation to try and refute executable
specifications.  Pioneered by \emph{QuickCheck} for functional
programming~\cite{claessen00icfp}, it has now spread to most major
proof assistants~\cite{BlanchetteBN11,QChick}.

% DM I revised the following paragraph.  Did I get the sense right?

In general, PBT does not extend well to coinductive specifications
(an exception being Isabelle's \emph{Nitpick},
% \url{https://isabelle.in.tum.de/dist/doc/nitpick.pdf}
which is, however, a counter-model generator).  A particular
challenge, for example, for \emph{QuickChick} is extending it to work
with Coq's notion of coinductive via \emph{guarded} recursion (which
is generally seen to be an unsatisfactory approach to coinduction). We
are not aware of applications of PBT to other form of coinduction, such as \emph{co-patterns}~\cite{AbelPTS13}.

While PBT originated in the functional programming community, we have
given in a previous paper (\cite{Blanco0M19}) a reconstruction of some
of its features (operational semantics, different flavors of
generation, shrinking) in purely proof-theoretic terms employing the
framework of \emph{Foundational Proof Certificates}~\cite{chihani17jar}: the
latter, in its full generality, defines a range of proof structures
used in various theorem provers such as resolution refutations,
Herbrand disjuncts, tableaux, etc.
%
In the context of PBT, the proof theory setup is much simpler.

\begin{center}
  [etc\dots]
\end{center}

The contributions of this paper are
\begin{itemize}
\item Tighten the link of our previous reconstruction of PBT to the
  proof-theoretic approach of model checking in~\cite{heath19jar} by casting it in the setting of $\mu$MALL
\item Extend PBT beyond the Horn-nabla fragment to Harrop spec, first using only case analysis, then also co-induction
\item Experiment with a multi-level architecture, where FPC only takes
  care of generation and testing  is delegated to a system such as Bedwyr
\item Hopefully provide some examples (CCS, $\pi$ calculus (?), co-evaluation)
\item
  \begin{center}
[others \dots]    
  \end{center}

\end{itemize}

\section{The story so far}
\label{sec:tecap}

\begin{center}
  [Here we recap what we did in PPDP, in the uniform proof style, see fig~\ref{fig:augmented}: the connection with linear logic comes in the next section]
\end{center}

Consider an attempt to find counter-examples to a conjecture of the
form \(\forall x [(\tau(x)\wedge P(x)) \supset Q(x)]\) where $\tau$ is
a typing predicate and $P$ and $Q$ are two other predicates defined
using Horn clause specifications.
%
By negating this conjecture, we attempt to find a (focused) proof of 
\(\exists x [(\tau(x)\land P(x)) \land \neg Q(x)]\).
%
In the focused proof setting, the \emph{positive phase} (where
test cases are generated) is represented by \(\exists x\) and
\((\tau(x)\land P(x))\). 
%
That phase is followed by the \emph{negative phase} (where conjectured
counter-examples are tested) and is represented by \(\neg Q(x)\).
%
%% turns out to be a search for a proof of \(\exists x [(\tau(x)\land
%%   P(x)) \land \neg Q(x)]\) in a focused sequent calculus for
%% (basically) Horn logic, and pre-condition phase, while proving $Q$ can
%% be relegated to deterministic logic programming-like computation (the
%% \emph{negative} phase), interpreting ``$\neg$'' as
%% Negation-as-failure..
%
%

\begin{figure}
\newcommand{\XXi}{{\color{blue}{\Xi}}}
\[
\infer{\XXi\vdash G_1\wedge G_2}
      {\XXi_1\vdash G_1\qquad \XXi_2\vdash G_2\qquad \andExpert{\XXi}{\XXi_1}{\XXi_2}}
\qquad
\infer{\XXi\vdash \true}
      {\trueExpert{\XXi}}
\]
\vskip -6pt
\[
\infer{\XXi\vdash G_1\vee G_2}
      {\XXi'\vdash G_i\qquad \orExpert{\XXi}{\XXi'}{i}}
\qquad
\infer{\XXi\vdash \exists x. G}
      {\XXi'\vdash G[t/x]\qquad \someExpert{\XXi}{\XXi'}{t}}
\]
\vskip -6pt
\[
\infer{\XXi\vdash t = t}
      {\eqExpert{\XXi}}
\qquad
\infer{\XXi\vdash A}
      {\XXi'\vdash G \quad (A~\hbox{\tt :-}~G)\in\instan\Pscr
                     \quad \unfoldExpert{\XXi}{\XXi'}}
\]
\caption{A proof system augmented with proof certificates and
  additional ``expert'' premises.}
\label{fig:augmented}
\end{figure}

\section{Linear logic as the proof theory of model checking}
\label{sec:ll}

A natural choice is linear logic cousin $\mu\mathrm{MALL}$~\cite{Baelde12}, which is associated to  the \emph{Bedwyr} model-checker. %  In fact, the latter
% has already been used for related aims~\cite{HeathM15}.

\subsection{Horn programs as fixed points}
\label{ssec:lpfixed}

\begin{center}
  [Here we introduce fixed points, and re-state the rules in fig~\ref{fig:augmented} with only unfolding rules. We may also want to introduce our first coinductive examples, probably \texttt{coeval} (it's Horn). 
  ]
\end{center}

\subsection{Onwards to linear logic}
\label{ssec:ll}

\begin{center}
  [
  \begin{itemize}
  \item Here we introduce polarization, focusing. I assume we will be
    using only $\mu$F rather than all muMALL; and if so, perhaps we
    can simplify the syntax of the focusing judgment
  \item Not sure we want to put clercks and experts everywhere, if they play a part only during generation and that should stay in the Horn+nabla fragment
  \item Now we can introduce non Horn examples such as simulation and points out that the left unfolding rules do case analysis.
\end{itemize}
  ]
\end{center}

\subsection{Full system}
\label{ssec:fullmumall}

\begin{center}
  [Here we add the rules for (co)induction, init (and cut?). In the end we take the muFlogic from the JAR paper and the ``completeness'' results thereof]
\end{center}

\section{Implementation and case studies}
\label{sec:case}

\begin{center}
  [Here we discuss the ``architecture'', namely using the lightweight FPC for generators and Bedwyr for testing]
\end{center}  
\subsection{Examples}
\label{ssec:exs}

\begin{center}
  [Which esamples? The ones which work]
\end{center}

\subsection{Refinements}
\label{ssec:refs}

\begin{center}
  [Anything else to make this interesting: integration with
  proof-outlines, pre-computation of terms 'a la Tarau, full
  ``post-mortem'' FPC, that is also for the testing phase, to
  elaborate further the results, for examples for explanations etc., wild ideas about generating infinite terms with observation functions as in \emph{Hipster} \cite{EinarsdottirJP18}]
\end{center}

\section{Related work}
\label{sec:rel}

\begin{center}
  [to much details, just for my own sake, only a small portion will make it in the paper]
\end{center}
  While coinductive logic programming,
see~\cite{Luke07} and~\cite{BasoldKL19} for a much more principled and
in depth treatment, may at first seem to fit the bill, the need to
model infinite \emph{behavior} rather than infinite objects, that is (ir)rational terms on the
domain of discourse, has lead us to adopt a much stronger logic (and
associated proof theory) with explicit rules for induction and
coinduction.
\input{long-related}
\section{Conclusion}
\label{sec:conc}

\bibliographystyle{abbrv}
\bibliography{colp,../ppdp19/l}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

%  LocalWords:  Coinductive QuickCheck utf inputenc fontenc Herbrand
%  LocalWords:  corecursion coinductive corecursive QuickChick CBV ir
%  LocalWords:  disjuncts PCFL LTS Harrop PBT Coq's coinductively CCS
%  LocalWords:  intuitionistic FPCs Bedwyr Bedwyr's nabla

### 2020-10-16 papers/coLP/summary.tex
Added another paper.

% Submissions must not exceed 12 pages ACM style 2-column (including
% figures, but excluding bibliography).

%\documentclass[preprint]{sigplanconf}
\documentclass[sigconf]{acmart}

%\settopmatter{printfolios=false,printccs=false,printacmref}
\settopmatter{printfolios=false,printccs=true,printacmref}

\usepackage{xspace}
% The following two fonts broke the typesetting of some math environments.
%\usepackage{txfonts}
%\usepackage{pifont}
\usepackage{amssymb,amsmath}
\usepackage{proof}
%\usepackage{url,breakurl}
%\usepackage{booktabs}
%\usepackage{graphics} % use for the Pitts-Gabbay quantifier

\usepackage{listings}
\usepackage{letltxmacro}
\input{../ppdp19/listing-macros}  % Use to get better highlighing of code
\lstset{language=lprolog}
\lstset{language=abella}

% \newcommand{\blue}[1]{{\color[rgb]{0,0,1} #1}}
% \newcommand{\tup}[1]{\langle #1\rangle}
% \newcommand{\tupp}[2]{\blue{\langle #1,}#2{\blue{\rangle}}}

\newcommand{\instan}[1]{\hbox{\sl grnd}~(#1)}
\newcommand{\Pscr}{{\mathcal P}}

\newcommand{\ok}{\checkmark}
% DM The use of pifont and txfonts broke typesetting for me: equality
% signs never appearred and some parentheses did not appear either.
%\newcommand{\noc}{\ding{56}}
\newcommand{\noc}{\ddag}
\newcommand{\lP}{$\lambda$Prolog\xspace}
\newcommand{\nat }{\hbox{\sl nat}\xspace}
\newcommand{\plus}{\hbox{\sl app}\xspace}
\newcommand{\lst}{\hbox{\sl lst}\xspace}
\newcommand{\sort}{\hbox{\sl sort}}
\newcommand{\rev}{\hbox{\sl rev}}
\newcommand{\ordered}{\hbox{\sl ordered}}

%\newcommand{\atac}{A2Tac\xspace}
\newcommand{\atac}{ACheck\xspace}
%\newcommand{\lra}{\mathrel{\longrightarrow}}
\newcommand{\lra}{\mathrel{\vdash}}
\newcommand{\seq}[2]{#1\lra #2}

\newcommand{\true}{tt}

%%%%%%%%%%%%%%%% LJF
\newcommand{\truen}{t^-}
\newcommand{\truep}{t^+}
\newcommand{\falsen}{f^-}
\newcommand{\falsep}{f^+}
\newcommand{\wedgep}{\wedge^{\!+}}
\newcommand{\wedgen}{\wedge^{\!-}}
\newcommand{\veep}{\vee^{\!+}}
\newcommand{\veen}{\vee^{\!-}}
\newcommand{\with}{\&}

\newcommand{\Nscr}{{\cal N}}
\newcommand{\Rscr}{{\cal R}}                                   % Used for an ambiguous rhs
%\newcommand{\Rscr}{\Delta_1\Downarrow\Delta_2}                % Used for an ambiguous rhs
\newcommand{\jUnf    }[4]{#1\mathbin\Uparrow#2\vdash#3\mathbin\Uparrow #4} % unfocused sequent
\newcommand{\jUnfG   }[2]{\jUnf{\Gamma}{#1}{#2}{{}}}           % unf sequ with \Gamma
\newcommand{\jUnfamb }[3]{#1\mathbin\Uparrow#2\vdash#3 \Rscr}  % unfocused sequent
\newcommand{\jUnfGamb}[1]{\Gamma\mathbin\Uparrow#1\vdash \Rscr}% unf sequ with \Gamma
\newcommand{\jLf     }[3]{#1\Downarrow#2\vdash#3}              % left focused sequent
\newcommand{\jLfG    }[1]{\jLf{\Gamma}{#1}{E}}                 % left foc seq with \Gamma 
\newcommand{\jRf     }[2]{#1\vdash #2\Downarrow}               % right focused sequent
\newcommand{\jRfG    }[1]{\jRf{\Gamma}{#1}}                    % right foc seq with \Gamma
%%%%%%%%%%%%%%%% 
\newcommand{\bxi}[1]{\blue{\Xi_{#1}}}
\newcommand{\bXi}[1]{\blue{\Xi_{#1} :\null}}

\newcommand{\andClerk}[3]{{\wedge_c}(#1,#2,#3)}
\newcommand{\falseClerk}[2]{f_c(#1,#2)}
\newcommand{\orClerk}[2]{{\vee_c}(#1,#2)}
\newcommand{\allClerk}[2]{\forall_c(#1,#2)}
\newcommand{\storeClerk}[3]{\hbox{\sl store}_c(#1,#2,#3)}

\newcommand{\trueExpert }[1]{{\true_e}(#1)}
\newcommand{\eqExpert }[1]{{=_e}(#1)}
\newcommand{\unfoldExpert}[2]{{\hbox{\sl unfold}_e}(#1,#2)}
\newcommand{\andExpert}[3]{{\wedge_e}(#1,#2,#3)}
\newcommand{\andExpertLJF}[6]{{\wedge_e}(#1,#2,#3,#4,#5,#6)}
\newcommand{\orExpert  }[3]{{\vee_e}(#1,#2,#3)}
\newcommand{\someExpert}[3]{\exists_e(#1,#2,#3)}
\newcommand{\initExpert}[2]{\hbox{\sl init}_e(#1,#2)}
\newcommand{\cutExpert}[4]{\hbox{\sl cut}_e(#1,#2,#3,#4)}
\newcommand{\decideExpert}[3]{\hbox{\sl decide}_e(#1,#2,#3)}
\newcommand{\releaseExpert}[2]{\hbox{\sl release}_e(#1,#2)}
%%%%%%%%%%%%%%%%

%
\newcommand{\step}{\longrightarrow}
\newcommand{\sstlc}{\textit{STLC}\xspace}
\newcommand{\vds}{\vdash_\Sigma}

\def\bnfas{\mathrel{::=}}
\def\bnfalt{\mid}

\def\lam{\lambda}
\def\Lam{\Lambda}
\def\arrow{\rightarrow}
\def\oftp{\mathord{:}}
\def\hastype{\mathrel{:}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Editorials
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\long\def\ednote#1{\footnote{[{\it #1\/}]}\message{ednote!}}
%\long\def\ednote#1{\begin{quote}[{\it #1\/}]\end{quote}\message{note!}}
\newenvironment{metanote}{\begin{quote}\message{note!}[\begingroup\it}%
                         {\endgroup]\end{quote}}
\long\def\ignore#1{}

\newcommand{\todo}[1]{\begin{metanote}TODO: #1\end{metanote}}
%
%\setlength{\textwidth}{13.2cm}
%\setlength{\textheight}{19.9cm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
%\conferenceinfo{}{}
%\CopyrightYear{}
%\copyrightdata{}
%\title{Property-Based Testing via Proof Reconstruction: Work-in-progress}

%\acmConference[PPDP'2019]{21th International Symposium on Principles and Practice of Declarative Programming}{October 7--9 }{Porto, Portugal}

\title{Notes on coinductive PBT}
\author{Roberto Blanco}
\affiliation{\institution{INRIA Paris, France}}
\email{roberto.blanco@inria.fr}
\author{Dale Miller}
\affiliation{\institution{INRIA Saclay \& LIX, \'Ecole Polytechnique, France}}
\email{dale.miller@inria.fr}
\author{Alberto Momigliano}
\affiliation{\institution{DI, Universit\`a degli Studi di Milano, Italy}}
\email{momigliano@di.unimi.it}

\begin{abstract}
  The \emph{linear } logic behind Bedwyr (``muFa'', according to
  Quentin) seems to be the proof-theory of coinductive PBT. Applications are mostly in separating
  equivalences in process/lambda calculi.
\end{abstract}
\maketitle

\section{The proposal}
\label{sec:p}

Our starting point is to see whether the FPC approach in the previous
paper extends to richer set of specifications. Originally, we wanted
to prove formulas of the form
%
\(\exists x [(\tau(x)\land P(x)) \land \neg Q(x)]\)
%
whether $P,Q$ were essentially Horn logic programs plus nabla. We
argued that the logic behind it was basically minimal logic, since we
only provided certificates for the positive phase, while the
negative/testing one was just exhaustive search.

Now, we want to PBT specs over \emph{infinite behavior} (more than infinite
objects). Here's an example: it is well known that in CCS there are processes that are similar to each other but not bisimilar,
for example \texttt{a.b + a} and \texttt{a.b}:

\begin{lstlisting}
Theorem sim_not_bisim: exists P Q,
 sim P Q /\ sim Q P /\ (bisim P Q -> false).
\end{lstlisting}
My Abella proof uses reflexivity of similarity in the positive phase,
but there is also a direct coinductive proof with the following
simulation:
\begin{enumerate}
\item $R_1 = \{\mathtt{(a.b+a, a.b ),(b,b),(0,b),(0,0)}\}$
  \item $R_2 = \{\mathtt{(a.b,a.b+a),(b,b), (0,0)}\}$ 
\end{enumerate}
%

More examples: in \cite{2007-Leroy-Grall}, it is suggested to take the usual rules for CBV in the lambda-calculus with constants, but \emph{coinductively}:
\begin{lstlisting}
CoDefine coeval: tm -> tm -> prop by
 coeval (con C) (con C);
 coeval (fun R) (fun R);
 coeval (app M N) V := exists R W, coeval M (fun R) /\ coeval N W /\ coeval (R W) V.
\end{lstlisting}

Whether this notion of co-evaluation makes sense is debatable, but the
following holds: it is not deterministic anymore, since a divergent
term such as $\Omega$ co-evaluates to anything
\begin{lstlisting}
Theorem fail_det: exists E V1 V2, coeval E V1
/\ coeval E V2 /\ (V1 = V2 -> false).
\end{lstlisting}
The proof uses coinduction twice with the trivial simulation
$\Omega,0$, and $\Omega, 1$ say.
%

Consider now \emph{ground} vs.\ \emph{applicative} similarity in
PCFL. The latter is finer than the former:

\begin{lstlisting}
CoDefine sim : tm -> tm -> ty -> prop by
sim  M1 M2 (arr S S') :=
 (forall M, {eval M1 (abs M)} -> 
 exists M' , {eval M2 (abs M')} /\ 
 (forall x, {of x S} -> sim  (M x) (M' x) S'));  ...
CoDefine gsim : tm -> tm -> ty -> prop by
 gsim  M1 M2 (arr S S') :=
 (forall x, {of x S} -> gsim  (app M1 x) (app M2 x) S');  ...
Theorem cex_asim_gsim: exists M1 M2 T, gsim M1 M2 T /\ (sim M1 M2 T -> false).
\end{lstlisting}
Two such separating terms are: $\lambda x.\ \bot$ and $\bot$, which
are not applicative similar since $\bot$ does not converge to a
lambda. The proof uses several lemmas such as transitivity of
\texttt{asim}, Kleene order being included in similarity, $\bot$ being
the least element in the similarity order etc.

Our angle: Can PBT find those processes/lambda terms?
All of this begs for a richer logic:
\begin{itemize}
\item   \texttt{(bi)sim} are not Horn clauses
\item they require some form of coinduction for proof search to succeed (and case analysis for the negative phase)
\end{itemize}

The combination of those two points rules out falling back on
coinductive lp as in
\url{https://www.swi-prolog.org/pldoc/doc/_SWI_/library/coinduction.pl},
based on \cite{Luke07}. Even $\lambda$Prolog seems out as we need to
do case analysis on eigenvariables and more crucially we cannot
simulate nabla with pi for non-Horn specs such as similarity (of
course we would need tables, but that's doable). We need a logic with
(co)induction and nabla. Sounds familiar?

\subsection{Shall we use Bedwyr as is?}
\label{sec:bed}

That's the first instinct, since it has automation for search and case analysis, the correct distinction
between eigenvariables and scoped constants, tabling and even tabling modulo lemmas (up to version 1.3). And in fact,  we
can investigate co-evaluation immediately, with hand-made height generators:
\begin{lstlisting}
 Define coinductive  coeval : tm -> tm -> prop by ...
 Define wt : num -> env -> tm -> prop by ... % generator
 ?= wt (s (s (s z))) void M /\ wt z void M1 /\ wt z void M2 /\ coeval M M1 /\ coeval M M2 /\ (M1 = M2 -> false))

 Found a solution:  2195msa
 M2 = con one
 M1 = con zero
 M  = app (fun (x1\ app x1 x1)) (fun (x1\ app x1 x1))
\end{lstlisting}
Note that we need generators for \texttt{M1,M2} since standard mode
analysis guaranteeing groundness of the output variable does not
immediately apply to coinductive LP.

For the sim/not bisim example, checking \texttt{bisim P Q -> false}
runs afoul of the limitations of the left proved that does not handle
universal and implications.  However, as noted in \cite{HeathM15}, the
logic behind Bedwyr is essentially linear and therefore we can rely on
negation normal forms and simply push negation inside to define
\emph{non}-bisimulation, yielding an interesting polarization
structure:

\begin{lstlisting}
Define inductive n_bisim : proc -> proc -> prop by
 n_bisim P Q :=
 (exists A P1, step P A P1 /\ forall Q1, step Q A Q1 -> n_bisim P1 Q1) \/
 (exists A Q1, step Q A Q1 /\ forall P1, step P A P1 -> n_bisim Q1 P1).

 ?= N = (s (s z)) /\ is_proc N P /\ is_proc N Q /\ sim P Q /\ sim Q P /\ n_bisim P Q.
Found a solution: 301ms
 Q = par (out tau null) (out tau null)
 P = out tau (out tau null)
\end{lstlisting}

A similar problem: find a process that is not bisimilar to its \texttt{par}

\begin{lstlisting}
?=is_proc (s z)) P /\ n_bisim (par P P) P).
  P = out a null
\end{lstlisting}

Question:
\begin{itemize}
\item What's going on recursion-theoretically? The complement of a
  r.e.\ coinductive relation must be inductive, but what is its
  degree? Conversely, the complement of an r.e.\ inductive relation
  such as evaluation is coinductive, certainly not r.e., so what we
  would get by negation normal form would be badly incomplete.
\end{itemize}

To recover the flexibility of different generating strategy, we can
port the FPC approach of the PPDP paper to Bedwyr. Current attempt is
doing this \emph{only for generation} and rely on Bedwyr for everything
else. 
\begin{lstlisting}
?= check  (qheight 2) (isProc (p2i P)) /\ (n_bisim (par P P)  P).
 \end{lstlisting}
From first experiments this is feasible, so far and allows us to use native tabling and the
L0/L1 search strategy. It's less efficient that I was expecting: a query like the above works, but
\begin{lstlisting}
(check (qheight 4 (and (isProc (p2i P)) (isProc (p2i Q)))
	/\ sim P Q /\ sim Q P /\ (n_bisim P Q)
\end{lstlisting}
takes some 20 secs, because we are building up two processes $P$ and $Q$
independently. This is different from previous examples in PPDP, where
generation was performed over judgments such as typing (see
preservation/progress). A possibility is to generate $P$ and $Q$ together, following
ideas in \cite{TNIQ}, whereby we fist generate $P$ and then obtain $Q$ by locally mutating the former.

This architecture may be also  less flexible, since we have to
live with the limitation of the left prover. And we do not have
certificates for the rest of the query. This may be limiting if we
need to steer both the satisfaction of the premises and the testing
phase, which is not purely negative anymore.

Or we can go all the way two-level and use a kernel for muFa, pretty
much as in Quentin's code
\url{http://slimmer.gforge.inria.fr/bedwyr/pcmc/}. It's going to be
slow, especially if we use it for generation as well and it uses
standard (co)induction rules via invariants, which is not exactly
user-friendly. We could add support for tables, but we run the risk of
re-inventing the wheel (read Bedwyr). However, it is as flexible as we
program it, in the sense that we can add tabling modulo (sort of
testing ``up-to'' previously established theorems).

As a refinement of the first approach, the logic of Bedwyr not being
completely clear, one could take the cex found by Bedwyr and re-run it
in Quentin's kernel to have an honest-to-goodness proof in $\mu$mall
(although we need to turn tables into invariants).

\subsection{HM logic}
\label{sec:hml}
In \cite{HeathM15}, certificates for non (bi)similarity in LTS are assertions
in a fragment of Hennessy-Milner logic that separates two
processes. This begs the question if it makes sense to formulate such
a logic for PCFL and simulation. In other terms, show two (lazy) programs not contextually equivalent by generating an HML formula that separates them.

The presentation in \cite{Gordon95}
where small step semantics is annotated with observations (essentially
destructors) seems particularly apt and corresponds to ground
bisimilarity --- it's possible that by changing the dichotomy between
passive/active types it could be applied to the applicative case as
well.  See also\cite{LagoR15} for a similar LTS for applicative.

First question: is HML enough or do we have to go to the modal $\nu$-calculus to accommodate infinite computations? Example: \texttt{iterate f x} does not simulate
\texttt{map f (iterate (f x))}. Let \texttt{f} be the square function. The programs have transitions (using Gordon's LTS):
\begin{lstlisting}
iterate f 3 >> 
  3 :: iterate (f (f 3)) --hd->  3 --3-> stuck 
  3 :: iterate (f (f 3)) --tl->  iterate (f (f 3)) --?? ->

map f (iterate (f 3)) >>
  f 3 :: map f (iterate f (f 3)) --hd-> 9 etc
  f 3 :: map f (iterate f (f 3)) --tl->  map f (iterate f (f 3))
\end{lstlisting}
so a formula satisfying the former but not the latter should be something like
\texttt{(<hd> <3> true)}. In the first case we observe here 3 and  9 in the other. This suggests that we
can extract a separating \emph{evaluation context} from a HM formula, here essentially \texttt{hd \_}.

Some problems:
\begin{itemize}
\item Differently from LTS (and from the equational theory of the lambda
  calculus), our operational semantics is deterministic in the sense
  that while we can make different observations on the same term, we
  cannot make different transitions with the same observation. Hence
  the modalities seems to collapse and I see box as only useful for saying that no observations can be made. But are they still dual?
\item If we use Gordon's active/passive approach, we need a satisfaction clause for \texttt{Trans  Red}:
  $$
  P \models F
  \mbox{ if for type of $P$ active, when } P\Downarrow V, V \models F
  $$
\item We may have to add fixed points to account for formulae satisfying bisimilar terms such as \texttt{ones} and \texttt{ones'}
\begin{lstlisting}
zeros = 0 :: zeros
ones  = 1 :: ones
ones' = map (fun x -> x + 1) zeros 
\end{lstlisting}
\end{itemize}

Other questions
\begin{itemize}
\item HM logic is presented as parameterized by
the underlying LTS, but can we also vary the motion of bisimulation:
strong vs weak, up-to? There is some generalization of LTS into synchronization trees that may do that.

\item  What are the relations with
trace semantics and related notions such as in
\cite{2007-Leroy-Grall}?

\item In a sense to be made precise could FPC
induce a related HM logic?
\end{itemize}
In the meantime, we can use PBT HM logic to find formul\ae\ that separates
two non-similar LTS, here the example in Section 5.2 of
\cite{HeathM15}, where \texttt{hsat} is HM forcing of a modal formula
by a process:
\begin{lstlisting}
?= N = (s (s (s z))) /\ is_hform N F /\ hsat (proc 6) F  /\ (hsat (proc 1) F -> false).

F = hdiam a (hconj (hdiam b tt) (hdiam c tt))
N = s (s (s z))
\end{lstlisting}
Note that we cannot directly use box, as the left prover
complains. However, we may get around it defining non-forcing.

\subsection{Other  examples}

Because of the immense body of work in model checking (finite) process
calculi such as CCS, if we want to sell our approach, we may want as
usual to insist on domains that are hard for weaker frameworks, so
stick to binders. This means the $\pi$ and $\lambda$ calculi.

From the meta-theory of $\pi$ calculus
\begin{itemize}
\item  early and late bisim not preserved by inputs
\item  ground bisim not a congruence
\item ground  bisim not preserved by name substitution
\item model checking with the HML described in \cite{Tiu:2010} . Note that for the
  encoding of satisfaction to be adequate we need to compute the list
  of possible free new names beside those occurring in the process and
  formula. Still `` One can thus calculate the number of new names
  needed based on the number of bound input modalities in A.'' (ibidem page 21). See also  some of the examples in the Bedwyr example suite \verb+{pi/pi_modal.def}+
\end{itemize}

 More from \cite{2007-Leroy-Grall}:
\begin{itemize}
  \item There are raw terms that diverge, but not coevaluate;
\begin{lstlisting}
?= wt (s (s (s (s z)))) void M  /\ divrg M /\ (forall V, coeval M V -> false).

Found a solution: + 2011
 M = app (con zero) (app (fun (x1\ app x1 x1)) (fun (x1\ app x1 x1)))
\end{lstlisting}
  \item Let \texttt{costeps} be the kleene closure of the step relation, coinductively. Then the are terms such that
    \texttt{costeps E V} but not \texttt{coeval E V} --- the reverse of Lemma 36.
  \item type soundness fails for co-evaluation (the Filinski cex, based on the $Y$ combinator, seems hard, see Section 8.2).
  \end{itemize}

  More from the lazy lambda-calculus:
  \begin{itemize}
  \item The eta  and surjective pairing laws hold for ground similarity but not for applicative.
  \end{itemize}

  Other PL examples that we could mutate and see what happens:
\begin{itemize}

\item Milner-Tofte type preservation result for circular closures~\cite{milner91tcs}
\item Transitivity of subtyping in recursive types (seen as rational trees over simple types)
  \cite{kozen_silva_2017}
\end{itemize}

\subsection{Related work}

Some precursors to coinductive LP \cite{JaffarSV04,CharatonikMNPW98},
all that much ado about nothing following \cite{Luke07} such as
\cite{DBLP:journals/cl/Ancona13,MantadelisRM14}. Note: the latter
seems intertwined with infinite terms rather than infinite behaviors
and does not have enough abstractions to tackle the examples we care
about.

Gupta says it can prove two Buchi automata bisimlar if the accepts the
same infinite strings (probably with some trick to express universal
quantification). For non equivalence one can show that there's a
string that the other does not accept, but the status of NAF wrt coLP
is totally unsettled. Anyhow, the automata example does not work in
SWI (but it does with the meta-interpreter).

NOTE: the CWB
\url{https://homepages.inf.ed.ac.uk/perdita/cwb/summary.html} can do a
bunch of stuff such as ``derive automatically logical formulae which
distinguish nonequivalent'' processes (of course, it is finite-state
systems). Perhaps try the new versions \url{http://caal.cs.aau.dk/}
\cite{CAAL} or \url{https://sourceforge.net/projects/cwb-nc/}. Also
the usual XSB tabled approach to simulation checking~\cite{symbolic}. Thus, we need to stress the pi calculus angle (binders, infinite state)

TO READ: stuff from Katya Komendantskaya \cite{KomendantskayaL17,BasoldKL19} (probably not that relevant). Stuff about testing equivalences in LTS (Sangiogi Book 1)
\bibliographystyle{ACM-Reference-Format}
\bibliography{colp}

\end{document}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

%  LocalWords:  coinductive Blanco INRIA Saclay Ecole Polytechnique
%  LocalWords:  Universit degli Studi di Milano nabla CBV PCFL asim
%  LocalWords:  Kleene lp eigenvariables bisim Milner Tofte Katya HML
%  LocalWords:  Komendantskaya colp muFa PPDP Bedwyr formul LTS hsat
%  LocalWords:  coevaluate costeps kleene Filinski cex surjective CWB
%  LocalWords:  XSB bisimulation bisimilarity hd

### 2020-10-16 papers/coLP/types20/ccs.tex
Added another paper.

\RequirePackage[utf8]{inputenc}
\RequirePackage[T1]{fontenc}
\documentclass[a4paper]{easychair}
\usepackage{url,xcolor}
\usepackage[xcolor]{changebar}
\usepackage{listings}
\usepackage{letltxmacro}
\input{listing-macros}  % Use to get better highlighing of code
\lstset{language=lprolog}
\lstset{language=abella}

\title{Notes on PBT and CCS}
\authorrunning{Momigliano}
\titlerunning{Notes on PBT and CCS}
\author{Alberto Momigliano}

\institute{
DI, Universit\`a degli Studi di Milano, Italy
}
\begin{document}

\maketitle

While PBT originated in the functional programming community, we have
given in a previous paper (\cite{Blanco0M19}) a reconstruction of some
of its features (operational semantics, different flavors of
generation, shrinking) in purely proof-theoretic terms employing the
framework of \emph{Foundational Proof Certificates}~\cite{chihani17jar}.
%
In the context of PBT, the FPC setup is quite simple.
Consider an attempt to find counter-examples to a conjecture of the
form \(\forall x [(\tau(x)\wedge P(x)) \supset Q(x)]\) where $\tau$ is
a typing predicate and $P$ and $Q$ are two other predicates defined
using Horn clause specifications.
%
By negating this conjecture, we attempt to find a (focused\footnote{
For Horn specs, focusing is immaterial, but when we do proof search over logic with fixed points, it helps to structure the search itself. Still, the rules are heavy-handed and we can just work with base sequents for the time being}) proof of 
\(\exists x [(\tau(x)\land P(x)) \land \neg Q(x)]\).
%
In the focused proof setting, the \emph{positive phase} (where
test cases are generated) is represented by \(\exists x\) and
\((\tau(x)\land P(x))\). 
%
That phase is followed by the \emph{negative phase} (where conjectured
counter-examples are tested) and is represented by \(\neg Q(x)\).
%
%% turns out to be a search for a proof of \(\exists x [(\tau(x)\land
%%   P(x)) \land \neg Q(x)]\) in a focused sequent calculus for
%% (basically) Horn logic, and pre-condition phase, while proving $Q$ can
%% be relegated to deterministic logic programming-like computation (the
%% \emph{negative} phase), interpreting ``$\neg$'' as
%% Negation-as-failure..
%

As explained in~\cite{Blanco0M19}, the standard PBT setup needs little
more than Horn logic. % specifications.
%
However, when addressing infinite computations, we need richer
specifications.  While coinductive logic programming,
see~\cite{Luke07} may at first seem to fit the bill, the need to model
infinite \emph{behavior} rather than infinite objects\footnote{I find
  the combination of Logic Programming and infinite terms difficult to handle (but I need to read~\cite{BasoldKL19}), so I suggest to restrict to finite objects that have meaningful infinite behaviors}, that is any infinite
terms on the domain of discourse, has lead us to adopt a much stronger
logic (and associated proof theory) with explicit rules for induction
and coinduction.

A natural choice for such a logic is the fixed point logic $\cal
G$~\cite{Gacek2012} and its linear logic cousin $\mu\mathrm{MALL}$~\cite{Baelde12}, which are associated to the \emph{Abella} proof
assistant (\url{http://abella-prover.org}) and the \emph{Bedwyr} model-checker (\url{http://slimmer.gforge.inria.fr/bedwyr/}).  In fact, the latter
has already been used for related aims~\cite{HeathM15}.

In our current setup, we attempt to find counter-examples, using
Bedwyr to execute both the generation of test cases  and the testing phase.
%
Such an implementation of PBT has the advantages of allowing us to
piggyback on Bedwyr's facilities for efficient proof 
search via tabling for (co)inductive predicates.

%and supports reasoning over lambda-tree syntax via the $\nabla$ quantifier.

To make things more concrete, consider CCS. While analogous goals
using, for example, the \emph{Concurrency Workbench}, it is a
remarkable feature of the proof-theoretic account that is easy to
generalizes PBT from a system without bindings (say, CCS) to a system
with bindings (say, the $\pi$-calculus).
%

%
Now, for the syntax of CCS:
%\begin{changebar}

  Wrt a standard presentation, we use \texttt{null} for $0$,
  \texttt{plus} for binary sums, \texttt{par} for parallel
  composition, \texttt{out} for the prefix process. We do not account
  for relabeling or restriction (but it's easy with an operator
  \texttt{nu : (act -> proc) -> proc}). Recursion (aka process
  definitions) could be easily accounted for with a constant \texttt{mu
    : (proc -> proc ) -> proc}, but this has some issues when defining
  a generator for it, so we postpone it. Basically, the generator
  clause would be \verb|is_proc (s N) (mu P) := is_proc N (P (mu P))|
  and this is out of the HO pattern unification fragment that Bedwyr
  handles. We can fix that, making the substitution explicit, but I
  just have not done it yet.
%  \end{changebar}

\begin{lstlisting}
Kind    proc, act,num  type.
Type    a,  b, tau     act .
Type    bar            act -> act .
Type    null           proc.
Type    out            act -> proc -> proc.
Type    plus, par      proc -> proc -> proc.
\end{lstlisting}

  The transition system:
  \begin{changebar}
    Co-names are obtained via \texttt{bar} and the \texttt{comp}
    predicates mediates among them. \texttt{step} is just the
    translation of the operational semantics rules of the above
    fragment of CCS (no value-passing, restriction, and recursion).
  \end{changebar}
\begin{lstlisting}
Define comp : act -> act -> prop by
comp A (bar A);
comp (bar A) A.

Define step : proc -> act -> proc -> prop by
step (out A P) A P;
step (plus P _) A P1 := step P A P1;
step (plus _ Q) A Q1 := step Q A Q1;
step (par P Q) A (par P1 Q) := step P A P1;
step (par P Q) A (par P Q1) := step Q A Q1;
step (par P Q) tau (par P1 Q1) :=
  exists A B, comp A B /\ step P A P1 /\ step Q A Q1.
\end{lstlisting}

  Now for \emph{strong} (bi)similarity (coinductive) and its negation,
  which we obtain by pushing negation inside. This is informally
  justified by the fact that Bedwyr's proof theory can be seen as
  multiplicative additive linear logic with fixed points, where
  negation normal forms exist in the same sense of classical logic\footnote{Th story is a bit more complicated since to handle open queries, we also need disunification, but again this is relevant only wrt the current implementation of Bedwyr.}:
\begin{lstlisting}
Define coinductive sim : proc -> proc -> prop by
sim P Q :=
	forall A P1,  step P A P1  -> exists Q1,  step Q A Q1  /\ sim P1 Q1.

Define coinductive bisim : proc -> proc -> prop by
bisim P Q :=
    (forall A P1, step P A P1 -> exists Q1, step Q A Q1 /\ bisim P1 Q1) /\
    (forall A Q1, step Q A Q1 -> exists P1, step P A P1 /\ bisim P1 Q1).

Define inductive n_bisim : proc -> proc -> prop by
  n_bisim P Q :=
	(exists A P1, step P A P1 /\ forall Q1, step Q A Q1 -> n_bisim P1 Q1) \/
	(exists A Q1, step Q A Q1 /\ forall P1, step P A P1 -> n_bisim Q1 P1).
\end{lstlisting}

Generators: these are here hard-wired exhaustive depth-bounded
ones, but using the FPC architecture of~\cite{Blanco0M19}, we can mix
and match different strategies (even random ones); for the sake of
these notes, these generators will do.
\begin{lstlisting}
% indices for gens
Type z num.
Type s num -> num.

Define is_act : num -> act -> prop by
is_act  _ a ; is_act _ b ; is_act _ tau;
is_act (s N) (bar A) := is_act N A.

Define is_proc :          num -> proc  -> prop by
is_proc _ null;
is_proc (s N) (out A P) := is_act N A /\ is_proc N P;
is_proc (s N) (plus P Q) := is_proc N P /\ is_proc N Q;
is_proc (s N) (par P Q) := is_proc N P /\ is_proc N Q.
\end{lstlisting}

Now, we can use directly Bedwyr, for example asking it to find two
terms that are similar but not bisimilar:
\begin{lstlisting}
?= N = (s (s z)) /\ is_proc N P /\ is_proc N Q /\ sim P Q /\ sim Q P /\ n_bisim P Q.
 % Q = par (out tau null) (out tau null),  that is t.0 | t.0
 % P = out tau (out tau null), that is t.t.0
\end{lstlisting}

The generation of a process is just logic programming search, while
testing their similarity is coinductive and it is implemented in
Bedwyr with tables, which basically encode the required
simulation. There is a meta-theoretic result that translates a tabled
logic programming derivation into one in a meta-logic such as $\cal G$
with explicit induction and coinduction rules.  Bedwyr carries out a
standard uniform proof of a goal such as \verb|P -> Q|, with one
exception: once \texttt{P} is assumed, we carry out a case analysis on
it by searching for a proof of it and apply the stream of computed
answer substitutions from \texttt{P} to the search of a proof of
\texttt{Q}. This case analysis has some technical limitations to make
it sound, which explains why we need to explicitly encode the negation
of bisimulation rather then simply call \verb|bisim P Q -> false|.

\begin{changebar}
  Another query: is there a process that is not bisimilar to the
  parallel composition with itself? In other terms, is this true:
  $\forall P, bisim\ P ( P | P)$? Yes, namely \texttt{a.0}.
\end{changebar}

\begin{lstlisting}
 ? N = (s z) /\ is_proc N P /\ n_bisim (par P P) P.
%  P = out a null
\end{lstlisting}

For a final example, consider \emph{trace} equivalence:
\begin{lstlisting}
Kind trace   type.
Type emp     trace.
Type tr      act -> trace -> trace.

Define trace : proc -> trace -> prop by
       trace _ emp ;
       trace P (tr A Trace) := step P A Q /\ trace Q Trace.

Define trace_equiv : proc -> proc -> prop by
       trace_equiv P Q := (forall Trace, trace P Trace -> trace Q Trace) /\
                          (forall Trace, trace Q Trace -> trace P Trace).
\end{lstlisting}
It's possible to separate trace equivalence and bi-similarity:
\begin{changebar}
\begin{lstlisting}
Define trace_eq_not_bisim : proc -> proc -> prop by
 trace_eq_not_bisim P Q :=
  N =  (s (s z)) /\ N1 =  (s (s (s z))) /\ is_proc N P /\ is_proc N1 Q
	 /\  trace_equiv P Q /\ n_bisim P Q.
% Q = plus (out a null) (out a (out a null))
% P = out a (out a null)
\end{lstlisting}
but this takes some 10 minutes, probably because we are building up two processes $P$ and $Q$
independently.
\end{changebar}
This is different from previous examples
in~\cite{Blanco0M19}, where generation was performed over judgments
such as typing (see preservation/progress). A possibility is to
generate $P$ and $Q$ together, following ideas in \cite{TNIQ}, whereby
we first generate $P$ and then obtain $Q$ by locally mutating the
former under some strong constraints. This is where
\emph{constrain-and-generate} could play a role somehow.
\begin{changebar}
  Note however that while \texttt{trace} has mode \texttt{+,-}, it is
  non-deterministic. Moreover, a predicate such as \texttt{trace\_equiv}
  has to be called with mode \texttt{+,+}, by the operational
  semantics of Bedwyr
\end{changebar}
Another avenue to investigate would be to \emph{pre-compute}
partitions of processes, possibly removing isomorphic ones (maybe
under the equational theory of process equivalence -- certainly
\texttt{P + Q} and \texttt{Q + P} ``are'' the same process). Then at
run time we do the testing bit. Papers by Tarau here may be relevant.

\begin{center}
  \begin{huge}
  \textbf{WARNING}
  \end{huge}
\end{center}

The previous examples DO NOT use coinduction, but follows only by case
analysis, that is would hold even if (bi)similarity is just \emph{a}
fixed point, not the greatest.

To find an example where we do use that, we move to the lambda-calculus:
 consider the usual rules for CBV
evaluation in the $\lambda$-calculus with constants, but define it
\emph{coinductively}, following see~\cite{2007-Leroy-Grall}:
\begin{lstlisting}
Type    zero, one    const.
Type    con          const -> tm.
Type    app          tm -> tm -> tm.
Type    fun          (tm -> tm) -> tm.

Define coinductive coeval: tm -> tm -> prop by
 coeval (con C) (con C);
 coeval (fun R) (fun R);
 coeval (app M N) V := 
   exists R W, coeval M (fun R) /\ coeval N W /\ coeval (R W) V.
% generators
Define wt : num -> env -> tm -> prop by
    wt _ _ (con zero) ;
    wt _ _ (con one) ;
    wt _ E (A)              := mem E A ;
    wt (s N) E (app X Y)    := wt N E X /\ wt N E Y ;
    wt (s N) E (fun F)      := nabla x, wt N (bind x E) (F (x)).

\end{lstlisting}
Is evaluation still deterministic?  And if not, can we find terms
\lstinline{M}, \lstinline{M1}, and \lstinline{M2} such that
\mbox{\lstinline{coeval M M1 /\\ coeval E M2 /\\ (M1 = M2 -> false)}}?
\begin{lstlisting}
 ?= wt (s (s (s z))) void M /\ wt z void M1 /\ wt z void M2 /\ coeval M M1 /\ coeval M M2 /\ (M1 = M2 -> false))

 Found a solution:  2195msa
 M2 = con one
 M1 = con zero
 M  = app (fun (x1\ app x1 x1)) (fun (x1\ app x1 x1))
\end{lstlisting}

Note that we need generators for \texttt{M1,M2}, since a coinductive
definition of evaluation requires mode \texttt{+,+}, and I do not
see how co-routining would help. The inductive version has mode
\texttt{+,-}, so we need to generate only \texttt{M}; in fact, we can
use \texttt{coeval M M1} \emph{directly} as a generator.

\bibliographystyle{abbrv} \bibliography{../colp,../../ppdp19/l}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

%  LocalWords:  Coinductive QuickCheck utf inputenc fontenc Herbrand
%  LocalWords:  corecursion coinductive corecursive QuickChick CBV ir
%  LocalWords:  disjuncts PCFL LTS Harrop PBT Coq's coinductively CCS
%  LocalWords:  intuitionistic FPCs Bedwyr Bedwyr's proc finitary
%  LocalWords:  bisimulation bisim Tarau sequents disunification
%  LocalWords:  routining

### 2020-10-16 papers/coLP/types20/t20.tex
Added another paper.

--- 2020-10-16 papers/coLP/outline.tex
@@
 \usepackage{url,xcolor}
-\usepackage{proof}
 \usepackage{listings}
 \usepackage{letltxmacro}
-
-\newcommand{\instan}[1]{\hbox{\sl grnd}~(#1)}
-\newcommand{\Pscr}{{\mathcal P}}
-
-\newcommand{\ok}{\checkmark}
-% DM The use of pifont and txfonts broke typesetting for me: equality
-% signs never appearred and some parentheses did not appear either.
-%\newcommand{\noc}{\ding{56}}
-\newcommand{\noc}{\ddag}
-\newcommand{\lP}{$\lambda$Prolog\xspace}
-\newcommand{\nat }{\hbox{\sl nat}\xspace}
-\newcommand{\plus}{\hbox{\sl app}\xspace}
-\newcommand{\lst}{\hbox{\sl lst}\xspace}
-\newcommand{\sort}{\hbox{\sl sort}}
-\newcommand{\rev}{\hbox{\sl rev}}
-\newcommand{\ordered}{\hbox{\sl ordered}}
-
-%\newcommand{\atac}{A2Tac\xspace}
-\newcommand{\atac}{ACheck\xspace}
-%\newcommand{\lra}{\mathrel{\longrightarrow}}
-\newcommand{\lra}{\mathrel{\vdash}}
-\newcommand{\seq}[2]{#1\lra #2}
-
-\newcommand{\true}{tt}
-
-%%%%%%%%%%%%%%%% LJF
-\newcommand{\truen}{t^-}
-\newcommand{\truep}{t^+}
-\newcommand{\falsen}{f^-}
-\newcommand{\falsep}{f^+}
-\newcommand{\wedgep}{\wedge^{\!+}}
-\newcommand{\wedgen}{\wedge^{\!-}}
-\newcommand{\veep}{\vee^{\!+}}
-\newcommand{\veen}{\vee^{\!-}}
-\newcommand{\with}{\&}
-
-\newcommand{\Nscr}{{\cal N}}
-\newcommand{\Rscr}{{\cal R}}                                   % Used for an ambiguous rhs
-%\newcommand{\Rscr}{\Delta_1\Downarrow\Delta_2}                % Used for an ambiguous rhs
-\newcommand{\jUnf    }[4]{#1\mathbin\Uparrow#2\vdash#3\mathbin\Uparrow #4} % unfocused sequent
-\newcommand{\jUnfG   }[2]{\jUnf{\Gamma}{#1}{#2}{{}}}           % unf sequ with \Gamma
-\newcommand{\jUnfamb }[3]{#1\mathbin\Uparrow#2\vdash#3 \Rscr}  % unfocused sequent
-\newcommand{\jUnfGamb}[1]{\Gamma\mathbin\Uparrow#1\vdash \Rscr}% unf sequ with \Gamma
-\newcommand{\jLf     }[3]{#1\Downarrow#2\vdash#3}              % left focused sequent
-\newcommand{\jLfG    }[1]{\jLf{\Gamma}{#1}{E}}                 % left foc seq with \Gamma 
-\newcommand{\jRf     }[2]{#1\vdash #2\Downarrow}               % right focused sequent
-\newcommand{\jRfG    }[1]{\jRf{\Gamma}{#1}}                    % right foc seq with \Gamma
-%%%%%%%%%%%%%%%% 
-\newcommand{\bxi}[1]{\blue{\Xi_{#1}}}
-\newcommand{\bXi}[1]{\blue{\Xi_{#1} :\null}}
-
-\newcommand{\andClerk}[3]{{\wedge_c}(#1,#2,#3)}
-\newcommand{\falseClerk}[2]{f_c(#1,#2)}
-\newcommand{\orClerk}[2]{{\vee_c}(#1,#2)}
-\newcommand{\allClerk}[2]{\forall_c(#1,#2)}
-\newcommand{\storeClerk}[3]{\hbox{\sl store}_c(#1,#2,#3)}
-
-\newcommand{\trueExpert }[1]{{\true_e}(#1)}
-\newcommand{\eqExpert }[1]{{=_e}(#1)}
-\newcommand{\unfoldExpert}[2]{{\hbox{\sl unfold}_e}(#1,#2)}
-\newcommand{\andExpert}[3]{{\wedge_e}(#1,#2,#3)}
-\newcommand{\andExpertLJF}[6]{{\wedge_e}(#1,#2,#3,#4,#5,#6)}
-\newcommand{\orExpert  }[3]{{\vee_e}(#1,#2,#3)}
-\newcommand{\someExpert}[3]{\exists_e(#1,#2,#3)}
-\newcommand{\initExpert}[2]{\hbox{\sl init}_e(#1,#2)}
-\newcommand{\cutExpert}[4]{\hbox{\sl cut}_e(#1,#2,#3,#4)}
-\newcommand{\decideExpert}[3]{\hbox{\sl decide}_e(#1,#2,#3)}
-\newcommand{\releaseExpert}[2]{\hbox{\sl release}_e(#1,#2)}
-%%%%%%%%%%%%%%%%
-
-%
-\newcommand{\step}{\longrightarrow}
-\newcommand{\sstlc}{\textit{STLC}\xspace}
-\newcommand{\vds}{\vdash_\Sigma}
-
-\def\bnfas{\mathrel{::=}}
-\def\bnfalt{\mid}
-
-\def\lam{\lambda}
-\def\Lam{\Lambda}
-\def\arrow{\rightarrow}
-\def\oftp{\mathord{:}}
-\def\hastype{\mathrel{:}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Editorials
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\long\def\ednote#1{\footnote{[{\it #1\/}]}\message{ednote!}}
-%\long\def\ednote#1{\begin{quote}[{\it #1\/}]\end{quote}\message{note!}}
-\newenvironment{metanote}{\begin{quote}\message{note!}[\begingroup\it}%
-                         {\endgroup]\end{quote}}
-\long\def\ignore#1{}
-
-\newcommand{\todo}[1]{\begin{metanote}TODO: #1\end{metanote}}
-%
-%\setlength{\textwidth}{13.2cm}
-%\setlength{\textheight}{19.9cm}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\input{types20/listing-macros}  % Use to get better highlighing of code
+\input{listing-macros}  % Use to get better highlighing of code
 \lstset{language=lprolog}
@@
 \maketitle
-
-\section{Introduction}
-\label{sec:intro}
-
-\begin{center}
-  [For the moment taken from the TYPES abstract, to be revised]
-\end{center}
 
@@
 In the context of PBT, the proof theory setup is much simpler.
-
-\begin{center}
-  [etc\dots]
-\end{center}
-
-The contributions of this paper are
-\begin{itemize}
-\item Tighten the link of our previous reconstruction of PBT to the
-  proof-theoretic approach of model checking in~\cite{heath19jar} by casting it in the setting of $\mu$MALL
-\item Extend PBT beyond the Horn-nabla fragment to Harrop spec, first using only case analysis, then also co-induction
-\item Experiment with a multi-level architecture, where FPC only takes
-  care of generation and testing  is delegated to a system such as Bedwyr
-\item Hopefully provide some examples (CCS, $\pi$ calculus (?), co-evaluation)
-\item
-  \begin{center}
-[others \dots]    
-  \end{center}
-
-\end{itemize}
-
-\section{The story so far}
-\label{sec:tecap}
-
-\begin{center}
-  [Here we recap what we did in PPDP, in the uniform proof style, see fig~\ref{fig:augmented}: the connection with linear logic comes in the next section]
-\end{center}
-
 Consider an attempt to find counter-examples to a conjecture of the
@@
 %
+FPCs are simple logic programs that guide the search for potential
+counter-examples using different generation strategies; they further
+capture diverse features such as $\delta$-debugging, fault isolation,
+explanation, etc.  Such a range of features can be
+programmed as the \emph{clerks and experts} predicates that decorate
+the sequent rules used in a FPC proof checking kernel: the kernel
+is also able to do a limited amount of proof reconstruction.
+
+As explained in~\cite{Blanco0M19}, the standard PBT setup needs little
+more than Horn logic. % specifications.
 %
-
-\begin{figure}
-\newcommand{\XXi}{{\color{blue}{\Xi}}}
-\[
-\infer{\XXi\vdash G_1\wedge G_2}
-      {\XXi_1\vdash G_1\qquad \XXi_2\vdash G_2\qquad \andExpert{\XXi}{\XXi_1}{\XXi_2}}
-\qquad
-\infer{\XXi\vdash \true}
-      {\trueExpert{\XXi}}
-\]
-\vskip -6pt
-\[
-\infer{\XXi\vdash G_1\vee G_2}
-      {\XXi'\vdash G_i\qquad \orExpert{\XXi}{\XXi'}{i}}
-\qquad
-\infer{\XXi\vdash \exists x. G}
-      {\XXi'\vdash G[t/x]\qquad \someExpert{\XXi}{\XXi'}{t}}
-\]
-\vskip -6pt
-\[
-\infer{\XXi\vdash t = t}
-      {\eqExpert{\XXi}}
-\qquad
-\infer{\XXi\vdash A}
-      {\XXi'\vdash G \quad (A~\hbox{\tt :-}~G)\in\instan\Pscr
-                     \quad \unfoldExpert{\XXi}{\XXi'}}
-\]
-\caption{A proof system augmented with proof certificates and
-  additional ``expert'' premises.}
-\label{fig:augmented}
-\end{figure}
-
-\section{Linear logic as the proof theory of model checking}
-\label{sec:ll}
-
-A natural choice is linear logic cousin $\mu\mathrm{MALL}$~\cite{Baelde12}, which is associated to  the \emph{Bedwyr} model-checker. %  In fact, the latter
-% has already been used for related aims~\cite{HeathM15}.
-
-\subsection{Horn programs as fixed points}
-\label{ssec:lpfixed}
-
-\begin{center}
-  [Here we introduce fixed points, and re-state the rules in fig~\ref{fig:augmented} with only unfolding rules.
-We may also want to introduce our first coinductive examples, probably \texttt{coeval} (it's Horn).
-  ]
-\end{center}
-
-\subsection{Onwards to linear logic}
-\label{ssec:ll}
-
-\begin{center}
-  [
-  \begin{itemize}
-  \item Here we introduce polarization, focusing. I assume we will be
-    using only $\mu$F rather than all muMALL; and if so, perhaps we
-    can simplify the syntax of the focusing judgment
-  \item Not sure we want to put clercks and experts everywhere, if they play a part only during generation and that should stay in the Horn+nabla fragment
-  \item Now we can introduce non Horn examples such as simulation and points out that the left unfolding rules do case analysis.
-\end{itemize}
-  ]
-\end{center}
-
-\subsection{Full system}
-\label{ssec:fullmumall}
-
-\begin{center}
-  [Here we add the rules for (co)induction, init (and cut?). In the end we take the muFlogic from the JAR paper and the ``completeness'' results thereof]
-\end{center}
-
-\section{Implementation and case studies}
-\label{sec:case}
-
-\begin{center}
-  [Here we discuss the ``architecture'', namely using the lightweight FPC for generators and Bedwyr for testing]
-\end{center}  
-\subsection{Examples}
-\label{ssec:exs}
-
-\begin{center}
-  [Which esamples? The ones which work]
-\end{center}
-
-\subsection{Refinements}
-\label{ssec:refs}
-
-\begin{center}
-  [Anything else to make this interesting: integration with
-  proof-outlines, pre-computation of terms 'a la Tarau, full
-  ``post-mortem'' FPC, that is also for the testing phase, to
-  elaborate further the results, for examples for explanations etc., wild ideas about generating infinite terms with observation functions as in \emph{Hipster} \cite{EinarsdottirJP18}]
-\end{center}
-
-\section{Related work}
-\label{sec:rel}
-
-\begin{center}
-  [to much details, just for my own sake, only a small portion will make it in the paper]
-\end{center}
-  While coinductive logic programming,
+However, when addressing infinite computations, we need richer
+specifications.  While coinductive logic programming,
 see~\cite{Luke07} and~\cite{BasoldKL19} for a much more principled and
@@
 coinduction.
-\input{long-related}
-\section{Conclusion}
-\label{sec:conc}
+
+A natural choice for such a logic is the fixed point logic $\cal
+G$~\cite{Gacek2012} and its linear logic cousin $\mu\mathrm{MALL}$~\cite{Baelde12}, which are associated to the \emph{Abella} proof
+assistant and the \emph{Bedwyr} model-checker.  In fact, the latter
+has already been used for related aims~\cite{HeathM15}.
+
+%and supports reasoning over lambda-tree syntax via the $\nabla$ quantifier.
+
+To make things more concrete, consider the usual rules for CBV
+evaluation in the $\lambda$-calculus with constants, but define it
+\emph{coinductively}, following see~\cite{2007-Leroy-Grall}: using Bedwyr's
+concrete syntax, this is written as: 
+\begin{lstlisting}
+Define coinductive coeval: tm -> tm -> prop by
+ coeval (con C) (con C);
+ coeval (fun R) (fun R);
+ coeval (app M N) V := 
+   exists R W, coeval M (fun R) /\ coeval N W /\ coeval (R W) V.
+\end{lstlisting}
+Is evaluation still deterministic?  And if not, can we find terms
+\lstinline{E}, \lstinline{V1}, and \lstinline{V2} such that
+\mbox{\lstinline{coeval E V1 /\\ coeval E V2 /\\ (V1 = V2 -> false)}}?\footnote{In
+  this case equality is purely syntactical, since by construction terms
+  will be ground when compared, but the logic implements a richer
+  notion~\cite{GacekMN11}.} Indeed we can, since a divergent term such
+as $\Omega$ co-evaluates to anything.  In fact, co-evaluation is not
+even type sound in its generality. Our PBT approach aims to find such
+counter-examples.
+
+It can also be used to separate various notion of
+equivalences in lambda and process calculi: for example,
+separating applicative and ground similarity in PCFL~\cite{PITobtpe},
+or analogous standard results in the $\pi$-calculus.
+%
+While analogous goals have been achieved for labeled transition systems
+and for CCS (using, for example, the \emph{Concurrency Workbench}), it
+is a remarkable feature of the proof-theoretic account that is easy to
+generalizes PBT from a system without bindings (say, CCS) to a system
+with bindings (say, the $\pi$-calculus).
+%
+Such ease is possible since proof theory accommodates the
+\emph{$\lambda$-tree syntax} approach to treating bindings~\cite{miller18jar}: this approach includes the $\nabla$
+quantifier~\cite{miller05tocl} that appears in both Abella and Bedwyr.
+
+In our current setup, we attempt to find counter-examples, using
+Bedwyr to execute both the generation of test cases (controlled by
+using specific FPCs~\cite{Blanco0M19}) and the testing phase.
+%
+Such an implementation of PBT has the advantages of allowing us to
+piggyback on Bedwyr's facilities for efficient proof 
+search via tabling for (co)inductive predicates.
+%
+There are a couple of treatments of the negation in the testing phase.
+%
+One approach to eliminating negation from intuitionistic specifications
+can be based on the techniques in~\cite{Momigliano00}.
+%
+Another approach identifies the proof theory behind model checking as
+the linear logic $\mu$MALL~\cite{HeathM19} and in that setting,
+negations can be eliminated by using De Morgan duality (and
+disunification).
 
 \bibliographystyle{abbrv}
-\bibliography{colp,../ppdp19/l}
+\bibliography{../colp,../../ppdp19/l}
 
@@
 %  LocalWords:  disjuncts PCFL LTS Harrop PBT Coq's coinductively CCS
-%  LocalWords:  intuitionistic FPCs Bedwyr Bedwyr's nabla
+%  LocalWords:  intuitionistic FPCs Bedwyr Bedwyr's

### 2020-10-16 papers/coLP/code/haskell/l.hs

@@ -1,0 +1,26 @@
+import Test.QuickCheck
+
+eqi [] [] = True
+eqi (x : xs) (y : ys) = (x == y) && eqi xs ys
+eqi _ _ = False
+
+zeros = 0 : zeros
+ones = 1 : ones
+ones' = map succ zeros
+
+
+prop_eq :: [Integer] -> [Integer] -> Property
+prop_eq xs ys =
+  eqi xs ys ==> (xs == ys)
+
+
+prop_d :: [Int] -> Int -> Property
+prop_d xs n =
+  not (null xs) && n >= 0 ==>
+  take n (cycle xs) == take n (cycle $ xs ++ xs)
+
+main = quickCheck prop_d
+
+prop_o :: Int -> Property
+prop_o n =
+  n >= 0 ==> take n ones == take n zeros

### 2020-10-16 papers/coLP/code/prolog/aut.pl

@@ -1,0 +1,54 @@
+% based on history meta-interpreter
+:- ensure_loaded('meta-co.pl').
+
+:- discontiguous coinductive/2.
+
+:- dynamic automata/3.
+:- dynamic accepts/2.
+:- dynamic drop/3.
+:- dynamic trans/3.
+:- dynamic comember/2.
+:- dynamic final/1.
+
+% true is X occurs in Xs infinitely oftern. Fail if finte list, loops
+% if it does not.
+coinductive(comember,2).
+comember(X,L) :-
+    drop(X,L,L1), comember(X,L1).
+drop(H,[H|T],T).
+drop(H,[_|T],T1) :- drop(H,T,T1).
+
+% eg. Xs = [1,2,3|Xs], query(comember(2,Xs)).
+% Xs = [1, 2, 3|Xs]
+
+% Xs = [1,2,3], query(comember(2,Xs)).
+% false.
+
+
+coinductive(automata,3).
+automata([X|T], St, Sts) :-
+    trans(St, X, NewSt), automata(T, NewSt,[St|Sts]).
+trans(s0, a, s1).
+trans(s1, b, s2).
+trans(s2, c, s3).
+trans(s3, d, s0).
+trans(s2, e, s0).
+final(s2).
+
+
+% String accepted by fixed automata  starting with InitSt and with Final
+% occurring inf often
+
+% (abcd)* accepted and final s2 occurs infintely often
+
+accepts(String,InitSt) :-
+    automata(String,InitSt,Sts),
+    final(Final),
+    comember(Final,Sts).
+
+%  query((automata(X,s0, States ), comember(s2,States))).
+%  X = [a, b, c, d|X],
+% States = [s3, s2, s1, s0|States] ;
+
+%query(accepts(String,s0)).
+%String = [a, b, c, d|String]

### 2020-10-16 papers/coLP/code/prolog/auts.pl

@@ -1,0 +1,62 @@
+% automata over different LTS
+
+:- ensure_loaded('meta-co.pl').
+:- discontiguous final/2.
+:- discontiguous trans/4.
+:- discontiguous coinductive/2.
+
+:- dynamic automata/4.
+:- dynamic accepts/3.
+:- dynamic drop/3.
+:- dynamic trans/3.
+:- dynamic comember/2.
+:- dynamic final/2.
+:- dynamic is_aut/2.
+:- dynamic not_sim/2.
+
+coinductive(comember,2).
+comember(X,L) :-
+    drop(X,L,L1), comember(X,L1).
+drop(H,[H|T],T).
+drop(H,[_|T],T1) :- drop(H,T,T1).
+
+
+coinductive(automata,4).
+
+accepts(String,InitSt,Trans) :-
+    automata(String,InitSt,Trans,Sts),
+    final(Trans,Final),
+    comember(Final,Sts).
+
+
+automata([X|T], St, Trans, Sts) :-
+    trans(Trans, St, X, NewSt),
+    automata(T, NewSt,Trans,[St|Sts]).
+
+% graph g1
+trans(g1, s0, a, s1).
+trans(g1, s1, b, s2).
+trans(g1, s2, c, s3).
+trans(g1, s3, d, s0).
+trans(g1, s2, e, s0).
+final(g1, s2).
+
+% graph g2, 
+trans(g2, s0, a, s1).
+trans(g2, s1, b, s2).
+%trans(g2, s2, c, s3).
+%trans(g2, s3, d, s0).
+trans(g2, s2, e, s0).
+final(g2, s2).
+
+is_aut(g1,s0).
+is_aut(g2,s0).
+
+not_sim(A1,A2) :-
+    is_aut(g1,InitSt1),
+    is_aut(g2,InitSt2),
+    accepts(String1,InitSt1,A1),
+%    not(accepts(String,InitSt2,A2)). % not not-handled by the meta
+    accepts(String2,InitSt2,A2),
+    not(String1 =  String2).				
+    

### 2020-10-16 papers/coLP/code/prolog/co.pl

@@ -1,0 +1,73 @@
+:- use_module(library(coinduction)).
+
+:- coinductive p/1.
+
+p([1|T]) :- p(T).
+
+
+:- coinductive from/2.
+from(N, [N|T]) :- from(s(N), T).
+
+% :- from(0,_). %loops, why?
+:- coinductive stream/1.
+stream([H | T]) :-
+    num(H), stream(T).
+
+num(0).
+num(s(N)) :- num(N).
+
+:- coinductive cmem/2.
+cmem( H, [ H | _ ] ).
+cmem( H, [ _ | T ] ) :- cmem( H, T ).
+
+%  cmem(0,[s(z),0]). % true
+%  X = [0|X],cmem(a,X). %true: anything member of a inf list
+
+% called comember/2, which is true if and only if the desired element
+% occurs an infnite number of times in the list.
+
+drop( H, [ H | T ], T ).
+drop( H, [ _ | T ], T1 ) :- drop( H, T, T1 ).
+
+:- coinductive comember/2.
+comember( X, L ) :-
+    drop( X, L, L1 ), comember( X, L1 ).
+
+% X = [ 1, 2, 3 | X ], comember( 2, X ). yes
+% X = [ 1, 2, 3 | X ], comember( 4, X ). loops
+%   X = [ 1, 2, 3], comember( E, X ). fails
+
+%--------------------------------
+
+
+%:-  stream([0, s(0), s(s(0)) | T ]).
+% TODO: some of the bewdyr examples
+
+% from Luke's thesis p.77
+% does not seem to work in SWI
+:- coinductive automata/2.
+automata([X|T], St) :- trans(St, X, NewSt), automata(T, NewSt).
+trans(s0, a, s1).
+trans(s1, b, s2).
+trans(s2, c, s3).
+trans(s3, d, s0).
+trans(s2, e, s0).
+final(s2).
+
+% :- automata(X, s0), comember(s2, X).
+:- coinductive bin/1.
+bin([0|T]) :- bin(T).
+bin([1|T]) :- bin(T).
+
+:-  X = [1|X] , bin(X).
+% contrary to paper yap bin(X) does not retun  X = [1|X]. (df)
+
+				% maxlist, example from Ancona should prove - Os = [1|Os], maxList(Os,2). but it does not
+
+max(A,B,Max) :-
+	(A < B -> Max = B; Max = A).
+mmax(1,2,2).
[…]

### 2020-10-16 papers/coLP/code/prolog/meta-co.pl

@@ -1,0 +1,44 @@
+% metainterpreter with depth-first history mechanism for "coinduction"
+:- discontiguous coinductive/2.
+query( Goal ) :-
+    solve( [], Goal ).
+solve( Hypothesis, ( Goal1, Goal2 ) ) :-
+    !,
+    solve( Hypothesis, Goal1 ),
+    solve( Hypothesis, Goal2 ).
+solve( _ , Atom ) :-
+    builtin( Atom ),
+    !,
+    Atom.
+solve( Hypothesis, Atom ) :-
+    coinductive( Atom ),
+    member( Atom, Hypothesis ).
+solve( Hypothesis, Atom ) :-
+    coinductive( Atom ),
+    !,
+    clause( Atom, Atoms ),
+    solve( [ Atom | Hypothesis ], Atoms ).
+
+solve( Hypothesis, Atom ) :-
+    %inductive( Atom ),
+    clause( Atom, Atoms ),
+    solve( Hypothesis, Atoms ).
+
+inductive( Atom ) :-
+    Atom \= ( _, _ ),
+    \+ builtin( Atom ),
+    functor( Atom, Predicate, Arity ),
+    inductive( Predicate, Arity ).
+coinductive( Atom ) :-
+    Atom \= ( _, _ ),
+    \+ builtin( Atom ),
+    functor( Atom, Predicate, Arity ),
+    coinductive( Predicate, Arity ).
+
+builtin( _ = _ ).
+builtin( true ).
+builtin( _ is _ ).
+builtin( _ > _ ).
+builtin( _ < _ ).
+builtin(\+ _).
+builtin(not(_G)).

### 2020-10-20 papers/types2020/main.tex
A little more progress on an outline.

\documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
%This is a template for producing LIPIcs articles. 
%See lipics-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Interfacing logic programming and Coq} % Other suggestions?
\titlerunning{Interfacing logic programming and Coq}

%TODO mandatory, please use full name; only 1 author per \author
%macro; first two parameters are mandatory, other parameters can be
%empty. Please provide at least the name of the affiliation and the
%country. The full address is optional

\author{Dale Miller}
       {Inria, Palaiseau, France \and LIX, Ecole Polytechnique, France
         \and \url{http://www.lix.polytechnique.fr/Labo/Dale.Miller/} }
       {dale.miller@inria.fr}
       {}%https://orcid.org/0000-0003-0274-4954
       {}

\author{Matteo Manighetti}
       {Inria, Palaiseau, France \and LIX, Ecole Polytechnique, France
         \and \url{http://www.lix.polytechnique.fr/Labo/Matteo.Manighetti/}}
       {matteo.manighetti@inria.fr}
       {}
       {}

\author{Alberto Momigliano}
       {Dipartimento di Informatica, Universit\`a degli studi di Milano, Italy \and
       \url{http://momigliano.di.unimi.it/}}
       {momigliano@di.unimi.it}
       {}
       {}

\authorrunning{D. Miller, M. Manighetti, and A. Momigliano}
 %TODO mandatory. First: Use abbreviated first/middle names. Second
 %(only in severe cases): Use first author plus 'et al.'

\Copyright{D. Miller, M. Manighetti, and A. Momigliano}
%TODO mandatory, please use full first names. LIPIcs license is
%"CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords

\category{} %optional, e.g. invited paper

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversion{A full version of the paper is available at \url{...}.}

\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\acknowledgements{I want to thank \dots}%optional

%\nolinenumbers %uncomment to disable line numbering

%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

%TODO mandatory: add short abstract of the document
\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Praesent convallis orci arcu, eu mollis dolor. Aliquam eleifend suscipit lacinia. Maecenas quam mi, porta ut lacinia sed, convallis ac dui. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse potenti. 
\end{abstract}

\section{Typesetting instructions -- Summary}
\label{sec:typesetting-summary}

LIPIcs is a series of open access high-quality conference proceedings across all fields in informatics established in cooperation with Schloss Dagstuhl. 
In order to do justice to the high scientific quality of the conferences that publish their proceedings in the LIPIcs series, which is ensured by the thorough review process of the respective events, we believe that LIPIcs proceedings must have an attractive and consistent layout matching the standard of the series.
Moreover, the quality of the metadata, the typesetting and the layout must also meet the requirements of other external parties such as indexing service, DOI registry, funding agencies, among others. The guidelines contained in this document serve as the baseline for the authors, editors, and the publisher to create documents that meet as many different requirements as possible.

Please comply with the following instructions when preparing your article for a LIPIcs proceedings volume. 
\paragraph*{Minimum requirements}

\begin{itemize}
\item Use pdflatex and an up-to-date \LaTeX{} system.
\item Use further \LaTeX{} packages and custom made macros carefully and only if required.
\item Use the provided sectioning macros: \verb+\section+, \verb+\subsection+, \verb+\subsubsection+, \linebreak \verb+\paragraph+, \verb+\paragraph*+, and \verb+\subparagraph*+.
\item Provide suitable graphics of at least 300dpi (preferably in PDF format).
\item Use BibTeX and keep the standard style (\verb+plainurl+) for the bibliography.
\item Please try to keep the warnings log as small as possible. Avoid overfull \verb+\hboxes+ and any kind of warnings/errors with the referenced BibTeX entries.
\item Use a spellchecker to correct typos.
\end{itemize}

\paragraph*{Mandatory metadata macros}
Please set the values of the metadata macros carefully since the information parsed from these macros will be passed to publication servers, catalogues and search engines.
Avoid placing macros inside the metadata macros. The following metadata macros/environments are mandatory:
\begin{itemize}
\item \verb+\title+ and, in case of long titles, \verb+\titlerunning+.
\item \verb+\author+, one for each author, even if two or more authors have the same affiliation.
\item \verb+\authorrunning+ and \verb+\Copyright+ (concatenated author names)\\
The \verb+\author+ macros and the \verb+\Copyright+ macro should contain full author names (especially with regard to the first name), while \verb+\authorrunning+ should contain abbreviated first names.
\item \verb+\ccsdesc+ (ACM classification, see \url{https://www.acm.org/publications/class-2012}).
\item \verb+\keywords+ (a comma-separated list of keywords).
\item \verb+\relatedversion+ (if there is a related version, typically the ``full version''); please make sure to provide a persistent URL, e.\,g., at arXiv.
\item \verb+\begin{abstract}...\end{abstract}+ .
\end{itemize}

\paragraph*{Please do not \ldots} %Do not override the \texttt{\seriesstyle}-defaults}
Generally speaking, please do not override the \texttt{lipics-v2019}-style defaults. To be more specific, a short checklist also used by Dagstuhl Publishing during the final typesetting is given below.
In case of \textbf{non-compliance} with these rules Dagstuhl Publishing will remove the corresponding parts of \LaTeX{} code and \textbf{replace it with the \texttt{lipics-v2019} defaults}. In serious cases, we may reject the LaTeX-source and expect the corresponding author to revise the relevant parts.
\begin{itemize}
\item Do not use a different main font. (For example, the \texttt{times} package is forbidden.)
\item Do not alter the spacing of the \texttt{lipics-v2019.cls} style file.
\item Do not use \verb+enumitem+ and \verb+paralist+. (The \texttt{enumerate} package is preloaded, so you can use
 \verb+\begin{enumerate}[(a)]+ or the like.)
\item Do not use ``self-made'' sectioning commands (e.\,g., \verb+\noindent{\bf My+ \verb+Paragraph}+).
\item Do not hide large text blocks using comments or \verb+\iffalse+ $\ldots$ \verb+\fi+ constructions. 
\item Do not use conditional structures to include/exclude content. Instead, please provide only the content that should be published -- in one file -- and nothing else.
\item Do not wrap figures and tables with text. In particular, the package \texttt{wrapfig} is not supported.
\item Do not change the bibliography style. In particular, do not use author-year citations. (The
\texttt{natbib} package is not supported.)
\end{itemize}

\enlargethispage{\baselineskip}

This is only a summary containing the most relevant details. Please read the complete document ``LIPIcs: Instructions for Authors and the \texttt{lipics-v2019} Class'' for all details and don't hesitate to contact Dagstuhl Publishing (\url{mailto:publishing@dagstuhl.de}) in case of questions or comments:
\href{http://drops.dagstuhl.de/styles/lipics-v2019/lipics-v2019-authors/lipics-v2019-authors-guidelines.pdf}{\texttt{http://drops.dagstuhl.de/styles/lipics-v2019/\newline lipics-v2019-authors/lipics-v2019-authors-guidelines.pdf}}

\section{Lorem ipsum dolor sit amet}

Lorem ipsum dolor sit amet, consectetur adipiscing elit \cite{DBLP:journals/cacm/Knuth74}. Praesent convallis orci arcu, eu mollis dolor. Aliquam eleifend suscipit lacinia. Maecenas quam mi, porta ut lacinia sed, convallis ac dui. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse potenti. Donec eget odio et magna ullamcorper vehicula ut vitae libero. Maecenas lectus nulla, auctor nec varius ac, ultricies et turpis. Pellentesque id ante erat. In hac habitasse platea dictumst. Curabitur a scelerisque odio. Pellentesque elit risus, posuere quis elementum at, pellentesque ut diam. Quisque aliquam libero id mi imperdiet quis convallis turpis eleifend.

\begin{lemma}[Lorem ipsum]
\label{lemma:lorem}
Vestibulum sodales dolor et dui cursus iaculis. Nullam ullamcorper purus vel turpis lobortis eu tempus lorem semper. Proin facilisis gravida rutrum. Etiam sed sollicitudin lorem. Proin pellentesque risus at elit hendrerit pharetra. Integer at turpis varius libero rhoncus fermentum vitae vitae metus.
\end{lemma}

\begin{proof}
Cras purus lorem, pulvinar et fermentum sagittis, suscipit quis magna.

\begin{claim}
content...
\end{claim}
\begin{claimproof}
content...
\end{claimproof}

\end{proof}

\begin{corollary}[Curabitur pulvinar, \cite{DBLP:books/mk/GrayR93}]
\label{lemma:curabitur}
Nam liber tempor cum soluta nobis eleifend option congue nihil imperdiet doming id quod mazim placerat facer possim assum. Lorem ipsum dolor sit amet, consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna aliquam erat volutpat.
\end{corollary}

\begin{proposition}\label{prop1}
This is a proposition
\end{proposition}

\autoref{prop1} and \cref{prop1} \ldots

\subsection{Curabitur dictum felis id sapien}

Curabitur dictum \cref{lemma:curabitur} felis id sapien \autoref{lemma:curabitur} mollis ut venenatis tortor feugiat. Curabitur sed velit diam. Integer aliquam, nunc ac egestas lacinia, nibh est vehicula nibh, ac auctor velit tellus non arcu. Vestibulum lacinia ipsum vitae nisi ultrices eget gravida turpis laoreet. Duis rutrum dapibus ornare. Nulla vehicula vulputate iaculis. Proin a consequat neque. Donec ut rutrum urna. Morbi scelerisque turpis sed elit sagittis eu scelerisque quam condimentum. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas. Aenean nec faucibus leo. Cras ut nisl odio, non tincidunt lorem. Integer purus ligula, venenatis et convallis lacinia, scelerisque at erat. Fusce risus libero, convallis at fermentum in, dignissim sed sem. Ut dapibus orci vitae nisl viverra nec adipiscing tortor condimentum \cite{DBLP:journals/cacm/Dijkstra68a}. Donec non suscipit lorem. Nam sit amet enim vitae nisl accumsan pretium.

\begin{lstlisting}[caption={Useless code},label=list:8-6,captionpos=t,float,abovecaptionskip=-\medskipamount]
for i:=maxint to 0 do 
begin 
    j:=square(root(i));
end;
\end{lstlisting}

\subsection{Proin ac fermentum augue}

Proin ac fermentum augue. Nullam bibendum enim sollicitudin tellus egestas lacinia euismod orci mollis. Nulla facilisi. Vivamus volutpat venenatis sapien, vitae feugiat arcu fringilla ac. Mauris sapien tortor, sagittis eget auctor at, vulputate pharetra magna. Sed congue, dui nec vulputate convallis, sem nunc adipiscing dui, vel venenatis mauris sem in dui. Praesent a pretium quam. Mauris non mauris sit amet eros rutrum aliquam id ut sapien. Nulla aliquet fringilla sagittis. Pellentesque eu metus posuere nunc tincidunt dignissim in tempor dolor. Nulla cursus aliquet enim. Cras sapien risus, accumsan eu cursus ut, commodo vel velit. Praesent aliquet consectetur ligula, vitae iaculis ligula interdum vel. Integer faucibus faucibus felis.

\begin{itemize}
\item Ut vitae diam augue. 
\item Integer lacus ante, pellentesque sed sollicitudin et, pulvinar adipiscing sem. 
\item Maecenas facilisis, leo quis tincidunt egestas, magna ipsum condimentum orci, vitae facilisis nibh turpis et elit. 
\end{itemize}

\begin{remark}
content...
\end{remark}

\section{Pellentesque quis tortor}

Nec urna malesuada sollicitudin. Nulla facilisi. Vivamus aliquam tempus ligula eget ornare. Praesent eget magna ut turpis mattis cursus. Aliquam vel condimentum orci. Nunc congue, libero in gravida convallis \cite{DBLP:conf/focs/HopcroftPV75}, orci nibh sodales quam, id egestas felis mi nec nisi. Suspendisse tincidunt, est ac vestibulum posuere, justo odio bibendum urna, rutrum bibendum dolor sem nec tellus.

\begin{lemma} [Quisque blandit tempus nunc]
Sed interdum nisl pretium non. Mauris sodales consequat risus vel consectetur. Aliquam erat volutpat. Nunc sed sapien ligula. Proin faucibus sapien luctus nisl feugiat convallis faucibus elit cursus. Nunc vestibulum nunc ac massa pretium pharetra. Nulla facilisis turpis id augue venenatis blandit. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus.
\end{lemma}

Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.

\section{Morbi eros magna}

Morbi eros magna, vestibulum non posuere non, porta eu quam. Maecenas vitae orci risus, eget imperdiet mauris. Donec massa mauris, pellentesque vel lobortis eu, molestie ac turpis. Sed condimentum convallis dolor, a dignissim est ultrices eu. Donec consectetur volutpat eros, et ornare dui ultricies id. Vivamus eu augue eget dolor euismod ultrices et sit amet nisi. Vivamus malesuada leo ac leo ullamcorper tempor. Donec justo mi, tempor vitae aliquet non, faucibus eu lacus. Donec dictum gravida neque, non porta turpis imperdiet eget. Curabitur quis euismod ligula.

%%
%% Bibliography
%%

%% Please use bibtex,

\bibliography{lipics-v2019-sample-article}

\appendix

\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}

List of different predefined enumeration styles:

\begin{itemize}
\item \verb|\begin{itemize}...\end{itemize}|
\item \dots
\item \dots
%\item \dots
\end{itemize}

\begin{enumerate}
\item \verb|\begin{enumerate}...\end{enumerate}|
\item \dots
\item \dots
%\item \dots
\end{enumerate}

\begin{alphaenumerate}
\item \verb|\begin{alphaenumerate}...\end{alphaenumerate}|
\item \dots
\item \dots
%\item \dots
\end{alphaenumerate}

\begin{romanenumerate}
\item \verb|\begin{romanenumerate}...\end{romanenumerate}|
\item \dots
\item \dots
%\item \dots
\end{romanenumerate}

\begin{bracketenumerate}
\item \verb|\begin{bracketenumerate}...\end{bracketenumerate}|
\item \dots
\item \dots
%\item \dots
\end{bracketenumerate}

\begin{description}
\item[Description 1] \verb|\begin{description} \item[Description 1]  ...\end{description}|
\item[Description 2] Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\item[Description 3]  \dots
%\item \dots
\end{description}

\cref{testenv-proposition} and \autoref{testenv-proposition} ...

\section{Theorem-like environments}\label{sec:theorem-environments}

List of different predefined enumeration styles:

\begin{theorem}\label{testenv-theorem}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{theorem}

\begin{lemma}\label{testenv-lemma}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{lemma}

\begin{corollary}\label{testenv-corollary}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{corollary}

\begin{proposition}\label{testenv-proposition}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{proposition}

\begin{exercise}\label{testenv-exercise}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{exercise}

\begin{definition}\label{testenv-definition}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{definition}

\begin{example}\label{testenv-example}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{example}

\begin{note}\label{testenv-note}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{note}

\begin{note*}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{note*}

\begin{remark}\label{testenv-remark}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{remark}

\begin{remark*}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{remark*}

\begin{claim}\label{testenv-claim}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{claim}

\begin{claim*}\label{testenv-claim2}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{claim*}

\begin{proof}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{proof}

\begin{claimproof}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{claimproof}

\end{document}

### 2020-10-21 papers/types2020/main.tex
Added more text to the introduction of the paper.  I'm still thinking about this and whether I have the right focus in mind.

@@
 \documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
-%This is a template for producing LIPIcs articles. 
-%See lipics-manual.pdf for further information.
-%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-%for section-numbered lemmas etc., use "numberwithinsect"
-%for enabling cleveref support, use "cleveref"
-%for enabling autoref support, use "autoref"
-%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
-%for enabling thm-restate support, use "thm-restate"
-
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
-
-\bibliographystyle{plainurl}% the mandatory bibstyle
+\input preamble
+\input head
 
@@
 
-%TODO mandatory, please use full name; only 1 author per \author
-%macro; first two parameters are mandatory, other parameters can be
-%empty. Please provide at least the name of the affiliation and the
-%country. The full address is optional
-
-\author{Dale Miller}
-       {Inria, Palaiseau, France \and LIX, Ecole Polytechnique, France
-         \and \url{http://www.lix.polytechnique.fr/Labo/Dale.Miller/} }
-       {dale.miller@inria.fr}
-       {}%https://orcid.org/0000-0003-0274-4954
-       {}
-
-\author{Matteo Manighetti}
-       {Inria, Palaiseau, France \and LIX, Ecole Polytechnique, France
-         \and \url{http://www.lix.polytechnique.fr/Labo/Matteo.Manighetti/}}
-       {matteo.manighetti@inria.fr}
-       {}
-       {}
-
-\author{Alberto Momigliano}
-       {Dipartimento di Informatica, Universit\`a degli studi di Milano, Italy \and
-       \url{http://momigliano.di.unimi.it/}}
-       {momigliano@di.unimi.it}
-       {}
-       {}
-
-\authorrunning{D. Miller, M. Manighetti, and A. Momigliano}
- %TODO mandatory. First: Use abbreviated first/middle names. Second
- %(only in severe cases): Use first author plus 'et al.'
-
-\Copyright{D. Miller, M. Manighetti, and A. Momigliano}
-%TODO mandatory, please use full first names. LIPIcs license is
-%"CC-BY";  http://creativecommons.org/licenses/by/3.0/
-
-\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
-
-\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
-
-\category{} %optional, e.g. invited paper
-
-\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
-%\relatedversion{A full version of the paper is available at \url{...}.}
-
-\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
-
-%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
-Please enter author specific funding statements as fifth argument of the \author macro.
-
-\acknowledgements{I want to thank \dots}%optional
-
-%\nolinenumbers %uncomment to disable line numbering
-
-%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \begin{document}
-
 \maketitle
 
-%TODO mandatory: add short abstract of the document
 \begin{abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Praesent convallis orci arcu, eu mollis dolor.
-Aliquam eleifend suscipit lacinia.
-Maecenas quam mi, porta ut lacinia sed, convallis ac dui.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse potenti.
+We do stuff
 \end{abstract}
 
-\section{Typesetting instructions -- Summary}
-\label{sec:typesetting-summary}
+\section{Introduction}
+\label{sec:intro}
 
-LIPIcs is a series of open access high-quality conference proceedings across all fields in informatics established in cooperation with Schloss Dagstuhl. 
-In order to do justice to the high scientific quality of the conferences that publish their proceedings in the LIPIcs series, which is ensured by the thorough review process of the respective events, we believe that LIPIcs proceedings must have an attractive and consistent layout matching the standard of the series.
-Moreover, the quality of the metadata, the typesetting and the layout must also meet the requirements of other external parties such as indexing service, DOI registry, funding agencies, among others.
-The guidelines contained in this document serve as the baseline for the authors, editors, and the publisher to create documents that meet as many different requirements as possible.
+Recently, C. Sacerdoti Coen and E. Tassi developed the ELPI
+implementation \cite{dunchev15lpar} of \lP 
+\cite{miller12proghol}.  More recently, Tassi has made ELPI available
+as a plugin to the Coq proof assistant \cite{tassi18coqpl}.
+%
+This implementation of \lP extends earlier implementations in
+primarily two directions.  First, ELPI adds a notion of contraints,
+which makes it more expressive than, say, the Teyjus implementation of
+\lP \cite{nadathur99cade}.  Second, the plugin version of ELPI comes
+equip with an API for accessing the Coq environment and a quotation
+and anti-quotation syntax allowing for Coq expressions to be mixed in
+with \lP program elements.
 
-Please comply with the following instructions when preparing your article for a LIPIcs proceedings volume. 
-\paragraph*{Minimum requirements}
+ELPI provides a means for treating bindings and unification (of
+structures containing bindings) that are convenient for various
+meta-programming tasks in logic programming all things useful for
+tactic programming in Coq (proof refinement).  For examples,
+ELPI programmers are spared from dealing with low-level aspects of
+representing bindings (e.g., De Bruijn indexes) while still having
+clean and effective ways to manipulate binding structures.
+%
+In short, ELPI appears to be a useful meta-language for Coq
+\cite{coen19mscs,tassi18coqpl,tassi19itp}.
+%
+In this paper, we outline two applications of the Coq plugin for ELPI
+that support this claim.
+%
+With these examples, we shall illustrate that ELPI is a useful
+meta-language not just because for its treatment of bindings and
+syntactic structures in general but also because it is actually based
+on a sound implementation of a higher-order intuitionistic logic.
+%
+Type inference is a well-known example of a meta-programming task for
+which logic programming often provides immediate and elegant
+implementations via the sound implementation of logic.
+%
+In this paper, we present two additional project in which an
+implementation of proof search provides direct, elegant, and compact
+implementations of meta-programming tasks.
+%
+Before we present these two examples, we first discuss the different
+world views that the integration of ELPI and Coq focuses us to
+confront.
 
-\begin{itemize}
-\item Use pdflatex and an up-to-date \LaTeX{} system.
-\item Use further \LaTeX{} packages and custom made macros carefully and only if required.
-\item Use the provided sectioning macros: \verb+\section+, \verb+\subsection+, \verb+\subsubsection+, \linebreak \verb+\paragraph+, \verb+\paragraph*+, and \verb+\subparagraph*+.
-\item Provide suitable graphics of at least 300dpi (preferably in PDF format).
-\item Use BibTeX and keep the standard style (\verb+plainurl+) for the bibliography.
-\item Please try to keep the warnings log as small as possible. Avoid overfull \verb+\hboxes+ and any kind of warnings/errors with the referenced BibTeX entries.
-\item Use a spellchecker to correct typos.
-\end{itemize}
+\section{Two cultures}
 
-\paragraph*{Mandatory metadata macros}
-Please set the values of the metadata macros carefully since the information parsed from these macros will be passed to publication servers, catalogues and search engines.
-Avoid placing macros inside the metadata macros. The following metadata macros/environments are mandatory:
-\begin{itemize}
-\item \verb+\title+ and, in case of long titles, \verb+\titlerunning+.
-\item \verb+\author+, one for each author, even if two or more authors have the same affiliation.
-\item \verb+\authorrunning+ and \verb+\Copyright+ (concatenated author names)\\
-The \verb+\author+ macros and the \verb+\Copyright+ macro should contain full author names (especially with regard to the first name), while \verb+\authorrunning+ should contain abbreviated first names.
-\item \verb+\ccsdesc+ (ACM classification, see \url{https://www.acm.org/publications/class-2012}).
-\item \verb+\keywords+ (a comma-separated list of keywords).
-\item \verb+\relatedversion+ (if there is a related version, typically the ``full version''); please make sure to provide a persistent URL, e.\,g., at arXiv.
-\item \verb+\begin{abstract}...\end{abstract}+ .
-\end{itemize}
+{\color{red} The rest of this section are just notes.}
 
-\paragraph*{Please do not \ldots} %Do not override the \texttt{\seriesstyle}-defaults}
-Generally speaking, please do not override the \texttt{lipics-v2019}-style defaults.
-To be more specific, a short checklist also used by Dagstuhl Publishing during the final typesetting is given below.
-In case of \textbf{non-compliance} with these rules Dagstuhl Publishing will remove the corresponding parts of \LaTeX{} code and \textbf{replace it with the \texttt{lipics-v2019} defaults}.
-In serious cases, we may reject the LaTeX-source and expect the corresponding author to revise the relevant parts.
-\begin{itemize}
-\item Do not use a different main font. (For example, the \texttt{times} package is forbidden.)
-\item Do not alter the spacing of the \texttt{lipics-v2019.cls} style file.
-\item Do not use \verb+enumitem+ and \verb+paralist+. (The \texttt{enumerate} package is preloaded, so you can use
- \verb+\begin{enumerate}[(a)]+ or the like.)
-\item Do not use ``self-made'' sectioning commands (e.\,g., \verb+\noindent{\bf My+ \verb+Paragraph}+).
-\item Do not hide large text blocks using comments or \verb+\iffalse+ $\ldots$ \verb+\fi+ constructions. 
-\item Do not use conditional structures to include/exclude content. Instead, please provide only the content that should be published -- in one file -- and nothing else.
-\item Do not wrap figures and tables with text. In particular, the package \texttt{wrapfig} is not supported.
-\item Do not change the bibliography style. In particular, do not use author-year citations. (The
-\texttt{natbib} package is not supported.)
-\end{itemize}
+These applications are not necessarily focused on the underlying
+declarative nature of lambda Prolog.  For example, ELPI has various
+extensions, such as ``spilling'' that are designed to make some
+aspects of logic programming appear to be more functional.
 
-\enlargethispage{\baselineskip}
+In this paper, we limit our attention mostly to using lambda Prolog
+for declarative processing: i.e., where logic plays a central role in
+describing operations.  (Type checking is an example of such
+processing, but we do not focus on type checking here).
 
-This is only a summary containing the most relevant details.
-Please read the complete document ``LIPIcs: Instructions for Authors and the \texttt{lipics-v2019} Class'' for all details and don't hesitate to contact Dagstuhl Publishing (\url{mailto:publishing@dagstuhl.de}) in case of questions or comments:
-\href{http://drops.dagstuhl.de/styles/lipics-v2019/lipics-v2019-authors/lipics-v2019-authors-guidelines.pdf}{\texttt{http://drops.dagstuhl.de/styles/lipics-v2019/\newline lipics-v2019-authors/lipics-v2019-authors-guidelines.pdf}}
+Of course, being able to do declarative and not-so-declarative
+operations within the same language makes that language all the more
+useful.
 
-\section{Lorem ipsum dolor sit amet}
+  Given that lP and Coq are both examples of mixing the
+  lambda-calculus with logic, it is important to understand their
+  difference, something that can be very confusing.  Take the
+  confusion around the term HOAS.  In some circles, this is a
+  appropriate and sensible approach.  In other circles, it is
+  semantically questionable.  But, in fact, this concept (using
+  meta-level binding for object-level binding) plays our 
+  differently depending on the meta language!
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit \cite{DBLP:journals/cacm/Knuth74}.
-Praesent convallis orci arcu, eu mollis dolor.
-Aliquam eleifend suscipit lacinia.
-Maecenas quam mi, porta ut lacinia sed, convallis ac dui.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse potenti.
-Donec eget odio et magna ullamcorper vehicula ut vitae libero.
-Maecenas lectus nulla, auctor nec varius ac, ultricies et turpis.
-Pellentesque id ante erat.
-In hac habitasse platea dictumst.
-Curabitur a scelerisque odio.
-Pellentesque elit risus, posuere quis elementum at, pellentesque ut diam.
-Quisque aliquam libero id mi imperdiet quis convallis turpis eleifend.
+  The origin story to ML (Meta language) is that it supported
+  LFC-style tactics and tacticals.  The strong typing of ML provided
+  some guarantees: only theorems were ever given type thm.  But many
+  steps of evolution have now taken place since then.  Isabelle's
+  different approach to tactics.  Logic and proof theory, when used
+  with Church's simple theory of types and Gentzen's sequent calculus,
+  provides a far richer meta language: that provides an immediate and
+  declarative treatment of bindings as well as for proof search
+  (without the need of exception handling, etc).  [I'm making the case
+  that FP is not a particularly natural framework for proof
+  assistants.  Of course, lots of libraries have now been built in the
+  FP style.]
 
-\begin{lemma}[Lorem ipsum]
-\label{lemma:lorem}
-Vestibulum sodales dolor et dui cursus iaculis.
-Nullam ullamcorper purus vel turpis lobortis eu tempus lorem semper.
-Proin facilisis gravida rutrum.
-Etiam sed sollicitudin lorem.
-Proin pellentesque risus at elit hendrerit pharetra.
-Integer at turpis varius libero rhoncus fermentum vitae vitae metus.
-\end{lemma}
+  - Various dichotomies we find in this work.
+    - proof theory vs type theory
+    - logic programming vs functional programming
+    - sequent calculus vs natural deduction
 
-\begin{proof}
-Cras purus lorem, pulvinar et fermentum sagittis, suscipit quis magna.
+  - While both lambda Prolog and Coq are based on an amalgamation of
+    $\lambda$-calculus and logic, these two languages take strikingly
+    different approaches to bindings.  In lambda Prolog, bindings are
+    syntactic and extensionality at any type is not available.  In
+    Coq, however, higher-order typed quantification encodes
+    extensional functions.  In lambda Prolog, (exists w. [lambda x. x]
+    = [lambda x. w]) fails, no matter the type of w.  If this query
+    was about functions, then this equation would be true when the
+    type of w is a singleton.  But in lP, this is a query only about
+    binding structures, and this fails to be proved since there is no
+    capture-avoiding substitution (the only substitution supported by
+    Church's STT) for w that makes the resulting two terms alpha,
+    beta, eta convertible.  [Is this true in Coq?  What is equality
+    here?]
 
-\begin{claim}
-content...
-\end{claim}
-\begin{claimproof}
-content...
-\end{claimproof}
+  - On the proof theory side of this work, we use two different proof
+    systems.
 
-\end{proof}
+    - LJF, here we have undefined predicates, no induction
+    - muMALL, were we have inductive definitions of predicates
 
-\begin{corollary}[Curabitur pulvinar, \cite{DBLP:books/mk/GrayR93}]
-\label{lemma:curabitur}
-Nam liber tempor cum soluta nobis eleifend option congue nihil imperdiet doming id quod mazim placerat facer possim assum.
-Lorem ipsum dolor sit amet, consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna aliquam erat volutpat.
-\end{corollary}
+    The Coq world provides for both of these kinds of predicate using
+    Prop (corresponds to undefined) and Set (inductively defined).
 
-\begin{proposition}\label{prop1}
-This is a proposition
-\end{proposition}
+\section{Elaboration of external proof certificates for the Coq
+  kernel}
 
-\autoref{prop1} and \cref{prop1} \ldots
+Follow the text in \cite{blanco20coq}.
 
-\subsection{Curabitur dictum felis id sapien}
+\section{Property based testing for Coq}
 
-Curabitur dictum \cref{lemma:curabitur} felis id sapien \autoref{lemma:curabitur} mollis ut venenatis tortor feugiat.
-Curabitur sed velit diam.
-Integer aliquam, nunc ac egestas lacinia, nibh est vehicula nibh, ac auctor velit tellus non arcu.
-Vestibulum lacinia ipsum vitae nisi ultrices eget gravida turpis laoreet.
-Duis rutrum dapibus ornare.
-Nulla vehicula vulputate iaculis.
-Proin a consequat neque.
-Donec ut rutrum urna.
-Morbi scelerisque turpis sed elit sagittis eu scelerisque quam condimentum.
-Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.
-Aenean nec faucibus leo.
-Cras ut nisl odio, non tincidunt lorem.
-Integer purus ligula, venenatis et convallis lacinia, scelerisque at erat.
-Fusce risus libero, convallis at fermentum in, dignissim sed sem.
-Ut dapibus orci vitae nisl viverra nec adipiscing tortor condimentum \cite{DBLP:journals/cacm/Dijkstra68a}.
-Donec non suscipit lorem.
-Nam sit amet enim vitae nisl accumsan pretium.
+\section{Conclusion}
 
-\begin{lstlisting}[caption={Useless code},label=list:8-6,captionpos=t,float,abovecaptionskip=-\medskipamount]
-for i:=maxint to 0 do 
-begin 
-    j:=square(root(i));
-end;
-\end{lstlisting}
-
-\subsection{Proin ac fermentum augue}
-
-Proin ac fermentum augue.
-Nullam bibendum enim sollicitudin tellus egestas lacinia euismod orci mollis.
-Nulla facilisi.
-Vivamus volutpat venenatis sapien, vitae feugiat arcu fringilla ac.
-Mauris sapien tortor, sagittis eget auctor at, vulputate pharetra magna.
-Sed congue, dui nec vulputate convallis, sem nunc adipiscing dui, vel venenatis mauris sem in dui.
-Praesent a pretium quam.
-Mauris non mauris sit amet eros rutrum aliquam id ut sapien.
-Nulla aliquet fringilla sagittis.
-Pellentesque eu metus posuere nunc tincidunt dignissim in tempor dolor.
-Nulla cursus aliquet enim.
-Cras sapien risus, accumsan eu cursus ut, commodo vel velit.
-Praesent aliquet consectetur ligula, vitae iaculis ligula interdum vel.
-Integer faucibus faucibus felis.
-
-\begin{itemize}
-\item Ut vitae diam augue. 
-\item Integer lacus ante, pellentesque sed sollicitudin et, pulvinar adipiscing sem. 
-\item Maecenas facilisis, leo quis tincidunt egestas, magna ipsum condimentum orci, vitae facilisis nibh turpis et elit. 
-\end{itemize}
-
-\begin{remark}
-content...
-\end{remark}
-
-\section{Pellentesque quis tortor}
-
-Nec urna malesuada sollicitudin.
-Nulla facilisi.
-Vivamus aliquam tempus ligula eget ornare.
-Praesent eget magna ut turpis mattis cursus.
-Aliquam vel condimentum orci.
-Nunc congue, libero in gravida convallis \cite{DBLP:conf/focs/HopcroftPV75}, orci nibh sodales quam, id egestas felis mi nec nisi.
-Suspendisse tincidunt, est ac vestibulum posuere, justo odio bibendum urna, rutrum bibendum dolor sem nec tellus.
-
-\begin{lemma} [Quisque blandit tempus nunc]
-Sed interdum nisl pretium non.
-Mauris sodales consequat risus vel consectetur.
-Aliquam erat volutpat.
-Nunc sed sapien ligula.
-Proin faucibus sapien luctus nisl feugiat convallis faucibus elit cursus.
-Nunc vestibulum nunc ac massa pretium pharetra.
-Nulla facilisis turpis id augue venenatis blandit.
-Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus.
-\end{lemma}
-
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-
-\section{Morbi eros magna}
-
-Morbi eros magna, vestibulum non posuere non, porta eu quam.
-Maecenas vitae orci risus, eget imperdiet mauris.
-Donec massa mauris, pellentesque vel lobortis eu, molestie ac turpis.
-Sed condimentum convallis dolor, a dignissim est ultrices eu.
-Donec consectetur volutpat eros, et ornare dui ultricies id.
-Vivamus eu augue eget dolor euismod ultrices et sit amet nisi.
-Vivamus malesuada leo ac leo ullamcorper tempor.
-Donec justo mi, tempor vitae aliquet non, faucibus eu lacus.
-Donec dictum gravida neque, non porta turpis imperdiet eget.
-Curabitur quis euismod ligula.
-
-%%
-%% Bibliography
-%%
-
-%% Please use bibtex,
-
-\bibliography{lipics-v2019-sample-article}
-
-\appendix
-
-\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}
-
-List of different predefined enumeration styles:
-
-\begin{itemize}
-\item \verb|\begin{itemize}...\end{itemize}|
-\item \dots
-\item \dots
-%\item \dots
-\end{itemize}
-
-\begin{enumerate}
-\item \verb|\begin{enumerate}...\end{enumerate}|
-\item \dots
-\item \dots
-%\item \dots
-\end{enumerate}
-
-\begin{alphaenumerate}
-\item \verb|\begin{alphaenumerate}...\end{alphaenumerate}|
-\item \dots
-\item \dots
-%\item \dots
-\end{alphaenumerate}
-
-\begin{romanenumerate}
-\item \verb|\begin{romanenumerate}...\end{romanenumerate}|
-\item \dots
-\item \dots
-%\item \dots
-\end{romanenumerate}
-
-\begin{bracketenumerate}
-\item \verb|\begin{bracketenumerate}...\end{bracketenumerate}|
-\item \dots
-\item \dots
-%\item \dots
-\end{bracketenumerate}
-
-\begin{description}
-\item[Description 1] \verb|\begin{description} \item[Description 1]  ...\end{description}|
-\item[Description 2] Fusce eu leo nisi.
-Cras eget orci neque, eleifend dapibus felis.
-Duis et leo dui.
-Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\item[Description 3]  \dots
-%\item \dots
-\end{description}
-
-\cref{testenv-proposition} and \autoref{testenv-proposition} ...
-
-\section{Theorem-like environments}\label{sec:theorem-environments}
-
-List of different predefined enumeration styles:
-
-\begin{theorem}\label{testenv-theorem}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{theorem}
-
-\begin{lemma}\label{testenv-lemma}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{lemma}
-
-\begin{corollary}\label{testenv-corollary}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{corollary}
-
-\begin{proposition}\label{testenv-proposition}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{proposition}
-
-\begin{exercise}\label{testenv-exercise}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{exercise}
-
-\begin{definition}\label{testenv-definition}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{definition}
-
-\begin{example}\label{testenv-example}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{example}
-
-\begin{note}\label{testenv-note}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{note}
-
-\begin{note*}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{note*}
-
-\begin{remark}\label{testenv-remark}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{remark}
-
-\begin{remark*}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{remark*}
-
-\begin{claim}\label{testenv-claim}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{claim}
-
-\begin{claim*}\label{testenv-claim2}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{claim*}
-
-\begin{proof}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{proof}
-
-\begin{claimproof}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{claimproof}
+% Not part of git repo.  Need to find a fix
+\bibliography{../../../../repo/parsifal/papers/references/master}
 

### 2020-10-21 papers/types2020/preamble.tex
Added more text to the introduction of the paper.  I'm still thinking about this and whether I have the right focus in mind.

%This is a template for producing LIPIcs articles. 
%See lipics-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

%TODO mandatory, please use full name; only 1 author per \author
%macro; first two parameters are mandatory, other parameters can be
%empty. Please provide at least the name of the affiliation and the
%country. The full address is optional

\author{Matteo Manighetti}
       {Inria, Palaiseau, France \and LIX, Ecole Polytechnique, France
         \and \url{http://www.lix.polytechnique.fr/Labo/Matteo.Manighetti/}}
       {matteo.manighetti@inria.fr}
       {}
       {}

\author{Dale Miller}
       {Inria, Palaiseau, France \and LIX, Ecole Polytechnique, France
         \and \url{http://www.lix.polytechnique.fr/Labo/Dale.Miller/} }
       {dale.miller@inria.fr}
       {}%https://orcid.org/0000-0003-0274-4954
       {}

\author{Alberto Momigliano}
       {Dipartimento di Informatica, Universit\`a degli studi di Milano, Italy \and
       \url{http://momigliano.di.unimi.it/}}
       {momigliano@di.unimi.it}
       {}
       {}

\authorrunning{M. Manighetti, D. Miller, and A. Momigliano}
 %TODO mandatory. First: Use abbreviated first/middle names. Second
 %(only in severe cases): Use first author plus 'et al.'

\Copyright{M. Manighetti, D. Miller, and A. Momigliano}
%TODO mandatory, please use full first names. LIPIcs license is
%"CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} 
%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Proof assistant, logic programming, Coq, \lP}
%TODO mandatory; please add comma-separated list of keywords

\category{} %optional, e.g. invited paper

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversion{A full version of the paper is available at \url{...}.}

\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

%\acknowledgements{I want to thank \dots}%optional

\nolinenumbers %uncomment to disable line numbering

\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-10-22 papers/types2020/main.tex
Some progress on introduction and motivation.   We need to discuss.

@@
+%% Important dates:
+%%  Paper submission: 19 October 31 October 2020
+%%  Author notification: 18 January 2021
+%%  Final version:  15 February 2021
+%%  Publication (presumably): end of March 2021
+
+%% Style requirements of LIPIcs.  The recommended length of a paper is
+%% 12-15 pages, excluding front-page(s) (authors, affiliation, keywords,
+%% abstract, ...), bibliography and an appendix of max 5 pages. If you
+%% need more pages, please ask the editors.
+
+%% Papers have to be submitted in pdf through EasyChair:
+%% https://easychair.org/conferences/?conf=types2020postproceed
+
 \documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
+\input head
 \input preamble
-\input head
 
@@
 Recently, C. Sacerdoti Coen and E. Tassi developed the ELPI
-implementation \cite{dunchev15lpar} of \lP 
-\cite{miller12proghol}.  More recently, Tassi has made ELPI available
-as a plugin to the Coq proof assistant \cite{tassi18coqpl}.
-%
-This implementation of \lP extends earlier implementations in
-primarily two directions.  First, ELPI adds a notion of contraints,
-which makes it more expressive than, say, the Teyjus implementation of
-\lP \cite{nadathur99cade}.  Second, the plugin version of ELPI comes
-equip with an API for accessing the Coq environment and a quotation
-and anti-quotation syntax allowing for Coq expressions to be mixed in
-with \lP program elements.
+implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol}.
+More recently, Tassi has made ELPI available as a plugin to the Coq
+proof assistant \cite{tassi18coqpl}.  This implementation of \lP
+extends earlier implementations in primarily two directions.  First,
+ELPI adds a notion of constraints, which makes it more expressive than,
+say, the Teyjus implementation of \lP \cite{nadathur99cade}.  Second,
+the plugin version of ELPI comes equip with an API for accessing the
+Coq environment and a quotation and anti-quotation syntax allowing for
+Coq expressions to be mixed in with \lP program elements.
 
@@
 meta-programming tasks in logic programming all things useful for
-tactic programming in Coq (proof refinement).  For examples,
-ELPI programmers are spared from dealing with low-level aspects of
+tactic programming in Coq (proof refinement).  For examples, ELPI
+programmers are spared from dealing with low-level aspects of
 representing bindings (e.g., De Bruijn indexes) while still having
-clean and effective ways to manipulate binding structures.
-%
-In short, ELPI appears to be a useful meta-language for Coq
-\cite{coen19mscs,tassi18coqpl,tassi19itp}.
-%
-In this paper, we outline two applications of the Coq plugin for ELPI
-that support this claim.
-%
+clean and effective ways to manipulate binding structures.  In short,
+ELPI appears to be a useful meta-language for Coq
+\cite{coen19mscs,tassi18coqpl,tassi19itp}.  In this paper, we outline
+two applications of the Coq plugin for ELPI that support this claim.
 With these examples, we shall illustrate that ELPI is a useful
@@
 on a sound implementation of a higher-order intuitionistic logic.
-%
 Type inference is a well-known example of a meta-programming task for
 which logic programming often provides immediate and elegant
-implementations via the sound implementation of logic.
-%
-In this paper, we present two additional project in which an
-implementation of proof search provides direct, elegant, and compact
-implementations of meta-programming tasks.
-%
-Before we present these two examples, we first discuss the different
-world views that the integration of ELPI and Coq focuses us to
-confront.
+implementations via the sound implementation of logic.  In this paper,
+we present two additional project in which an implementation of proof
+search provides direct, elegant, and compact implementations of
+meta-programming tasks.  Before we present these two examples, we
+first discuss the different world views that the integration of ELPI
+and Coq forces us to confront.
 
@@
 
-{\color{red} The rest of this section are just notes.}
+In proof theory, one learns quickly that things always come in pairs:
+negative/positive, left/right, bottom-up/top-down,
+premises/conclusion, introduction/elimination,
+% determinism/nondeterminism,
+% invertible/noninvertible, 
+etc.
+%
+When we examine the larger setting of this project of linking a logic
+programming engine with Coq and its kernel, we find a large number of
+pairing that are valuable to explicitly describe.
 
-These applications are not necessarily focused on the underlying
-declarative nature of lambda Prolog.  For example, ELPI has various
-extensions, such as ``spilling'' that are designed to make some
-aspects of logic programming appear to be more functional.
+\subsection{Proof theory vs type theory}
 
-In this paper, we limit our attention mostly to using lambda Prolog
-for declarative processing: i.e., where logic plays a central role in
-describing operations.  (Type checking is an example of such
-processing, but we do not focus on type checking here).
+In many ways, (structural) proof theory is more elementary and
+low-level than most approaches to type theory.  For example, type
+theories usually answer the question ``What is a proof?'' with the
+response ``a dependently typed $\lambda$-term''.  That is, when
+describing a type theory, one usually decides that a proof is a
+certain kind of term within the system.  In contrast, the proof theory
+treats logical propositions and proofs as separate.  For example,
+proof theory does not assume that there are terms within the logic
+that describe proofs.
 
-Of course, being able to do declarative and not-so-declarative
-operations within the same language makes that language all the more
-useful.
+Gentzen's discovery that the key to treating classical and
+intuitionistic logics in the same proof system was the invention of
+the weakening and contraction structural rules on the right-side of
+sequents~\cite{gentzen35}.  This discovery lead him away from natural
+deduction to the invention of multiple-conclusion sequent calculus, a
+very important kind of proof that has not been given a satisfactory
+representation in type theory.  This same innovation of Gentzen also
+opened the way to the invention of another key proof theoretic
+discovery, that of linear logic \cite{girard87tcs}.
+Similarly, sequent calculus provides an elegant presentation for
+linear logic while most treatments of linear logic in type theory
+usually only capture linearity in a partial fashion.
 
-  Given that lP and Coq are both examples of mixing the
-  lambda-calculus with logic, it is important to understand their
-  difference, something that can be very confusing.  Take the
-  confusion around the term HOAS.  In some circles, this is a
-  appropriate and sensible approach.  In other circles, it is
-  semantically questionable.  But, in fact, this concept (using
-  meta-level binding for object-level binding) plays our 
-  differently depending on the meta language!
+Proof theory also has rather immediate and natural treatments for
+co-induction (as well as induction: see
+\cite{baelde12tocl,momigliano12jal}) whereas the treatment of
+coinduction within type theory remains more challenging and is well
+developed only for various syntactic conditions (see, for example,
+\cite{bertot08entcs}).
 
-  The origin story to ML (Meta language) is that it supported
-  LFC-style tactics and tacticals.  The strong typing of ML provided
-  some guarantees: only theorems were ever given type thm.  But many
-  steps of evolution have now taken place since then.  Isabelle's
-  different approach to tactics.  Logic and proof theory, when used
-  with Church's simple theory of types and Gentzen's sequent calculus,
-  provides a far richer meta language: that provides an immediate and
-  declarative treatment of bindings as well as for proof search
-  (without the need of exception handling, etc).  [I'm making the case
-  that FP is not a particularly natural framework for proof
-  assistants.  Of course, lots of libraries have now been built in the
-  FP style.]
+\subsection{Proof search vs proof normalization}
 
-  - Various dichotomies we find in this work.
-    - proof theory vs type theory
-    - logic programming vs functional programming
-    - sequent calculus vs natural deduction
+Gentzen-style proofs are used to model computation in at least two
+different ways.  The functional programming paradigm, following the
+Curry-Howard correspondence, models computation abstractly as the
+$\beta$-reduction of natural deduction proof: that is, computation is
+modeled using \emph{proof normalization}.  On the other hand, the
+logic programming paradigm, following the notion of goal-directed
+proof search, models computation abstractly as a regimented search of
+cut-free sequent calculus proofs: that is, computation is modeled
+using \emph{proof search}.
 
-  - While both lambda Prolog and Coq are based on an amalgamation of
-    $\lambda$-calculus and logic, these two languages take strikingly
-    different approaches to bindings.  In lambda Prolog, bindings are
-    syntactic and extensionality at any type is not available.  In
-    Coq, however, higher-order typed quantification encodes
-    extensional functions.  In lambda Prolog, (exists w. [lambda x. x]
-    = [lambda x. w]) fails, no matter the type of w.  If this query
-    was about functions, then this equation would be true when the
-    type of w is a singleton.  But in lP, this is a query only about
-    binding structures, and this fails to be proved since there is no
-    capture-avoiding substitution (the only substitution supported by
-    Church's STT) for w that makes the resulting two terms alpha,
-    beta, eta convertible.  [Is this true in Coq?  What is equality
-    here?]
+Proof search has proved to have some richness that is hard to capture
+in the proof normalization setting.  In particular, Gentzen's
+invention of \emph{eigenvariables} are a kind of a proof-level
+binder.  In the proof normalization setting, such eigenvariables
+are instantiated during proof normalization.  In contrast, during the
+search for cut-free proofs, eigenvariables are part of the syntax of
+terms and formulas.  As a result, they can be used in the treatment of
+bindings in data structures more generally.  Hence, the logic
+programming paradigm provides a completely natural, elegant, and
+powerful method, called $\lambda$-tree syntax, for treating bindings
+within syntax.
 
-  - On the proof theory side of this work, we use two different proof
-    systems.
+In this paper, we shall use two different proof systems.  The first
+logic is first-order intuitionistic logic.  Here, predicates
+are not inductively defined and quantification will be limited to
+range over first-order term.  Actually, we shall examine only the Horn
+clause fragment of that logic.
+%
+The second logic will allow for inductively defined predicates using a
+fixed point notation.   Given again our restriction to essentially
+Horn clause specifications, it is possible to view intuitionistic
+reasoning with such specification as if they are actually
+specification in $\mu$MALL, an extension of multiplicative additive
+linear logic with the least and greatest fixed point operators
+\cite{baelde12tocl,baelde07lpar,heath19jar}.
+%
+The difference between these two logics and associate proof theories
+is reminiscent to the difference between the type \lsti{Prop}
+(corresponding to undefined propositions) and Sets (corresponding to
+inductively defined propositions).
 
-    - LJF, here we have undefined predicates, no induction
-    - muMALL, were we have inductive definitions of predicates
+\subsection{\lP vs Coq}
 
-    The Coq world provides for both of these kinds of predicate using
-    Prop (corresponds to undefined) and Set (inductively defined).
+Given that \lP and Coq are both examples of mixing the
+$\lambda$-calculus with logic, it is important to understand their
+differences.  The confusion around the term \emph{higher-order
+abstract syntax} (HOAS), for example, helps to illustrate their
+differences.  In the functional programming setting, including the Coq
+system, the HOAS approach leads to encoding binding structures within
+terms using functions.  Such encodings have results in surprisingly
+complicated encodings, allowing, for example, for \emph{exotic terms}
+\cite{despeyroux95tlca} and to structures on which induction is not
+possible \cite{roeckl01fossacs}.  In the logic programming setting,
+particularly \lP, HOAS is well supported in both the availability of
+binders allowed with terms ($\lambda$-bindings), in formulas
+(quantifiers), and in proofs (eigenvariables).  For this reason, the
+term $\lambda$-tree syntax was introduced to name this particular take
+on HOAS \cite{miller19jar}.  The Abella proof assistant
+\cite{baelde14jfr} was designed, in part, to allow for inductive and
+co-inductive proofs for specifications using the $\lambda$-tree syntax
+approach.
+
+Another difference between \lP and functional programming can be
+illustrated by considering how they are used in the specification of
+tactics that are popular in modern proof assistants.  In fact, the
+origin story for the ML functional programming language was that ML
+was the meta-language for implementing the LCF suite of tactics and
+tacticals \cite{gordon79}.  In order to implement tactics, ML adopted
+not only novel features such as polymorphically typed higher-order
+functional programming but also the non-functional mechanisms of
+failure and exception handling.  While \lP is also based on
+polymorphically typed higher-order relational programming, it also
+comes with a completely declarative version of failure and
+backtracking search.  Combining those features along with its support
+of unification (even in the presence of term-level bindings), \lP
+provides a rather different approach to the specification of tactics
+\cite{felty93jar}.
+
+While \lP is a typed language, it only uses simple types and these are
+more often used to indicate syntactic categories.  Any dependency
+information needs to be captured using predicates and quantifiers.
+
+\section{Proof theory and proof certificates}
+
+{\color{red} The goal in this section is to say a bit about focusing
+  generally, and to give a bit of detail about FPC for Horn clauses.
+  Initial draft of text taken from \cite{blanco19ppdp}.
+}
+
+\subsection{Focused proof systems}
+\label{ssec:focused}
+
+The proof search approach to encoding Horn clause computation
+results in the structuring of proofs with repeated
+switching between a \emph{goal-reduction} phase and a
+\emph{backchaining} phase~\cite{miller91apal}.
+%
+The notion of \emph{focused proof systems}~\cite{andreoli92jlc,liang09tcs}
+generalizes this view of proof construction by identifying the
+following two phases. % of proof construction.
+\begin{enumerate}
+\item The \emph{negative}%
+\footnote{The terminology of negative and positive phases is a bit
+  unfortunate: historically, these terms do not refer to positive or negative
+  subformula occurrences but rather to certain semantic models used in
+  the study of linear logic~\cite{girard87tcs}.}
+%
+  phase corresponds to  goal-reduction: in this phase,
+  inference rules that involve \emph{don't-care-non\-de\-ter\-min\-ism} are
+  applied.
+%
+  As a result, there is no need to consider backtracking over choices
+  made in building this phase.
+
+\item The \emph{positive} phase corresponding to  backchaining: in this phase, inference rules that involve
+  \emph{don't-know-non\-de\-ter\-min\-ism} are applied: here,
+  inference rules need to be supplied with information in order to
+  ensure that a completed proof can be built.
+%
+  That information can be items such as which term is needed to
+  instantiate a universally quantified formula and which disjunct of a
+  disjunctive goal formula should be proved.
+\end{enumerate}
+%
+Thus, when building a proof tree (in the sequent calculus) from the
+conclusion to its leaves, the negative phase corresponds to a simple
+computation that needs no external information, while the positive
+phase may need such external information to be supplied.
+%
+In the literature, it is common to refer to a repository of such
+external information as either an \emph{oracle} or a \emph{proof
+certificate}.
+
+When using a focused proof system for logic extended with fixed
+points, such as  employed in Bedwyr~\cite{baelde07cade} and described
+in~\cite{baelde12tocl,heath17linearity}, proofs of formulas such as
+\[
+  \exists x [(\tau(x)\wedge P(x)) \wedge \neg Q(x)]
+  \tag{*}\label{eq:full}
+\]
+are a single \emph{bipole}: that is, when reading a proof bottom up, a
+positive phase is followed on all its premises by a single negative
+phase that completes the proof.%
+\footnote{The reader familiar with focusing will understand that there
+are two ``polarized'' conjunctions, written in linear logic as
+$\otimes$ and $\with$ or in classical and intuitionistic logics as
+$\wedgep$ and $\wedgen$, respectively.  In this paper, we use simply
+$\wedge$ to denote the positive biased conjunction.}
+%
+In particular, the positive phase corresponds to the \emph{generation} phase
+and the negative phase corresponds to the \emph{testing} phase.
+%
+
+Instead of giving a full focused proof system of a logic including
+fixed points (since, as we will argue, that proof system will not, in
+fact, be needed to account for PBT ), we offer the following analogy.
+%
+Suppose that we are given a finite search tree and we are asked to
+prove that there is a secret located in one of the nodes of that
+tree.
+%
+A proof that we have found that secret can be taken to be a
+description of the path to that node from the root of the tree: that
+path can be seen as the proof certificate for that claim.
+%
+On the other hand, a proof that no node contains the secret is a
+rather different thing: here, one expects to use a blind and
+exhaustive search (via, say, depth-first or breath-first search) and
+that the result of that search never discovers the secret.
+%
+A proof of this fact requires \emph{no} external information: instead it
+requires a lot of computation involved with exploring the tree until
+exhaustion.
+%
+This follows the familiar pattern where the positive (generate) phase
+requires external information while the negative (testing) phase
+requires none.
+%
+Thus, a proof certificate for $(\ref{eq:full})$ is also a proof
+certificate for 
+\[
+  \exists x [\tau(x)\wedge P(x)].
+  \tag{**}\label{eq:short}
+\]
+Such a certificate would contain the witness (closed) term $t$ for the
+existential quantifier and sufficient information to confirm that
+$P(t)$ can be proved (a proof of a typing judgment such as $\tau(t)$
+is usually trivial).
+%
+%In short\ednote{AM: a bit repetitious}, a proof certificate for
+%$(\ref{eq:short})$ can play the part 
+%of a proof certificate also for $(\ref{eq:full})$, if we also use
+%negation-as-finite-failure  with $Q(t)$.
+%  ****
+Since a proof certificate for the existence-of-a-counterexample formula
+$(\ref{eq:full})$ can be taken as a proof certificate of
+$(\ref{eq:short})$, then we only need to consider proof certificates
+for Horn clause programs.
+%
+%Since the logic of Horn clauses is rather simple,
+We illustrate next what such proofs and proof certificates look like
+for the rather simple logic of Horn clauses.
+
+\subsection{Certificate checking with expert predicates}
+%\subsection{Adding certificates and experts to a proof system}
+\label{ssec:fpc}
+
+%% \begin{metanote}
+%%   AM: I know this is well known and boring, but using \texttt{prog A
+%%     G} in Fig \ref{fig:augmented} is imprecise --- in the comment of
+%%   next figure we mention \texttt{A :- G} and we gloss over the
+%%   instantiation of \texttt{prog A' G} to match the goal \texttt{A},
+%%   which we assume to be ground, given the exists rule
+%% \end{metanote}
+
+%% DM I've modified the text a bit to hopefully improve the state of
+%% things here.  The formal proof theory is all about closed
+%% formulas.  The prolog implementation allows open formulas.  This
+%% difference invades most aspects of proof checking, not just the
+%% prog predicate.
+
+% DM We should find a place to smooth over this issue (which is not
+% hard, I think).
+
+\begin{figure}
+\newcommand{\XXi}{{\color{blue}{\Xi}}}
+\[
+\infer{\XXi\vdash t = t}
+      {\eqExpert{\XXi}}
+\qquad
+\infer{\XXi\vdash \true}
+      {\trueExpert{\XXi}}
+\qquad
+\infer{\XXi\vdash G_1\vee G_2}
+      {\XXi'\vdash G_i\qquad \orExpert{\XXi}{\XXi'}{i}}
+\qquad
+\infer{\XXi\vdash \exists x. G}
+      {\XXi'\vdash G[t/x]\qquad \someExpert{\XXi}{\XXi'}{t}}
+\]
+\vskip -6pt
+\[
+\infer{\XXi\vdash G_1\wedge G_2}
+      {\XXi_1\vdash G_1\quad \XXi_2\vdash G_2\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}}
+\qquad
+\infer{\XXi\vdash A}
+      {\XXi'\vdash G \quad (A~\hbox{\tt :-}~G)\in\instan\Pscr
+                     \quad \unfoldExpert{\XXi}{\XXi'}}
+\]
+\caption{A proof system augmented with proof certificates and
+  additional ``expert'' premises.}
+\label{fig:augmented}
+\lstinputlisting[linerange={sigs-end}]{code/kernel.sig}
+\lstinputlisting[linerange={check-end}]{code/kernel.mod}
+\caption{A simple proof checking kernel.}
+\label{fig:kernel}
+\end{figure}
+
+Figure~\ref{fig:augmented} contains a simple proof system for Horn
+clause provability in which each inference rule is augmented with an
+additional premise involving an \emph{expert predicate}, a
+certificate $\Xi$, and possibly continuations of certificates ($\Xi'$,
+$\Xi_1$, $\Xi_2$) with extracted information from certificates (in the
+case of $\vee$ and $\exists$).
+%
+%
+The premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$ in the last
+inference rule of Figure~\ref{fig:augmented} states that the logic
+programming clause $(A~\hbox{\tt :-}~G)$ is a ground instance of some
+clause in a fixed program $\Pscr$.
+%
+Although this proof system is a focused proof system, the
+richness of focusing is not apparent in this simplified 
+setting: thus, we drop the adjective ``focused'' from this point forward.
+
+Figure~\ref{fig:kernel} contains the \lP implementation of the
+inference rules in Figure~\ref{fig:augmented}: here the infix
+turnstile $\vdash$ symbol is replaced by the \lst<i{check} predicate.
+%
+Notice that it is easy to show that no matter how the expert predicates
+are defined, if the goal \lsti{check Cert B} is provable in \lP then
+\lsti{B} must be a sound consequence of the program clauses stored in
+the \lsti{prog} predicate (which provides a natural implementation of
+the premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$.
+% AM: not there
+% in Figure~\ref{fig:kernel}).
+
+As we mentioned in the introduction, our notion of proof certificates
+ is taken from the general setting of \emph{foundational
+  proof certificates}~\cite{chihani17jar}.
+%
+In our case here, an FPC is a collection of \lP clauses that
+provide the remaining details not supplied in Figure~\ref{fig:kernel}:
+that is, the exact set of constructors for the \lsti{cert} type as
+well as the exact specification of the six expert predicates listed in
+that figure.
+%
+Figure~\ref{fig:resources} displays two such FPCs,
+both of which can be used to describe proofs where we bound
+the number of occurrences of unfoldings in a proof.
+%
+For example, the first FPC
+provides the experts for treating certificates that are constructed
+using the \lsti{qheight} constructor.
+%
+As is easy to verify, the query \mbox{\lsti{(check (qheight 5) B)}} (for the
+encoding \lsti{B} of a goal formula) is provable in \lP using the
+clauses in Figures~\ref{fig:kernel} and~\ref{fig:resources} if and
+only if the height of that proof is 5 or less.
+%
+Similarly, the second FPC uses the constructor \lsti{qsize} (with two
+integers) and can be used to bound the total number of instances of
+unfoldings in a proof.
+%
+In particular, the query \mbox{\lsti{sigma H\ (check (qsize 5 H) B)}}
+% \begin{lstlisting}
+% sigma H\ (check (qsize 5 H) B)
+% \end{lstlisting}
+is provable if and only if the total number of unfoldings of that
+proof is 5 or less.
+
+\begin{figure}
+\lstinputlisting[linerange={resources-end}]{code/fpcs.sig}
+\lstinputlisting[linerange={resources-end}]{code/fpcs.mod}
+\caption{Two FPCs that describe proofs that are limited in either
+  height or in size.}
+\label{fig:resources}
+\end{figure}
+
+Finally, Figure~\ref{fig:max} contains the FPC based on the constructor
+\lsti{max} that is used to record explicitly all information within a
+proof: in particular, all disjunctive choices and all substitution
+instances for existential quantifiers are collected into a binary tree
+structure of type \lsti{max}.
+%
+In this sense, proof certificates built with the \lsti{max}
+constructor are \emph{maximally} explicit.
+%
+Such proof certificates are used, for example, in~\cite{Pair}; it is
+important to note that proof checking with such maximally explicit
+certificates can be done with much simpler proof-checkers than those
+used in logic programming since backtracking search and unification
+are not needed.
+%
+
+\begin{figure}
+\lstinputlisting[linerange={max-end}]{code/fpcs.sig}
+\lstinputlisting[linerange={max-end}]{code/fpcs.mod}
+\caption{The \lsti{max} FPC}
+\label{fig:max}
+\end{figure}
+
+Further, if we view a particular FPC as a means of restricting proofs, it is
+possible to build an FPC that \emph{restricts} proofs satisfying two FPCs
+simultaneously.
+%
+In particular, Figure~\ref{fig:pairing} defines an FPC based on the
+(infix) constructor \lsti{<c>}, which \emph{pairs} two terms of type
+\lsti{cert}.
+%
+The pairing experts for the certificate \lsti{Cert1 <c> Cert2} simply
+request that the corresponding experts also succeed for both
+\lsti{Cert1} and \lsti{Cert2} and, in the case of the \lsti{orE} and
+\lsti{someE}, also return the same choice and substitution term,
+respectively.
+%
+Thus, the query 
+\begin{lstlisting}
+?- check ((qheight 4) <c> (qsize 10)) B
+\end{lstlisting}
+will succeed if there is a proof of \lsti{B} that has a height less
+than or equal to 4 while also being of size less than or equal to 10.
+%
+A related use of the pairing of two proof certificates is to %use it to
+\emph{distill} or \emph{elaborate} proof certificates. 
+%
+For example, the proof certificate \lsti{(qsize 5 0)} is rather implicit
+since it will match any proof that used unfold exactly 5 times.
+%
+However, the query
+\begin{lstlisting}
+?- check ((qsize 5 0) <c> (max Max)) B.
+\end{lstlisting}
+will store into the \lP variable \lsti{Max} more complete details of
+any proof that satisfies the \lsti{(qsize 5 0)} constraint.
+%
+In particular, this forms the infrastructure of 
+ an \emph{explanation} tool for attributing ``blame'' for
+the origin of a counterexample; these maximal
+certificates are an appropriate starting point for documenting
+both the counterexample and why it serves as a counterexample.
+
+\begin{figure}
+\lstinputlisting[linerange={pairing-end}]{code/fpcs.sig}
+\lstinputlisting[linerange={pairing-end}]{code/fpcs.mod}
+\caption{FPC for pairing}
+\label{fig:pairing}
+\end{figure}
+
+Various  additional examples and experiments using the pairing of FPCs
+can be found in~\cite{Pair}. Using similar techniques, it is possible to define FPCs that target
+specific types for special treatment: for example, when generating
+integers, only (user-defined) small integers could be inserted into
+counterexamples.
+
+\section{Property-based testing for Coq}
+
+Simplest setting since we are only sending back to Coq a term (a
+counterexample), not a general proof.
 
@@
 
-\section{Property based testing for Coq}
-
 \section{Conclusion}
 
-% Not part of git repo.  Need to find a fix
-\bibliography{../../../../repo/parsifal/papers/references/master}
-
+\bibliography{l}
 \end{document}

### 2020-10-22 papers/types2020/elab.tex
paper split into sections
some comments to Dale's "bold claims"

--- 2020-10-21 papers/types2020/main.tex
@@
-\documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
-\input preamble
-\input head
-
-\title{Interfacing logic programming and Coq} % Other suggestions?
-\titlerunning{Interfacing logic programming and Coq}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-We do stuff
-\end{abstract}
-
-\section{Introduction}
-\label{sec:intro}
-
-Recently, C. Sacerdoti Coen and E. Tassi developed the ELPI
-implementation \cite{dunchev15lpar} of \lP 
-\cite{miller12proghol}.  More recently, Tassi has made ELPI available
-as a plugin to the Coq proof assistant \cite{tassi18coqpl}.
-%
-This implementation of \lP extends earlier implementations in
-primarily two directions.  First, ELPI adds a notion of contraints,
-which makes it more expressive than, say, the Teyjus implementation of
-\lP \cite{nadathur99cade}.  Second, the plugin version of ELPI comes
-equip with an API for accessing the Coq environment and a quotation
-and anti-quotation syntax allowing for Coq expressions to be mixed in
-with \lP program elements.
-
-ELPI provides a means for treating bindings and unification (of
-structures containing bindings) that are convenient for various
-meta-programming tasks in logic programming all things useful for
-tactic programming in Coq (proof refinement).  For examples,
-ELPI programmers are spared from dealing with low-level aspects of
-representing bindings (e.g., De Bruijn indexes) while still having
-clean and effective ways to manipulate binding structures.
-%
-In short, ELPI appears to be a useful meta-language for Coq
-\cite{coen19mscs,tassi18coqpl,tassi19itp}.
-%
-In this paper, we outline two applications of the Coq plugin for ELPI
-that support this claim.
-%
-With these examples, we shall illustrate that ELPI is a useful
-meta-language not just because for its treatment of bindings and
-syntactic structures in general but also because it is actually based
-on a sound implementation of a higher-order intuitionistic logic.
-%
-Type inference is a well-known example of a meta-programming task for
-which logic programming often provides immediate and elegant
-implementations via the sound implementation of logic.
-%
-In this paper, we present two additional project in which an
-implementation of proof search provides direct, elegant, and compact
-implementations of meta-programming tasks.
-%
-Before we present these two examples, we first discuss the different
-world views that the integration of ELPI and Coq focuses us to
-confront.
-
-\section{Two cultures}
-
-{\color{red} The rest of this section are just notes.}
-
-These applications are not necessarily focused on the underlying
-declarative nature of lambda Prolog.  For example, ELPI has various
-extensions, such as ``spilling'' that are designed to make some
-aspects of logic programming appear to be more functional.
-
-In this paper, we limit our attention mostly to using lambda Prolog
-for declarative processing: i.e., where logic plays a central role in
-describing operations.  (Type checking is an example of such
-processing, but we do not focus on type checking here).
-
-Of course, being able to do declarative and not-so-declarative
-operations within the same language makes that language all the more
-useful.
-
-  Given that lP and Coq are both examples of mixing the
-  lambda-calculus with logic, it is important to understand their
-  difference, something that can be very confusing.  Take the
-  confusion around the term HOAS.  In some circles, this is a
-  appropriate and sensible approach.  In other circles, it is
-  semantically questionable.  But, in fact, this concept (using
-  meta-level binding for object-level binding) plays our 
-  differently depending on the meta language!
-
-  The origin story to ML (Meta language) is that it supported
-  LFC-style tactics and tacticals.  The strong typing of ML provided
-  some guarantees: only theorems were ever given type thm.  But many
-  steps of evolution have now taken place since then.  Isabelle's
-  different approach to tactics.  Logic and proof theory, when used
-  with Church's simple theory of types and Gentzen's sequent calculus,
-  provides a far richer meta language: that provides an immediate and
-  declarative treatment of bindings as well as for proof search
-  (without the need of exception handling, etc).  [I'm making the case
-  that FP is not a particularly natural framework for proof
-  assistants.  Of course, lots of libraries have now been built in the
-  FP style.]
-
-  - Various dichotomies we find in this work.
-    - proof theory vs type theory
-    - logic programming vs functional programming
-    - sequent calculus vs natural deduction
-
-  - While both lambda Prolog and Coq are based on an amalgamation of
-    $\lambda$-calculus and logic, these two languages take strikingly
-    different approaches to bindings.  In lambda Prolog, bindings are
-    syntactic and extensionality at any type is not available.  In
-    Coq, however, higher-order typed quantification encodes
-    extensional functions.  In lambda Prolog, (exists w. [lambda x. x]
-    = [lambda x. w]) fails, no matter the type of w.  If this query
-    was about functions, then this equation would be true when the
-    type of w is a singleton.  But in lP, this is a query only about
-    binding structures, and this fails to be proved since there is no
-    capture-avoiding substitution (the only substitution supported by
-    Church's STT) for w that makes the resulting two terms alpha,
-    beta, eta convertible.  [Is this true in Coq?  What is equality
-    here?]
-
-  - On the proof theory side of this work, we use two different proof
-    systems.
-
-    - LJF, here we have undefined predicates, no induction
-    - muMALL, were we have inductive definitions of predicates
-
-    The Coq world provides for both of these kinds of predicate using
-    Prop (corresponds to undefined) and Set (inductively defined).
-
 \section{Elaboration of external proof certificates for the Coq
@@
 
-\section{Property based testing for Coq}
-
-\section{Conclusion}
-
-% Not part of git repo.  Need to find a fix
-\bibliography{../../../../repo/parsifal/papers/references/master}
-
-\end{document}
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:

### 2020-10-22 papers/types2020/main.tex
paper split into sections
some comments to Dale's "bold claims"

--- 2020-10-21 papers/types2020/main.tex
@@
 meta-programming tasks in logic programming all things useful for
-tactic programming in Coq (proof refinement).  For examples,
-ELPI programmers are spared from dealing with low-level aspects of
+tactic programming in Coq (proof refinement).  For examples, ELPI
+programmers are spared from dealing with low-level aspects of
 representing bindings (e.g., De Bruijn indexes) while still having
-clean and effective ways to manipulate binding structures.
-%
-In short, ELPI appears to be a useful meta-language for Coq
+clean and effective ways to manipulate binding structures.  In short,
+ELPI appears to be a useful meta-language for Coq
 \cite{coen19mscs,tassi18coqpl,tassi19itp}.  
-%
-In this paper, we outline two applications of the Coq plugin for ELPI
-that support this claim.
-%
+Type inference is a well-known example of a meta-programming task for
+which logic programming often provides immediate and elegant
+implementations via the sound implementation of logic.  % In this paper,
+% we present two additional project in which an implementation of proof
+% search provides direct, elegant, and compact implementations of
+% meta-programming tasks.
+In this paper, we outline
+two applications of the Coq plugin for ELPI that support this claim.
 With these examples, we shall illustrate that ELPI is a useful
@@
 on a sound implementation of a higher-order intuitionistic logic.
-%
-Type inference is a well-known example of a meta-programming task for
-which logic programming often provides immediate and elegant
-implementations via the sound implementation of logic.
-%
-In this paper, we present two additional project in which an
-implementation of proof search provides direct, elegant, and compact
-implementations of meta-programming tasks.
-%
-Before we present these two examples, we first discuss the different
-world views that the integration of ELPI and Coq focuses us to
-confront.
+Before we present these two examples, we
+first discuss the different world views that the integration of ELPI
+and Coq forces us to confront.
 
@@
 
-{\color{red} The rest of this section are just notes.}
+In proof theory, one learns quickly that things always come in pairs:
+negative/positive, left/right, bottom-up/top-down,
+premises/conclusion, introduction/elimination,
+% determinism/nondeterminism,
+% invertible/noninvertible, 
+etc.
+%
+When we examine the larger setting of this project of linking a logic
+programming engine with Coq and its kernel, we find a large number of
+pairing that are valuable to explicitly describe.
 
-These applications are not necessarily focused on the underlying
-declarative nature of lambda Prolog.  For example, ELPI has various
-extensions, such as ``spilling'' that are designed to make some
-aspects of logic programming appear to be more functional.
+\subsection{Proof theory vs type theory}
 
-In this paper, we limit our attention mostly to using lambda Prolog
-for declarative processing: i.e., where logic plays a central role in
-describing operations.  (Type checking is an example of such
-processing, but we do not focus on type checking here).
+In many ways, (structural) proof theory is more elementary and
+low-level than most approaches to type theory.  For example, type
+theories usually answer the question ``What is a proof?'' with the
+response ``a dependently typed $\lambda$-term''.  That is, when
+describing a type theory, one usually decides that a proof is a
+certain kind of term within the system.  In contrast, the proof theory
+treats logical propositions and proofs as separate.  For example,
+proof theory does not assume that there are terms within the logic
+that describe proofs.
 
-Of course, being able to do declarative and not-so-declarative
-operations within the same language makes that language all the more
-useful.
+\begin{metanote}
+  This paragraph seem to identify proof and type theory with sequents and nat ded and while
+  I agree with the gist (although people such as Frank would argue that sequents are a way of organizing proof search for nat ded), we're not gonna make a lot of friends at types and more importantly is a big hammer to nail the point that logic programming is helpful with Coq --am
+\end{metanote}
 
-  Given that lP and Coq are both examples of mixing the
-  lambda-calculus with logic, it is important to understand their
-  difference, something that can be very confusing.  Take the
-  confusion around the term HOAS.  In some circles, this is a
-  appropriate and sensible approach.  In other circles, it is
-  semantically questionable.  But, in fact, this concept (using
-  meta-level binding for object-level binding) plays our 
-  differently depending on the meta language!
+Gentzen's discovery that the key to treating classical and
+intuitionistic logics in the same proof system was the invention of
+the weakening and contraction structural rules on the right-side of
+sequents~\cite{gentzen35}.  This discovery lead him away from natural
+deduction to the invention of multiple-conclusion sequent calculus, a
+very important kind of proof that has not been given a satisfactory
+representation in type theory.  This same innovation of Gentzen also
+opened the way to the invention of another key proof theoretic
+discovery, that of linear logic \cite{girard87tcs}.
+Similarly, sequent calculus provides an elegant presentation for
+linear logic while most treatments of linear logic in type theory
+usually only capture linearity in a partial fashion.
 
-  The origin story to ML (Meta language) is that it supported
-  LFC-style tactics and tacticals.  The strong typing of ML provided
-  some guarantees: only theorems were ever given type thm.  But many
-  steps of evolution have now taken place since then.  Isabelle's
-  different approach to tactics.  Logic and proof theory, when used
-  with Church's simple theory of types and Gentzen's sequent calculus,
-  provides a far richer meta language: that provides an immediate and
-  declarative treatment of bindings as well as for proof search
-  (without the need of exception handling, etc).  [I'm making the case
-  that FP is not a particularly natural framework for proof
-  assistants.  Of course, lots of libraries have now been built in the
-  FP style.]
+Proof theory also has rather immediate and natural treatments for
+co-induction (as well as induction: see
+\cite{baelde12tocl,momigliano12jal}) whereas the treatment of
+coinduction within type theory remains more challenging and is well
+developed only for various syntactic conditions (see, for example,
+\cite{bertot08entcs}).
+\begin{metanote}
+  The criticism applies to Coq, less to type theory, see co-patterns --am
+\end{metanote}
 
-  - Various dichotomies we find in this work.
-    - proof theory vs type theory
-    - logic programming vs functional programming
-    - sequent calculus vs natural deduction
+\subsection{Proof search vs proof normalization}
 
-  - While both lambda Prolog and Coq are based on an amalgamation of
-    $\lambda$-calculus and logic, these two languages take strikingly
-    different approaches to bindings.  In lambda Prolog, bindings are
-    syntactic and extensionality at any type is not available.  In
-    Coq, however, higher-order typed quantification encodes
-    extensional functions.  In lambda Prolog, (exists w. [lambda x. x]
-    = [lambda x. w]) fails, no matter the type of w.  If this query
-    was about functions, then this equation would be true when the
-    type of w is a singleton.  But in lP, this is a query only about
-    binding structures, and this fails to be proved since there is no
-    capture-avoiding substitution (the only substitution supported by
-    Church's STT) for w that makes the resulting two terms alpha,
-    beta, eta convertible.  [Is this true in Coq?  What is equality
-    here?]
+Gentzen-style proofs are used to model computation in at least two
+different ways.  The functional programming paradigm, following the
+Curry-Howard correspondence, models computation abstractly as the
+$\beta$-reduction of natural deduction proof: that is, computation is
+modeled using \emph{proof normalization}.  On the other hand, the
+logic programming paradigm, following the notion of goal-directed
+proof search, models computation abstractly as a regimented search of
+cut-free sequent calculus proofs: that is, computation is modeled
+using \emph{proof search}.
 
-  - On the proof theory side of this work, we use two different proof
-    systems.
+Proof search has proved to have some richness that is hard to capture
+in the proof normalization setting.  In particular, Gentzen's
+invention of \emph{eigenvariables} are a kind of a proof-level
+binder.  In the proof normalization setting, such eigenvariables
+are instantiated during proof normalization.  In contrast, during the
+search for cut-free proofs, eigenvariables are part of the syntax of
+terms and formulas.  As a result, they can be used in the treatment of
+bindings in data structures more generally.  Hence, the logic
+programming paradigm provides a completely natural, elegant, and
+powerful method, called $\lambda$-tree syntax, for treating bindings
+within syntax.
 
-    - LJF, here we have undefined predicates, no induction
-    - muMALL, were we have inductive definitions of predicates
+In this paper, we shall use two different proof systems.  The first
+logic is first-order intuitionistic logic.  Here, predicates
+are not inductively defined and quantification will be limited to
+range over first-order term.  Actually, we shall examine only the Horn
+clause fragment of that logic.
+%
+The second logic will allow for inductively defined predicates using a
+fixed point notation.   Given again our restriction to essentially
+Horn clause specifications, it is possible to view intuitionistic
+reasoning with such specification as if they are actually
+specification in $\mu$MALL, an extension of multiplicative additive
+linear logic with the least and greatest fixed point operators
+\cite{baelde12tocl,baelde07lpar,heath19jar}.
+%
+The difference between these two logics and associate proof theories
+is reminiscent to the difference between the type \lsti{Prop}
+(corresponding to undefined propositions) and Sets (corresponding to
+inductively defined propositions).
+\begin{metanote}
+  Still not onboard with this --am
+\end{metanote}
 
-    The Coq world provides for both of these kinds of predicate using
-    Prop (corresponds to undefined) and Set (inductively defined).
+\subsection{\lP vs Coq}
 
-\section{Elaboration of external proof certificates for the Coq
-  kernel}
+Given that \lP and Coq are both examples of mixing the
+$\lambda$-calculus with logic, it is important to understand their
+differences.  The confusion around the term \emph{higher-order
+abstract syntax} (HOAS), for example, helps to illustrate their
+differences.  In the functional programming setting, including the Coq
+system, the HOAS approach leads to encoding binding structures within
+terms using functions.  Such encodings have results in surprisingly
+complicated encodings, allowing, for example, for \emph{exotic terms}
+\cite{despeyroux95tlca} and to structures on which induction is not
+possible \cite{roeckl01fossacs}.  In the logic programming setting,
+particularly \lP, HOAS is well supported in both the availability of
+binders allowed with terms ($\lambda$-bindings), in formulas
+(quantifiers), and in proofs (eigenvariables).  For this reason, the
+term $\lambda$-tree syntax was introduced to name this particular take
+on HOAS \cite{miller19jar}.  The Abella proof assistant
+\cite{baelde14jfr} was designed, in part, to allow for inductive and
+co-inductive proofs for specifications using the $\lambda$-tree syntax
+approach.
 
-Follow the text in \cite{blanco20coq}.
+Another difference between \lP and functional programming can be
+illustrated by considering how they are used in the specification of
+tactics that are popular in modern proof assistants.  In fact, the
+origin story for the ML functional programming language was that ML
+was the meta-language for implementing the LCF suite of tactics and
+tacticals \cite{gordon79}.  In order to implement tactics, ML adopted
+not only novel features such as polymorphically typed higher-order
+functional programming but also the non-functional mechanisms of
+failure and exception handling.  While \lP is also based on
+polymorphically typed higher-order relational programming, it also
+comes with a completely declarative version of failure and
+backtracking search.  Combining those features along with its support
+of unification (even in the presence of term-level bindings), \lP
+provides a rather different approach to the specification of tactics
+\cite{felty93jar}.
 
-\section{Property based testing for Coq}
+While \lP is a typed language, it only uses simple types and these are
+more often used to indicate syntactic categories.  Any dependency
+information needs to be captured using predicates and quantifiers.
+
+\input sec3
+
+\input pbt
+
+\input elab
 
@@
 
-% Not part of git repo.  Need to find a fix
-\bibliography{../../../../repo/parsifal/papers/references/master}
+\bibliography{l}
+\end{document}
 
-\end{document}
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:
+
+% LocalWords:  Tassi ELPI Teyjus intuitionistic Gentzen's Gentzen LCF
+% LocalWords:  eigenvariables tacticals polymorphically

### 2020-10-22 papers/types2020/pbt.tex
paper split into sections
some comments to Dale's "bold claims"

\section{Property-based testing for Coq}

Simplest setting since we are only sending back to Coq a term (a
counterexample), not a general proof.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-10-22 papers/types2020/sec3.tex
paper split into sections
some comments to Dale's "bold claims"

\section{Proof theory and proof certificates}

{\color{red} The goal in this section is to say a bit about focusing
  generally, and to give a bit of detail about FPC for Horn clauses.
  Initial draft of text taken from \cite{blanco19ppdp}.
}

\subsection{Focused proof systems}
\label{ssec:focused}

The proof search approach to encoding Horn clause computation
results in the structuring of proofs with repeated
switching between a \emph{goal-reduction} phase and a
\emph{backchaining} phase~\cite{miller91apal}.
%
The notion of \emph{focused proof systems}~\cite{andreoli92jlc,liang09tcs}
generalizes this view of proof construction by identifying the
following two phases. % of proof construction.
\begin{enumerate}
\item The \emph{negative}%
\footnote{The terminology of negative and positive phases is a bit
  unfortunate: historically, these terms do not refer to positive or negative
  subformula occurrences but rather to certain semantic models used in
  the study of linear logic~\cite{girard87tcs}.}
%
  phase corresponds to  goal-reduction: in this phase,
  inference rules that involve \emph{don't-care-non\-de\-ter\-min\-ism} are
  applied.
%
  As a result, there is no need to consider backtracking over choices
  made in building this phase.

\item The \emph{positive} phase corresponding to  backchaining: in this phase, inference rules that involve
  \emph{don't-know-non\-de\-ter\-min\-ism} are applied: here,
  inference rules need to be supplied with information in order to
  ensure that a completed proof can be built.
%
  That information can be items such as which term is needed to
  instantiate a universally quantified formula and which disjunct of a
  disjunctive goal formula should be proved.
\end{enumerate}
%
Thus, when building a proof tree (in the sequent calculus) from the
conclusion to its leaves, the negative phase corresponds to a simple
computation that needs no external information, while the positive
phase may need such external information to be supplied.
%
In the literature, it is common to refer to a repository of such
external information as either an \emph{oracle} or a \emph{proof
certificate}.

When using a focused proof system for logic extended with fixed
points, such as  employed in Bedwyr~\cite{baelde07cade} and described
in~\cite{baelde12tocl,heath17linearity}, proofs of formulas such as
\[
  \exists x [(\tau(x)\wedge P(x)) \wedge \neg Q(x)]
  \tag{*}\label{eq:full}
\]
are a single \emph{bipole}: that is, when reading a proof bottom up, a
positive phase is followed on all its premises by a single negative
phase that completes the proof.%
\footnote{The reader familiar with focusing will understand that there
are two ``polarized'' conjunctions, written in linear logic as
$\otimes$ and $\with$ or in classical and intuitionistic logics as
$\wedgep$ and $\wedgen$, respectively.  In this paper, we use simply
$\wedge$ to denote the positive biased conjunction.}
%
In particular, the positive phase corresponds to the \emph{generation} phase
and the negative phase corresponds to the \emph{testing} phase.
%

Instead of giving a full focused proof system of a logic including
fixed points (since, as we will argue, that proof system will not, in
fact, be needed to account for PBT ), we offer the following analogy.
%
Suppose that we are given a finite search tree and we are asked to
prove that there is a secret located in one of the nodes of that
tree.
%
A proof that we have found that secret can be taken to be a
description of the path to that node from the root of the tree: that
path can be seen as the proof certificate for that claim.
%
On the other hand, a proof that no node contains the secret is a
rather different thing: here, one expects to use a blind and
exhaustive search (via, say, depth-first or breath-first search) and
that the result of that search never discovers the secret.
%
A proof of this fact requires \emph{no} external information: instead it
requires a lot of computation involved with exploring the tree until
exhaustion.
%
This follows the familiar pattern where the positive (generate) phase
requires external information while the negative (testing) phase
requires none.
%
Thus, a proof certificate for $(\ref{eq:full})$ is also a proof
certificate for 
\[
  \exists x [\tau(x)\wedge P(x)].
  \tag{**}\label{eq:short}
\]
Such a certificate would contain the witness (closed) term $t$ for the
existential quantifier and sufficient information to confirm that
$P(t)$ can be proved (a proof of a typing judgment such as $\tau(t)$
is usually trivial).
%
%In short\ednote{AM: a bit repetitious}, a proof certificate for
%$(\ref{eq:short})$ can play the part 
%of a proof certificate also for $(\ref{eq:full})$, if we also use
%negation-as-finite-failure  with $Q(t)$.
%  ****
Since a proof certificate for the existence-of-a-counterexample formula
$(\ref{eq:full})$ can be taken as a proof certificate of
$(\ref{eq:short})$, then we only need to consider proof certificates
for Horn clause programs.
%
%Since the logic of Horn clauses is rather simple,
We illustrate next what such proofs and proof certificates look like
for the rather simple logic of Horn clauses.

\subsection{Certificate checking with expert predicates}
%\subsection{Adding certificates and experts to a proof system}
\label{ssec:fpc}

%% \begin{metanote}
%%   AM: I know this is well known and boring, but using \texttt{prog A
%%     G} in Fig \ref{fig:augmented} is imprecise --- in the comment of
%%   next figure we mention \texttt{A :- G} and we gloss over the
%%   instantiation of \texttt{prog A' G} to match the goal \texttt{A},
%%   which we assume to be ground, given the exists rule
%% \end{metanote}

%% DM I've modified the text a bit to hopefully improve the state of
%% things here.  The formal proof theory is all about closed
%% formulas.  The prolog implementation allows open formulas.  This
%% difference invades most aspects of proof checking, not just the
%% prog predicate.

% DM We should find a place to smooth over this issue (which is not
% hard, I think).

\begin{figure}
\newcommand{\XXi}{{\color{blue}{\Xi}}}
\[
\infer{\XXi\vdash t = t}
      {\eqExpert{\XXi}}
\qquad
\infer{\XXi\vdash \true}
      {\trueExpert{\XXi}}
\qquad
\infer{\XXi\vdash G_1\vee G_2}
      {\XXi'\vdash G_i\qquad \orExpert{\XXi}{\XXi'}{i}}
\qquad
\infer{\XXi\vdash \exists x. G}
      {\XXi'\vdash G[t/x]\qquad \someExpert{\XXi}{\XXi'}{t}}
\]
\vskip -6pt
\[
\infer{\XXi\vdash G_1\wedge G_2}
      {\XXi_1\vdash G_1\quad \XXi_2\vdash G_2\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}}
\qquad
\infer{\XXi\vdash A}
      {\XXi'\vdash G \quad (A~\hbox{\tt :-}~G)\in\instan\Pscr
                     \quad \unfoldExpert{\XXi}{\XXi'}}
\]
\caption{A proof system augmented with proof certificates and
  additional ``expert'' premises.}
\label{fig:augmented}
TODO
%\lstinputlisting[linerange={sigs-end}]{code/kernel.sig}
%\lstinputlisting[linerange={check-end}]{code/kernel.mod}
\caption{A simple proof checking kernel.}
\label{fig:kernel}
\end{figure}

Figure~\ref{fig:augmented} contains a simple proof system for Horn
clause provability in which each inference rule is augmented with an
additional premise involving an \emph{expert predicate}, a
certificate $\Xi$, and possibly continuations of certificates ($\Xi'$,
$\Xi_1$, $\Xi_2$) with extracted information from certificates (in the
case of $\vee$ and $\exists$).
%
%
The premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$ in the last
inference rule of Figure~\ref{fig:augmented} states that the logic
programming clause $(A~\hbox{\tt :-}~G)$ is a ground instance of some
clause in a fixed program $\Pscr$.
%
Although this proof system is a focused proof system, the
richness of focusing is not apparent in this simplified 
setting: thus, we drop the adjective ``focused'' from this point forward.

Figure~\ref{fig:kernel} contains the \lP implementation of the
inference rules in Figure~\ref{fig:augmented}: here the infix
turnstile $\vdash$ symbol is replaced by the \lst<i{check} predicate.
%
Notice that it is easy to show that no matter how the expert predicates
are defined, if the goal \lsti{check Cert B} is provable in \lP then
\lsti{B} must be a sound consequence of the program clauses stored in
the \lsti{prog} predicate (which provides a natural implementation of
the premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$.
% AM: not there
% in Figure~\ref{fig:kernel}).

As we mentioned in the introduction, our notion of proof certificates
 is taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:kernel}:
that is, the exact set of constructors for the \lsti{cert} type as
well as the exact specification of the six expert predicates listed in
that figure.
%
Figure~\ref{fig:resources} displays two such FPCs,
both of which can be used to describe proofs where we bound
the number of occurrences of unfoldings in a proof.
%
For example, the first FPC
provides the experts for treating certificates that are constructed
using the \lsti{qheight} constructor.
%
As is easy to verify, the query \mbox{\lsti{(check (qheight 5) B)}} (for the
encoding \lsti{B} of a goal formula) is provable in \lP using the
clauses in Figures~\ref{fig:kernel} and~\ref{fig:resources} if and
only if the height of that proof is 5 or less.
%
Similarly, the second FPC uses the constructor \lsti{qsize} (with two
integers) and can be used to bound the total number of instances of
unfoldings in a proof.
%
In particular, the query \mbox{\lsti{sigma H\ (check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
is provable if and only if the total number of unfoldings of that
proof is 5 or less.

\begin{figure}
\lstinputlisting[linerange={resources-end}]{code/fpcs.sig}
\lstinputlisting[linerange={resources-end}]{code/fpcs.mod}
\caption{Two FPCs that describe proofs that are limited in either
  height or in size.}
\label{fig:resources}
\end{figure}

% Finally, Figure~\ref{fig:max} contains the FPC based on the constructor
% \lsti{max} that is used to record explicitly all information within a
% proof: in particular, all disjunctive choices and all substitution
% instances for existential quantifiers are collected into a binary tree
% structure of type \lsti{max}.
% %
% In this sense, proof certificates built with the \lsti{max}
% constructor are \emph{maximally} explicit.
% %
% Such proof certificates are used, for example, in~\cite{Pair}; it is
% important to note that proof checking with such maximally explicit
% certificates can be done with much simpler proof-checkers than those
% used in logic programming since backtracking search and unification
% are not needed.
%

% \begin{figure}
% \lstinputlisting[linerange={max-end}]{code/fpcs.sig}
% \lstinputlisting[linerange={max-end}]{code/fpcs.mod}
% \caption{The \lsti{max} FPC}
% \label{fig:max}
% \end{figure}

Further, if we view a particular FPC as a means of restricting proofs, it is
possible to build an FPC that \emph{restricts} proofs satisfying two FPCs
simultaneously.
%
In particular, Figure~\ref{fig:pairing} defines an FPC based on the
(infix) constructor \lsti{<c>}, which \emph{pairs} two terms of type
\lsti{cert}.
%
The pairing experts for the certificate \lsti{Cert1 <c> Cert2} simply
request that the corresponding experts also succeed for both
\lsti{Cert1} and \lsti{Cert2} and, in the case of the \lsti{orE} and
\lsti{someE}, also return the same choice and substitution term,
respectively.
%
Thus, the query 
\begin{lstlisting}
?- check ((qheight 4) <c> (qsize 10)) B
\end{lstlisting}
will succeed if there is a proof of \lsti{B} that has a height less
than or equal to 4 while also being of size less than or equal to 10.
%
% A related use of the pairing of two proof certificates is to %use it to
% \emph{distill} or \emph{elaborate} proof certificates. 
% %
% For example, the proof certificate \lsti{(qsize 5 0)} is rather implicit
% since it will match any proof that used unfold exactly 5 times.
% %
% However, the query
% \begin{lstlisting}
% ?- check ((qsize 5 0) <c> (max Max)) B.
% \end{lstlisting}
% will store into the \lP variable \lsti{Max} more complete details of
% any proof that satisfies the \lsti{(qsize 5 0)} constraint.
% %
% In particular, this forms the infrastructure of 
%  an \emph{explanation} tool for attributing ``blame'' for
% the origin of a counterexample; these maximal
% certificates are an appropriate starting point for documenting
% both the counterexample and why it serves as a counterexample.

\begin{figure}
\lstinputlisting[linerange={pairing-end}]{code/fpcs.sig}
\lstinputlisting[linerange={pairing-end}]{code/fpcs.mod}
\caption{FPC for pairing}
\label{fig:pairing}
\end{figure}

Various  additional examples and experiments using the pairing of FPCs
can be found in~\cite{Pair}. Using similar techniques, it is possible to define FPCs that target
specific types for special treatment: for example, when generating
integers, only (user-defined) small integers could be inserted into
counterexamples.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-10-23 papers/types2020/pbt.tex
some metanotes on how what to write in sec 3 and 4

@@
 
+\begin{metanote}
+  dump from previous paper
+\end{metanote}
+
+When using a focused proof system for logic extended with fixed
+points, such as  employed in Bedwyr~\cite{baelde07cade} and described
+in~\cite{baelde12tocl,heath17linearity}, proofs of formulas such as
+\[
+  \exists x [(\tau(x)\wedge P(x)) \wedge \neg Q(x)]
+  \tag{*}\label{eq:full}
+\]
+are a single \emph{bipole}: that is, when reading a proof bottom up, a
+positive phase is followed on all its premises by a single negative
+phase that completes the proof.%
+\footnote{The reader familiar with focusing will understand that there
+are two ``polarized'' conjunctions, written in linear logic as
+$\otimes$ and $\with$ or in classical and intuitionistic logics as
+$\wedgep$ and $\wedgen$, respectively.  In this paper, we use simply
+$\wedge$ to denote the positive biased conjunction.}
+%
+In particular, the positive phase corresponds to the \emph{generation} phase
+and the negative phase corresponds to the \emph{testing} phase.
+%
+
+Instead of giving a full focused proof system of a logic including
+fixed points (since, as we will argue, that proof system will not, in
+fact, be needed to account for PBT ), we offer the following analogy.
+%
+Suppose that we are given a finite search tree and we are asked to
+prove that there is a secret located in one of the nodes of that
+tree.
+%
+A proof that we have found that secret can be taken to be a
+description of the path to that node from the root of the tree: that
+path can be seen as the proof certificate for that claim.
+%
+On the other hand, a proof that no node contains the secret is a
+rather different thing: here, one expects to use a blind and
+exhaustive search (via, say, depth-first or breath-first search) and
+that the result of that search never discovers the secret.
+%
+A proof of this fact requires \emph{no} external information: instead it
+requires a lot of computation involved with exploring the tree until
+exhaustion.
+%
+This follows the familiar pattern where the positive (generate) phase
+requires external information while the negative (testing) phase
+requires none.
+%
+Thus, a proof certificate for $(\ref{eq:full})$ is also a proof
+certificate for 
+\[
+  \exists x [\tau(x)\wedge P(x)].
+  \tag{**}\label{eq:short}
+\]
+Such a certificate would contain the witness (closed) term $t$ for the
+existential quantifier and sufficient information to confirm that
+$P(t)$ can be proved (a proof of a typing judgment such as $\tau(t)$
+is usually trivial).
+%
+%In short\ednote{AM: a bit repetitious}, a proof certificate for
+%$(\ref{eq:short})$ can play the part 
+%of a proof certificate also for $(\ref{eq:full})$, if we also use
+%negation-as-finite-failure  with $Q(t)$.
+%  ****
+Since a proof certificate for the existence-of-a-counterexample formula
+$(\ref{eq:full})$ can be taken as a proof certificate of
+$(\ref{eq:short})$, then we only need to consider proof certificates
+for Horn clause programs.
+%
+%Since the logic of Horn clauses is rather simple,
+We illustrate next what such proofs and proof certificates look like
+for the rather simple logic of Horn clauses.
+
 %%% Local Variables:

### 2020-10-23 papers/types2020/sec3.tex
some metanotes on how what to write in sec 3 and 4

@@
 
+\begin{metanote}
+  (AM): I would start with the logical reconstruction of \texttt{check}, adding the LJT-style proof terms to fig \ref{fig:augmented} and then reverse engineer what we want/need to say about focusing and FPC.
+We \textbf{do} not talk about PBT yet, just logic programming (with proof terms).
+In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
+\end{metanote}
 \subsection{Focused proof systems}
@@
 
-When using a focused proof system for logic extended with fixed
-points, such as  employed in Bedwyr~\cite{baelde07cade} and described
-in~\cite{baelde12tocl,heath17linearity}, proofs of formulas such as
-\[
-  \exists x [(\tau(x)\wedge P(x)) \wedge \neg Q(x)]
-  \tag{*}\label{eq:full}
-\]
-are a single \emph{bipole}: that is, when reading a proof bottom up, a
-positive phase is followed on all its premises by a single negative
-phase that completes the proof.%
-\footnote{The reader familiar with focusing will understand that there
-are two ``polarized'' conjunctions, written in linear logic as
-$\otimes$ and $\with$ or in classical and intuitionistic logics as
-$\wedgep$ and $\wedgen$, respectively.  In this paper, we use simply
-$\wedge$ to denote the positive biased conjunction.}
-%
-In particular, the positive phase corresponds to the \emph{generation} phase
-and the negative phase corresponds to the \emph{testing} phase.
-%
-
-Instead of giving a full focused proof system of a logic including
-fixed points (since, as we will argue, that proof system will not, in
-fact, be needed to account for PBT ), we offer the following analogy.
-%
-Suppose that we are given a finite search tree and we are asked to
-prove that there is a secret located in one of the nodes of that
-tree.
-%
-A proof that we have found that secret can be taken to be a
-description of the path to that node from the root of the tree: that
-path can be seen as the proof certificate for that claim.
-%
-On the other hand, a proof that no node contains the secret is a
-rather different thing: here, one expects to use a blind and
-exhaustive search (via, say, depth-first or breath-first search) and
-that the result of that search never discovers the secret.
-%
-A proof of this fact requires \emph{no} external information: instead it
-requires a lot of computation involved with exploring the tree until
-exhaustion.
-%
-This follows the familiar pattern where the positive (generate) phase
-requires external information while the negative (testing) phase
-requires none.
-%
-Thus, a proof certificate for $(\ref{eq:full})$ is also a proof
-certificate for 
-\[
-  \exists x [\tau(x)\wedge P(x)].
-  \tag{**}\label{eq:short}
-\]
-Such a certificate would contain the witness (closed) term $t$ for the
-existential quantifier and sufficient information to confirm that
-$P(t)$ can be proved (a proof of a typing judgment such as $\tau(t)$
-is usually trivial).
-%
-%In short\ednote{AM: a bit repetitious}, a proof certificate for
-%$(\ref{eq:short})$ can play the part 
-%of a proof certificate also for $(\ref{eq:full})$, if we also use
-%negation-as-finite-failure  with $Q(t)$.
-%  ****
-Since a proof certificate for the existence-of-a-counterexample formula
-$(\ref{eq:full})$ can be taken as a proof certificate of
-$(\ref{eq:short})$, then we only need to consider proof certificates
-for Horn clause programs.
-%
-%Since the logic of Horn clauses is rather simple,
-We illustrate next what such proofs and proof certificates look like
-for the rather simple logic of Horn clauses.
-
 \subsection{Certificate checking with expert predicates}
@@
 
-%% \begin{metanote}
-%%   AM: I know this is well known and boring, but using \texttt{prog A
-%%     G} in Fig \ref{fig:augmented} is imprecise --- in the comment of
-%%   next figure we mention \texttt{A :- G} and we gloss over the
-%%   instantiation of \texttt{prog A' G} to match the goal \texttt{A},
-%%   which we assume to be ground, given the exists rule
-%% \end{metanote}
-
-%% DM I've modified the text a bit to hopefully improve the state of
-%% things here.  The formal proof theory is all about closed
-%% formulas.  The prolog implementation allows open formulas.  This
-%% difference invades most aspects of proof checking, not just the
-%% prog predicate.
-
-% DM We should find a place to smooth over this issue (which is not
-% hard, I think).
-
+\begin{metanote}
+  add proof terms and backchaining to sequents and code
+\end{metanote}
 \begin{figure}
@@
 %
-The premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$ in the last
-inference rule of Figure~\ref{fig:augmented} states that the logic
-programming clause $(A~\hbox{\tt :-}~G)$ is a ground instance of some
-clause in a fixed program $\Pscr$.
 % 
@@
 \lsti{B} must be a sound consequence of the program clauses stored in
-the \lsti{prog} predicate (which provides a natural implementation of
-the premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$.
+\dots
+% the \lsti{prog} predicate (which provides a natural implementation of
+% the premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$.
 % AM: not there
@@
 
+\begin{metanote}
+  Rethink and update (with the right connectives) these FPC
+\end{metanote}
 \begin{figure}
@@
 specific types for special treatment: for example, when generating
-integers, only (user-defined) small integers could be inserted into
-counterexamples.
+integers, only (user-defined) small integers could be produced.
 

### 2020-10-24 papers/types2020/sec3.tex
adding refs for LJT

@@
 \begin{metanote}
-  (AM): I would start with the logical reconstruction of \texttt{check}, adding the LJT-style proof terms to fig \ref{fig:augmented} and then reverse engineer what we want/need to say about focusing and FPC.
+  (AM): I would start with the logical reconstruction of \texttt{check}, adding the LJT-style proof terms~c\cite{Herbelin94,LengrandDM06} to fig \ref{fig:augmented} and then reverse engineer what we want/need to say about focusing and FPC.
 We \textbf{do} not talk about PBT yet, just logic programming (with proof terms).
@@
 \item The \emph{negative}%
-\footnote{The terminology of negative and positive phases is a bit
-  unfortunate: historically, these terms do not refer to positive or negative
-  subformula occurrences but rather to certain semantic models used in
-  the study of linear logic~\cite{girard87tcs}.}
+% \footnote{The terminology of negative and positive phases is a bit
+%   unfortunate: historically, these terms do not refer to positive or negative
+%   subformula occurrences but rather to certain semantic models used in
+%   the study of linear logic~\cite{girard87tcs}.}
 %
@@
 %
-In particular, Figure~\ref{fig:pairing} defines an FPC based on the
-(infix) constructor \lsti{<c>}, which \emph{pairs} two terms of type
-\lsti{cert}.
+% In particular, Figure~\ref{fig:pairing} defines an FPC based on the
+% (infix) constructor \lsti{<c>}, which \emph{pairs} two terms of type
+% \lsti{cert}.
 %
@@
 
-\begin{figure}
-\lstinputlisting[linerange={pairing-end}]{code/fpcs.sig}
-\lstinputlisting[linerange={pairing-end}]{code/fpcs.mod}
-\caption{FPC for pairing}
-\label{fig:pairing}
-\end{figure}
+% \begin{figure}
+% \lstinputlisting[linerange={pairing-end}]{code/fpcs.sig}
+% \lstinputlisting[linerange={pairing-end}]{code/fpcs.mod}
+% \caption{FPC for pairing}
+% \label{fig:pairing}
+% \end{figure}
 

### 2020-10-24 papers/types2020/sec3.tex
code of dep-kernel inserted, fpc updated.

@@
 \begin{metanote}
-  add proof terms and backchaining to sequents and code
+  add proof terms and backchaining to sequents and code -- now sig is commented out
 \end{metanote}
+\begin{footnotesize}
 \begin{figure}
@@
 \label{fig:augmented}
-TODO
-%\lstinputlisting[linerange={sigs-end}]{code/kernel.sig}
-%\lstinputlisting[linerange={check-end}]{code/kernel.mod}
+%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
+\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
 \caption{A simple proof checking kernel.}
@@
 \end{figure}
-
+\end{footnotesize}
 Figure~\ref{fig:augmented} contains a simple proof system for Horn

### 2020-10-26 papers/types2020/cultures.tex
Split out another two sections.

--- 2020-10-22 papers/types2020/main.tex
@@
-%% Important dates:
-%%  Paper submission: 19 October 31 October 2020
-%%  Author notification: 18 January 2021
-%%  Final version:  15 February 2021
-%%  Publication (presumably): end of March 2021
-
-%% Style requirements of LIPIcs.  The recommended length of a paper is
-%% 12-15 pages, excluding front-page(s) (authors, affiliation, keywords,
-%% abstract, ...), bibliography and an appendix of max 5 pages. If you
-%% need more pages, please ask the editors.
-
-%% Papers have to be submitted in pdf through EasyChair:
-%% https://easychair.org/conferences/?conf=types2020postproceed
-
-\documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
-\input head
-\input preamble
-
-\title{Interfacing logic programming and Coq} % Other suggestions?
-\titlerunning{Interfacing logic programming and Coq}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-We do stuff
-\end{abstract}
-
-\section{Introduction}
-\label{sec:intro}
-
-Recently, C. Sacerdoti Coen and E. Tassi developed the ELPI
-implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol}.
-More recently, Tassi has made ELPI available as a plugin to the Coq
-proof assistant \cite{tassi18coqpl}.  This implementation of \lP
-extends earlier implementations in primarily two directions.  First,
-ELPI adds a notion of constraints, which makes it more expressive than,
-say, the Teyjus implementation of \lP \cite{nadathur99cade}.  Second,
-the plugin version of ELPI comes equip with an API for accessing the
-Coq environment and a quotation and anti-quotation syntax allowing for
-Coq expressions to be mixed in with \lP program elements.
-
-ELPI provides a means for treating bindings and unification (of
-structures containing bindings) that are convenient for various
-meta-programming tasks in logic programming all things useful for
-tactic programming in Coq (proof refinement).  For examples, ELPI
-programmers are spared from dealing with low-level aspects of
-representing bindings (e.g., De Bruijn indexes) while still having
-clean and effective ways to manipulate binding structures.  In short,
-ELPI appears to be a useful meta-language for Coq
-\cite{coen19mscs,tassi18coqpl,tassi19itp}.  
-Type inference is a well-known example of a meta-programming task for
-which logic programming often provides immediate and elegant
-implementations via the sound implementation of logic.  % In this paper,
-% we present two additional project in which an implementation of proof
-% search provides direct, elegant, and compact implementations of
-% meta-programming tasks.
-In this paper, we outline
-two applications of the Coq plugin for ELPI that support this claim.
-With these examples, we shall illustrate that ELPI is a useful
-meta-language not just because for its treatment of bindings and
-syntactic structures in general but also because it is actually based
-on a sound implementation of a higher-order intuitionistic logic.
-Before we present these two examples, we
-first discuss the different world views that the integration of ELPI
-and Coq forces us to confront.
-
 \section{Two cultures}
@@
 information needs to be captured using predicates and quantifiers.
-
-\input sec3
-
-\input pbt
-
-\input elab
-
-\section{Conclusion}
-
-\bibliography{l}
-\end{document}
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: t
-%%% End:
-
-% LocalWords:  Tassi ELPI Teyjus intuitionistic Gentzen's Gentzen LCF
-% LocalWords:  eigenvariables tacticals polymorphically

### 2020-10-26 papers/types2020/intro.tex
Split out another two sections.

--- 2020-10-21 papers/types2020/main.tex
@@
-\documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
-\input preamble
-\input head
-
-\title{Interfacing logic programming and Coq} % Other suggestions?
-\titlerunning{Interfacing logic programming and Coq}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-We do stuff
-\end{abstract}
-
 \section{Introduction}
@@
 Recently, C. Sacerdoti Coen and E. Tassi developed the ELPI
-implementation \cite{dunchev15lpar} of \lP 
-\cite{miller12proghol}.  More recently, Tassi has made ELPI available
-as a plugin to the Coq proof assistant \cite{tassi18coqpl}.
-%
-This implementation of \lP extends earlier implementations in
-primarily two directions.  First, ELPI adds a notion of contraints,
-which makes it more expressive than, say, the Teyjus implementation of
-\lP \cite{nadathur99cade}.  Second, the plugin version of ELPI comes
-equip with an API for accessing the Coq environment and a quotation
-and anti-quotation syntax allowing for Coq expressions to be mixed in
-with \lP program elements.
+implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol}.
+More recently, Tassi has made ELPI available as a plugin to the Coq
+proof assistant \cite{tassi18coqpl}.  This implementation of \lP
+extends earlier implementations in primarily two directions.  First,
+ELPI adds a notion of constraints, which makes it more expressive than,
+say, the Teyjus implementation of \lP \cite{nadathur99cade}.  Second,
+the plugin version of ELPI comes equip with an API for accessing the
+Coq environment and a quotation and anti-quotation syntax allowing for
+Coq expressions to be mixed with \lP program elements.
 
@@
 on a sound implementation of a higher-order intuitionistic logic.
-%
-Type inference is a well-known example of a meta-programming task for
-which logic programming often provides immediate and elegant
-implementations via the sound implementation of logic.
-%
-In this paper, we present two additional project in which an
-implementation of proof search provides direct, elegant, and compact
-implementations of meta-programming tasks.
-%
-Before we present these two examples, we first discuss the different
-world views that the integration of ELPI and Coq focuses us to
-confront.
-
-\section{Two cultures}
-
-{\color{red} The rest of this section are just notes.}
-
-These applications are not necessarily focused on the underlying
-declarative nature of lambda Prolog.  For example, ELPI has various
-extensions, such as ``spilling'' that are designed to make some
-aspects of logic programming appear to be more functional.
-
-In this paper, we limit our attention mostly to using lambda Prolog
-for declarative processing: i.e., where logic plays a central role in
-describing operations.  (Type checking is an example of such
-processing, but we do not focus on type checking here).
-
-Of course, being able to do declarative and not-so-declarative
-operations within the same language makes that language all the more
-useful.
-
-  Given that lP and Coq are both examples of mixing the
-  lambda-calculus with logic, it is important to understand their
-  difference, something that can be very confusing.  Take the
-  confusion around the term HOAS.  In some circles, this is a
-  appropriate and sensible approach.  In other circles, it is
-  semantically questionable.  But, in fact, this concept (using
-  meta-level binding for object-level binding) plays our 
-  differently depending on the meta language!
-
-  The origin story to ML (Meta language) is that it supported
-  LFC-style tactics and tacticals.  The strong typing of ML provided
-  some guarantees: only theorems were ever given type thm.  But many
-  steps of evolution have now taken place since then.  Isabelle's
-  different approach to tactics.  Logic and proof theory, when used
-  with Church's simple theory of types and Gentzen's sequent calculus,
-  provides a far richer meta language: that provides an immediate and
-  declarative treatment of bindings as well as for proof search
-  (without the need of exception handling, etc).  [I'm making the case
-  that FP is not a particularly natural framework for proof
-  assistants.  Of course, lots of libraries have now been built in the
-  FP style.]
-
-  - Various dichotomies we find in this work.
-    - proof theory vs type theory
-    - logic programming vs functional programming
-    - sequent calculus vs natural deduction
-
-  - While both lambda Prolog and Coq are based on an amalgamation of
-    $\lambda$-calculus and logic, these two languages take strikingly
-    different approaches to bindings.  In lambda Prolog, bindings are
-    syntactic and extensionality at any type is not available.  In
-    Coq, however, higher-order typed quantification encodes
-    extensional functions.  In lambda Prolog, (exists w. [lambda x. x]
-    = [lambda x. w]) fails, no matter the type of w.  If this query
-    was about functions, then this equation would be true when the
-    type of w is a singleton.  But in lP, this is a query only about
-    binding structures, and this fails to be proved since there is no
-    capture-avoiding substitution (the only substitution supported by
-    Church's STT) for w that makes the resulting two terms alpha,
-    beta, eta convertible.  [Is this true in Coq?  What is equality
-    here?]
-
-  - On the proof theory side of this work, we use two different proof
-    systems.
-
-    - LJF, here we have undefined predicates, no induction
-    - muMALL, were we have inductive definitions of predicates
-
-    The Coq world provides for both of these kinds of predicate using
-    Prop (corresponds to undefined) and Set (inductively defined).
-
-\section{Elaboration of external proof certificates for the Coq
-  kernel}
-
-Follow the text in \cite{blanco20coq}.
-
-\section{Property based testing for Coq}
-
-\section{Conclusion}
-
-% Not part of git repo.  Need to find a fix
-\bibliography{../../../../repo/parsifal/papers/references/master}
-
-\end{document}
+Before we present these two examples, we
+first discuss the different world views that the integration of ELPI
+and Coq forces us to confront.

### 2020-10-26 papers/types2020/main.tex
some small progress on sec 4
Split out another two sections.

--- 2020-10-21 papers/types2020/main.tex
@@
 
-\section{Introduction}
-\label{sec:intro}
-
-Recently, C. Sacerdoti Coen and E. Tassi developed the ELPI
-implementation \cite{dunchev15lpar} of \lP 
-\cite{miller12proghol}.  More recently, Tassi has made ELPI available
-as a plugin to the Coq proof assistant \cite{tassi18coqpl}.
-%
-This implementation of \lP extends earlier implementations in
-primarily two directions.  First, ELPI adds a notion of contraints,
-which makes it more expressive than, say, the Teyjus implementation of
-\lP \cite{nadathur99cade}.  Second, the plugin version of ELPI comes
-equip with an API for accessing the Coq environment and a quotation
-and anti-quotation syntax allowing for Coq expressions to be mixed in
-with \lP program elements.
-
-ELPI provides a means for treating bindings and unification (of
-structures containing bindings) that are convenient for various
-meta-programming tasks in logic programming all things useful for
-tactic programming in Coq (proof refinement).  For examples,
-ELPI programmers are spared from dealing with low-level aspects of
-representing bindings (e.g., De Bruijn indexes) while still having
-clean and effective ways to manipulate binding structures.
-%
-In short, ELPI appears to be a useful meta-language for Coq
-\cite{coen19mscs,tassi18coqpl,tassi19itp}.
-%
-In this paper, we outline two applications of the Coq plugin for ELPI
-that support this claim.
-%
-With these examples, we shall illustrate that ELPI is a useful
-meta-language not just because for its treatment of bindings and
-syntactic structures in general but also because it is actually based
-on a sound implementation of a higher-order intuitionistic logic.
-%
-Type inference is a well-known example of a meta-programming task for
-which logic programming often provides immediate and elegant
-implementations via the sound implementation of logic.
-%
-In this paper, we present two additional project in which an
-implementation of proof search provides direct, elegant, and compact
-implementations of meta-programming tasks.
-%
-Before we present these two examples, we first discuss the different
-world views that the integration of ELPI and Coq focuses us to
-confront.
-
-\section{Two cultures}
-
-{\color{red} The rest of this section are just notes.}
-
-These applications are not necessarily focused on the underlying
-declarative nature of lambda Prolog.  For example, ELPI has various
-extensions, such as ``spilling'' that are designed to make some
-aspects of logic programming appear to be more functional.
-
-In this paper, we limit our attention mostly to using lambda Prolog
-for declarative processing: i.e., where logic plays a central role in
-describing operations.  (Type checking is an example of such
-processing, but we do not focus on type checking here).
-
-Of course, being able to do declarative and not-so-declarative
-operations within the same language makes that language all the more
-useful.
-
-  Given that lP and Coq are both examples of mixing the
-  lambda-calculus with logic, it is important to understand their
-  difference, something that can be very confusing.  Take the
-  confusion around the term HOAS.  In some circles, this is a
-  appropriate and sensible approach.  In other circles, it is
-  semantically questionable.  But, in fact, this concept (using
-  meta-level binding for object-level binding) plays our 
-  differently depending on the meta language!
-
-  The origin story to ML (Meta language) is that it supported
-  LFC-style tactics and tacticals.  The strong typing of ML provided
-  some guarantees: only theorems were ever given type thm.  But many
-  steps of evolution have now taken place since then.  Isabelle's
-  different approach to tactics.  Logic and proof theory, when used
-  with Church's simple theory of types and Gentzen's sequent calculus,
-  provides a far richer meta language: that provides an immediate and
-  declarative treatment of bindings as well as for proof search
-  (without the need of exception handling, etc).  [I'm making the case
-  that FP is not a particularly natural framework for proof
-  assistants.  Of course, lots of libraries have now been built in the
-  FP style.]
-
-  - Various dichotomies we find in this work.
-    - proof theory vs type theory
-    - logic programming vs functional programming
-    - sequent calculus vs natural deduction
-
-  - While both lambda Prolog and Coq are based on an amalgamation of
-    $\lambda$-calculus and logic, these two languages take strikingly
-    different approaches to bindings.  In lambda Prolog, bindings are
-    syntactic and extensionality at any type is not available.  In
-    Coq, however, higher-order typed quantification encodes
-    extensional functions.  In lambda Prolog, (exists w. [lambda x. x]
-    = [lambda x. w]) fails, no matter the type of w.  If this query
-    was about functions, then this equation would be true when the
-    type of w is a singleton.  But in lP, this is a query only about
-    binding structures, and this fails to be proved since there is no
-    capture-avoiding substitution (the only substitution supported by
-    Church's STT) for w that makes the resulting two terms alpha,
-    beta, eta convertible.  [Is this true in Coq?  What is equality
-    here?]
-
-  - On the proof theory side of this work, we use two different proof
-    systems.
-
-    - LJF, here we have undefined predicates, no induction
-    - muMALL, were we have inductive definitions of predicates
-
-    The Coq world provides for both of these kinds of predicate using
-    Prop (corresponds to undefined) and Set (inductively defined).
-
-\section{Elaboration of external proof certificates for the Coq
-  kernel}
-
-Follow the text in \cite{blanco20coq}.
-
-\section{Property based testing for Coq}
+\input intro
+\input cultures
+\input sec3
+\input pbt
+\input elab
 
@@
 
-% Not part of git repo.  Need to find a fix
-\bibliography{../../../../repo/parsifal/papers/references/master}
+\bibliography{l}
+\end{document}
 
-\end{document}
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2020-10-26 papers/types2020/pbt.tex

--- 2020-10-22 papers/types2020/pbt.tex
@@
-\section{Property-based testing for Coq}
+\section{Revisiting property-based testing for Coq}
 
-Simplest setting since we are only sending back to Coq a term (a
-counterexample), not a general proof.
+In a previous paper~\cite{blanco19ppdp}, we presented a
+proof-theoretical reconstruction of property-based testing (PBT) of
+relational specifications, adopting techniques from foundational proof
+certificates to account for several features of these testing
+paradigm: from various \emph{generation} strategies, to
+\emph{shrinking} and fault localization.
 
+Given the connection that \textsf{coq-elpi} offers between logic
+programming and the internals of Coq, it is natural to extend the FPC-driven logic programming interpreter of the previous section to perform PBT of \lstinline{Inductive} types.
+
+While nothing prevents us from porting all the features we discussed in~\cite{blanco19ppdp}, for the sake of this paper we will implement only FPC corresponding to different flavors \emph{exhaustive} generation and their combination, as adopted, e.g., in SmallCheck~\cite{smallcheck} and  $\alpha$Check~\cite{cheney_momigliano_2017}.
+
+Of course, \textsf{QuickChick}~\cite{QChick}
+(\url{https://softwarefoundations.cis.upenn.edu/qc-current}) is a
+sophisticated and far-reaching PBT tool for Coq, which is based on a
+different perspective: being a clone of Haskell's QuickCheck, it
+emphasizes testing executable (read \emph{decidable)} specifications
+with random generators.
+While current research~\cite{LampropoulosPP18} aims to increase automation, it is fair to say testing, in particular of  relational specifications, is still very labor intensive.
+We do not intend to compete with  \textsf{QuickChick}, but as we shall see, as far as \lsti{Inductive} definitions that corresponds to pure Horn programs, we can test them immediately without any decidability proof or configuration of generators and shrinkers.
+
+\subsection{PBT as proof reconstruction}
+\label{ssec:pbt-lp}
+
+\begin{metanote}
+  I'm just saying the bare minimum -am
+\end{metanote}
+
+ If we view  a property as a logical formula
+\(\forall x [(\tau(x)\wedge P(x)) \supset Q(x)]\) where $\tau$ is a
+typing predicate and $P$ and $Q$ are two other predicates defined
+using Horn clause specifications, providing a counter-example
+consists of  negating the property, and \emph{searching} for a proof of
+\[
+  \exists x [(\tau(x)\wedge P(x)) \wedge \neg Q(x)]
+  \tag{*}\label{eq:full}
+\]
+%\(\exists x [(\tau(x)\land P(x)) \land \neg Q(x)]\).
+
+In our current setting $P$ and $Q$ will be \lsti{Inductive} propositions, while $\tau$   are honest-to-goodness datatypes.
+We will treat the two quite differently, in so far as elements of those types will be \emph{generated}, while predicates will only be \emph{checked}.
+This distinction plays also a part in interpreting the negation sign.
+%
+When using a focused proof system for logic extended with fixed
+points~\cite{baelde12tocl}, negation corresponds to the usual intutionistic interpretation.  % proofs of formulas such as
+% \[
+%   \exists x [(\tau(x)\wedge P(x)) \wedge \neg Q(x)]
+%   \tag{*}\label{eq:full}
+% \]
+% are a single \emph{bipole}: that is, when reading a proof bottom up, a
+% positive phase is followed on all its premises by a single negative
+% phase that completes the proof.%
+% In particular, the positive phase corresponds to the \emph{generation} phase
+% and the negative phase corresponds to the \emph{testing} phase.
+% %
+
+% Instead of giving a full focused proof system of a logic including
+% fixed points (since, as we will argue, that proof system will not, in
+% fact, be needed to account for PBT ), we offer the following analogy.
+% %
+% Suppose that we are given a finite search tree and we are asked to
+% prove that there is a secret located in one of the nodes of that
+% tree.
+% %
+% A proof that we have found that secret can be taken to be a
+% description of the path to that node from the root of the tree: that
+% path can be seen as the proof certificate for that claim.
+% %
+% On the other hand, a proof that no node contains the secret is a
+% rather different thing: here, one expects to use a blind and
+% exhaustive search (via, say, depth-first or breath-first search) and
+% that the result of that search never discovers the secret.
+% %
+% A proof of this fact requires \emph{no} external information: instead it
+% requires a lot of computation involved with exploring the tree until
+% exhaustion.
+% %
+% This follows the familiar pattern where the positive (generate) phase
+% requires external information while the negative (testing) phase
+% requires none.
+%
+However, for the sake of PBT, we can identify a proof certificate for
+$(\ref{eq:full})$ with a proof certificate for
+\[
+  \exists x [\tau(x)\wedge P(x)].
+  \tag{**}\label{eq:short}
+\]
+and will resort to negation-as-failure to check that the conclusion does not hold.
+This also means that we will not be able to return a Coq proof terms for the refutation of our property --- and neither does \textsf{QuickChick}, which runs at the OCaml level --- although we can return the witness for the existential.
+% Such a certificate would contain the witness (closed) term $t$ for the
+% existential quantifier and sufficient information to confirm that
+% $P(t)$ can be proved (a proof of a typing judgment such as $\tau(t)$
+% is usually trivial).
+% Since a proof certificate for the existence-of-a-counterexample formula
+% $(\ref{eq:full})$ can be taken as a proof certificate of
+% $(\ref{eq:short})$, then we only need to consider proof certificates
+% for Horn clause programs.
+%
+%Since the logic of Horn clauses is rather simple,
+% We illustrate next what such proofs and proof certificates look like
+% for the rather simple logic of Horn clauses.
+
+\begin{metanote}
+  TODO next
+  \begin{itemize}
+  \item The elpi dep-pbt tactic
+  \item the simple reverse example
+  \item a bit of details about the \texttt{aexp} example
+  \end{itemize}
+\end{metanote}
 %%% Local Variables:
@@
 %%% End:
+
+% LocalWords:  Bedwyr bipole elpi QuickCheck OCaml TODO

### 2020-10-26 papers/types2020/pbt.tex
"ordered" example in pbt.tex

@@
   \item The elpi dep-pbt tactic
-  \item the simple reverse example
+  \item the simple ordered example
   \item a bit of details about the \texttt{aexp} example
@@
 \end{metanote}
+
+Let us consider a  relational specification of orered list, as well as insertion in such lists:
+\begin{lstlisting}
+Inductive insert (x:nat) : list nat -> list nat -> Prop :=
+| i_n : is_nat x -> insert x [] [x]
+| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
+| i_c : forall y: nat, forall ys rs, x > y -> insert x ys rs -> insert x (y :: ys) (x :: rs).
+
+Inductive ordered : list nat -> Prop :=
+| o_n : ordered []
+| o_s : forall x : nat, ordered [x]
+| o_c : forall (x y : nat), forall xs,  ordered xs  -> x <= y -> ordered (x:: y :: xs).     
+\end{lstlisting}
+%
+A property we may wish to checl before embaring in a formal proof is whether insertion preserves ordered-ness. Since we did slip in an error, our tactic reports a counter-example:
+\begin{lstlisting}
+Conjecture ins_ord:  forall x xs rs, ordered xs -> insert x xs rs -> ordered rs.
+\end{lstlisting}
 %%% Local Variables:

### 2020-10-26 papers/types2020/pbt.tex
finishing the ordered example

@@
   \item The elpi dep-pbt tactic
-  \item the simple ordered example
+  \item  the simple ordered example [DONE]
   \item a bit of details about the \texttt{aexp} example
@@
 
-Let us consider a  relational specification of orered list, as well as insertion in such lists:
+Let us consider a  relational specification of ordered list, as well as insertion in such lists:
 \begin{lstlisting}
@@
 %
-A property we may wish to checl before embaring in a formal proof is whether insertion preserves ordered-ness. Since we did slip in an error, our tactic reports a counter-example:
+A property we may wish to check before embarking in a formal proof is whether insertion preserves ordered-ness. Since we did slip in an error, our tactic reports a counter-example, e.g.,
+\verb|(* Proof Term: [0, [0; 1; 0]] *)|:
 \begin{lstlisting}
 Conjecture ins_ord:  forall x xs rs, ordered xs -> insert x xs rs -> ordered rs.
+intros x xs rs  Ho Hi.
+elpi dep_pbt height 5 (Ho /\ Hi ) (x) (xs).
+Abort.
 \end{lstlisting}
+In this query the tactic tests the hypotheses \lsti{Ho} and \lsti{Hi}
+against data \lsti{x,xs} generated exhaustively up to proofs bounded by
+at most $5$ unfoldings of the library \lsti{Inductive} definition of
+\lsti{nat} and \lsti{list}.
+
+Note that the same query in \textsf{QuickChick} would have required much more setup: if we wished to proceed relationally as above we would have had to provide a proof of decidability of the relevant notions.
+Were we to use functions, then we would have to implement a generator and shrinker for ordered lists.
 %%% Local Variables:

### 2020-10-26 papers/types2020/app.tex
appendix

\section{Appendix}
\label{sec:app}

Stuff we may not have space for in the paper
\begin{itemize}
\item The LJF sequent rules
\item the code for
  \lsti{dep_pbt}

\item \dots
\end{itemize}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

% LocalWords:  LJF

### 2020-10-26 papers/types2020/cultures.tex
Changing machine

@@
 programming paradigm provides a completely natural, elegant, and
-powerful method, called $\lambda$-tree syntax, for treating bindings
+powerful method for treating bindings
 within syntax.
@@
 (quantifiers), and in proofs (eigenvariables).  For this reason, the
-term $\lambda$-tree syntax was introduced to name this particular take
-on HOAS \cite{miller19jar}.  The Abella proof assistant
-\cite{baelde14jfr} was designed, in part, to allow for inductive and
-co-inductive proofs for specifications using the $\lambda$-tree syntax
-approach.
+term \emph{$\lambda$-tree syntax} was introduced to name this
+particular take on HOAS \cite{miller19jar}.  The Abella proof
+assistant \cite{baelde14jfr} was designed, in part, to allow for
+inductive and co-inductive proofs for specifications using the
+$\lambda$-tree syntax approach.
 

### 2020-10-26 papers/types2020/conc.tex
adding file for conclusion
some work on fig 1

\section{Conclusion and future work}

A mention to the use of \emph{delays} in \texttt{check}: instead of generating only ground terms, we could generate say lists with at least two elements as \texttt{[1;2|Tail]} or simply not commit to a base type. This  is similar to the use of needed narrowing in LazySmallCheck to generate partial trace or the idea of using CLP for data generation (see Dubois' FocalTest or the stuff by Fabio Fioravanti).
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-10-26 papers/types2020/sec3.tex
some small progress on sec 4
adding file for conclusion
some work on fig 1

@@
 
-\begin{metanote}
-  add proof terms and backchaining to sequents and code -- now sig is commented out
-\end{metanote}
 \begin{footnotesize}
 \begin{figure}
-\newcommand{\XXi}{{\color{blue}{\Xi}}}
+  \newcommand{\XXi}{}
+  \newcommand{\bc}[2]{#1 >\!> #2}
 \[
 \infer{\XXi\vdash t = t}
-      {\eqExpert{\XXi}}
+{%\eqExpert{\XXi}
+}
 \qquad
 \infer{\XXi\vdash \true}
-      {\trueExpert{\XXi}}
+{%\trueExpert{\XXi}
+}
 \qquad
 \infer{\XXi\vdash G_1\vee G_2}
-      {\XXi'\vdash G_i\qquad \orExpert{\XXi}{\XXi'}{i}}
+{\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
+}
 \qquad
 \infer{\XXi\vdash \exists x. G}
-      {\XXi'\vdash G[t/x]\qquad \someExpert{\XXi}{\XXi'}{t}}
-\]
-\vskip -6pt
-\[
+{\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
+}
+\qquad 
 \infer{\XXi\vdash G_1\wedge G_2}
-      {\XXi_1\vdash G_1\quad \XXi_2\vdash G_2\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}}
+{\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
+}
 \qquad
 \infer{\XXi\vdash A}
-      {\XXi'\vdash G \quad (A~\hbox{\tt :-}~G)\in\instan\Pscr
-                     \quad \unfoldExpert{\XXi}{\XXi'}}
+      {\bc D A \quad D \in \Pscr}
 \]
-\caption{A proof system augmented with proof certificates and
-  additional ``expert'' premises.}
+\vskip -18pt
+\[
+  \infer{\bc A A}{}{}\qquad
+  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
+  \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
+\]
+\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
+\caption{A proof system for Horn logic and its encoding in coq-elpi .}
 \label{fig:augmented}
@@
 \lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
-\caption{A simple proof checking kernel.}
+\caption{A dependent kernel for Coq terms.}
 \label{fig:kernel}
@@
 
-\begin{metanote}
-  Rethink and update (with the right connectives) these FPC
-\end{metanote}
 \begin{figure}

### 2020-10-27 papers/types2020/sec3.tex
after main merge

@@
 
-{\color{red} The goal in this section is to say a bit about focusing
-  generally, and to give a bit of detail about FPC for Horn clauses.
-  Initial draft of text taken from \cite{blanco19ppdp}.
-}
+% {\color{red} The goal in this section is to say a bit about focusing
+%   generally, and to give a bit of detail about FPC for Horn clauses.
+%   Initial draft of text taken from \cite{blanco19ppdp}.
+% }
 
 \begin{metanote}
-  (AM): I would start with the logical reconstruction of \texttt{check}, adding the LJT-style proof terms~c\cite{Herbelin94,LengrandDM06} to fig \ref{fig:augmented} and then reverse engineer what we want/need to say about focusing and FPC.
+  (AM): Start with the logical reconstruction of \texttt{check}, to and then reverse engineer what we want/need to say about focusing and FPC.
 We \textbf{do} not talk about PBT yet, just logic programming (with proof terms).
 In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
+
+  One way to proceed:
+
+  \begin{itemize}
+  \item One figure fig \ref{fig:interp} for Horn logic over Coq terms (no certificates) and its encoding with quotations 
+  \item Then the logic behind the dependent kernel with certificates and proof terms
+  \end{itemize}
 \end{metanote}
-\subsection{Focused proof systems}
-\label{ssec:focused}
+
+\subsection{Horn logic in coq-elpi}
 
@@
 \emph{backchaining} phase~\cite{miller91apal}.
+
+In Fig.~\ref{fig:interp} we give a proof system for Horn logic and its encoding in coq-elpi
+\todo{explain the interface, quotations etc}
+
+\begin{footnotesize}
+  \begin{figure}
+  \newcommand{\XXi}{}
+  \newcommand{\bc}[2]{#1 >\!> #2}
+\[
+\infer{\XXi\vdash t = t}
+{%\eqExpert{\XXi}
+}
+\qquad
+\infer{\XXi\vdash \true}
+{%\trueExpert{\XXi}
+}
+\qquad
+\infer{\XXi\vdash G_1\vee G_2}
+{\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
+}
+\qquad
+\infer{\XXi\vdash \exists x. G}
+{\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
+}
+\qquad 
+\infer{\XXi\vdash G_1\wedge G_2}
+{\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
+}
+\qquad
+\infer{\XXi\vdash A}
+      {\bc D A \quad D \in \Pscr}
+\]
+\vskip -18pt
+\[
+  \infer{\bc A A}{}{}\qquad
+  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
+  \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
+\]
+\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
+\caption{A proof system for Horn logic and its encoding in coq-elpi .}
+\label{fig:interp}
+%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
+%\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
+%\caption{A dependent kernel for Coq terms.}
+%\label{fig:kernel}
+\end{figure}
+\end{footnotesize}
+
+\subsection{Focused proof systems with certificates}
+\label{ssec:focused}
+
 %
@@
 
-\subsection{Certificate checking with expert predicates}
+\todo{Talk about the LJT thing: grammar for formulae and proof terms\cite{Herbelin94,LengrandDM06}?}
+
+%\subsection{Certificate checking with expert predicates}
 %\subsection{Adding certificates and experts to a proof system}
@@
 \begin{figure}
-  \newcommand{\XXi}{}
+    \newcommand{\XXi}{{\color{blue}{\Xi}}}
+
   \newcommand{\bc}[2]{#1 >\!> #2}
@@
 \lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
-\caption{A proof system for Horn logic and its encoding in coq-elpi .}
+\caption{dep kernel.}
 \label{fig:augmented}
@@
 \lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
-\caption{A dependent kernel for Coq terms.}
+\caption{Implementation of a  dependent kernel for Coq terms.}
 \label{fig:kernel}
@@
 \end{footnotesize}
+
 Figure~\ref{fig:augmented} contains a simple proof system for Horn
@@
 %%% End:
+
+% LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE

### 2020-10-27 papers/types2020/cultures.tex
Finished a revision of the first two sections.

--- 2020-10-26 papers/types2020/cultures.tex
@@
 
-In proof theory, one learns quickly that things always come in pairs:
-negative/positive, left/right, bottom-up/top-down,
-premises/conclusion, introduction/elimination,
-% determinism/nondeterminism,
-% invertible/noninvertible, 
-etc.
-%
-When we examine the larger setting of this project of linking a logic
-programming engine with Coq and its kernel, we find a large number of
-pairing that are valuable to explicitly describe.
+When studying structural proof theory, one learns quickly that many
+concepts come in pairs: negative/positive, left/right,
+bottom-up/top-down, premises/conclusion, introduction/elimination,
+etc.  When we examine the larger setting of this project of linking a
+logic programming engine with Coq and its kernel, we find a large
+number of new pairing that is valuable to explicitly discuss.
 
@@
 
-In many ways, (structural) proof theory is more elementary and
-low-level than most approaches to type theory.  For example, type
-theories usually answer the question ``What is a proof?'' with the
-response ``a dependently typed $\lambda$-term''.  That is, when
-describing a type theory, one usually decides that a proof is a
-certain kind of term within the system.  In contrast, the proof theory
-treats logical propositions and proofs as separate.  For example,
-proof theory does not assume that there are terms within the logic
-that describe proofs.
+In many ways, proof theory is more elementary and low-level than most
+approaches to type theory.  For example, type theories usually answer
+the question ``What is a proof?'' with the response ``a dependently
+typed $\lambda$-term''.  That is, when describing a type theory, one
+usually decides that a proof is a certain kind of $\lambda$-term
+within the system.  In contrast, the proof theory treats logical
+propositions and proofs as separate.  For example, proof theory does
+not assume that there are terms within the logic that describe proofs.
 
-\begin{metanote}
-  This paragraph seem to identify proof and type theory with sequents and nat ded and while
-  I agree with the gist (although people such as Frank would argue that sequents are a way of organizing proof search for nat ded), we're not gonna make a lot of friends at types and more importantly is a big hammer to nail the point that logic programming is helpful with Coq --am
-\end{metanote}
+%% \begin{metanote}
+%%   This paragraph seem to identify proof and type theory with sequents
+%%   and nat ded and while I agree with the gist (although people such as
+%%   Frank would argue that sequents are a way of organizing proof search
+%%   for nat ded), we're not gonna make a lot of friends at types and
+%%   more importantly is a big hammer to nail the point that logic
+%%   programming is helpful with Coq --am
+%% \end{metanote}
+
+%% DM: I've tried to address your concerns by adding an additional
+%% paragraph and references (below).
 
@@
 the weakening and contraction structural rules on the right-side of
-sequents~\cite{gentzen35}.  This discovery lead him away from natural
-deduction to the invention of multiple-conclusion sequent calculus, a
-very important kind of proof that has not been given a satisfactory
-representation in type theory.  This same innovation of Gentzen also
-opened the way to the invention of another key proof theoretic
-discovery, that of linear logic \cite{girard87tcs}.
-Similarly, sequent calculus provides an elegant presentation for
-linear logic while most treatments of linear logic in type theory
-usually only capture linearity in a partial fashion.
+sequents~\cite{gentzen35}.  This invention lead him away from natural
+deduction to multiple-conclusion sequent calculus, a very important
+kind of proof that has not been given a satisfactory representation in
+type theory.  This same innovation of Gentzen also opened the way to
+the invention of another key proof-theoretic discovery, that of linear
+logic \cite{girard87tcs}.  Similarly, sequent calculus provides an
+elegant presentation of linear logic while most treatments of linear
+logic in type theory usually only capture linearity in a partial
+fashion.
 
-Proof theory also has rather immediate and natural treatments for
-co-induction (as well as induction: see
-\cite{baelde12tocl,momigliano12jal}) whereas the treatment of
-coinduction within type theory remains more challenging and is well
-developed only for various syntactic conditions (see, for example,
-\cite{bertot08entcs}).
-\begin{metanote}
-  The criticism applies to Coq, less to type theory, see co-patterns --am
-\end{metanote}
+As is often observed, however, sequent calculus is too low-level to be
+used explicitly as capturing the ``essence of a proof''.  Recent
+developments in focused proof systems \cite{andreoli92jlc,liang09tcs},
+makes it possible to extract from sequent calculus not only natural
+deduction proofs \cite{pimentel16lsfa}, but also proof nets
+\cite{chaudhuri08tcs}, and Herbrand-style expansion trees
+\cite{chaudhuri16jlc}.  The role of focused proof systems to
+characterize classes of proof structures is described in the next
+section and used in our two example applications of ELPI with Coq.
+
+%% DM I no longer think this comment is useful for this paper, so I
+%% delete it instead of trying to fix it.
+
+%% Proof theory also has rather immediate and natural treatments for
+%% co-induction (as well as induction: see
+%% \cite{baelde12tocl,momigliano12jal}) whereas the treatment of
+%% coinduction within type theory remains more challenging and is well
+%% developed only for various syntactic conditions (see, for example,
+%% \cite{bertot08entcs}).
+%% \begin{metanote}
+%%   The criticism applies to Coq, less to type theory, see co-patterns --am
+%% \end{metanote}
 
@@
 different ways.  The functional programming paradigm, following the
-Curry-Howard correspondence, models computation abstractly as the
+Curry-Howard correspondence, models computation abstractly as 
 $\beta$-reduction of natural deduction proof: that is, computation is
-modeled using \emph{proof normalization}.  On the other hand, the
+modeled as \emph{proof normalization}.  On the other hand, the
 logic programming paradigm, following the notion of goal-directed
@@
 cut-free sequent calculus proofs: that is, computation is modeled
-using \emph{proof search}.
+as \emph{proof search}.
 
-Proof search has proved to have some richness that is hard to capture
-in the proof normalization setting.  In particular, Gentzen's
-invention of \emph{eigenvariables} are a kind of a proof-level
-binder.  In the proof normalization setting, such eigenvariables
-are instantiated during proof normalization.  In contrast, during the
-search for cut-free proofs, eigenvariables are part of the syntax of
-terms and formulas.  As a result, they can be used in the treatment of
-bindings in data structures more generally.  Hence, the logic
-programming paradigm provides a completely natural, elegant, and
-powerful method, called $\lambda$-tree syntax, for treating bindings
-within syntax.
+Proof search has features that are hard to capture in the proof
+normalization setting.  In particular, Gentzen's \emph{eigenvariables}
+are a kind of a proof-level binder.  In the proof normalization
+setting, such eigenvariables are instantiated during proof
+normalization.  However, during the search for cut-free proofs,
+eigenvariables do not vary, and they are part of the syntax of terms
+and formulas.  As a result, they can be used in the treatment of
+bindings in data structures more generally.  Such eigenvariables can
+be used by the logic programming paradigm to provide a natural,
+elegant, and powerful computing method with bindings within syntax.
 
-In this paper, we shall use two different proof systems.  The first
-logic is first-order intuitionistic logic.  Here, predicates
-are not inductively defined and quantification will be limited to
-range over first-order term.  Actually, we shall examine only the Horn
-clause fragment of that logic.
-%
-The second logic will allow for inductively defined predicates using a
-fixed point notation.   Given again our restriction to essentially
-Horn clause specifications, it is possible to view intuitionistic
-reasoning with such specification as if they are actually
-specification in $\mu$MALL, an extension of multiplicative additive
-linear logic with the least and greatest fixed point operators
-\cite{baelde12tocl,baelde07lpar,heath19jar}.
-%
-The difference between these two logics and associate proof theories
-is reminiscent to the difference between the type \lsti{Prop}
-(corresponding to undefined propositions) and Sets (corresponding to
-inductively defined propositions).
-\begin{metanote}
-  Still not onboard with this --am
-\end{metanote}
+It is worth noting that the role of the cut rule and of
+cut-elimination is different in these two approaches to computing.
+In the proof normalization paradigm, the cut rule can be used to model
+$\beta$ reduction, especially via the explicit substitution approach
+\cite{lengrand06phd}. In the proof search paradigm, since computing involves
+the search for cut-free proofs, the cut rule plays no part in the
+performance of computation.  However, cut and cut-elimination can be
+used to reason about computation: for example, cut-elimination can be
+used to prove ``substitution lemmas for free'' \cite{gacek12jar}.
+
+%% The following paragraph seems out-of-place so I delete it now.
+
+%% In this paper, we shall use two different proof systems.  The first
+%% logic is first-order intuitionistic logic.  Here, predicates
+%% are not inductively defined and quantification will be limited to
+%% range over first-order term.  Actually, we shall examine only the Horn
+%% clause fragment of that logic.
+%% %
+%% The second logic will allow for inductively defined predicates using a
+%% fixed point notation.   Given again our restriction to essentially
+%% Horn clause specifications, it is possible to view intuitionistic
+%% reasoning with such specification as if they are actually
+%% specification in $\mu$MALL, an extension of multiplicative additive
+%% linear logic with the least and greatest fixed point operators
+%% \cite{baelde12tocl,baelde07lpar,heath19jar}.
 
@@
 
-Given that \lP and Coq are both examples of mixing the
+Given that \lP and Coq both result from a mixing of the
 $\lambda$-calculus with logic, it is important to understand their
@@
 system, the HOAS approach leads to encoding binding structures within
-terms using functions.  Such encodings have results in surprisingly
-complicated encodings, allowing, for example, for \emph{exotic terms}
-\cite{despeyroux95tlca} and to structures on which induction is not
-possible \cite{roeckl01fossacs}.  In the logic programming setting,
-particularly \lP, HOAS is well supported in both the availability of
-binders allowed with terms ($\lambda$-bindings), in formulas
-(quantifiers), and in proofs (eigenvariables).  For this reason, the
-term $\lambda$-tree syntax was introduced to name this particular take
-on HOAS \cite{miller19jar}.  The Abella proof assistant
-\cite{baelde14jfr} was designed, in part, to allow for inductive and
-co-inductive proofs for specifications using the $\lambda$-tree syntax
-approach.
+terms using functions.  Such encodings are surprisingly complicated
+since they allow for \emph{exotic terms} \cite{despeyroux95tlca} and
+for structures on which induction is not possible
+\cite{roeckl01fossacs}.  In the logic programming setting,
+particularly \lP, HOAS is well supported since bindings are allowed
+with terms ($\lambda$-bindings), in formulas (quantifiers), and in
+proofs (eigenvariables).  For this reason, the term
+\emph{$\lambda$-tree syntax} was introduced to name this particular
+take on HOAS \cite{miller19jar}.  The Abella proof assistant
+\cite{baelde14jfr} was designed, in part, to provide inductive and
+co-inductive inference involving specifications using the
+$\lambda$-tree syntax approach.
 
@@
 illustrated by considering how they are used in the specification of
-tactics that are popular in modern proof assistants.  In fact, the
-origin story for the ML functional programming language was that ML
-was the meta-language for implementing the LCF suite of tactics and
-tacticals \cite{gordon79}.  In order to implement tactics, ML adopted
-not only novel features such as polymorphically typed higher-order
-functional programming but also the non-functional mechanisms of
-failure and exception handling.  While \lP is also based on
-polymorphically typed higher-order relational programming, it also
+tactics.  In fact, the origin story for the ML functional programming
+language was that it was the meta-language for implementing the LCF
+suite of tactics and tacticals \cite{gordon79}.  To implement tactics,
+ML adopted not only novel features such as polymorphically typed
+higher-order functional programming but also the non-functional
+mechanisms of failure and exception handling.  While \lP is also based
+on polymorphically typed higher-order relational programming, it also
 comes with a completely declarative version of failure and
@@
 
-While \lP is a typed language, it only uses simple types and these are
-more often used to indicate syntactic categories.  Any dependency
-information needs to be captured using predicates and quantifiers.
+While \lP is a typed language, it only uses simple, polymorphic types,
+and these are more often used to indicate syntactic categories.  Any
+dependency information needs to be captured using predicates and
+quantifiers.

### 2020-10-27 papers/types2020/intro.tex
Finished a revision of the first two sections.

@@
 Recently, C. Sacerdoti Coen and E. Tassi developed the ELPI
-implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol}.
-More recently, Tassi has made ELPI available as a plugin to the Coq
+implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
+more recently, Tassi has made ELPI available as a plugin to the Coq
 proof assistant \cite{tassi18coqpl}.  This implementation of \lP
 extends earlier implementations in primarily two directions.  First,
-ELPI adds a notion of constraints, which makes it more expressive than,
-say, the Teyjus implementation of \lP \cite{nadathur99cade}.  Second,
-the plugin version of ELPI comes equip with an API for accessing the
-Coq environment and a quotation and anti-quotation syntax allowing for
-Coq expressions to be mixed with \lP program elements.
+ELPI adds a notion of constraints, which makes it more expressive than
+the Teyjus implementation \cite{nadathur99cade} of \lP. Second, the
+plugin version of ELPI comes equipped with an API for accessing the
+Coq environment and a quotation and anti-quotation syntax for mixing
+for Coq expressions with \lP program elements.
 
-ELPI provides a means for treating bindings and unification (of
-structures containing bindings) that are convenient for various
-meta-programming tasks in logic programming all things useful for
-tactic programming in Coq (proof refinement).  For examples, ELPI
-programmers are spared from dealing with low-level aspects of
-representing bindings (e.g., De Bruijn indexes) while still having
-clean and effective ways to manipulate binding structures.  In short,
-ELPI appears to be a useful meta-language for Coq
-\cite{coen19mscs,tassi18coqpl,tassi19itp}.  
-Type inference is a well-known example of a meta-programming task for
-which logic programming often provides immediate and elegant
-implementations via the sound implementation of logic.  % In this paper,
-% we present two additional project in which an implementation of proof
-% search provides direct, elegant, and compact implementations of
-% meta-programming tasks.
-In this paper, we outline
-two applications of the Coq plugin for ELPI that support this claim.
-With these examples, we shall illustrate that ELPI is a useful
-meta-language not just because for its treatment of bindings and
-syntactic structures in general but also because it is actually based
-on a sound implementation of a higher-order intuitionistic logic.
-Before we present these two examples, we
-first discuss the different world views that the integration of ELPI
-and Coq forces us to confront.
+The logic programming interpreter that underlies ELPI provides several
+convenient features for the kind of meta-programming tasks that can
+arise within modern proof assistants.  For example, \lP provides a
+declarative and direct treatment of abstract syntax that contains
+bindings: that treatment includes capture-avoiding substitution,
+unification, and recursive programming.  The ELPI spares the
+programmer from dealing with low-level aspects of bindings
+representation (e.g., De Bruijn indexes) while still having clean and
+effective ways to manipulate binding structures.
+
+Since relations (not functions) play a central role in \lP, ELPI is
+capable to providing direct support for the many relations that arise
+in proof assistants.  Such relations include typing (e.g.,
+$M\colon\sigma$), evaluation (e.g., natural semantics specifications
+\cite{kahn87stacs,hannan93jfp}), and interaction (e.g., structured
+operational semantics \cite{plotkin81,milner89book}).  Felty has also
+made the point that LCF-style tactics and tacticals can be given an
+elegant and natural specification using the higher-order relational
+specifications provided by \lP \cite{felty93jar}.  Some recent
+implementations built using ELPI
+\cite{coen19mscs,tassi18coqpl,tassi19itp} support the usefulness of
+higher-order logic programming as a meta-programming language useful
+for Coq.
+
+In this paper, we present two applications of ELPI within Coq. With
+these examples, we shall illustrate that ELPI is useful not only
+because it has some useful meta-programming features (e.g., a builtin
+treatment of bindings in syntax) but also because it soundly
+implements a higher-order intuitionistic logic.  The latter aspect of
+ELPI makes it possible to encode the proofs and proof theory of
+specific subsets of the Coq logic.
+
+Before we can present these examples, we first highlight the rather
+striking differences in notions of computing and reasoning that arise
+on each side of the ELPI-Coq API.   We will also present a quick
+summary of the key proof theory concepts that are used by our example
+applications.

### 2020-10-27 papers/types2020/sec3.tex
fig with dep kerned in sequent form added

@@
 
+Stuff in red is from old paper
 \subsection{Horn logic in coq-elpi}
@@
   \newcommand{\XXi}{}
-  \newcommand{\bc}[2]{#1 >\!> #2}
+  \newcommand{\bc}[2]{#1 \Downarrow #2}
 \[
@@
 \label{ssec:focused}
+\todo{Adapt the discussion}
+\todo{Talk about the LJT thing: grammar for formulae and proof terms\cite{Herbelin94,LengrandDM06}?}
 
 %
+{\color{red}
 The notion of \emph{focused proof systems}~\cite{andreoli92jlc,liang09tcs}
@@
 certificate}.
-
-\todo{Talk about the LJT thing: grammar for formulae and proof terms\cite{Herbelin94,LengrandDM06}?}
+}
 
@@
 %\subsection{Adding certificates and experts to a proof system}
-\label{ssec:fpc}
+%\label{ssec:fpc}
 
@@
 
-  \newcommand{\bc}[2]{#1 >\!> #2}
+  \newcommand{\bc}[5]{#1;#2\vdash #3 : #4 \Downarrow #5}
 \[
-\infer{\XXi\vdash t = t}
-{%\eqExpert{\XXi}
+% \infer{\XXi\vdash t = t}
+% {%\eqExpert{\XXi}
+% }
+% \qquad
+% \infer{\XXi\vdash \true}
+% {%\trueExpert{\XXi}
+% }
+% \qquad
+% \infer{\XXi\vdash G_1\vee G_2}
+% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
+% }
+% \qquad
+% \infer{\XXi\vdash \exists x. G}
+% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
+% }
+% \qquad 
+\infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
+{\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
 }
 \qquad
-\infer{\XXi\vdash \true}
-{%\trueExpert{\XXi}
-}
-\qquad
-\infer{\XXi\vdash G_1\vee G_2}
-{\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
-}
-\qquad
-\infer{\XXi\vdash \exists x. G}
-{\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
-}
-\qquad 
-\infer{\XXi\vdash G_1\wedge G_2}
-{\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
-}
-\qquad
-\infer{\XXi\vdash A}
-      {\bc D A \quad D \in \Pscr}
+\infer{\XXi;\Gamma\vdash t : A}
+      {\bc{\XXi}{\Gamma} t D A \quad D \in \Pscr\cup\Gamma}
 \]
@@
 \[
-  \infer{\bc A A}{}{}\qquad
-  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
-  \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
+  \infer{\bc {\XXi}{\Gamma} {[]} A A}{\trueExpert{\XXi}}{}\qquad
+%  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
+  \infer{\bc{\XXi}{\Gamma}{(t\ l)} {\forall x:B.\, D} A}
+  {\XXi_1;\Gamma\vdash t : A\qquad \bc{\XXi_2}{\Gamma}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
 \]
-\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
+%\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
 \caption{dep kernel.}
@@
 
-Figure~\ref{fig:augmented} contains a simple proof system for Horn
-clause provability in which each inference rule is augmented with an
+Figure~\ref{fig:augmented} contains a simple proof system \dots  in which each inference rule is augmented with an
 additional premise involving an \emph{expert predicate}, a
@@
 %
-Although this proof system is a focused proof system, the
-richness of focusing is not apparent in this simplified 
-setting: thus, we drop the adjective ``focused'' from this point forward.
+\todo{Explain relations with other term assignments.
+Note that we do not use explicit substitutions (we are cut free).
+Explain the choice of not having certificates represent proof terms.
+Explain that the unfold clause handles defined atoms and defined connectives (conjunction, disjunction, existential and also equality: this may be less efficient and proof-theoretically induce pointless phase changes, but it is arguably very elegant e simplifies proof terms that are inherited by Coq }
 
-Figure~\ref{fig:kernel} contains the \lP implementation of the
+Figure~\ref{fig:kernel} contains the elpi implementation of the
 inference rules in Figure~\ref{fig:augmented}: here the infix
-turnstile $\vdash$ symbol is replaced by the \lst<i{check} predicate.
+turnstile $\vdash$ symbol is replaced by the \lsti{check} predicate.
 %
-Notice that it is easy to show that no matter how the expert predicates
-are defined, if the goal \lsti{check Cert B} is provable in \lP then
-\lsti{B} must be a sound consequence of the program clauses stored in
-\dots
+% Notice that it is easy to show that no matter how the expert predicates
+% are defined, if the goal \lsti{check Cert B} is provable in \lP then
+% \lsti{B} must be a sound consequence of the program clauses stored in
+% \dots
 % the \lsti{prog} predicate (which provides a natural implementation of
@@
 
+\paragraph{Certificates}
+
+\todo{Shorten it}
+\todo{Show the dprolog tactic? have an example? could  be
+\lsti{Goal  exists R, insert 2 [0;1] R.} and we would move the code for insert here}
+{\color{red}
 As we mentioned in the introduction, our notion of proof certificates
@@
 %
-In particular, the query \mbox{\lsti{sigma H\ (check (qsize 5 H) B)}}
+In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
 % \begin{lstlisting}
@@
 integers, only (user-defined) small integers could be produced.
+}
 

### 2020-10-28 papers/types2020/cultures.tex
a metanote in culture

@@
 approaches to type theory.  For example, type theories usually answer
-the question ``What is a proof?'' with the response ``a dependently
+the question ``What is a proof?'' with the response ``a (dependently)
 typed $\lambda$-term''.  That is, when describing a type theory, one
@@
 type theory.  This same innovation of Gentzen also opened the way to
-the invention of another key proof-theoretic discovery, that of linear
-logic \cite{girard87tcs}.  Similarly, sequent calculus provides an
-elegant presentation of linear logic while most treatments of linear
-logic in type theory usually only capture linearity in a partial
-fashion.
+another key proof-theoretic discovery, that of linear logic
+\cite{girard87tcs}.  Similarly, sequent calculus provides an elegant
+presentation of linear logic while most treatments of linear logic in
+type theory usually only capture linearity in a partial fashion.
+\begin{metanote}
+  `` partial fashion.'' Isn't that a bit vague? Some references? -am
+\end{metanote}
 

### 2020-10-28 papers/types2020/intro.tex

@@
 proof assistant \cite{tassi18coqpl}.  This implementation of \lP
-extends earlier implementations in primarily two directions.  First,
-ELPI adds a notion of constraints, which makes it more expressive than
+extends earlier implementations in primarily two directions:  First,
+ELPI adds a notion of constraints and constraints handling rules, which makes it more expressive than
 the Teyjus implementation \cite{nadathur99cade} of \lP. Second, the
@@
 declarative and direct treatment of abstract syntax that contains
-bindings: that treatment includes capture-avoiding substitution,
-unification, and recursive programming.  The ELPI spares the
+bindings, including capture-avoiding substitution,
+unification, and recursive programming.  ELPI spares the
 programmer from dealing with low-level aspects of bindings
@@
 in proof assistants.  Such relations include typing (e.g.,
-$M\colon\sigma$), evaluation (e.g., natural semantics specifications
+$\Gamma\vdash M\colon\sigma$), evaluation (e.g., natural semantics specifications
 \cite{kahn87stacs,hannan93jfp}), and interaction (e.g., structured
@@
 these examples, we shall illustrate that ELPI is useful not only
-because it has some useful meta-programming features (e.g., a builtin
-treatment of bindings in syntax) but also because it soundly
+because of its meta-programming features,
+% (e.g., a builtin treatment of bindings in syntax)
+but also because it soundly
 implements a higher-order intuitionistic logic.  The latter aspect of
@@
 applications.
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:
+
+% LocalWords:  Sacerdoti Coen Tassi ELPI Teyjus Felty LCF tacticals
+% LocalWords:  intuitionistic

### 2020-10-28 papers/types2020/sec3.tex
Made a small adjustment to Figure 1.  I also wrote a new version of dep-kernel.mod that makes the interp code look more like the proof figure.  I'm not sure this new version is what we should use.  Matteo: can you double check the correctness of this new version, just in case we might use it.
insert example in sec3

@@
+\newpage
 \section{Proof theory and proof certificates}
@@
 In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
-
  One way to proceed:
-
   \begin{itemize}
@@
   \end{itemize}
+
+Stuff in red is from old paper
 \end{metanote}
 
-Stuff in red is from old paper
 \subsection{Horn logic in coq-elpi}
@@
 
-\begin{footnotesize}
 \begin{figure}
-  \newcommand{\XXi}{}
-  \newcommand{\bc}[2]{#1 \Downarrow #2}
+\newcommand{\XXi}{\Pscr}
+\newcommand{\bc}[2]{\Pscr\Downarrow #1 \vdash #2}
 \[
@@
 }
-\qquad
-\infer{\XXi\vdash A}
-      {\bc D A \quad D \in \Pscr}
 \]
@@
 \[
+\infer{\XXi\vdash A}
+      {\bc D A \quad D \in \Pscr}\qquad
   \infer{\bc A A}{}{}\qquad
@@
 \]
+In the judgment $\bc{D}{A}$, the variable $A$ ranges over only atomic
+formulas.
 \lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
-\caption{A proof system for Horn logic and its encoding in coq-elpi .}
+\caption{A proof system for Horn logic and its encoding in coq-elpi.}
 \label{fig:interp}
@@
 \end{figure}
-\end{footnotesize}
 
@@
 
+\paragraph{A prolog tactic}
+
+\begin{metanote}
+  Let's dump it here and see how it looks
+\end{metanote}
+
+\begin{lstlisting}
+Elpi Accumulate lp:{{
+  solve [str ''height'', int N] [goal _Ctx Ev G _Who] _OutGoals :-
+    coq.say "Goal:" {coq.term->string G},
+    check (qheight N) (go G Term),
+    Ev = Term,
+    coq.say "Proof:" {coq.term->string Ev}.
+...
+}}.
+\end{lstlisting}
+The following example show how we can do FPC-driven logic programming in Coq and return
+a Coq term
+\begin{lstlisting}
+Inductive insert (x:nat) : list nat -> list nat -> Prop :=
+| i_n : is_nat x -> insert x [] [x]
+| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
+| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
+Lemma i1:  exists R, insert 2 [0;1] R.
+elpi  dprolog 10.
+Qed.
+Print i1.
+\end{lstlisting}
+\begin{verbatim}
+ex_intro (fun R : list nat => insert 2 [0; 1] R) [2; 2; 2]
+  (i_c 2 0 [1] [2; 2] (le_S 0 1 (le_S 0 0 (le_n 0)))
+	 (i_c 2 1 [] [2] (le_S 1 1 (le_n 1)) (i_n 2)))
+     : exists R : list nat, insert 2 [0; 1] R
+\end{verbatim}
+%
 %%% Local Variables:

### 2020-10-28 papers/types2020/cultures.tex

@@
 presentation of linear logic while most treatments of linear logic in
-type theory usually only capture linearity in a partial fashion.
-\begin{metanote}
-  `` partial fashion.'' Isn't that a bit vague? Some references? -am
-\end{metanote}
+type theory have difficulty treating linear negation and the
+multiplicative disjunction $\lpar$ and its unit $\bot$.
+
+% DM I don't have a reference but I added more details.
+%\begin{metanote}
+%  `` partial fashion.'' Isn't that a bit vague? Some references? -am
+%\end{metanote}
 

### 2020-10-28 papers/types2020/pbt.tex

@@
 
+\subsection{Implementing the tactic}
+The working environment we expect when the tactic is invoked is a proof
+environment, where the goal one is trying to prove is the property that the
+specification should meet. This means that, after \lsti|intro| has been used to
+introduce the relevant hypotheses, we will have a context consisting of
+typing information for a bunch of variables that we should divide in two categories:
+dependent variables, that appear also in the types of other variables, and non-dependent
+variables, that don't appear elsewhere. Obviously, the two are the result of and \lsti|intro| on either
+a dependent or a non-dependent product; in our setting, however, this division acquires an additional reading:
+since specifications are encoded as pure Horn clauses, were no dependency is involved, non-dependent variables
+simply tag the specification clauses whereas dependent variables represent the typing information for the data variables that 
+the speficications use.
+
+In order to generate a counterexample, we will need the user to specify which variables of the environment should
+be used for generating data and which for executing the specification. In addition to this, the user should 
+specify all the certificate information that will guide the data generation phase. The goal of the current environment will
+directly provide the property for which we want to find a counterexample. Say we want to generate a counterexample for
+some specification represented by \lsti|H1| and \lsti|H2| in the environment, and we want to generate the data for
+some variables \lsti|v1| and \lsti|v2| with a height bounded to 10: the final call to the tactic will then 
+look like:
+\begin{lstlisting}
+  elpi pbt 10 (H1 /\ H2) (v1) (v2)
+\end{lstlisting}
+
+In order to generate the query to the \lsti|check| predicate, some pre-processing is needed. In particular, we need to
+turn the data variables into \lP logic variables when they appear inside a specification, and to generate querys to
+\lsti|check| for each of these logic variables in association with the type of the data variable it corresponds to.
+In order to code this preprocessing step, we leverage extensively \lP's higher order programming features. The substitutions
+are handled with the technique of \lsti|copy| clauses.
+
 Let us consider a  relational specification of ordered list, as well as insertion in such lists:

### 2020-10-28 papers/types2020/pbt.tex
commit before Wed's meeting

@@
 
-Let us consider a  relational specification of ordered list, as well as insertion in such lists:
+Let us consider a  relational specification of ordered list:
 \begin{lstlisting}
-Inductive insert (x:nat) : list nat -> list nat -> Prop :=
-| i_n : is_nat x -> insert x [] [x]
-| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
-| i_c : forall y: nat, forall ys rs, x > y -> insert x ys rs -> insert x (y :: ys) (x :: rs).
-
 Inductive ordered : list nat -> Prop :=

### 2020-10-28 papers/types2020/cultures.tex
massaging Matteo's text for dep_pbt in the pbt section

@@
 characterize classes of proof structures is described in the next
-section and used in our two example applications of ELPI with Coq.
+section and used in our two example applications of Elpi with Coq.
 

### 2020-10-28 papers/types2020/intro.tex
massaging Matteo's text for dep_pbt in the pbt section

@@
 
-Recently, C. Sacerdoti Coen and E. Tassi developed the ELPI
+Recently, C. Sacerdoti Coen and E. Tassi developed the Elpi
 implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
-more recently, Tassi has made ELPI available as a plugin to the Coq
+more recently, Tassi has made Elpi available as a plugin to the Coq
 proof assistant \cite{tassi18coqpl}.  This implementation of \lP
 extends earlier implementations in primarily two directions:  First,
-ELPI adds a notion of constraints and constraints handling rules, which makes it more expressive than
+Elpi adds a notion of constraints and constraints handling rules, which makes it more expressive than
 the Teyjus implementation \cite{nadathur99cade} of \lP. Second, the
-plugin version of ELPI comes equipped with an API for accessing the
+plugin version of Elpi comes equipped with an API for accessing the
 Coq environment and a quotation and anti-quotation syntax for mixing
@@
 
-The logic programming interpreter that underlies ELPI provides several
+The logic programming interpreter that underlies Elpi provides several
 convenient features for the kind of meta-programming tasks that can
@@
 bindings, including capture-avoiding substitution,
-unification, and recursive programming.  ELPI spares the
+unification, and recursive programming.  Elpi spares the
 programmer from dealing with low-level aspects of bindings
@@
 
-Since relations (not functions) play a central role in \lP, ELPI is
+Since relations (not functions) play a central role in \lP, Elpi is
 capable to providing direct support for the many relations that arise
@@
 specifications provided by \lP \cite{felty93jar}.  Some recent
-implementations built using ELPI
+implementations built using Elpi
 \cite{coen19mscs,tassi18coqpl,tassi19itp} support the usefulness of
@@
 
-In this paper, we present two applications of ELPI within Coq. With
-these examples, we shall illustrate that ELPI is useful not only
+In this paper, we present two applications of Elpi within Coq. With
+these examples, we shall illustrate that Elpi is useful not only
 because of its meta-programming features,
@@
 implements a higher-order intuitionistic logic.  The latter aspect of
-ELPI makes it possible to encode the proofs and proof theory of
+Elpi makes it possible to encode the proofs and proof theory of
 specific subsets of the Coq logic.
@@
 striking differences in notions of computing and reasoning that arise
-on each side of the ELPI-Coq API.   We will also present a quick
+on each side of the Elpi-Coq API.   We will also present a quick
 summary of the key proof theory concepts that are used by our example
@@
 
-% LocalWords:  Sacerdoti Coen Tassi ELPI Teyjus Felty LCF tacticals
+% LocalWords:  Sacerdoti Coen Tassi Elpi Teyjus Felty LCF tacticals
 % LocalWords:  intuitionistic

### 2020-10-28 papers/types2020/pbt.tex
massaging Matteo's text for dep_pbt in the pbt section

--- 2020-10-26 papers/types2020/pbt.tex
@@
 
-Let us consider a  relational specification of ordered list, as well as insertion in such lists:
+\subsection{Implementing the tactic}
+We will invoke the tactic in a
+proof environment  where the goal is % one is trying to prove is the
+property that the specification should meet. This means that, after
+\lsti|intro| has been used to introduce the relevant hypotheses, we
+will have a context % consisting of typing information for a bunch of variables
+that we  divide in two categories: dependent
+variables, which appear also in the types of other variables, and
+non-dependent variables tat do not % Obviously, the
+% two are the result of and \lsti|intro| on either a dependent or a
+% non-dependent product
+; in our setting this division acquires an additional reading beyond
+what the property looks like: since specifications are encoded as pure
+Horn clauses where no dependency is involved non-dependent variables
+simply tag the specification clauses, whereas dependent variables
+represent the typing information for the data variables that the
+speficications use.
+
+Whithin the tactic the user
+specifies which variables of the environment should be used for
+generating data and which for executing the specification. In addition
+to this, the user should specify all the certificate information that
+will guide the data generation phase. The goal of the current
+environment will directly provide the property for which we want to
+find a counterexample. % Say we want to generate a counterexample for
+% some specification represented by \lsti|H1| and \lsti|H2| in the
+% environment, and we want to generate the data for some variables
+% \lsti|v1| and \lsti|v2| with a height bounded to 10: the final call to
+% the tactic will then look like:
+% \begin{lstlisting}
+%   elpi pbt 10 (H1 /\ H2) (v1) (v2)
+% \end{lstlisting}
+
+To exemplify, let us add to the previous specification a definition of ordered list:
 \begin{lstlisting}
-Inductive insert (x:nat) : list nat -> list nat -> Prop :=
-| i_n : is_nat x -> insert x [] [x]
-| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
-| i_c : forall y: nat, forall ys rs, x > y -> insert x ys rs -> insert x (y :: ys) (x :: rs).
-
 Inductive ordered : list nat -> Prop :=
@@
 %
-A property we may wish to check before embarking in a formal proof is whether insertion preserves ordered-ness. Since we did slip in an error, our tactic reports a counter-example, e.g.,
-\verb|(* Proof Term: [0, [0; 1; 0]] *)|:
+A property we may wish to check before embarking in a formal proof is whether insertion preserves ordered-ness:
 \begin{lstlisting}
@@
 In this query the tactic tests the hypotheses \lsti{Ho} and \lsti{Hi}
-against data \lsti{x,xs} generated exhaustively up to proofs bounded by
-at most $5$ unfoldings of the library \lsti{Inductive} definition of
-\lsti{nat} and \lsti{list}.
+against data \lsti{x,xs} generated exhaustively up to proofs bounded
+by at most $5$ unfoldings of the library \lsti{Inductive} definition
+of \lsti{nat} and \lsti{list}. Since we did slip in an error (where?), our
+tactic reports a counter-example, namely
+\verb|(* Proof Term: [0, [0; 1; 0]] *)|
 
-Note that the same query in \textsf{QuickChick} would have required much more setup: if we wished to proceed relationally as above we would have had to provide a proof of decidability of the relevant notions.
-Were we to use functions, then we would have to implement a generator and shrinker for ordered lists.
+In order to generate the query to the \lsti|check| predicate, some
+pre-processing is needed. In particular, we need to turn the data
+variables into \lP logic variables when they appear inside a
+specification, and to generate querys to \lsti|check| for each of
+these logic variables in association with the type of the data
+variable it corresponds to.  In order to code this preprocessing step,
+we leverage extensively \lP's higher order programming features. The
+substitutions are handled with the technique of \lsti|copy| clauses.
+
+Note that the same query in \textsf{QuickChick} would have required
+much more setup: if we wished to proceed relationally as above we
+would have had to provide a proof of decidability of the relevant
+notions. Were we to use functions, then we would have to implement a
+generator and shrinker for ordered lists.
 %%% Local Variables:

### 2020-10-29 papers/types2020/app.tex
some work on sec 3 and 4

@@
 Stuff we may not have space for in the paper
-\begin{itemize}
-\item The LJF sequent rules
-\item the code for
-  \lsti{dep_pbt}
 
-\item \dots
-\end{itemize}
+In Fig.~\ref{fig:interp} we give a proof system for Horn logic and its encoding in coq-elpi
+\todo{explain the interface, quotations etc}
+
+\begin{figure}
+\newcommand{\XXi}{\Pscr}
+\newcommand{\bc}[2]{\Pscr\Downarrow #1 \vdash #2}
+\[
+\infer{\XXi\vdash t = t}
+{%\eqExpert{\XXi}
+}
+\qquad
+\infer{\XXi\vdash \true}
+{%\trueExpert{\XXi}
+}
+\qquad
+\infer{\XXi\vdash G_1\vee G_2}
+{\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
+}
+\qquad
+\infer{\XXi\vdash \exists x. G}
+{\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
+}
+\qquad 
+\infer{\XXi\vdash G_1\wedge G_2}
+{\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
+}
+\]
+\vskip -18pt
+\[
+\infer{\XXi\vdash A}
+      {\bc D A \quad D \in \Pscr}\qquad
+  \infer{\bc A A}{}{}\qquad
+  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
+  \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
+\]
+In the judgment $\bc{D}{A}$, the variable $A$ ranges over only atomic
+formulas.
+\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
+\caption{A proof system for Horn logic and its encoding in coq-elpi.}
+\label{fig:interp}
+%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
+%\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
+%\caption{A dependent kernel for Coq terms.}
+%\label{fig:kernel}
+\end{figure}
+
 %%% Local Variables:

### 2020-10-29 papers/types2020/pbt.tex
some work on sec 3 and 4

@@
 
-\begin{metanote}
-  I'm just saying the bare minimum -am
-\end{metanote}
+% \begin{metanote}
+%   I'm just saying the bare minimum -am
+% \end{metanote}
 
@@
 
-\begin{metanote}
-  TODO next
-  \begin{itemize}
-  \item The elpi dep-pbt tactic
-  \item  the simple ordered example [DONE]
-  \item a bit of details about the \texttt{aexp} example
-  \end{itemize}
-\end{metanote}
+% \begin{metanote}
+%   TODO next
+%   \begin{itemize}
+%   \item The elpi dep-pbt tactic
+%   \item  the simple ordered example [DONE]
+%   \item a bit of details about the \texttt{aexp} example
+%   \end{itemize}
+% \end{metanote}
 
@@
 % \end{lstlisting}
+Formally, for a property
+\[
+\forall x_1 \colon \tau_1\dots x_n \colon \tau_m, A_1\land\dots\land A_m \supset B
+\]
+and the specification of a FPC, the call will be:
+\begin{lstlisting}
+elpi dep_pbt <fpc> $(A_1\land\dots\land A_m) \ (x_1) \dots (x_n)$.
+\end{lstlisting}
 
-To exemplify, let us add to the previous specification a definition of ordered list:
+To exemplify, let us add to the previous specification of list
+insertion a definition of ordered list:
 \begin{lstlisting}
@@
 variables into \lP logic variables when they appear inside a
-specification, and to generate querys to \lsti|check| for each of
+specification, and to generate queries to \lsti|check| for each of
 these logic variables in association with the type of the data
-variable it corresponds to.  In order to code this preprocessing step,
+variable it corresponds to.  In order to realize this preprocessing step,
 we leverage extensively \lP's higher order programming features. The
@@
 Note that the same query in \textsf{QuickChick} would have required
-much more setup: if we wished to proceed relationally as above we
+much more setup: if we wished to proceed relationally as above, we
 would have had to provide a proof of decidability of the relevant
@@
 generator and shrinker for ordered lists.
+
+\todo{the \texttt{aexp} example}
+
 %%% Local Variables:

### 2020-10-29 papers/types2020/sec3.tex
some work on sec 3 and 4

@@
  One way to proceed:
-  \begin{itemize}
-  \item One figure fig \ref{fig:interp} for Horn logic over Coq terms (no certificates) and its encoding with quotations 
-  \item Then the logic behind the dependent kernel with certificates and proof terms
-  \end{itemize}
+  % \begin{itemize}
+  % \item One figure fig \ref{fig:interp} for Horn logic over Coq terms (no certificates) and its encoding with quotations 
+  % \item Then the logic behind the dependent kernel with certificates and proof terms
+  % \end{itemize}
 
@@
 
-\subsection{Horn logic in coq-elpi}
+%\subsection{Horn logic in coq-elpi}
+
+\subsection{Focused proof systems with certificates}
+\label{ssec:focused}
+\todo{Adapt the discussion}
+\todo{grammar for formulae and proof terms\cite{Herbelin94,LengrandDM06}?}
 
@@
 \emph{backchaining} phase~\cite{miller91apal}.
-
-In Fig.~\ref{fig:interp} we give a proof system for Horn logic and its encoding in coq-elpi
-\todo{explain the interface, quotations etc}
-
-\begin{figure}
-\newcommand{\XXi}{\Pscr}
-\newcommand{\bc}[2]{\Pscr\Downarrow #1 \vdash #2}
-\[
-\infer{\XXi\vdash t = t}
-{%\eqExpert{\XXi}
-}
-\qquad
-\infer{\XXi\vdash \true}
-{%\trueExpert{\XXi}
-}
-\qquad
-\infer{\XXi\vdash G_1\vee G_2}
-{\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
-}
-\qquad
-\infer{\XXi\vdash \exists x. G}
-{\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
-}
-\qquad 
-\infer{\XXi\vdash G_1\wedge G_2}
-{\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
-}
-\]
-\vskip -18pt
-\[
-\infer{\XXi\vdash A}
-      {\bc D A \quad D \in \Pscr}\qquad
-  \infer{\bc A A}{}{}\qquad
-  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
-  \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
-\]
-In the judgment $\bc{D}{A}$, the variable $A$ ranges over only atomic
-formulas.
-\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
-\caption{A proof system for Horn logic and its encoding in coq-elpi.}
-\label{fig:interp}
-%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
-%\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
-%\caption{A dependent kernel for Coq terms.}
-%\label{fig:kernel}
-\end{figure}
-
-\subsection{Focused proof systems with certificates}
-\label{ssec:focused}
-\todo{Adapt the discussion}
-\todo{Talk about the LJT thing: grammar for formulae and proof terms\cite{Herbelin94,LengrandDM06}?}
 
@@
 
-\paragraph{Certificates}
+\subsection{Certificates}
 
 \todo{Shorten it}
-\todo{Show the dprolog tactic? have an example? could  be
-\lsti{Goal  exists R, insert 2 [0;1] R.} and we would move the code for insert here}
+% \todo{Show the dprolog tactic? have an example? could  be
+% \lsti{Goal  exists R, insert 2 [0;1] R.} and we would move the code for insert here}
 {\color{red}
@@
 
-\paragraph{A prolog tactic}
+\subsection{A prolog tactic}
 
@@
 Lemma i1:  exists R, insert 2 [0;1] R.
+eexists.
 elpi  dprolog 10.

### 2020-10-29 papers/types2020/elab.tex
Some modifications to section 5 and the bib file.  I'm not thinking about adding some small bits of code to this section.

@@
+\newpage
 \section{Elaboration of external proof certificates for the Coq
@@
 
-Follow the text in \cite{blanco20coq}.
+% Follow the text in \cite{blanco20coq}.
+
+The trusted base of Coq is its kernel, which is a type-checking
+program that certifies that a dependently typed $\lambda$-term has a
+given type.  If type checking succeeds, the formula corresponding to
+that type is, in fact, accepted by Coq users as a theorem of
+intuitionistic logic.  The rest of the Coq system, especially its
+tactic language, is designed to help a human user build
+proofs-as-$\lambda$-terms that can be checked by the kernel.
+
+There are many theorem provers for intuitionistic logic for which a
+successful proof is not the kind of detailed $\lambda$-term required
+by the Coq kernel.  Most prover provide, instead, a trace of some key
+aspects of a proof: some details are often not included in such a
+trace.  For example:
+\begin{enumerate}
+
+\item Substitution instances of quantifiers might not be recorded in a
+  proof since such instances can, in principle, be reconstructed using
+  unification.
+
+\item Detailed typing information might not need to be stored within a
+  proof since types can often be reconstructed during proof checking.
+
+\item Some simplifications steps might be applied within a proof
+  without recording which rewrites were used.  A simple 
+  non-deterministic proof-search engine might be expected to
+  reconstruct an equivalent simplification.
+\end{enumerate}
+
+As we have describe above, the Foundational Proof Certificate (FPC)
+framework has been designed for formally defining a rich collection of
+proof certificates that do not necessarily contain full details.
+Proof checkers for FPC certificates can be written using simple logic
+programs: the standard implementation mechanisms of unification and
+backtracking search can then be used to reconstruct missing proof
+details \cite{miller17fac}.  Such proof checking kernels can also be
+used to elaborate such certificates into fully detailed proofs
+\cite{blanco17phd,blanco17cade}.
+
+We describe here briefly how to use the technology behind FPCs and
+logic programming in order to provide a flexible approach to
+connecting external provers of intuitionistic logic to Coq.
+%
+Following the general outline that has been described in
+\cite{blanco17cade}, we assume that the following step are taken.
+%
+\begin{enumerate}
+\item Have an external prover output some form of proof evidence
+  (proof certificate) for formulas it claims are theorems.
+
+\item Develop a formal definition of the semantics behind such proof
+  certificates using the FPC framework.
+
+\item Check proofs by executing the logic programming checker that is
+  parametrized by the particular FPC definition.
+\end{enumerate}
+%
+It is also possible to pair two different proof certificates,
+following two different FPC definitions and to do the proof checking
+in parallel.  The logic programming setting allows such parallel
+checking to also be checking-and-synthesizing.  That is, during the
+checking of one certificate, it is possible to synthesize, for
+example, a fully detailed term that is appropriate for handing to the
+Coq kernel.
+
+If one is interested only in building Coq kernel proof structures, we
+can bypass the use of an explicit pairing operation and build the
+synthesis of such proof structures directly into the FPC proof
+checker.  If one is interested in checking only one kind of external
+proof structure, then the FPC for that structure could also be built
+into the checker (via, say, partial evaluation of logic programs
+\cite{lloyd91jlp}).
+
+To illustrate the possibilities ***
+
+** We limit our attention here to first-order intuitionistic logic:
+that is, we do not consider here inductive definitions.
+
+** Note that with intuitionistic logic, there is much less ambiguity
+of polarities possible.  In particular, only the conjunction and the
+atomic formulas can have ambiguous polarizations.  The two
+conjunctions are familiar to those working with typed
+$\lambda$-calculus:
+
+the negative conjunction (corresponding to $\with$ in linear logic)
+corresponds to surjective pairing.
+
+the positive conjunction (corresponding to $\with$ in linear logic)
+corresponds to ***.
+
+The choice between negated atoms and positive atoms is similar to the
+choice in representing $\beta$-normal terms using either head-normal
+form or administrative normal form.
+
+We built the fpccoq system to demonstrate the feasibility of using
+logic programming to check proof certificates produced by external
+provers and to elaborate such certificates into proofs with sufficient
+details that they can be given directly to Coq's kernel to
+check.  Note that one does not need to trust the external prover nor
+the fpccoq tool-chain: one only needs to trust the Coq kernel.
+
+De Bruijn
+decide depth
+simply typed lambda terms
+
+\newpage
 

### 2020-10-29 papers/types2020/app.tex
added some related and future work, suggestions for title, revisions to sec 4

--- 2020-10-26 papers/types2020/app.tex
@@
 
-Stuff we may not have space for in the paper
-\begin{itemize}
-\item The LJF sequent rules
-\item the code for
-  \lsti{dep_pbt}
+In this appendix we list some definitions and pieces of code that may be helpful
 
-\item \dots
-\end{itemize}
+\subsection{Interp}
+
+In Fig.~\ref{fig:interp} we the encoding of the vanilla meta-interpreter used in
+the testing phase of the \lsti{dep_pbt} tactic.
+Differently from Fig.~\ref{fig:augmented} we appeal, via quotations, to Coq's defined connectives --- note the use of a green cut to avoid overlapping with the \lsti{Inductive} atomic case and the absence of certificates.
+
+\begin{figure}
+% \newcommand{\XXi}{\Pscr}
+% \newcommand{\bc}[2]{\Pscr\Downarrow #1 \vdash #2}
+% \[
+% \infer{\XXi\vdash t = t}
+% {%\eqExpert{\XXi}
+% }
+% \qquad
+% \infer{\XXi\vdash \true}
+% {%\trueExpert{\XXi}
+% }
+% \qquad
+% \infer{\XXi\vdash G_1\vee G_2}
+% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
+% }
+% \qquad
+% \infer{\XXi\vdash \exists x. G}
+% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
+% }
+% \qquad 
+% \infer{\XXi\vdash G_1\wedge G_2}
+% {\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
+% }
+% \]
+% \vskip -18pt
+% \[
+% \infer{\XXi\vdash A}
+%       {\bc D A \quad D \in \Pscr}\qquad
+%   \infer{\bc A A}{}{}\qquad
+%   \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
+%   \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
+% \]
+% In the judgment $\bc{D}{A}$, the variable $A$ ranges over only atomic
+% formulas.
+ \lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
+\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
+\caption{Encoding in coq-elpi of the vanilla meta-interpreter}
+\label{fig:interp}
+%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
+%\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
+%\caption{A dependent kernel for Coq terms.}
+%\label{fig:kernel}
+\end{figure}
+
 %%% Local Variables:

### 2020-10-29 papers/types2020/conc.tex
added some related and future work, suggestions for title, revisions to sec 4

@@
 \section{Conclusion and future work}
+
+Link to the repo
+\subsection{Related work}
+\label{ssec:rel}
+
+\begin{itemize}
+\item Brief discussion/comparison with \texttt{auto}
+\item Mention other papers by Enrico as applications of coq-elpi
+\item Say we \textbf{do} not talk about Coq metaprogramming (Ltac, Mtac), do say that doing our to application in say MetaCoq would be a disaster (reimplement unfication, backtracking, binders)
+\end{itemize}
+
+\subsection{Future work}
+\label{ssec:fut}
 
@@
 This  is similar to the use of needed narrowing in LazySmallCheck to generate partial trace or the idea of using CLP for data generation (see Dubois' FocalTest or the stuff by Fabio Fioravanti).
+
+Extending our tactic to handle computations (\texttt{simp})
+
+Extending the LJF kernel to handle defined atoms
 %%% Local Variables:

### 2020-10-29 papers/types2020/pbt.tex
added some related and future work, suggestions for title, revisions to sec 4

@@
 
-In a previous paper~\cite{blanco19ppdp}, we presented a
+We have presented n a previous paper~\cite{blanco19ppdp} a
 proof-theoretical reconstruction of property-based testing (PBT) of
 relational specifications, adopting techniques from foundational proof
-certificates to account for several features of these testing
+certificates to account for several features of this testing
 paradigm: from various \emph{generation} strategies, to
@@
 Given the connection that \textsf{coq-elpi} offers between logic
-programming and the internals of Coq, it is natural to extend the FPC-driven logic programming interpreter of the previous section to perform PBT of \lstinline{Inductive} types.
+programming and the internals of Coq, it is natural to extend the
+FPC-driven logic programming interpreter of the previous section to
+perform PBT over \lstinline{Inductive} types.
 
-While nothing prevents us from porting all the features we discussed in~\cite{blanco19ppdp}, for the sake of this paper we will implement only FPC corresponding to different flavors \emph{exhaustive} generation and their combination, as adopted, e.g., in SmallCheck~\cite{smallcheck} and  $\alpha$Check~\cite{cheney_momigliano_2017}.
+While nothing prevents us from porting all the features we discussed
+in~\cite{blanco19ppdp}, for the sake of this paper we will implement
+only FPC corresponding to different flavors \emph{exhaustive}
+generation and their combination, as adopted, e.g., in
+SmallCheck~\cite{smallcheck} and
+$\alpha$Check~\cite{cheney_momigliano_2017}.
 
-Of course, \textsf{QuickChick}~\cite{QChick}
+Of course, Coq already features \textsf{QuickChick}~\cite{QChick}
 (\url{https://softwarefoundations.cis.upenn.edu/qc-current}) is a
-sophisticated and far-reaching PBT tool for Coq, which is based on a
-different perspective: being a clone of Haskell's QuickCheck, it
-emphasizes testing executable (read \emph{decidable)} specifications
-with random generators.
-While current research~\cite{LampropoulosPP18} aims to increase automation, it is fair to say testing, in particular of  relational specifications, is still very labor intensive.
-We do not intend to compete with  \textsf{QuickChick}, but as we shall see, as far as \lsti{Inductive} definitions that corresponds to pure Horn programs, we can test them immediately without any decidability proof or configuration of generators and shrinkers.
+sophisticated and well-supported PBT tool, based on a different
+perspective: being a clone of Haskell's QuickCheck, it emphasizes
+testing executable (read \emph{decidable)} specifications with random
+generators. While current research~\cite{LampropoulosPP18} aims to
+increase automation, it is fair to say that testing, in particular of
+relational specifications, is still very labor intensive. We do not
+intend to compete with \textsf{QuickChick}, but we shall see that we
+can test them immediately \lsti{Inductive} definitions that
+corresponds to pure Horn programs, without any decidability proof nor
+commitment to the random generation strategy, and related
+configuration of generators and shrinkers.
 
@@
 
-\subsection{Implementing the tactic}
+\subsection{An Elpi tactic for PBT}
 We will invoke the tactic in a
 proof environment  where the goal is % one is trying to prove is the
-property that the specification should meet. This means that, after
+the property that the system-under-test should meet. This means that, after
 \lsti|intro| has been used to introduce the relevant hypotheses, we
-will have a context % consisting of typing information for a bunch of variables
-that we  divide in two categories: dependent
-variables, which appear also in the types of other variables, and
-non-dependent variables tat do not % Obviously, the
+ have a proof context % consisting of typing information for a bunch of variables
+where eigenvariables  play different roles:  dependent
+variables, that is the ones appearing also in the types of other variables, are subject to generation, while
+non-dependent ones tags assumptions that are only to tested. This is reminiscent of bi.directional type-checking. % Obviously, the
 % two are the result of and \lsti|intro| on either a dependent or a
 % non-dependent product
-; in our setting this division acquires an additional reading beyond
-what the property looks like: since specifications are encoded as pure
-Horn clauses where no dependency is involved non-dependent variables
-simply tag the specification clauses, whereas dependent variables
-represent the typing information for the data variables that the
-speficications use.
+% ; in our setting this division acquires an additional reading beyond
+% what the property looks like: since specifications are encoded as pure
+% Horn clauses where no dependency is involved, non-dependent variables
+% simply tag the specification clauses, whereas dependent variables
+% represent the typing information for the data variables that the
+% specifications use.
 
-Whithin the tactic the user
+Within the tactic the user therefore
 specifies which variables of the environment should be used for
@@
 % \end{lstlisting}
-Formally, for a property
+Concretely, for a property
 \[
@@
 \]
-and the specification of a FPC, the call will be:
+and the specification of a FPC, the call to the tactic will be:
 \begin{lstlisting}
@@
 \end{lstlisting}
+The tactic will call \lsti{check} with the given FPC on the dependent variables and will simply call a vanilla meta-interpreter (see Fig.~\ref{fig:interp} in the appendix) to test
+\begin{lstlisting}
+interp $(A_1\land\dots\land A_m$), not (interp $B$)}
+\end{lstlisting}
+where \lsti{not} in \lP's negation-as-failure operator.
 
@@
 %
-A property we may wish to check before embarking in a formal proof is whether insertion preserves ordered-ness:
+A property we may wish to check before embarking in a formal proof is
+whether insertion preserves ordered-ness:
 \begin{lstlisting}
-Conjecture ins_ord:  forall x xs rs, ordered xs -> insert x xs rs -> ordered rs.
+Conjecture ins_ord:  forall (x : nat) xs rs, ordered xs -> insert x xs rs -> ordered rs.
 intros x xs rs Ho Hi.
@@
 these logic variables in association with the type of the data
-variable it corresponds to.  In order to realize this preprocessing step,
+variable it corresponds to.  In order to realize this pre-processing step,
 we leverage extensively \lP's higher order programming features. The

### 2020-10-29 papers/types2020/sec3.tex
added some related and future work, suggestions for title, revisions to sec 4

@@
-\newpage
+%\newpage
 \section{Proof theory and proof certificates}
@@
 \label{fig:augmented}
+\end{figure}
+\begin{figure}
 %\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
@@
 Various  additional examples and experiments using the pairing of FPCs
-can be found in~\cite{Pair}. Using similar techniques, it is possible to define FPCs that target
+can be found in~\cite{blanco17cade}. Using similar techniques, it is possible to define FPCs that target
 specific types for special treatment: for example, when generating
@@
 \begin{verbatim}
-ex_intro (fun R : list nat => insert 2 [0; 1] R) [2; 2; 2]
-  (i_c 2 0 [1] [2; 2] (le_S 0 1 (le_S 0 0 (le_n 0)))
-	 (i_c 2 1 [] [2] (le_S 1 1 (le_n 1)) (i_n 2)))
+ex_intro (fun R : list nat => insert 2 [0; 1] R) [0; 1; 2]
+  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2)))
 	 : exists R : list nat, insert 2 [0; 1] R

### 2020-10-30 papers/types2020/elab.tex
Minor rewording

@@
 %
-It is also possible to pair two different proof certificates,
-following two different FPC definitions and to do the proof checking
-in parallel.  The logic programming setting allows such parallel
-checking to also be checking-and-synthesizing.  That is, during the
+It is also possible to pair two different proof certificates, defined
+by two different FPC definitions, and do the proof checking in
+parallel.  The logic programming setting allows such parallel checking
+to also serve as checking-and-synthesizing.  That is, during the
 checking of one certificate, it is possible to synthesize, for

### 2020-10-30 papers/types2020/main.tex
"ordered" example in pbt.tex
adding file for conclusion
some work on fig 1
after main merge
added some related and future work, suggestions for title, revisions to sec 4
title changed and temporary abstract added

@@
 
-\title{Interfacing logic programming and Coq} % Other suggestions?
-\titlerunning{Interfacing logic programming and Coq}
+\title{Two applications of logic programming to Coq} % Other suggestions?
+% bit too general -- elpi-coq is such an interface -a,
+% on the usefulness of logic programming in Coq
+% two applications of logic programming to Coq
+\titlerunning{Two applications of logic programming to Coq}
 
@@
 \begin{abstract}
-We do stuff
+The \emph{foundational proof certificate} (FPC) framework is capable
+of checking and elaborating a wide range of proof evidence and can easily
+be implemented using the logic programming language \lP.
+%
+% The checking of a proof certificate in this higher-order logic
+% programming language can yield a fully detailed, dependently
+% typed $\lambda$-terms that can be checked by the Coq kernel.
+%
+We use the recently released Coq-Elpi plugin for Coq in which the Elpi
+implementation of \lP can send and retrive information to and from the Coq kernel.
+%
+We provide two examples: first we implement a FPC-driven sequent calculus for a fragment of the Calculus of Inductive Constructions and we package it into a tactic to perform property-based testing of Inductive types corresponding to pure (higher-order) Horn programs.
+Secondly, we implement a proof checker for first-order LJF to demonstrate how  proof certificates
+can be supplied by external (to Coq) provers and elaborated into the
+fully detailed proof terms that can be checked by the Coq kernel.
 \end{abstract}
@@
 
-\section{Conclusion}
-
+\input conc
 \bibliography{l}
+\appendix
+\input app
 \end{document}
@@
 %%% End:
+
+% LocalWords:  pbt elab conc Elpi LJF

### 2020-10-30 papers/types2020/app.tex
progress on 4

@@
 
+\subsection{Semantics of the typed arithmetic language}
+\label{ssec:types}
+
+\begin{lstlisting}
+Inductive has_type : tm -> typ -> Prop :=
+   | T_Tru : has_type ttrue TBool
+   | T_Fls : has_type tfalse TBool
+    | T_Test : forall t1 t2 t3 T,
+        has_type t1 TBool -> has_type t2 T -> has_type t3 T ->  has_type (tif  t1 t2 t3) T 
+   | T_Zro : has_type tzero TNat
+   | T_Scc : forall t1,
+         has_type t1 TNat -> has_type (tsucc t1) TNat
+   | T_Prd : forall t1,
+        has_type t1 TNat -> has_type (tpred t1 ) TNat
+   | T_Iszro : forall t1,
+        has_type t1 TNat ->  has_type (tiszero t1) TBool.
+Inductive nvalue : tm -> Prop :=
+| nv_zero : nvalue tzero   | nv_succ : forall t, nvalue t -> nvalue (tsucc t).
+Inductive bvalue : tm -> Prop :=
+| bv_t : bvalue ttrue   | bv_f : bvalue tfalse. 
+ Reserved Notation "t1 '===>' t2" (at level 40).
+ Inductive step : tm -> tm -> Prop :=
+   | ST_IfTrue : forall t1 t2,
+       (tif ttrue t1 t2) ===> t1
+   | ST_IfFalse : forall t1 t2,
+       (tif tfalse t1 t2) ===> t2
+   | ST_If : forall t1 t1' t2 t3,
+       t1 ===> t1' -> (tif t1 t2 t3) ===> (tif t1' t2 t3)
+   | ST_Succ : forall t1 t1',
+       t1 ===> t1' -> (tsucc t1) ===> (tsucc t1')
+   | ST_PredZero : (tpred tzero) ===> tzero
+   | ST_PredSucc : forall t1,
+       nvalue t1 -> (tpred (tsucc t1)) ===> t1
+   | ST_Pred : forall t1 t1',
+       t1 ===> t1' -> (tpred t1) ===> (tpred t1')
+   | ST_IszeroZero : (tiszero tzero) ===> ttrue
+   | ST_IszeroSucc : forall t1,
+        nvalue t1 ->  (tiszero (tsucc t1)) ===> tfalse
+   | ST_Iszero : forall t1 t1',
+       t1 ===> t1' -> (tiszero t1) ===> (tiszero t1')
+ where "t1 '===>' t2" := (step t1 t2).
+Inductive notstuck (e : tm) (Step : tm -> tm -> Prop) : Prop :=
+  | pn : nvalue e  -> notstuck e Step
+  | pb : bvalue e -> notstuck e Step
+  | ps e' : Step e e' -> notstuck e Step.
+\end{lstlisting}
 %%% Local Variables:
@@
 
-% LocalWords:  LJF
+% LocalWords:  LJF Interp dep pbt coq elpi

### 2020-10-30 papers/types2020/conc.tex
progress on 4

@@
 \item Brief discussion/comparison with \texttt{auto}
-\item Mention other papers by Enrico as applications of coq-elpi
-\item Say we \textbf{do} not talk about Coq metaprogramming (Ltac, Mtac), do say that doing our to application in say MetaCoq would be a disaster (reimplement unfication, backtracking, binders)
+\item Mention other papers by Enrico as applications of coq-elpi~\cite{coen19mscs,tassi19itp}\dots
+\item Say we \textbf{do} not talk about Coq metaprogramming (Ltac, Mtac), do say that doing our to application in MetaCoq~\cite{sozeau2020metacoq} would be a disaster (reimplement unfication, backtracking, binders)
 \end{itemize}
@@
 A mention to the use of \emph{delays} in \texttt{check}: instead of generating only ground terms, we could generate say lists with at least two elements as \texttt{[1;2|Tail]} or simply not commit to a base type.
-This  is similar to the use of needed narrowing in LazySmallCheck to generate partial trace or the idea of using CLP for data generation (see Dubois' FocalTest or the stuff by Fabio Fioravanti).
+This  is similar to the use of needed narrowing in LazySmallCheck to generate partial trace or the idea of using CLP for data generation (see Dubois' FocalTest or~\cite{FioravantiPS15}).
 

### 2020-10-30 papers/types2020/main.tex
progress on 4

@@
 be implemented using the logic programming language \lP.
-%
-% The checking of a proof certificate in this higher-order logic
-% programming language can yield a fully detailed, dependently
-% typed $\lambda$-terms that can be checked by the Coq kernel.
-%
 We use the recently released Coq-Elpi plugin for Coq in which the Elpi
-implementation of \lP can send and retrive information to and from the Coq kernel.
-%
+implementation of \lP can send and retrieve information to and from the Coq kernel.
 We provide two examples: first we implement a FPC-driven sequent calculus for a fragment of the Calculus of Inductive Constructions and we package it into a tactic to perform property-based testing of Inductive types corresponding to pure (higher-order) Horn programs.
@@
 \bibliography{l}
+\newpage
 \appendix

### 2020-10-30 papers/types2020/pbt.tex
progress on 4

@@
 
-\todo{the \texttt{aexp} example}
+%\todo{the \texttt{aexp} example}
 
+For a more significant case study, let us turn to the semantic of programming languages, where PBT has been used extensively and successfully~\cite{Klein12}.
+Here we will use a far simple example, the a typed arithmetic language featuring numerals with predecessor and Booleans with if-then-else and test-for-zero, which comes from (\url{https://softwarefoundations.cis.upenn.edu/plf-current/Types.html}):
+\begin{lstlisting}
+Inductive tm : Type :=
+ | ttrue : tm | tfalse : tm | tif : tm -> tm -> tm -> tm | tzero : tm | tsucc : tm -> tm
+ | tpred : tm -> tm | tiszero : tm -> tm.
+ Inductive typ : Type := |TBool : typ |TNat : typ.
+\end{lstlisting}
+The static and small step dynamic semantics are reported in appendix~\ref{ssec:types}.
+
+While it is obvious that \emph{subject expansion} fails for this calculus, it is gratifying to have it confirmed, with \lsti{e = tif ttrue tzero ttrue}:
+\begin{lstlisting}
+Conjecture subexp:  forall e e' t, step e e' -> has_type e' t -> has_type e t. 
+ intros e e' t HS HT.
+ elpi dep_pbt height 2 (HS /\ HT) (e).
+\end{lstlisting}
+
+Another way to asses the fault detection capability of our PBT setup is via \emph{mutation analysis}, whereby localized bugs are voluntarily inserted, with the view that they should caught (``killed'') by the testing suite.
+Following on  an exercise if the aforementioned chapter of SF, we modify the typing relation by adding the following additional clause:
+
+\begin{lstlisting}
+Module M1.
+Inductive has_type : tm -> typ -> Prop :=
+$\dots$
+  | T_SuccBool : forall t, has_type t TBool -> has_type (tsucc t) TBool.
+end M1.
+\end{lstlisting}
+Some of the required properties for a typing system now fails, not
+only type uniqueness, but also progress with counterexample \lsti{e = tsucc ttrue}:
+\begin{lstlisting}
+Definition progress (e :tm) (Has_type : tm -> typ -> Prop) (Step : tm -> tm -> Prop):= 
+     forall t, Has_type e t -> notstuck e Step.
+Conjecture progress_m1:  forall e, progress e M1.has_type step.
+ unfold progress.
+ intros e t Ht.    
+ elpi dep_pbt 2 height (Ht) (e) .  
+\end{lstlisting}
+Whereas the \emph{Types} examples is admittedly quite simple-minded,
+it has been used as a benchmark for evaluating \textsf{QuickChick}'s automation of generators under invariants~\cite{LampropoulosPP18}, and to be amenable to the tool, the specification had to be massaged non-insignificantly.
 %%% Local Variables:
@@
 
-% LocalWords:  Bedwyr bipole elpi QuickCheck OCaml TODO
+% LocalWords:  Bedwyr bipole elpi QuickCheck OCaml TODO coq xs nat
+% LocalWords:  tif ttrue tzero tsucc

### 2020-10-30 papers/types2020/sec3.tex
progress on 4

@@
 Explain the choice of not having certificates represent proof terms.
-Explain that the unfold clause handles defined atoms and defined connectives (conjunction, disjunction, existential and also equality: this may be less efficient and proof-theoretically induce pointless phase changes, but it is arguably very elegant e simplifies proof terms that are inherited by Coq }
+Explain that the unfold clause handles defined atoms and defined connectives (conjunction, disjunction, existential and also equality): this may be less efficient and proof-theoretically induce pointless phase changes, but it is arguably very elegant e simplifies proof terms that arand inherited by Coq }
 
-Figure~\ref{fig:kernel} contains the elpi implementation of the
+Figure~\ref{fig:kernel} contains the Elpi implementation of the
 inference rules in Figure~\ref{fig:augmented}: here the infix
@@
 Print i1.
+ex_intro (fun R : list nat => insert 2 [0; 1] R) [0; 1; 2]
+  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2)) : exists R : list nat, insert 2 [0; 1] R
 \end{lstlisting}
-\begin{verbatim}
-ex_intro (fun R : list nat => insert 2 [0; 1] R) [0; 1; 2]
-  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2)))
-	 : exists R : list nat, insert 2 [0; 1] R
-\end{verbatim}
+
 %

### 2020-10-31 papers/types2020/conc.tex
sec 4 ready some work on conclusion

--- 2020-10-26 papers/types2020/conc.tex
@@
 
-A mention to the use of \emph{delays} in \texttt{check}: instead of generating only ground terms, we could generate say lists with at least two elements as \texttt{[1;2|Tail]} or simply not commit to a base type.
-This  is similar to the use of needed narrowing in LazySmallCheck to generate partial trace or the idea of using CLP for data generation (see Dubois' FocalTest or the stuff by Fabio Fioravanti).
+\todo{sum up}
+
+\begin{metanote}
+  Decided against a related work section: most is already discussed in the intro, the rest can be mentioned in passing --am
+\end{metanote}
+% \subsection{Related work}
+% \label{ssec:rel}
+
+% \begin{itemize}
+% \item Brief discussion/comparison with \texttt{auto}
+% \item Mention other papers by Enrico as applications of coq-elpi~\cite{coen19mscs,tassi19itp}\dots
+% \item Say we \textbf{do} not talk about Coq metaprogramming (Ltac, Mtac), do say that doing our to application in MetaCoq~\cite{sozeau2020metacoq} would be a disaster (reimplement unfication, backtracking, binders)
+% \end{itemize}
+
+\subsection{Future work}
+\label{ssec:fut}
+
+The code reported in Fig.~\ref{fig:augmented} is somewhat idealized in
+so far that we omitted a rule performing weak-head reduction (allowing
+to handle directly existential goals) and more importantly that used
+the \emph{delay} mechanism. This a distinguished feature of Elpi, whose
+usefulness has been advocated in ~\cite{DunchevCT16,coen19mscs}.  The
+use of constraints for data generation is well
+developed~\cite{FioravantiPS15} and we could try to leverage to
+improve PBT using partially instantiated terms: for example generating
+say lists with at least two elements as \texttt{[1,2|Tail]}, without
+recurring to needed narrowing as in in
+LazySmallCheck~\cite{smallcheck}.
+
+\todo{Link to the repo}
+Extending our tactic to handle computations (\texttt{simp}) and/or to return the partial proof state, unsolved goals in particular. Mention \emph{Harpoon}, Beluga's new tactic language?
+
+Extending the LJF kernel to handle defined atoms
 %%% Local Variables:
@@
 %%% End:
+
+% LocalWords:  Elpi

### 2020-10-31 papers/types2020/intro.tex
sec 4 ready some work on conclusion

@@
 implementations built using Elpi
-\cite{coen19mscs,tassi18coqpl,tassi19itp} support the usefulness of
-higher-order logic programming as a meta-programming language useful
-for Coq.
+support the usefulness of
+higher-order logic programming as a meta-programming language 
+for proof assistants in general~\cite{DunchevCT16,coen19mscs} and in particular
+for Coq via the \texttt{coq-elpi} plug in~\cite{tassi19itp,CohenST20} .
 
@@
 % LocalWords:  Sacerdoti Coen Tassi Elpi Teyjus Felty LCF tacticals
-% LocalWords:  intuitionistic
+% LocalWords:  intuitionistic elpi

### 2020-10-31 papers/types2020/pbt.tex
sec 4 ready some work on conclusion

--- 2020-10-29 papers/types2020/pbt.tex
@@
 When using a focused proof system for logic extended with fixed
-points~\cite{baelde12tocl}, negation corresponds to the usual intutionistic interpretation.  % proofs of formulas such as
+points~\cite{baelde12tocl}, negation corresponds to the usual
+intutionistic interpretation, which is close to what Coq supports.  % proofs of formulas such as
 % \[
@@
 However, for the sake of PBT, we can identify a proof certificate for
-$(\ref{eq:full})$ with a proof certificate for
-\[
+$(\ref{eq:full})$ simply with a proof certificate for
+$
   \exists x [\tau(x)\wedge P(x)].
-  \tag{**}\label{eq:short}
-\]
-and will resort to negation-as-failure to check that the conclusion does not hold.
-This also means that we will not be able to return a Coq proof terms for the refutation of our property --- and neither does \textsf{QuickChick}, which runs at the OCaml level --- although we can return the witness for the existential.
+$
+and we can resort to negation-as-failure to check that the conclusion does not hold without caring for any evidence for the latter.
+This also means that we do not return a Coq proof terms for the refutation of our property --- and neither does \textsf{QuickChick}, which runs at the OCaml level --- although we can return the witness for the existential.
 % Such a certificate would contain the witness (closed) term $t$ for the
@@
 \subsection{An Elpi tactic for PBT}
-We will invoke the tactic in a
-proof environment  where the goal is % one is trying to prove is the
-the property that the system-under-test should meet. This means that, after
-\lsti|intro| has been used to introduce the relevant hypotheses, we
- have a proof context % consisting of typing information for a bunch of variables
-where eigenvariables  play different roles:  dependent
-variables, that is the ones appearing also in the types of other variables, are subject to generation, while
-non-dependent ones tags assumptions that are only to tested. This is reminiscent of bi.directional type-checking. % Obviously, the
+We will invoke the tactic in a proof environment where the overall
+goal is % one is trying to prove is the
+the property that the system-under-test should meet. This means that,
+after \lsti|intro| has been used to introduce the relevant hypotheses,
+we have a proof
+context % consisting of typing information for a bunch of variables
+where eigenvariables play different roles: dependent variables, i.e., the ones appearing also in the types of other variables, are
+subject to \emph{generation}, while non-dependent ones tag assumptions that
+are only to tested. This is reminiscent of bi-directional
+type-checking. % Obviously, the
 % two are the result of and \lsti|intro| on either a dependent or a
@@
 to this, the user should specify all the certificate information that
-will guide the data generation phase. The goal of the current
-environment will directly provide the property for which we want to
-find a counterexample. % Say we want to generate a counterexample for
+will guide the data generation phase. % The goal of the current
+% environment will directly provide the property for which we want to
+% find a counterexample.
+% Say we want to generate a counterexample for
 % some specification represented by \lsti|H1| and \lsti|H2| in the
@@
 \end{lstlisting}
-The tactic will call \lsti{check} with the given FPC on the dependent variables and will simply call a vanilla meta-interpreter (see Fig.~\ref{fig:interp} in the appendix) to test
+The tactic will call \lsti{check} with the given FPC on the dependent variables and will simply call a vanilla meta-interpreter (see Fig.~\ref{fig:interp} in the appendix) to test the negation of the property:
 \begin{lstlisting}
@@
 \end{lstlisting}
-where \lsti{not} in \lP's negation-as-failure operator.
+where \lsti{not} is \lP's negation-as-failure operator.
 
@@
 In this query the tactic tests the hypotheses \lsti{Ho} and \lsti{Hi}
-against data \lsti{x,xs} generated exhaustively up to proofs bounded
-by at most $5$ unfoldings of the library \lsti{Inductive} definition
-of \lsti{nat} and \lsti{list}. Since we did slip in an error (where?), our
-tactic reports a counter-example, namely
-\verb|(* Proof Term: [0, [0; 1; 0]] *)|
+against data \lsti{x,xs} generated exhaustively via proofs bounded by
+at most $5$ unfoldings of the library \lsti{Inductive} definitions of
+\lsti{nat} and \lsti{list}. We do not generate values for \lsti{rs},
+since by mode information we know that it will be computed.
+%
+Since we did slip in an error (where?), our tactic reports a
+counter-example, namely \verb|Proof Term: [0, [0; 1; 0]]|, and
+we therefore abort the proof.
 
-In order to generate the query to the \lsti|check| predicate, some
+In order to generate the PBT query, some
 pre-processing is needed. In particular, we need to turn the data
-variables into \lP logic variables when they appear inside a
-specification, and to generate queries to \lsti|check| for each of
+eigenvariables into \lP logic variables when they appear inside a
+specification, and to generate queries %to \lsti|check|
+for each of
 these logic variables in association with the type of the data
@@
 we leverage extensively \lP's higher order programming features. The
-substitutions are handled with the technique of \lsti|copy| clauses.
+substitutions are handled with the technique of \lsti|copy| clauses~\cite{miller91jlc}.
 
-Note that the same query in \textsf{QuickChick} would have required
-much more setup: if we wished to proceed relationally as above, we
-would have had to provide a proof of decidability of the relevant
-notions. Were we to use functions, then we would have to implement a
-generator and shrinker for ordered lists.
+%\todo{Matteo: we're doing some preprocessing for interp as well, right?}
+Note that testing the above conjecture with \textsf{QuickChick} would
+have required much more setup: if we wished to proceed relationally as
+above, we would have had to provide a proof of decidability of the
+relevant notions. Were we to use functions, then we would have to
+implement a generator and shrinker for ordered lists, since automatic
+derivation of the former does not (yet) work for this kind of specification.
 
-\todo{the \texttt{aexp} example}
+%\todo{the \texttt{aexp} example}
 
+For a more significant case study, let us turn to the semantics of
+programming languages, where PBT has been used extensively and
+successfully~\cite{Klein12}. Here we will consider a far simple example,
+a typed arithmetic language featuring numerals with predecessor
+and Booleans with if-then-else and test-for-zero, which comes from
+(\url{https://softwarefoundations.cis.upenn.edu/plf-current/Types.html}):
+\begin{lstlisting}
+Inductive tm : Type :=
+ | ttrue : tm | tfalse : tm | tif : tm -> tm -> tm -> tm | tzero : tm | tsucc : tm -> tm
+ | tpred : tm -> tm | tiszero : tm -> tm.
+ Inductive typ : Type := |TBool : typ |TNat : typ.
+\end{lstlisting}
+The static and small step dynamic semantics are reported in appendix~\ref{ssec:types}.
+
+While it is obvious that \emph{subject expansion} fails for this calculus, it is gratifying to have it confirmed by our tactic, with counterexample \lsti{e = tif ttrue tzero ttrue}:
+\begin{lstlisting}
+Conjecture subexp:  forall e e' t, step e e' -> has_type e' t -> has_type e t. 
+ intros e e' t HS HT.
+ elpi dep_pbt height 2 (HS /\ HT) (e).
+\end{lstlisting}
+
+Another way to asses the fault detection capability of our PBT setup is via \emph{mutation analysis}~\cite{CavadaCM20}, whereby localized bugs are voluntarily inserted, with the view that they should caught (``killed'') by the testing suite.
+Following on  an exercise if the aforementioned chapter of SF, we modify the typing relation by adding the following (nonsensical) additional clause:
+
+\begin{lstlisting}
+Module M1.
+Inductive has_type : tm -> typ -> Prop :=
+$\dots$
+  | T_SuccBool : forall t, has_type t TBool -> has_type (tsucc t) TBool.
+end M1.
+\end{lstlisting}
+Some of the required properties for our SUT now fails: not
+only type uniqueness, but also progress with counterexample \lsti{e = tsucc ttrue}:
+\begin{lstlisting}
+Definition progress (e :tm) (Has_type : tm -> typ -> Prop) (Step : tm -> tm -> Prop):= 
+     forall t, Has_type e t -> notstuck e Step.
+Conjecture progress_m1:  forall e, progress e M1.has_type step.
+ unfold progress.
+ intros e t Ht.    
+ elpi dep_pbt 2 height (Ht) (e) .  
+\end{lstlisting}
+
+Whereas examples based on the \emph{Types} chapter of SF are
+admittedly quite simple-minded, they have been used as a benchmark for
+evaluating \textsf{QuickChick}'s automation of generators under
+invariants~\cite{LampropoulosPP18}, and to be amenable to the tool,
+the specification had to be massaged non-insignificantly.
 %%% Local Variables:
@@
 
-% LocalWords:  Bedwyr bipole elpi QuickCheck OCaml TODO
+% LocalWords:  Bedwyr bipole elpi QuickCheck OCaml TODO coq xs nat
+% LocalWords:  tif ttrue tzero tsucc SUT

### 2020-11-01 papers/types2020/app.tex
related work bits spear thru the paper
ported the appendix and sec3 to kernel-v2

sec3 is still a mess

@@
 
-In Fig.~\ref{fig:interp} we the encoding of the vanilla meta-interpreter used in
+In Fig.~\ref{fig:interp} we report the encoding of the vanilla meta-interpreter used in
 the testing phase of the \lsti{dep_pbt} tactic.
-Differently from Fig.~\ref{fig:augmented} we appeal, via quotations, to Coq's defined connectives --- note the use of a green cut to avoid overlapping with the \lsti{Inductive} atomic case and the absence of certificates.
+Differently from Fig.~\ref{fig:augmented} we appeal, via quotations, to Coq's defined connectives.
+An \lsti{atomic} proposition is one defined \lsti{Inductive}ly.
+The predicate \lsti{definition} hides the API call to the constructors of such a type.
 
@@
  \lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
-\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
+\lstinputlisting[linerange={interp-end}]{code/dep-kernel-v2.mod}
 \caption{Encoding in coq-elpi of the vanilla meta-interpreter}
@@
 
+We list the rules for static and dynamic semantics of the language mentioned in section~\ref{ssec:dep_pbt} and related notions:
 \begin{lstlisting}

### 2020-11-01 papers/types2020/conc.tex
related work bits spear thru the paper
ported the appendix and sec3 to kernel-v2

sec3 is still a mess

@@
 % \item Brief discussion/comparison with \texttt{auto}
-% \item Mention other papers by Enrico as applications of coq-elpi~\cite{coen19mscs,tassi19itp}\dots
-% \item Say we \textbf{do} not talk about Coq metaprogramming (Ltac, Mtac), do say that doing our to application in MetaCoq~\cite{sozeau2020metacoq} would be a disaster (reimplement unfication, backtracking, binders)
+% \item Mention other papers by Enrico as applications of coq-elpi~\cite{coen19mscs,tassi19itp}\dots [done]
+% \item Say we \textbf{do} not talk about Coq metaprogramming (Ltac, Mtac), do say that doing our to application in MetaCoq~\cite{sozeau2020metacoq} would be a disaster (reimplement unfication, backtracking, binders) [done]
 % \end{itemize}
@@
 
-\todo{Link to the repo}
-Extending our tactic to handle computations (\texttt{simp}) and/or to return the partial proof state, unsolved goals in particular. Mention \emph{Harpoon}, Beluga's new tactic language?
+\todo{Link to the repo? \url{https://github.com/proofcert/fpc-elpi}}
+
+Extending our tactic to handle computations (\texttt{simp}) and/or to
+return the partial proof state, unsolved goals in particular. Mention
+\emph{Harpoon}, Beluga's new tactic language?
 

### 2020-11-01 papers/types2020/intro.tex
related work bits spear thru the paper
ported the appendix and sec3 to kernel-v2

sec3 is still a mess

@@
 implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
-more recently, Tassi has made Elpi available as a plugin to the Coq
-proof assistant \cite{tassi18coqpl}.  This implementation of \lP
+more recently, Tassi has made Elpi available as the \texttt{coq-elpi} plugin to the Coq
+proof assistant \cite{tassi18coqpl} (\url{https://github.com/LPCIC/coq-elpi}).  This implementation of \lP
 extends earlier implementations in primarily two directions:  First,
@@
 the Teyjus implementation \cite{nadathur99cade} of \lP. Second, the
-plugin version of Elpi comes equipped with an API for accessing the
-Coq environment and a quotation and anti-quotation syntax for mixing
-for Coq expressions with \lP program elements.
+plugin version of Elpi comes equipped with  a quotation and anti-quotation syntax for mixing
+Coq expressions with \lP program elements and an API for accessing the
+Coq environment, typechecking included.
 
@@
 Since relations (not functions) play a central role in \lP, Elpi is
-capable to providing direct support for the many relations that arise
-in proof assistants.  Such relations include typing (e.g.,
-$\Gamma\vdash M\colon\sigma$), evaluation (e.g., natural semantics specifications
-\cite{kahn87stacs,hannan93jfp}), and interaction (e.g., structured
-operational semantics \cite{plotkin81,milner89book}).  Felty has also
+capable to providing direct support for the many relations that have a
+role in proof assistants implementation and usage.  Such relations include typing (e.g.,
+$\Gamma\vdash M\colon\sigma$), evaluation (e.g., natural semantics
+specifications \cite{kahn87stacs,hannan93jfp}), and interaction (e.g.,
+structured operational semantics \cite{plotkin81,milner89book}).
+
+Felty has also
 made the point that LCF-style tactics and tacticals can be given an
@@
 for Coq via the \texttt{coq-elpi} plug in~\cite{tassi19itp,CohenST20} .
+%
 
@@
 
+While other meta-programming frameworks based on functional
+programming such as \emph{MetaCoq}~\cite{sozeau2020metacoq} can and have been used for related endeavors, we believe (together with~\cite{DunchevCT16}) that they would require much more boilerplate.
+
 Before we can present these examples, we first highlight the rather
 striking differences in notions of computing and reasoning that arise
-on each side of the Elpi-Coq API.   We will also present a quick
+on each side of the \texttt{coq-elpi} API.   We will also present a quick
 summary of the key proof theory concepts that are used by our example
@@
 % LocalWords:  Sacerdoti Coen Tassi Elpi Teyjus Felty LCF tacticals
-% LocalWords:  intuitionistic elpi
+% LocalWords:  intuitionistic elpi coq MetaCoq

### 2020-11-01 papers/types2020/pbt.tex
related work bits spear thru the paper
ported the appendix and sec3 to kernel-v2

sec3 is still a mess

@@
 \subsection{An Elpi tactic for PBT}
+\label{ssec:dep_pbt}
 We will invoke the tactic in a proof environment where the overall

### 2020-11-01 papers/types2020/sec3.tex
sec 4 ready some work on conclusion
related work bits spear thru the paper
ported the appendix and sec3 to kernel-v2

sec3 is still a mess

@@
 We \textbf{do} not talk about PBT yet, just logic programming (with proof terms).
-In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
- One way to proceed:
+% In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
+% One way to proceed:
   % \begin{itemize}
@@
   % \end{itemize}
-
+%
 Stuff in red is from old paper
 \end{metanote}
-
-%\subsection{Horn logic in coq-elpi}
 
@@
 \todo{Adapt the discussion}
-\todo{grammar for formulae and proof terms\cite{Herbelin94,LengrandDM06}?}
+\todo{grammar for formulae and proof terms \cite{Herbelin94,LengrandDM06}  or just the figure? Mention spine calculus}
 
@@
 
+\todo{Figure now uses Dale kernel-v2 --- needs further abstraction and explanation, see appendix}
 \begin{footnotesize}
@@
 %\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
-\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
+\lstinputlisting[linerange={check-end}]{code/dep-kernel-v2.mod}
 \caption{Implementation of a  dependent kernel for Coq terms.}
@@
 
-\subsection{Certificates}
+%\subsection{Certificates}
 
-\todo{Shorten it}
+\todo{What happened to the unfold certificate???}
 % \todo{Show the dprolog tactic? have an example? could  be
 % \lsti{Goal  exists R, insert 2 [0;1] R.} and we would move the code for insert here}
-{\color{red}
-As we mentioned in the introduction, our notion of proof certificates
+{\color{red}}
+The notion of proof certificates
  is taken from the general setting of \emph{foundational
@@
 that is, the exact set of constructors for the \lsti{cert} type as
-well as the exact specification of the six expert predicates listed in
+well as the exact specification of the expert predicates listed in
 that figure.
@@
 %
-As is easy to verify, the query \mbox{\lsti{(check (qheight 5) B)}} (for the
-encoding \lsti{B} of a goal formula) is provable in \lP using the
-clauses in Figures~\ref{fig:kernel} and~\ref{fig:resources} if and
-only if the height of that proof is 5 or less.
+As is easy to verify, the query \mbox{\lsti{(check (qheight 5) B)}} is
+provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
+and~\ref{fig:resources} if and only if the height of that proof is 5
+or less.
 %
@@
 %
-In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
+%In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
 % \begin{lstlisting}
@@
 % \end{lstlisting}
-is provable if and only if the total number of unfoldings of that
-proof is 5 or less.
+%is provable if and only if the total number of unfoldings of that
+%proof is 5 or less.
 
@@
 \end{figure}
-
-% Finally, Figure~\ref{fig:max} contains the FPC based on the constructor
-% \lsti{max} that is used to record explicitly all information within a
-% proof: in particular, all disjunctive choices and all substitution
-% instances for existential quantifiers are collected into a binary tree
-% structure of type \lsti{max}.
-% %
-% In this sense, proof certificates built with the \lsti{max}
-% constructor are \emph{maximally} explicit.
-% %
-% Such proof certificates are used, for example, in~\cite{Pair}; it is
-% important to note that proof checking with such maximally explicit
-% certificates can be done with much simpler proof-checkers than those
-% used in logic programming since backtracking search and unification
-% are not needed.
-%
-
-% \begin{figure}
-% \lstinputlisting[linerange={max-end}]{code/fpcs.sig}
-% \lstinputlisting[linerange={max-end}]{code/fpcs.mod}
-% \caption{The \lsti{max} FPC}
-% \label{fig:max}
-% \end{figure}
 
@@
 request that the corresponding experts also succeed for both
-\lsti{Cert1} and \lsti{Cert2} and, in the case of the \lsti{orE} and
-\lsti{someE}, also return the same choice and substitution term,
-respectively.
+\lsti{Cert1} and \lsti{Cert2} 
 %
-Thus, the query 
-\begin{lstlisting}
-?- check ((qheight 4) <c> (qsize 10)) B
-\end{lstlisting}
+Thus, the query \lsti{check ((qheight 4) <c> (qsize 10)) B}
 will succeed if there is a proof of \lsti{B} that has a height less
 than or equal to 4 while also being of size less than or equal to 10.
-%
-% A related use of the pairing of two proof certificates is to %use it to
-% \emph{distill} or \emph{elaborate} proof certificates. 
-% %
-% For example, the proof certificate \lsti{(qsize 5 0)} is rather implicit
-% since it will match any proof that used unfold exactly 5 times.
-% %
-% However, the query
-% \begin{lstlisting}
-% ?- check ((qsize 5 0) <c> (max Max)) B.
-% \end{lstlisting}
-% will store into the \lP variable \lsti{Max} more complete details of
-% any proof that satisfies the \lsti{(qsize 5 0)} constraint.
-% %
-% In particular, this forms the infrastructure of 
-%  an \emph{explanation} tool for attributing ``blame'' for
-% the origin of a counterexample; these maximal
-% certificates are an appropriate starting point for documenting
-% both the counterexample and why it serves as a counterexample.
 
-% \begin{figure}
-% \lstinputlisting[linerange={pairing-end}]{code/fpcs.sig}
-% \lstinputlisting[linerange={pairing-end}]{code/fpcs.mod}
-% \caption{FPC for pairing}
-% \label{fig:pairing}
-% \end{figure}
+% Various  additional examples and experiments using the pairing of FPCs
+% can be found in~\cite{blanco17cade}. Using similar techniques, it is possible to define FPCs that target
+% specific types for special treatment: for example, when generating
+% integers, only (user-defined) small integers could be produced.
 
-Various  additional examples and experiments using the pairing of FPCs
-can be found in~\cite{blanco17cade}. Using similar techniques, it is possible to define FPCs that target
-specific types for special treatment: for example, when generating
-integers, only (user-defined) small integers could be produced.
-}
-
-\subsection{A prolog tactic}
+\subsection{A prolog-like tactic}
 
@@
 
+The \lsti{dprolog} tactic can be seen as a programmable version of
+Coq's \lsti{auto}: it currently lacks the latter capability of solving
+goals by \lsti{reflexivity} and the modularity capability of using
+\lsti{Hints}, but it is not compelled to use depth-first search, since it follows the dictate of the given FPC.
+For example the FPC can provide a \emph{trace}, which may be more customable that the one offered by the \lsti{Debug} option.
 %
@@
 % LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE
+% LocalWords:  Elpi modularity

### 2020-11-02 papers/types2020/sec3.tex
Comments on the code, some text on proof system

@@
 % \qquad 
+\infer{\XXi;\Gamma\vdash A : s}
+{(A,s) \in \mathcal{A}
+}
+\qquad
 \infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
@@
 \qquad
-\infer{\XXi;\Gamma\vdash t : A}
-      {\bc{\XXi}{\Gamma} t D A \quad D \in \Pscr\cup\Gamma}
+\infer{\XXi;\Gamma\vdash k l : A}
+      {\bc{\XXi}{\Gamma} l D A \quad k : D \in \Pscr_A\cup\Gamma}
 \]
@@
 
+The proof system we use is a sequent calculus inspired by the calculus for
+proof search in Pure Type Systems introduced in \cite{LengrandDM06}, based on
+its turn on ideas stemming from focusing (in particular, the LJT calculus of
+Herbelin \cite{Herbelin94}). Similarly to that calculus, we have a term
+language that includes terms and lists of terms, and two distinct typing
+judgements for the two categories. In our case, where proof search steps are
+viewed as computation, the term judgement corresponds to a goal-reduction
+step while the list judgement corresponds to a backchaining step.
+
+Since we are only dealing with types that correspond to Horn clauses, and
+atomic types are inductively defined, we will never have occurrences of
+applications of hypotetic assumptions. This means, at the term level, that
+we don't need free variables in our grammar of terms.
+
+\begin{align*}
+t,u :=& k l \;|\; t t \;|\; \forall x : t.\, u \\
+l :=& [] \;|\; (t l)
+\end{align*}
+
+The syntax will depend on the external set of inductive definitions available
+in the context: we denote by $\Pscr_A$ the set of type declarations for the
+constructors of type $A$, of the form $k : D$ where
+
 Figure~\ref{fig:augmented} contains a simple proof system \dots  in which each inference rule is augmented with an
@@
 
-Figure~\ref{fig:kernel} contains the Elpi implementation of the
+Figure~\ref{fig:kernel} contains the elpi implementation of the
 inference rules in Figure~\ref{fig:augmented}: here the infix

### 2020-11-02 papers/types2020/elab.tex
Finished a draft of section 5.

@@
-\newpage
+%\newpage
 \section{Elaboration of external proof certificates for the Coq
@@
 
-There are many theorem provers for intuitionistic logic for which a
-successful proof is not the kind of detailed $\lambda$-term required
-by the Coq kernel.  Most prover provide, instead, a trace of some key
-aspects of a proof: some details are often not included in such a
-trace.  For example:
+There are many theorem provers for intuitionistic logic
+\cite{raths07jar} for which a successful proof is not the kind of
+detailed $\lambda$-term required by the Coq kernel.  Often, such
+provers provide no information about the proofs they discover.  To the
+extent that some prove evidence is output after a successful run, such
+evidence is usually just a trace of some key aspects of a proof: some
+details are often not included in such a trace.  For example:
 \begin{enumerate}
@@
 
-As we have describe above, the Foundational Proof Certificate (FPC)
-framework has been designed for formally defining a rich collection of
-proof certificates that do not necessarily contain full details.
-Proof checkers for FPC certificates can be written using simple logic
-programs: the standard implementation mechanisms of unification and
-backtracking search can then be used to reconstruct missing proof
-details \cite{miller17fac}.  Such proof checking kernels can also be
-used to elaborate such certificates into fully detailed proofs
-\cite{blanco17phd,blanco17cade}.
+As we have describe in Section~\ref{sec:three}, the Foundational Proof
+Certificate (FPC) framework has been designed for formally defining a
+rich collection of proof certificates that do not necessarily contain
+full details.  Proof checkers for FPC certificates can be written
+using simple logic programs: the standard implementation mechanisms of
+unification and backtracking search can then be used to reconstruct
+missing proof details \cite{miller17fac}.  Such proof checking kernels
+can also be used to elaborate such certificates into fully detailed
+proofs \cite{blanco17phd,blanco17cade}.
 
@@
 logic programming in order to provide a flexible approach to
-connecting external provers of intuitionistic logic to Coq.
+connecting external provers of first-order intuitionistic logic to
+Coq.
 %
@@
 
-To illustrate the possibilities ***
+To illustrate this approach to elaborating externally provided proof
+structures into terms the Coq kernel can check, we consider here
+examples take from the file \verb+coq_fpc.v+, which is part of the
+repository \url{https://github.com/proofcert/fpc-elpi/}.
+Figure~\ref{fig:debruijn} presents the specific FPC that defines the
+proof semantics of using the De Briujn notions for untyped
+$\lambda$-terms as proof evidence of a proof in intuitionistic
+propositional logic.  That FPC definition, along with a proof checker
+capable of synthesizing Coq proof terms during proof checker, makes it
+possible to use the term denoting the untyped $\lambda$-term
+$\lambda x (x (\lambda y (\lambda z (x (\lambda u z)))))$
 
-** We limit our attention here to first-order intuitionistic logic:
-that is, we do not consider here inductive definitions.
+\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
+(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 
+                                                    [lambda (apply 1 [])])])]))
+\end{lstlisting}
+as a proof of following theorem of Coq.
+\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
+Theorem dneg_peirce_mid : forall P Q: Prop, (((((P -> Q) -> P) -> P) -> Q) -> Q).
+\end{lstlisting}
 
-** Note that with intuitionistic logic, there is much less ambiguity
-of polarities possible.  In particular, only the conjunction and the
-atomic formulas can have ambiguous polarizations.  The two
-conjunctions are familiar to those working with typed
-$\lambda$-calculus:
+\begin{figure}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.sig}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.mod}
+\caption{The FPC definition of De Bruijn notation as proof evidence.}
+\label{fig:debruijn}
+\end{figure}
 
-the negative conjunction (corresponding to $\with$ in linear logic)
-corresponds to surjective pairing.
-
-the positive conjunction (corresponding to $\with$ in linear logic)
-corresponds to ***.
-
-The choice between negated atoms and positive atoms is similar to the
-choice in representing $\beta$-normal terms using either head-normal
-form or administrative normal form.
-
-We built the fpccoq system to demonstrate the feasibility of using
-logic programming to check proof certificates produced by external
-provers and to elaborate such certificates into proofs with sufficient
-details that they can be given directly to Coq's kernel to
-check.  Note that one does not need to trust the external prover nor
-the fpccoq tool-chain: one only needs to trust the Coq kernel.
-
-De Bruijn
-decide depth
-simply typed lambda terms
-
-\newpage
+%\newpage
 

### 2020-11-02 papers/types2020/sec3.tex
Added label to section 3.
More about implementation

@@
 \section{Proof theory and proof certificates}
+\label{sec:three}
 
@@
   \newcommand{\XXi}{{\color{blue}{\Xi}}}
-
   \newcommand{\bc}[5]{#1;#2\vdash #3 : #4 \Downarrow #5}
@@
 }
+% \qquad
+% \infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
+% {\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
+% }
 \qquad
-\infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
-{\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
-}
-\qquad
-\infer{\XXi;\Gamma\vdash k l : A}
-      {\bc{\XXi}{\Gamma} l D A \quad k : D \in \Pscr_A\cup\Gamma}
+\infer{\XXi_1;\Gamma\vdash k\ l : A}
+      {\bc{\XXi_2}{\Gamma} l D A \quad k : D \in \Pscr_A\cup\Gamma\qquad \unfoldExpert{\XXi_1}{\XXi_2}} 
 \]
@@
 %  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
-  \infer{\bc{\XXi}{\Gamma}{(t\ l)} {\forall x:B.\, D} A}
+  \infer{\bc{\XXi}{\Gamma}{(t\ ::\ l)} {\forall x:B.\, D} A}
   {\XXi_1;\Gamma\vdash t : A\qquad \bc{\XXi_2}{\Gamma}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
@@
 atomic types are inductively defined, we will never have occurrences of
-applications of hypotetic assumptions. This means, at the term level, that
-we don't need free variables in our grammar of terms.
+applications of hypotetic assumptions. This means, at the term level, that we
+don't need to have variables in our grammar of terms. Terms are always
+applied to a (possibly empty) list of arguments, and as is usual in Pure Type
+Systems we have the same grammar for terms and types, as shown in figure~\ref{fig:terms}.
 
+\begin{figure}
 \begin{align*}
-t,u :=& k l \;|\; t t \;|\; \forall x : t.\, u \\
-l :=& [] \;|\; (t l)
+t,u :=& k\ l \;|\; \forall x : t.\, u \\
+l :=& [] \;|\; (t\ ::\ l)
 \end{align*}
+\caption{Grammar of terms}
+\label{fig:terms}
+\end{figure}
 
-The syntax will depend on the external set of inductive definitions available
-in the context: we denote by $\Pscr_A$ the set of type declarations for the
-constructors of type $A$, of the form $k : D$ where
+The syntax depends on the external set of inductive definitions available in
+the context: we denote by $\Pscr_A$ the set of type declarations for the
+constructors of type $A$, of the form $k : D$ where $D$ is of the form
+$\forall x_1 : t_1.\, \dots \forall x_n : t_n.\, D\ u_1 \dots u_m$. We denote
+by $\mathcal{A}$ the set of arity declaration for the inductive types.
 
-Figure~\ref{fig:augmented} contains a simple proof system \dots  in which each inference rule is augmented with an
-additional premise involving an \emph{expert predicate}, a
-certificate $\Xi$, and possibly continuations of certificates ($\Xi'$,
-$\Xi_1$, $\Xi_2$) with extracted information from certificates (in the
-case of $\vee$ and $\exists$).
+Figure~\ref{fig:augmented} contains a simple proof system \dots in which each
+inference rule is augmented with an additional premise involving an
+\emph{expert predicate}, a certificate $\Xi$, and possibly continuations of
+certificates ($\Xi'$, $\Xi_1$, $\Xi_2$) with extracted information from
+certificates.% (in the case of $\vee$ and $\exists$).
 %
@@
 %
-\todo{Explain relations with other term assignments.
-Note that we do not use explicit substitutions (we are cut free).
-Explain the choice of not having certificates represent proof terms.
-Explain that the unfold clause handles defined atoms and defined connectives (conjunction, disjunction, existential and also equality): this may be less efficient and proof-theoretically induce pointless phase changes, but it is arguably very elegant e simplifies proof terms that arand inherited by Coq }
+\todo{Explain relations with other term assignments. Note that we do not use
+explicit substitutions (we are cut free). Explain the choice of not having
+certificates represent proof terms. Explain that the unfold clause handles
+defined atoms and defined connectives (conjunction, disjunction, existential
+and also equality): this may be less efficient and proof-theoretically induce
+pointless phase changes, but it is arguably very elegant e simplifies proof
+terms that arand inherited by Coq }
 
-Figure~\ref{fig:kernel} contains the elpi implementation of the
+Figure~\ref{fig:kernel} contains the Elpi implementation of the
 inference rules in Figure~\ref{fig:augmented}: here the infix
-turnstile $\vdash$ symbol is replaced by the \lsti{check} predicate.
+turnstile $\vdash$ symbol is replaced by the \lsti{check} predicate, and every rule
+is directly mapped to a clause in the code. The argument of the clause is
+either given by a \lsti{go} or a \lsti{bc} constructor, signaling whether the clause
+is a goal-reduction or a backchaining one.
+
+Coq terms are accessed through the Coq-Elpi API, and their representation
+in \lP takes advantage of the native \lP constructs such as lists and binders.
+The following is the signature of the term constructors we use:
+\begin{lstlisting}
+  kind term   type.
+  type prod   name -> term -> (term -> term) -> term.
+  type app    list term -> term.
+  type global gref -> term.
+  type indt   inductive -> gref.
+  type indc   constructor -> gref.
+\end{lstlisting}
+
+The constructor for products forms a term from a term and a \lP function from
+terms to terms; when, in the implementation of the product-left rule, we
+apply the second argument of \lsti{prod _ Ty1 Ty2} to the variable \lsti{Tm},
+we get a term that can be used to continue backchaining. The meaning of this
+application is meta-level substitution, in the style of Higher Order Abstract
+Syntax: one can think as if something of type \lsti{term -> term} were a
+\lsti{term} with a hole in it for another \lsti{term}.
+
+The unfolding rule makes use of the Coq-Elpi primitives
+\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
+nested) application, and \lsti{coq.env.indt} in order to access the global
+environment of inductive definitions and query for information about them.
+
 %
@@
 
-\todo{What happened to the unfold certificate???}
 % \todo{Show the dprolog tactic? have an example? could  be

### 2020-11-02 papers/types2020/elab.tex

@@
 possible to use the term denoting the untyped $\lambda$-term
-$\lambda x (x (\lambda y (\lambda z (x (\lambda u z)))))$
+$\lambda x (x (\lambda y (y (\lambda z (x (\lambda u\,z))))))$.
 

### 2020-11-03 papers/types2020/cultures.tex
Fixed some typos in Section 4.  Left some comments in footnotes where I wasn't sure what was the correct fix.

@@
 used to reason about computation: for example, cut-elimination can be
-used to prove ``substitution lemmas for free'' \cite{gacek12jar}.
+used to prove ``substitution lemmas for free'' that arise in the study
+of operational semantics \cite{gacek12jar}.
 
@@
 illustrated by considering how they are used in the specification of
-tactics.  In fact, the origin story for the ML functional programming
+tactics.  The origin story for the ML functional programming
 language was that it was the meta-language for implementing the LCF

### 2020-11-03 papers/types2020/pbt.tex
Fixed some typos in Section 4.  Left some comments in footnotes where I wasn't sure what was the correct fix.

@@
 
-We have presented n a previous paper~\cite{blanco19ppdp} a
+We have presented in a previous paper~\cite{blanco19ppdp} a
 proof-theoretical reconstruction of property-based testing (PBT) of
@@
 Of course, Coq already features \textsf{QuickChick}~\cite{QChick}
-(\url{https://softwarefoundations.cis.upenn.edu/qc-current}) is a
+(\url{https://softwarefoundations.cis.upenn.edu/qc-current}), which is a
 sophisticated and well-supported PBT tool, based on a different
@@
 commitment to the random generation strategy, and related
-configuration of generators and shrinkers.
+configuration of generators and shrinkers.\footnote{DM I don't
+  understand this sentence.  Rewrite?}
 
@@
 
- If we view  a property as a logical formula
+Consider a conjecture of the form
 \(\forall x [(\tau(x)\wedge P(x)) \supset Q(x)]\) where $\tau$ is a
 typing predicate and $P$ and $Q$ are two other predicates defined
-using Horn clause specifications, providing a counter-example
-consists of  negating the property, and \emph{searching} for a proof of
+using Horn clause specifications.  A counter-example to this conjecture
+consists of a witness to the negated formula
 \[
@@
 %\(\exists x [(\tau(x)\land P(x)) \land \neg Q(x)]\).
-
-In our current setting $P$ and $Q$ will be \lsti{Inductive} propositions, while $\tau$   are honest-to-goodness datatypes.
-We will treat the two quite differently, in so far as elements of those types will be \emph{generated}, while predicates will only be \emph{checked}.
+%
+In our current setting $P$ and $Q$ will be \lsti{Inductive}
+propositions, while $\tau$   are honest-to-goodness datatypes. We will
+treat the properties and type quite differently, in so far as elements of those types will be \emph{generated}, while predicates will only be \emph{checked}.
 This distinction plays also a part in interpreting the negation sign.
@@
 points~\cite{baelde12tocl}, negation corresponds to the usual
-intutionistic interpretation, which is close to what Coq supports.  % proofs of formulas such as
+intuitionistic interpretation, which is close to what Coq supports.  % proofs of formulas such as
 % \[
@@
 $
-  \exists x [\tau(x)\wedge P(x)].
+  \exists x [\tau(x)\wedge P(x)]
 $
@@
 subject to \emph{generation}, while non-dependent ones tag assumptions that
-are only to tested. This is reminiscent of bi-directional
+are only to be tested. This is reminiscent of bi-directional
 type-checking. % Obviously, the
@@
 \begin{lstlisting}
-interp $(A_1\land\dots\land A_m$), not (interp $B$)}
+interp $(A_1\land\dots\land A_m$), not (interp $B$)
 \end{lstlisting}
@@
 programming languages, where PBT has been used extensively and
-successfully~\cite{Klein12}. Here we will consider a far simple example,
+successfully~\cite{Klein12}. Here we will consider a far simpler example,
 a typed arithmetic language featuring numerals with predecessor
-and Booleans with if-then-else and test-for-zero, which comes from
+and booleans with if-then-else and test-for-zero, which comes from
 (\url{https://softwarefoundations.cis.upenn.edu/plf-current/Types.html}):
@@
 \end{lstlisting}
+\footnote{in the elpi dep-pbt call, we have (height 2) here but (2
+  height) below.  Which is it?}
 
-Another way to asses the fault detection capability of our PBT setup is via \emph{mutation analysis}~\cite{CavadaCM20}, whereby localized bugs are voluntarily inserted, with the view that they should caught (``killed'') by the testing suite.
+Another way to asses the fault detection capability of our PBT setup is via \emph{mutation analysis}~\cite{CavadaCM20}, whereby localized bugs are voluntarily inserted, with the view that they should be caught (``killed'') by the testing suite.
 Following on  an exercise if the aforementioned chapter of SF, we modify the typing relation by adding the following (nonsensical) additional clause:
@@
 \end{lstlisting}
-Some of the required properties for our SUT now fails: not
+Some of the required properties for our SUT\footnote{DM What is SUT?} now fails: not
 only type uniqueness, but also progress with counterexample \lsti{e = tsucc ttrue}:

### 2020-11-03 papers/types2020/cultures.tex
comments to s 1 2 5

--- 2020-10-28 papers/types2020/cultures.tex
@@
 usually decides that a proof is a certain kind of $\lambda$-term
-within the system.  In contrast, the proof theory treats logical
+within the system.  In contrast,  proof theory treats logical
 propositions and proofs as separate.  For example, proof theory does
@@
 
-Gentzen's discovery that the key to treating classical and
-intuitionistic logics in the same proof system was the invention of
+Gentzen's discovered that the key to treating classical and
+intuitionistic logics in the same proof system was %the invention of
 the weakening and contraction structural rules on the right-side of
@@
 As is often observed, however, sequent calculus is too low-level to be
-used explicitly as capturing the ``essence of a proof''.  Recent
-developments in focused proof systems \cite{andreoli92jlc,liang09tcs},
+used explicitly as capturing the ``essence of a proof''.  
+Developments in focused proof systems \cite{andreoli92jlc,liang09tcs},
 makes it possible to extract from sequent calculus not only natural
@@
 
-Given that \lP and Coq both result from a mixing of the
-$\lambda$-calculus with logic, it is important to understand their
-differences.  The confusion around the term \emph{higher-order
-abstract syntax} (HOAS), for example, helps to illustrate their
-differences.  In the functional programming setting, including the Coq
-system, the HOAS approach leads to encoding binding structures within
-terms using functions.  Such encodings are surprisingly complicated
-since they allow for \emph{exotic terms} \cite{despeyroux95tlca} and
-for structures on which induction is not possible
-\cite{roeckl01fossacs}.  In the logic programming setting,
-particularly \lP, HOAS is well supported since bindings are allowed
-with terms ($\lambda$-bindings), in formulas (quantifiers), and in
-proofs (eigenvariables).  For this reason, the term
-\emph{$\lambda$-tree syntax} was introduced to name this particular
-take on HOAS \cite{miller19jar}.  The Abella proof assistant
-\cite{baelde14jfr} was designed, in part, to provide inductive and
-co-inductive inference involving specifications using the
-$\lambda$-tree syntax approach.
+Given that \lP and Coq both result from combining the
+$\lambda$-calculus with logic, it is important to understand some of
+their their differences.  The confusion around the term
+\emph{higher-order abstract syntax} (HOAS), is a case in point.  In
+the functional programming setting, including the Coq system, the HOAS
+approach leads to encoding binding structures within terms using
+functions.  Such encodings are surprisingly complicated since they
+allow for \emph{exotic terms} \cite{despeyroux95tlca} and for
+structures on which induction is not possible \cite{roeckl01fossacs}.
+\begin{metanote}
+  AM: I find this a bit unbalanced. Exotic terms can be avoided with
+  Honsell's theory of context and/or Adam's PHOAS -- not that I like
+  it.. And of course there is Beluga and friends.  On the other hand
+  Coq users are happy with library-based approaches such as Autosubt or well-typed DB
+  that allows for sphisticated treatments of binders.
+I would say something more generic such as that mixing functions and HOAS is not immediate, may require additional machinery both in terms of type theories than of code generation, while in lp\dots
+\end{metanote}
+In the logic programming setting, particularly \lP, HOAS is well
+supported since bindings are allowed with terms ($\lambda$-bindings),
+in formulas (quantifiers), and in proofs (eigenvariables).  For this
+reason, the term \emph{$\lambda$-tree syntax} was introduced to name
+this particular take on HOAS \cite{miller19jar}.  The Abella proof
+assistant \cite{baelde14jfr} was designed, in part, to provide
+inductive and co-inductive inference involving specifications using
+the $\lambda$-tree syntax approach.
 
@@
 While \lP is a typed language, it only uses simple, polymorphic types,
+\begin{metanote}
+AM:  simple are not poly: can we say ML-style polymorphism?
+\end{metanote}
 and these are more often used to indicate syntactic categories.  Any

### 2020-11-03 papers/types2020/elab.tex
comments to s 1 2 5

@@
 that type is, in fact, accepted by Coq users as a theorem of
-intuitionistic logic.  The rest of the Coq system, especially its
+intuitionistic logic.
+\begin{metanote}
+  Small point: the Coq kernel can be soundly extended with axioms and will also validate classical tautologies
+\end{metanote}
+The rest of the Coq system, especially its
 tactic language, is designed to help a human user build
@@
 provers provide no information about the proofs they discover.  To the
-extent that some prove evidence is output after a successful run, such
-evidence is usually just a trace of some key aspects of a proof: some
-details are often not included in such a trace.  For example:
+extent that some  evidence is output after a successful run, such
+evidence is usually just a trace of some key aspects of a proof, where some
+details are often not included.  For example:
 \begin{enumerate}
@@
 \item Check proofs by executing the logic programming checker that is
-  parametrized by the particular FPC definition.
+  parameterized by the particular FPC definition.
 \end{enumerate}
@@
 Figure~\ref{fig:debruijn} presents the specific FPC that defines the
-proof semantics of using the De Briujn notions for untyped
+proof semantics of using the De Bruijn notions for untyped
 $\lambda$-terms as proof evidence of a proof in intuitionistic
@@
 
+\begin{metanote}
+  AM: I think we need more: the FPC may go to the appendix, we may want to say that there is a tactic that proves these theorems etc. 
+\end{metanote}
+
 %\newpage
@@
 %%% End:
+
+% LocalWords:  intuitionistic

### 2020-11-03 papers/types2020/intro.tex
comments to s 1 2 5

@@
 proof assistant \cite{tassi18coqpl} (\url{https://github.com/LPCIC/coq-elpi}).  This implementation of \lP
-extends earlier implementations in primarily two directions:  First,
+extends earlier ones in primarily two directions:  First,
 Elpi adds a notion of constraints and constraints handling rules, which makes it more expressive than
@@
 
-Since relations (not functions) play a central role in \lP, Elpi is
+Since relations (not functions) are central in \lP, Elpi is
 capable to providing direct support for the many relations that have a
@@
 but also because it soundly
-implements a higher-order intuitionistic logic.  The latter aspect of
+implements a higher-order intuitionistic logic.
+\todo{Explain better? Our kernels are deep embedding, aren't they? -am}The latter aspect of
 Elpi makes it possible to encode the proofs and proof theory of

### 2020-11-03 papers/types2020/pbt.tex
Ran spell checker
pass to sec 4

--- 2020-11-01 papers/types2020/pbt.tex
@@
 
-We have presented n a previous paper~\cite{blanco19ppdp} a
-proof-theoretical reconstruction of property-based testing (PBT) of
-relational specifications, adopting techniques from foundational proof
+We have presented in a previous paper~\cite{blanco19ppdp} a
+proof-theoretical reconstruction of property-based testing~\cite{fink97sen} (PBT) of
+relational specifications adopting techniques from foundational proof
 certificates to account for several features of this testing
@@
 only FPC corresponding to different flavors \emph{exhaustive}
-generation and their combination, as adopted, e.g., in
-SmallCheck~\cite{smallcheck} and
-$\alpha$Check~\cite{cheney_momigliano_2017}.
+generation, as adopted, e.g., in SmallCheck~\cite{smallcheck} and
+$\alpha$Check~\cite{cheney_momigliano_2017}, and their combination.
+Note however that it would take additional two lines of code to the \lsti{unfold} expert to implement a form of random data generation in the sense of randomized backtracking~\cite{pltredexconstraintlogic}.
 
 Of course, Coq already features \textsf{QuickChick}~\cite{QChick}
-(\url{https://softwarefoundations.cis.upenn.edu/qc-current}) is a
+(\url{https://softwarefoundations.cis.upenn.edu/qc-current}),  a
 sophisticated and well-supported PBT tool, based on a different
@@
 generators. While current research~\cite{LampropoulosPP18} aims to
-increase automation, it is fair to say that testing, in particular of
+increase automation, it is fair to say that testing with ~\cite{QChick}, in particular of
 relational specifications, is still very labor intensive. We do not
 intend to compete with \textsf{QuickChick}, but we shall see that we
-can test them immediately \lsti{Inductive} definitions that
+can test immediately \lsti{Inductive} definitions that
 corresponds to pure Horn programs, without any decidability proof nor
@@
 
+\todo{Drop the subsections?}
 \subsection{PBT as proof reconstruction}
@@
 
- If we view  a property as a logical formula
-\(\forall x [(\tau(x)\wedge P(x)) \supset Q(x)]\) where $\tau$ is a
-typing predicate and $P$ and $Q$ are two other predicates defined
+If we view  a property as a many-sorted logical formula:
+\[
+  \forall x_1 \colon \tau_1\dots x_n \colon \tau_m, A_1\land\dots\land A_m \supset B\tag{*}\label{eq:prop}
+  \]
+%\(\forall x [(\tau(x)\wedge P(x)) \supset Q(x)]\) where $\tau$ is a
+%typing predicate and
+  where the $A_i$ and $B$ are  predicates defined
 using Horn clause specifications, providing a counter-example
@@
 \[
-  \exists x [(\tau(x)\wedge P(x)) \wedge \neg Q(x)]
-  \tag{*}\label{eq:full}
+  \exists x_1 \colon \tau_1\dots x_n \colon \tau_n [A_1\land\dots\land A_m\wedge \neg B]
+  \tag{**}\label{eq:full}
 \]
@@
 
-In our current setting $P$ and $Q$ will be \lsti{Inductive} propositions, while $\tau$   are honest-to-goodness datatypes.
-We will treat the two quite differently, in so far as elements of those types will be \emph{generated}, while predicates will only be \emph{checked}.
-This distinction plays also a part in interpreting the negation sign.
+In our Coq current setting $A_i$ and $B$ will be \lsti{Inductive}
+propositions, while the $\tau_j$ are honest-to-goodness datatypes. We
+will treat the two quite differently, in so far as elements of those
+types will be \emph{generated}, while predicates will only be
+\emph{checked}. This distinction is reminiscent of bi-directional
+type-checking and  plays also a part in interpreting the
+negation sign.
 %
@@
 points~\cite{baelde12tocl}, negation corresponds to the usual
-intutionistic interpretation, which is close to what Coq supports.  % proofs of formulas such as
+intutionistic interpretation, which is  what Coq supports.  % proofs of formulas such as
 % \[
@@
 %
-However, for the sake of PBT, we can identify a proof certificate for
+However, for the sake of PBT and as we argued in~\cite{blanco19ppdp}  , we can identify a proof certificate for
 $(\ref{eq:full})$ simply with a proof certificate for
 $
-  \exists x [\tau(x)\wedge P(x)].
+ \exists x_1 \colon \tau_1\dots x_n \colon \tau_n [A_1\land\dots\land A_m].
 $
 and we can resort to negation-as-failure to check that the conclusion does not hold without caring for any evidence for the latter.
-This also means that we do not return a Coq proof terms for the refutation of our property --- and neither does \textsf{QuickChick}, which runs at the OCaml level --- although we can return the witness for the existential.
+This also means that we do not produce a Coq proof terms for the refutation of our property --- and neither does \textsf{QuickChick}, which runs at the OCaml level --- although we can return the witness for the existential.
 % Such a certificate would contain the witness (closed) term $t$ for the
@@
 
-% \begin{metanote}
-%   TODO next
-%   \begin{itemize}
-%   \item The elpi dep-pbt tactic
-%   \item  the simple ordered example [DONE]
-%   \item a bit of details about the \texttt{aexp} example
-%   \end{itemize}
-% \end{metanote}
-
 \subsection{An Elpi tactic for PBT}
@@
 goal is % one is trying to prove is the
-the property that the system-under-test should meet. This means that,
+the property that the system-under-test (SUT) should meet. This means that,
 after \lsti|intro| has been used to introduce the relevant hypotheses,
-we have a proof
-context % consisting of typing information for a bunch of variables
-where eigenvariables play different roles: dependent variables, i.e., the ones appearing also in the types of other variables, are
-subject to \emph{generation}, while non-dependent ones tag assumptions that
-are only to tested. This is reminiscent of bi-directional
-type-checking. % Obviously, the
-% two are the result of and \lsti|intro| on either a dependent or a
-% non-dependent product
-% ; in our setting this division acquires an additional reading beyond
-% what the property looks like: since specifications are encoded as pure
-% Horn clauses where no dependency is involved, non-dependent variables
-% simply tag the specification clauses, whereas dependent variables
-% represent the typing information for the data variables that the
-% specifications use.
+% we have a proof
+% context % consisting of typing information for a bunch of variables
+% where eigenvariables play different roles: dependent variables, i.e., the ones appearing also in the types of other variables, are
+% subject to \emph{generation}, while non-dependent ones tag assumptions that
+% are only to tested.  % Obviously, the
+% % two are the result of and \lsti|intro| on either a dependent or a
+% % non-dependent product
+% % ; in our setting this division acquires an additional reading beyond
+% % what the property looks like: since specifications are encoded as pure
+% % Horn clauses where no dependency is involved, non-dependent variables
+% % simply tag the specification clauses, whereas dependent variables
+% % represent the typing information for the data variables that the
+% % specifications use.
 
-Within the tactic the user therefore
+% Within the tactic
+the user therefore
 specifies which variables of the environment should be used for
@@
 % \end{lstlisting}
-Concretely, for a property
-\[
-\forall x_1 \colon \tau_1\dots x_n \colon \tau_m, A_1\land\dots\land A_m \supset B
-\]
+Concretely, for the property~$(\ref{eq:prop})$
+% \[
+% \forall x_1 \colon \tau_1\dots x_n \colon \tau_m, A_1\land\dots\land A_m \supset B
+% \]
 and the specification of a FPC, the call to the tactic will be:
@@
 \end{lstlisting}
-The tactic will call \lsti{check} with the given FPC on the dependent variables and will simply call a vanilla meta-interpreter (see Fig.~\ref{fig:interp} in the appendix) to test the negation of the property:
+The tactic calls \lsti{check} with the given FPC on the dependent
+variables and will simply call a vanilla meta-interpreter (see
+Fig.~\ref{fig:interp} in the appendix) to test the hypotheses and the negation of the
+goal:
 \begin{lstlisting}
@@
 In this query the tactic tests the hypotheses \lsti{Ho} and \lsti{Hi}
-against data \lsti{x,xs} generated exhaustively via proofs bounded by
-at most $5$ unfoldings of the library \lsti{Inductive} definitions of
+against data \lsti{x,xs} generated exhaustively up to height
+at most $5$  out of the library \lsti{Inductive} definitions of
 \lsti{nat} and \lsti{list}. We do not generate values for \lsti{rs},
@@
 a typed arithmetic language featuring numerals with predecessor
-and Booleans with if-then-else and test-for-zero, which comes from
+and Booleans with if-then-else and test-for-zero, which comes from the \emph{Software Foundations} book series
 (\url{https://softwarefoundations.cis.upenn.edu/plf-current/Types.html}):
@@
 \end{lstlisting}
-The static and small step dynamic semantics are reported in appendix~\ref{ssec:types}.
+The completely standard static and small step dynamic semantics are
+reported in appendix~\ref{ssec:types}.
 
@@
 
-Another way to asses the fault detection capability of our PBT setup is via \emph{mutation analysis}~\cite{CavadaCM20}, whereby localized bugs are voluntarily inserted, with the view that they should caught (``killed'') by the testing suite.
-Following on  an exercise if the aforementioned chapter of SF, we modify the typing relation by adding the following (nonsensical) additional clause:
+Another way to asses the fault detection capability of a PBT setup is via \emph{mutation analysis}~\cite{CavadaCM20}, whereby localized bugs are voluntarily inserted, with the view that they should caught (``killed'') by a ``good enough'' testing suite.
+Following on  an exercise if the aforementioned chapter of SF, we modify the typing relation by adding the following (nonsensical)  clause:
 

### 2020-11-03 papers/types2020/pbt.tex

@@
 We have presented in a previous paper~\cite{blanco19ppdp} a
-proof-theoretical reconstruction of property-based testing~\cite{fink97sen} (PBT) of
-relational specifications adopting techniques from foundational proof
+proof-theoretical reconstruction of property-based testing (PBT) of
+relational specifications, adopting techniques from foundational proof
 certificates to account for several features of this testing
@@
 Of course, Coq already features \textsf{QuickChick}~\cite{QChick}
-(\url{https://softwarefoundations.cis.upenn.edu/qc-current}),  a
-sophisticated and well-supported PBT tool, based on a different
+(\url{https://softwarefoundations.cis.upenn.edu/qc-current}), which is
+a sophisticated and well-supported PBT tool, based on a different
 perspective: being a clone of Haskell's QuickCheck, it emphasizes
@@
 generators. While current research~\cite{LampropoulosPP18} aims to
-increase automation, it is fair to say that testing with ~\cite{QChick}, in particular of
-relational specifications, is still very labor intensive. We do not
-intend to compete with \textsf{QuickChick}, but we shall see that we
-can test immediately \lsti{Inductive} definitions that
-corresponds to pure Horn programs, without any decidability proof nor
-commitment to the random generation strategy, and related
-configuration of generators and shrinkers.
+increase automation, it is fair to say that testing with
+~\cite{QChick}, in particular of relational specifications, is still
+very labor intensive. We do not intend to compete with
+\textsf{QuickChick}, but we shall see that we can test immediately
+\lsti{Inductive} definitions that corresponds to pure Horn
+programs. We can do that without having to provide a decidability
+proof for those definitions. Furthermore, we are not committed to a
+fixed random generation strategy, which in gnereal requires additional
+work in the configuration of generators and shrinkers.
 
@@
   where the $A_i$ and $B$ are  predicates defined
-using Horn clause specifications, providing a counter-example
-consists of  negating the property, and \emph{searching} for a proof of
+using Horn clause specifications, a counter-example to this conjecture
+consists of a witness to the negated formula
 \[
@@
 $
- \exists x_1 \colon \tau_1\dots x_n \colon \tau_n [A_1\land\dots\land A_m].
+ \exists x_1 \colon \tau_1\dots x_n \colon \tau_n [A_1\land\dots\land A_m]
 $
@@
 % % specifications use.
-
 % Within the tactic
@@
 programming languages, where PBT has been used extensively and
-successfully~\cite{Klein12}. Here we will consider a far simple example,
+successfully~\cite{Klein12}. Here we will consider a far simpler example,
 a typed arithmetic language featuring numerals with predecessor
-and Booleans with if-then-else and test-for-zero, which comes from the \emph{Software Foundations} book series
-(\url{https://softwarefoundations.cis.upenn.edu/plf-current/Types.html}):
+and Booleans with if-then-else and test-for-zero, which comes from the \emph{Software Foundations} book series (\url{https://softwarefoundations.cis.upenn.edu/plf-current/Types.html}):
 \begin{lstlisting}
@@
  intros e t Ht.    
- elpi dep_pbt 2 height (Ht) (e) .  
+ elpi dep_pbt height 2 (Ht) (e) .  
 \end{lstlisting}
@@
 
-% LocalWords:  Bedwyr bipole elpi QuickCheck OCaml TODO coq xs nat
-% LocalWords:  tif ttrue tzero tsucc SUT
+% LocalWords:  Bedwyr bipole elpi QuickCheck OCaml TODO coq xs nat DM
+% LocalWords:  tif ttrue tzero tsucc SUT PBT SmallCheck QuickChick
+% LocalWords:  decidability shrinkers dep pbt fpc interp forall ness
+% LocalWords:  ord unfoldings pre relationally shrinker booleans typ
+% LocalWords:  tfalse tpred tiszero TBool TNat subexp SuccBool
+% LocalWords:  notstuck

### 2020-11-03 papers/types2020/conc.tex
first pass to sec 3

@@
 
-\todo{sum up}
+\todo{sum up contribution}
 
-\begin{metanote}
-  Decided against a related work section: most is already discussed in the intro, the rest can be mentioned in passing --am
-\end{metanote}
+% \begin{metanote}
+%   Decided against a related work section: most is already discussed in the intro, the rest can be mentioned in passing --am
+% \end{metanote}
 % \subsection{Related work}
@@
 
-\subsection{Future work}
-\label{ssec:fut}
-
+% \subsection{Future work}
+% \label{ssec:fut}
+\todo{finish future work}
+There are many avenues of developments for thisline of research.
 The code reported in Fig.~\ref{fig:augmented} is somewhat idealized in
@@
 
-\todo{Link to the repo? \url{https://github.com/proofcert/fpc-elpi}}
-
 Extending our tactic to handle computations (\texttt{simp}) and/or to
@@
 Extending the LJF kernel to handle defined atoms
+
+\todo{Link to the repo. Which one? \url{https://github.com/proofcert/fpc-elpi} is provate and full of stuff}
 %%% Local Variables:

### 2020-11-03 papers/types2020/elab.tex
first pass to sec 3

@@
 
+\todo{Mention that we did \textbf{not} use pairing as well for the check predicate in sec 3}
 If one is interested only in building Coq kernel proof structures, we

### 2020-11-03 papers/types2020/sec3.tex
first pass to sec 3

--- 2020-11-01 papers/types2020/sec3.tex
@@
 \section{Proof theory and proof certificates}
+\label{sec:three}
+  \newcommand{\XXi}{{\color{blue}{\Xi}}}
 
-% {\color{red} The goal in this section is to say a bit about focusing
-%   generally, and to give a bit of detail about FPC for Horn clauses.
-%   Initial draft of text taken from \cite{blanco19ppdp}.
-% }
-
-\begin{metanote}
-  (AM): Start with the logical reconstruction of \texttt{check}, to and then reverse engineer what we want/need to say about focusing and FPC.
+\todo{Anything here?}
+% \begin{metanote}
+%   (AM): Start with the logical reconstruction of \texttt{check}, to and then reverse engineer what we want/need to say about focusing and FPC.
 We \textbf{do} not talk about PBT yet, just logic programming (with proof terms).
 % In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
-% One way to proceed:
-  % \begin{itemize}
-  % \item One figure fig \ref{fig:interp} for Horn logic over Coq terms (no certificates) and its encoding with quotations 
-  % \item Then the logic behind the dependent kernel with certificates and proof terms
-  % \end{itemize}
-%
-Stuff in red is from old paper
-\end{metanote}
+% \end{metanote}
 
@@
 \label{ssec:focused}
-\todo{Adapt the discussion}
-\todo{grammar for formulae and proof terms \cite{Herbelin94,LengrandDM06}  or just the figure? Mention spine calculus}
+\todo{Change title? }
+\todo{Use same names in Fig and 3: A for Atom, L for ListArgs etc.}
 
-The proof search approach to encoding Horn clause computation
-results in the structuring of proofs with repeated
-switching between a \emph{goal-reduction} phase and a
-\emph{backchaining} phase~\cite{miller91apal}.
+% The proof search approach to encoding Horn clause computation
+% results in the structuring of proofs with repeated
+% switching between a \emph{goal-reduction} phase and a
+% \emph{backchaining} phase~\cite{miller91apal}.
 
 %
-{\color{red}
-The notion of \emph{focused proof systems}~\cite{andreoli92jlc,liang09tcs}
-generalizes this view of proof construction by identifying the
-following two phases. % of proof construction.
-\begin{enumerate}
-\item The \emph{negative}%
-% \footnote{The terminology of negative and positive phases is a bit
-%   unfortunate: historically, these terms do not refer to positive or negative
-%   subformula occurrences but rather to certain semantic models used in
-%   the study of linear logic~\cite{girard87tcs}.}
-%
-  phase corresponds to  goal-reduction: in this phase,
-  inference rules that involve \emph{don't-care-non\-de\-ter\-min\-ism} are
-  applied.
-%
-  As a result, there is no need to consider backtracking over choices
-  made in building this phase.
-
-\item The \emph{positive} phase corresponding to  backchaining: in this phase, inference rules that involve
-  \emph{don't-know-non\-de\-ter\-min\-ism} are applied: here,
-  inference rules need to be supplied with information in order to
-  ensure that a completed proof can be built.
-%
-  That information can be items such as which term is needed to
-  instantiate a universally quantified formula and which disjunct of a
-  disjunctive goal formula should be proved.
-\end{enumerate}
-%
-Thus, when building a proof tree (in the sequent calculus) from the
-conclusion to its leaves, the negative phase corresponds to a simple
-computation that needs no external information, while the positive
-phase may need such external information to be supplied.
-%
-In the literature, it is common to refer to a repository of such
-external information as either an \emph{oracle} or a \emph{proof
-certificate}.
-}
 
@@
 
-\todo{Figure now uses Dale kernel-v2 --- needs further abstraction and explanation, see appendix}
+% \todo{Figure now uses Dale kernel-v2 --- needs further abstraction and explanation, see appendix}
 \begin{footnotesize}
-\begin{figure}
-    \newcommand{\XXi}{{\color{blue}{\Xi}}}
-
-  \newcommand{\bc}[5]{#1;#2\vdash #3 : #4 \Downarrow #5}
+\begin{figure}[t]
+  %
+  \renewcommand{\Gamma}{}
+    \newcommand{\bc}[5]{#1#2\vdash #3 : #4 \Downarrow #5}
 \[
@@
 % \qquad 
-\infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
-{\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
+\infer{\XXi\Gamma\vdash A : s}
+{(A,s) \in \mathcal{A}
 }
+% \qquad
+% \infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
+% {\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
+% }
 \qquad
-\infer{\XXi;\Gamma\vdash t : A}
-      {\bc{\XXi}{\Gamma} t D A \quad D \in \Pscr\cup\Gamma}
+\infer{\XXi_1\Gamma\vdash k\ l : A}
+      {\bc{\XXi_2}{\Gamma} l D A \quad k : D \in \Pscr_A\cup\Gamma\qquad \unfoldExpert{\XXi_1}{\XXi_2}} 
 \]
@@
 \begin{figure}
-%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
+\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
 \lstinputlisting[linerange={check-end}]{code/dep-kernel-v2.mod}
@@
 
-Figure~\ref{fig:augmented} contains a simple proof system \dots  in which each inference rule is augmented with an
-additional premise involving an \emph{expert predicate}, a
-certificate $\Xi$, and possibly continuations of certificates ($\Xi'$,
-$\Xi_1$, $\Xi_2$) with extracted information from certificates (in the
-case of $\vee$ and $\exists$).
+Figure~\ref{fig:augmented} contains an implicitely focused proof system  for a small fragment of the Calculus of Inductive Constructions that corresponds to Horn logic programming over pure \lsti{Inductive} definitions.
+%
+It is a sequent calculus inspired by the calculus for
+proof search in Pure Type Systems introduced in \cite{LengrandDM06}, based in
+ turn on ideas stemming from focusing (in particular, the LJT calculus of
+Herbelin \cite{Herbelin94}). Similarly to that calculus, we have a term
+language that includes terms and lists of terms, and two distinct typing
+judgements for the two categories. In our case, where proof search steps are
+viewed as computation, the term judgement corresponds to a goal-reduction
+step while the list judgement corresponds to a backchaining step. This coincides
+with the idea behind the \emph{spine calculus}~\cite{Cervesato97tr}.
+
+Since we are only dealing with types that correspond to Horn clauses, and
+atomic types are inductively defined, we do not need a $\forall$ rule on the right, although the proof theory would gladly allow it. This means, at the term level, that we
+don't need to have variables in our grammar of terms. Terms are always
+applied to a (possibly empty) list of arguments, and as is usual in Pure Type
+Systems we have the same grammar for terms and types:
+%, as shown in figure~\ref{fig:terms}.
+\todo{Let's think about that: we need A at least, but A is also a metavariable for atoms}
+%\begin{figure}
+\begin{align*}
+t,u ::= & k\ l \;|\; \forall x : t.\, u \\
+l ::= & [] \;|\; (t\ ::\ l)
+\end{align*}
+% \caption{Grammar of terms}
+% \label{fig:terms}
+% \end{figure}
+
+\todo{This does not see quite right: D vs A, consider say even}
+The proof system is parametrized by  external set of inductive definitions available in
+the (implict) global context: we denote by $\Pscr_A$ the set of type declarations for the
+constructors of type $A$, of the form $k : D$ where $D$ is of the form
+$\forall x_1 : t_1.\, \dots \forall x_n : t_n.\, D\ u_1 \dots u_m$. We denote
+by $\mathcal{A}$ the set of arity declaration for the inductive types.
+
 %
 %
-%
-\todo{Explain relations with other term assignments.
-Note that we do not use explicit substitutions (we are cut free).
-Explain the choice of not having certificates represent proof terms.
-Explain that the unfold clause handles defined atoms and defined connectives (conjunction, disjunction, existential and also equality): this may be less efficient and proof-theoretically induce pointless phase changes, but it is arguably very elegant e simplifies proof terms that arand inherited by Coq }
+%done Explain relations with other term assignments. Note that we do not use
+% explicit substitutions (we are cut free)
+% Explain the choice of not having certificates represent proof terms.
+\todo{  Explain that the unfold clause handles
+defined atoms and defined connectives (conjunction, disjunction, existential
+and also equality): this may be less efficient and proof-theoretically induce
+pointless phase changes, but it is arguably very elegant e simplifies proof
+terms that arand inherited by Coq }
 
 Figure~\ref{fig:kernel} contains the Elpi implementation of the
-inference rules in Figure~\ref{fig:augmented}: here the infix
-turnstile $\vdash$ symbol is replaced by the \lsti{check} predicate.
-%
-% Notice that it is easy to show that no matter how the expert predicates
-% are defined, if the goal \lsti{check Cert B} is provable in \lP then
-% \lsti{B} must be a sound consequence of the program clauses stored in
-% \dots
-% the \lsti{prog} predicate (which provides a natural implementation of
-% the premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$.
-% AM: not there
-% in Figure~\ref{fig:kernel}).
+inference rules in Figure~\ref{fig:augmented}: $\XXi\vdash t \colon A$ corresponds to
+\lsti{check $\XXi$ (go A T)} and $\XXi\vdash l \colon D \Downarrow A$ corresponds to
+\lsti{check $\XXi$ (bc D A L)}.
 
-%\subsection{Certificates}
+% here the infix
+% turnstile $\vdash$ symbol is replaced by the \lsti{check} predicate, and every rule
+% is directly mapped to a clause in the code. The argument of the clause is
+% either given by a \lsti{go} or a \lsti{bc} constructor, signaling whether the clause
+% is a goal-reduction or a backchaining one.
 
-\todo{What happened to the unfold certificate???}
-% \todo{Show the dprolog tactic? have an example? could  be
-% \lsti{Goal  exists R, insert 2 [0;1] R.} and we would move the code for insert here}
-{\color{red}}
-The notion of proof certificates
- is taken from the general setting of \emph{foundational
+Coq terms are accessed through the Coq-Elpi API, and their representation
+in \lP takes advantage of the native \lP constructs such as lists and binders.
+The following is a list of some of signatures of the notions that we use:
+\begin{lstlisting}
+  kind term   type.   % reification of Coq terms
+  kind gref    type     % reification of refernces to global terms
+  type global gref -> term.
+  type indt   inductive -> gref.  % reification of inductive types
+  type indc   constructor -> gref. % reification of their constructors 
+  type app    list term -> term.  % reification of nary application
+  type prod   name -> term -> (term -> term) -> term.
+\end{lstlisting}
+
+Note that \lsti{prod} encodes dependent products taking a name for pretty printing,  a term and a \lP function from
+terms to terms, namely $\forall x : A, B$ is encoded by \lsti{prod ``x'' A (x\ B x)}; when, in the implementation of the product-left rule, we
+apply $B$ to the variable \lsti{Tm},
+we get a term that can be used to continue backchaining. This
+application is obtained via meta-level substitution, in the style of HOAS.
+In this sense, our calculus uses \emph{implict} substitutions, rather than explicit ones as in LJT and PTSC's tradition, the more since proof search in our application is cut-free.
+% Syntax: one can think as if something of type \lsti{term -> term} were a
+% \lsti{term} with a hole in it for another \lsti{term}.
+\todo{Now safe-dest-app is hiiden inside \texttt{definition}. Put the rest of the code in the appendix?}
+The unfolding rule makes use of the Coq-Elpi primitives
+\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
+nested) application, and \lsti{coq.env.indt} in order to access the global
+environment of inductive definitions and query for information about them.
+
+\smallskip
+In figure~\ref{fig:augmented}  each
+inference rule is augmented with an additional premise involving an
+\emph{expert predicate}, a certificate $\Xi$, and possibly continuations of
+certificates ($\Xi'$, $\Xi_1$, $\Xi_2$) with extracted information from
+These notions are taken from the general setting of \emph{foundational
   proof certificates}~\cite{chihani17jar}.
@@
 provide the remaining details not supplied in Figure~\ref{fig:kernel}:
-that is, the exact set of constructors for the \lsti{cert} type as
-well as the exact specification of the expert predicates listed in
-that figure.
+that is, the exact set of constructors for the  type of certificates as
+well as the specification of the expert predicates listed \emph{ibidem}.
 %
-Figure~\ref{fig:resources} displays two such FPCs,
+The top of Figure~\ref{fig:resources} displays two such FPCs,
 both of which can be used to describe proofs where we bound
-the number of occurrences of unfoldings in a proof.
+the dimension of  a proof.
 %
-For example, the first FPC
-provides the experts for treating certificates that are constructed
-using the \lsti{qheight} constructor.
-%
-As is easy to verify, the query \mbox{\lsti{(check (qheight 5) B)}} is
+For example, the first FPC dictates that the query \mbox{\lsti{(check (qheight 5) B)}} is
 provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
@@
 %
-Similarly, the second FPC uses the constructor \lsti{qsize} (with two
-integers) and can be used to bound the total number of instances of
+Similarly, the second FPC can be used to bound the total number of instances of
 unfoldings in a proof.
@@
 
-\begin{figure}
+\begin{figure}[t]
 \lstinputlisting[linerange={resources-end}]{code/fpcs.sig}
 \lstinputlisting[linerange={resources-end}]{code/fpcs.mod}
-\caption{Two FPCs that describe proofs that are limited in either
-  height or in size.}
+\caption{Sample FPCs}
 \label{fig:resources}
@@
 
+\todo{We could move here from elab the paragraph ``It is also possible to pair two different proof certificates, defined
+by two different FPC definitions,'' and/or reduce redundancy}
 Further, if we view a particular FPC as a means of restricting proofs, it is
@@
 
+\begin{metanote}
+  AM stops here
+\end{metanote}
 \subsection{A prolog-like tactic}

### 2020-11-04 papers/types2020/conc.tex
first reasonable draft of sec 3

@@
 \todo{finish future work}
-There are many avenues of developments for thisline of research.
+There are many avenues of developments for this line of research.
 The code reported in Fig.~\ref{fig:augmented} is somewhat idealized in
@@
 
-Extending our tactic to handle computations (\texttt{simp}) and/or to
+Extending our tactic to handle computations (\texttt{simpl}) and/or to
 return the partial proof state, unsolved goals in particular. Mention
@@
 
-\todo{Link to the repo. Which one? \url{https://github.com/proofcert/fpc-elpi} is provate and full of stuff}
+\todo{Link to the repo. Which one? \url{https://github.com/proofcert/fpc-elpi} is private and full of stuff}
 %%% Local Variables:
@@
 
-% LocalWords:  Elpi
+% LocalWords:  Elpi simp LJF

### 2020-11-04 papers/types2020/sec3.tex
first reasonable draft of sec 3

@@
 
-Figure~\ref{fig:augmented} contains an implicitely focused proof system  for a small fragment of the Calculus of Inductive Constructions that corresponds to Horn logic programming over pure \lsti{Inductive} definitions.
+Figure~\ref{fig:augmented} contains an implicitly focused proof system  for a small fragment of the Calculus of Inductive Constructions that corresponds to Horn logic programming over pure \lsti{Inductive} definitions.
 %
@@
 language that includes terms and lists of terms, and two distinct typing
-judgements for the two categories. In our case, where proof search steps are
-viewed as computation, the term judgement corresponds to a goal-reduction
-step while the list judgement corresponds to a backchaining step. This coincides
+judgments for the two categories. In our case, where proof search steps are
+viewed as computation, the term judgment corresponds to a goal-reduction
+step while the list judgment corresponds to a backchaining step. This coincides
 with the idea behind the \emph{spine calculus}~\cite{Cervesato97tr}.
+
+It may be at first surprising that there are no introduction rules for
+propositional connectives, nor equality for that matter.
+However, one of the beauty of the Calculus of Inductive Construction is that they are in fact defined inductively and therefore the \emph{unfold} rule will handle those; this also simplifies proof-terms.
 
@@
 atomic types are inductively defined, we do not need a $\forall$ rule on the right, although the proof theory would gladly allow it. This means, at the term level, that we
-don't need to have variables in our grammar of terms. Terms are always
+do not have variables in our grammar of terms. Terms are always
 applied to a (possibly empty) list of arguments, and as is usual in Pure Type
@@
 \todo{This does not see quite right: D vs A, consider say even}
-The proof system is parametrized by  external set of inductive definitions available in
-the (implict) global context: we denote by $\Pscr_A$ the set of type declarations for the
+The proof system is parameterized by  external set of inductive definitions available in
+the (implicit) global context: we denote by $\Pscr_A$ the set of type declarations for the
 constructors of type $A$, of the form $k : D$ where $D$ is of the form
@@
 % Explain the choice of not having certificates represent proof terms.
-\todo{  Explain that the unfold clause handles
-defined atoms and defined connectives (conjunction, disjunction, existential
-and also equality): this may be less efficient and proof-theoretically induce
-pointless phase changes, but it is arguably very elegant e simplifies proof
-terms that arand inherited by Coq }
+% \todo{  Explain that the unfold clause handles
+% defined atoms and defined connectives (conjunction, disjunction, existential
+% and also equality): this may be less efficient and proof-theoretically induce
+% pointless phase changes, but it is arguably very elegant e simplifies proof
+% terms that are inherited by Coq }
 
@@
 Note that \lsti{prod} encodes dependent products taking a name for pretty printing,  a term and a \lP function from
-terms to terms, namely $\forall x : A, B$ is encoded by \lsti{prod ``x'' A (x\ B x)}; when, in the implementation of the product-left rule, we
+terms to terms, namely $\forall x : A, B$ is encoded by \texttt{prod ``x'' A (x\ B x)};
+when, in the implementation of the product-left rule, we
 apply $B$ to the variable \lsti{Tm},
@@
 application is obtained via meta-level substitution, in the style of HOAS.
-In this sense, our calculus uses \emph{implict} substitutions, rather than explicit ones as in LJT and PTSC's tradition, the more since proof search in our application is cut-free.
+In this sense, our calculus uses \emph{implicit} substitutions, rather than explicit ones as in LJT and PTSC's tradition, the more since proof search in our application is cut-free.
 % Syntax: one can think as if something of type \lsti{term -> term} were a
 % \lsti{term} with a hole in it for another \lsti{term}.
-\todo{Now safe-dest-app is hiiden inside \texttt{definition}. Put the rest of the code in the appendix?}
+\todo{Now safe-dest-app is hidden inside \texttt{definition}. Put the rest of the code in the appendix?}
 The unfolding rule makes use of the Coq-Elpi primitives
@@
 
-\begin{metanote}
-  AM stops here
-\end{metanote}
 \subsection{A prolog-like tactic}
@@
 \begin{metanote}
-  Let's dump it here and see how it looks
+Keep this? Perhaps explain Ev = Term?
 \end{metanote}
 
+Thanks to the Coq-Elpi interface, in particular to the ``main''
+proocedure \lsti{solve}, we can package the \lP's code for the checker
+as a tactic that can be called as any other tactic in a Coq script.
+
 \begin{lstlisting}
+Elpi Tactic dprolog.
 Elpi Accumulate lp:{{
-  solve [str ''height'', int N] [goal _Ctx Ev G _Who] _OutGoals :-
+  solve [str ''height'', int N] [goal _  Ev G _] _ :-
     coq.say "Goal:" {coq.term->string G},
@@
 ...
+/* Other clauses for different fpc omitted */
 }}.
 \end{lstlisting}
-The following example show how we can do FPC-driven logic programming in Coq and return
-a Coq term
+The following example show how we can use the above tactic to do
+FPC-driven logic programming in Coq and return a Coq proof-term:
 \begin{lstlisting}
@@
 eexists.
-elpi  dprolog 10.
+elpi  dprolog height 10.
 Qed.
@@
 The \lsti{dprolog} tactic can be seen as a programmable version of
-Coq's \lsti{auto}: it currently lacks the latter capability of solving
-goals by \lsti{reflexivity} and the modularity capability of using
-\lsti{Hints}, but it is not compelled to use depth-first search, since it follows the dictate of the given FPC.
-For example the FPC can provide a \emph{trace}, which may be more customable that the one offered by the \lsti{Debug} option.
+Coq's \lsti{auto}: it currently lacks the latter's capability of
+solving goals by \lsti{reflexivity} as well as the modularity of using
+\lsti{Hints} as introduction rules; on the bright side, it is not
+restricted to depth-first search, since it follows the dictate of the
+given FPC. For example the FPC can provide a \emph{trace} that may be
+more customable that the one offered by the \lsti{auto}'s hard-wired
+\lsti{Debug} facility.
 %
@@
 % LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE
-% LocalWords:  Elpi modularity
+% LocalWords:  Elpi modularity Herbelin bc PTSC's dprolog

### 2020-11-04 papers/types2020/sec3.tex
updated todo after chat

@@
 
-\subsection{Focused proof systems with certificates}
+\subsection{Proof systems with certificates}
 \label{ssec:focused}
@@
 \infer{\XXi\Gamma\vdash A : s}
-{(A,s) \in \mathcal{A}
+{%(A,s) \in \mathcal{A}
+  \mbox{call to Coq?}
 }
@@
 \infer{\XXi_1\Gamma\vdash k\ l : A}
-      {\bc{\XXi_2}{\Gamma} l D A \quad k : D \in \Pscr_A\cup\Gamma\qquad \unfoldExpert{\XXi_1}{\XXi_2}} 
+      {\bc{\XXi_2}{\Gamma} l D A \quad k : D \in \Pscr_A\qquad \unfoldExpert{\XXi_1}{\XXi_2}} 
 \]
@@
 do not have variables in our grammar of terms. Terms are always
-applied to a (possibly empty) list of arguments, and as is usual in Pure Type
-Systems we have the same grammar for terms and types:
+applied to a (possibly empty) list of arguments% , and as is usual in Pure Type
+% Systems we have the same grammar for terms and types
+.
 %, as shown in figure~\ref{fig:terms}.
-\todo{Let's think about that: we need A at least, but A is also a metavariable for atoms}
-%\begin{figure}
-\begin{align*}
-t,u ::= & k\ l \;|\; \forall x : t.\, u \\
-l ::= & [] \;|\; (t\ ::\ l)
-\end{align*}
-% \caption{Grammar of terms}
-% \label{fig:terms}
-% \end{figure}
+% \todo{Let's think about that: we need A at least, but A is also a metavariable for atoms}
+% %\begin{figure}
+% \begin{align*}
+% t,u ::= & k\ l \;|\; \forall x : t.\, u \\
+% l ::= & [] \;|\; (t\ ::\ l)
+% \end{align*}
+% % \caption{Grammar of terms}
+% % \label{fig:terms}
+% % \end{figure}
 
-\todo{This does not see quite right: D vs A, consider say even}
+\todo{Lookup Coq rules for unfold rule}
 The proof system is parameterized by  external set of inductive definitions available in
@@
 % \lsti{term} with a hole in it for another \lsti{term}.
-\todo{Now safe-dest-app is hidden inside \texttt{definition}. Put the rest of the code in the appendix?}
+\todo{Reput safe-dest-app in code, a bit more details about the unfold case}
 The unfolding rule makes use of the Coq-Elpi primitives
@@
 
-\begin{metanote}
-Keep this? Perhaps explain Ev = Term?
-\end{metanote}
+\todo{explain Ev = Term as an Elpi thing}
 
@@
 ...
-/* Other clauses for different fpc omitted */
+(* Other clauses for different fpc omitted *)
 }}.

### 2020-11-04 papers/types2020/conc.tex
updated todo after chat
future work finished (aha, aha)

@@
 
+There are many avenues of developments for this line of research.
+The code reported in Fig.~\ref{fig:augmented} is somewhat idealized in
+so far that we omitted a rule performing weak-head reduction (allowing
+to handle directly existential goals) and more importantly one that uses
+ Elpi's \emph{delay} mechanism. % This a distinguished feature of Elpi, whose
+% usef has been advocated in ~\cite{DunchevCT16,coen19mscs}.
+The
+use of constraints is of course for data generation is well
+developed~\cite{FioravantiPS15} and we could try to leverage it to
+improve PBT using partially instantiated terms: for example generating
+say lists with at least two elements as \texttt{[1,2|Tail]}, without
+recurring to needed narrowing as in in
+LazySmallCheck~\cite{smallcheck}. On the more practical side, it would be worthwhile to investigate \emph{random} generation, following the ideas in~\cite{pltredexconstraintlogic,blanco19ppdp}.
+
+Another obvious limitation of our PBT tactic is that is does not 
+ handle any computation within an inductive relation.
+We plan to tackle this either at the Ltac level by co-routining it with simplification tactics and/or by adding rewriting features to our FPC system~\cite{ChihaniM16}.
+% Mention \emph{Harpoon}, Beluga's new tactic language?
+
+ Finally it makes sense to tie together the two threads of this paper
+ and provide a way checking and elaborating proof evidence for
+ intutionistic logic \emph{with} (inductively) defined atoms. This
+ can go as far as FPC for inductive proofs~\cite{blanco15wof}.
+
+ \noindent
+ The source files mentioned in this paper are available at \url{https://github.com/proofcert/fpc-elpi}.
  % \begin{metanote}
@@
 % \begin{itemize}
-% \item Brief discussion/comparison with \texttt{auto}
+% \item Brief discussion/comparison with \texttt{auto} [done]
 % \item Mention other papers by Enrico as applications of coq-elpi~\cite{coen19mscs,tassi19itp}\dots [done]
@@
 % \label{ssec:fut}
-\todo{finish future work}
-There are many avenues of developments for this line of research.
-The code reported in Fig.~\ref{fig:augmented} is somewhat idealized in
-so far that we omitted a rule performing weak-head reduction (allowing
-to handle directly existential goals) and more importantly that used
-the \emph{delay} mechanism. This a distinguished feature of Elpi, whose
-usefulness has been advocated in ~\cite{DunchevCT16,coen19mscs}.  The
-use of constraints for data generation is well
-developed~\cite{FioravantiPS15} and we could try to leverage to
-improve PBT using partially instantiated terms: for example generating
-say lists with at least two elements as \texttt{[1,2|Tail]}, without
-recurring to needed narrowing as in in
-LazySmallCheck~\cite{smallcheck}.
+%\todo{finish future work}
 
-Extending our tactic to handle computations (\texttt{simpl}) and/or to
-return the partial proof state, unsolved goals in particular. Mention
-\emph{Harpoon}, Beluga's new tactic language?
+ % \todo{Link to the repo. Use \url{https://github.com/proofcert/fpc-elpi}, make it public andmove papers to parsifal}
 
-Extending the LJF kernel to handle defined atoms
-
-\todo{Link to the repo. Which one? \url{https://github.com/proofcert/fpc-elpi} is private and full of stuff}
 %%% Local Variables:

### 2020-11-05 papers/types2020/cultures.tex
Revised sections 1 and 2, responding to Alberto's comments.  Moving now to sections 3 and 5.

--- 2020-11-03 papers/types2020/cultures.tex
@@
 usually decides that a proof is a certain kind of $\lambda$-term
-within the system.  In contrast, the proof theory treats logical
+within the system.  In contrast, proof theory treats logical
 propositions and proofs as separate.  For example, proof theory does
@@
 
-%% \begin{metanote}
-%%   This paragraph seem to identify proof and type theory with sequents
-%%   and nat ded and while I agree with the gist (although people such as
-%%   Frank would argue that sequents are a way of organizing proof search
-%%   for nat ded), we're not gonna make a lot of friends at types and
-%%   more importantly is a big hammer to nail the point that logic
-%%   programming is helpful with Coq --am
-%% \end{metanote}
-
-%% DM: I've tried to address your concerns by adding an additional
-%% paragraph and references (below).
-
-Gentzen's discovery that the key to treating classical and
-intuitionistic logics in the same proof system was the invention of
-the weakening and contraction structural rules on the right-side of
-sequents~\cite{gentzen35}.  This invention lead him away from natural
+Gentzen's discovered that the key to treating classical and
+intuitionistic logics in the same proof system was the weakening and
+contraction structural rules on the right-side of
+sequents~\cite{gentzen35}.  This discovery lead him away from natural
 deduction to multiple-conclusion sequent calculus, a very important
-kind of proof that has not been given a satisfactory representation in
-type theory.  This same innovation of Gentzen also opened the way to
-another key proof-theoretic discovery, that of linear logic
-\cite{girard87tcs}.  Similarly, sequent calculus provides an elegant
-presentation of linear logic while most treatments of linear logic in
-type theory have difficulty treating linear negation and the
-multiplicative disjunction $\lpar$ and its unit $\bot$.
-
-% DM I don't have a reference but I added more details.
-%\begin{metanote}
-%  `` partial fashion.'' Isn't that a bit vague? Some references? -am
-%\end{metanote}
+kind of proof system that has not been given a satisfactory
+representation in type theory.  This same innovation of Gentzen also
+opened the way to another key proof-theoretic discovery, that of
+linear logic \cite{girard87tcs}.  Similarly, sequent calculus provides
+an elegant presentation of linear logic while most treatments of
+linear logic in type theory have difficulty treating linear negation
+and the multiplicative disjunction $\lpar$ and its unit $\bot$.
 
 As is often observed, however, sequent calculus is too low-level to be
-used explicitly as capturing the ``essence of a proof''.  Recent
-developments in focused proof systems \cite{andreoli92jlc,liang09tcs},
-makes it possible to extract from sequent calculus not only natural
-deduction proofs \cite{pimentel16lsfa}, but also proof nets
+used explicitly as capturing the ``essence of a proof''.  Fortunately,
+the notion of \emph{focused proof systems}
+\cite{andreoli92jlc,liang09tcs}, makes it possible to collect and join
+the micro-level inference rules of the sequent calculus into large
+scale, synthetic inference rules.  As a result, using such proof
+systems, it is possible to extract from sequent calculus not only
+natural deduction proofs \cite{pimentel16lsfa}, but also proof nets
 \cite{chaudhuri08tcs}, and Herbrand-style expansion trees
@@
 section and used in our two example applications of Elpi with Coq.
-
-%% DM I no longer think this comment is useful for this paper, so I
-%% delete it instead of trying to fix it.
-
-%% Proof theory also has rather immediate and natural treatments for
-%% co-induction (as well as induction: see
-%% \cite{baelde12tocl,momigliano12jal}) whereas the treatment of
-%% coinduction within type theory remains more challenging and is well
-%% developed only for various syntactic conditions (see, for example,
-%% \cite{bertot08entcs}).
-%% \begin{metanote}
-%%   The criticism applies to Coq, less to type theory, see co-patterns --am
-%% \end{metanote}
 
@@
 $\beta$-reduction of natural deduction proof: that is, computation is
-modeled as \emph{proof normalization}.  On the other hand, the
-logic programming paradigm, following the notion of goal-directed
-proof search, models computation abstractly as a regimented search of
-cut-free sequent calculus proofs: that is, computation is modeled
-as \emph{proof search}.
+modeled as \emph{proof normalization}.  On the other hand, the logic
+programming paradigm, following the notion of goal-directed proof
+search, models computation abstractly as a regimented search for
+cut-free sequent calculus proofs: that is, computation is modeled as
+\emph{proof search}.
 
@@
 normalization setting.  In particular, Gentzen's \emph{eigenvariables}
-are a kind of a proof-level binder.  In the proof normalization
-setting, such eigenvariables are instantiated during proof
-normalization.  However, during the search for cut-free proofs,
-eigenvariables do not vary, and they are part of the syntax of terms
-and formulas.  As a result, they can be used in the treatment of
-bindings in data structures more generally.  Such eigenvariables can
-be used by the logic programming paradigm to provide a natural,
-elegant, and powerful computing method with bindings within syntax.
+are a kind of proof-level binder.  In the proof normalization setting,
+such eigenvariables are instantiated during proof normalization.
+However, during the search for cut-free proofs, eigenvariables do not
+vary, and they are part of the syntax of terms and formulas.  As a
+result, they can be used in the treatment of bindings in data
+structures more generally.  Such eigenvariables can be used by the
+logic programming paradigm to provide a natural, elegant, and powerful
+computing approach to bindings within syntax.
 
@@
 
-%% The following paragraph seems out-of-place so I delete it now.
-
-%% In this paper, we shall use two different proof systems.  The first
-%% logic is first-order intuitionistic logic.  Here, predicates
-%% are not inductively defined and quantification will be limited to
-%% range over first-order term.  Actually, we shall examine only the Horn
-%% clause fragment of that logic.
-%% %
-%% The second logic will allow for inductively defined predicates using a
-%% fixed point notation.   Given again our restriction to essentially
-%% Horn clause specifications, it is possible to view intuitionistic
-%% reasoning with such specification as if they are actually
-%% specification in $\mu$MALL, an extension of multiplicative additive
-%% linear logic with the least and greatest fixed point operators
-%% \cite{baelde12tocl,baelde07lpar,heath19jar}.
-
 \subsection{\lP vs Coq}
 
-Given that \lP and Coq both result from a mixing of the
-$\lambda$-calculus with logic, it is important to understand their
-differences.  The confusion around the term \emph{higher-order
-abstract syntax} (HOAS), for example, helps to illustrate their
-differences.  In the functional programming setting, including the Coq
-system, the HOAS approach leads to encoding binding structures within
-terms using functions.  Such encodings are surprisingly complicated
-since they allow for \emph{exotic terms} \cite{despeyroux95tlca} and
-for structures on which induction is not possible
-\cite{roeckl01fossacs}.  In the logic programming setting,
-particularly \lP, HOAS is well supported since bindings are allowed
-with terms ($\lambda$-bindings), in formulas (quantifiers), and in
-proofs (eigenvariables).  For this reason, the term
-\emph{$\lambda$-tree syntax} was introduced to name this particular
-take on HOAS \cite{miller19jar}.  The Abella proof assistant
-\cite{baelde14jfr} was designed, in part, to provide inductive and
-co-inductive inference involving specifications using the
-$\lambda$-tree syntax approach.
+Given that \lP and Coq both result from combining the
+$\lambda$-calculus with logic, it is important to understand some of
+their their differences.  The confusion around the term
+\emph{higher-order abstract syntax} (HOAS), is a case in point.  In
+the functional programming setting, including the Coq system, the HOAS
+approach leads to encoding binding structures within terms using
+functions.  The earliest such encodings were unsatisfactory since they
+would allow for \emph{exotic terms} \cite{despeyroux95tlca} and for
+structures on which induction was not possible \cite{roeckl01fossacs}.
+Later approaches yielded non-canonical and complex encodings, although
+these could support inductive and coinductive reasoning
+\cite{chlipala08icfp,honsell01tcs,pientka10ijcar}. 
+% DM I'm hoping that the additional sentence and references are enough
+% here. 
+%% \begin{metanote}
+%%   AM: I find this a bit unbalanced. Exotic terms can be avoided with
+%%   Honsell's theory of context and/or Adam's PHOAS -- not that I like
+%%   it.. And of course there is Beluga and friends.  On the other hand
+%%   Coq users are happy with library-based approaches such as Autosubt
+%%   or well-typed DB that allows for sphisticated treatments of
+%%   binders. I would say something more generic such as that mixing
+%%   functions and HOAS is not immediate, may require additional
+%%   machinery both in terms of type theories than of code generation,
+%%   while in lp\dots
+%% \end{metanote}
+In the logic programming setting, particularly \lP, HOAS is well
+supported since bindings are allowed with terms ($\lambda$-bindings),
+in formulas (quantifiers), and in proofs (eigenvariables).  (In fact,
+the original paper on HOAS \cite{pfenning88pldi} was inspired by \lP.)
+For this reason, the term \emph{$\lambda$-tree syntax} was introduced
+to name this particular take on HOAS \cite{miller19jar}.  The Abella
+proof assistant \cite{baelde14jfr} was designed, in part, to provide
+inductive and co-inductive inference involving specifications using
+the $\lambda$-tree syntax approach.
 
@@
 
-While \lP is a typed language, it only uses simple, polymorphic types,
-and these are more often used to indicate syntactic categories.  Any
-dependency information needs to be captured using predicates and
-quantifiers.
+While \lP is a typed language, it only uses ML-style polymorphism
+types, and these are more often used to indicate syntactic categories.
+Any dependency information needs to be captured using predicates and
+logical connectives and quantifiers.
+
+% LocalWords:  Elpi polymorphically

### 2020-11-05 papers/types2020/intro.tex
Revised sections 1 and 2, responding to Alberto's comments.  Moving now to sections 3 and 5.

@@
 implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
-more recently, Tassi has made Elpi available as the \texttt{coq-elpi} plugin to the Coq
-proof assistant \cite{tassi18coqpl} (\url{https://github.com/LPCIC/coq-elpi}).  This implementation of \lP
-extends earlier ones in primarily two directions:  First,
-Elpi adds a notion of constraints and constraints handling rules, which makes it more expressive than
-the Teyjus implementation \cite{nadathur99cade} of \lP. Second, the
-plugin version of Elpi comes equipped with  a quotation and anti-quotation syntax for mixing
-Coq expressions with \lP program elements and an API for accessing the
-Coq environment, typechecking included.
+more recently, Tassi has made Elpi available as the \texttt{coq-elpi}
+plugin to the Coq proof assistant \cite{tassi18coqpl}
+(\url{https://github.com/LPCIC/coq-elpi}).  This implementation of \lP
+extends earlier ones in primarily two directions: First, Elpi adds a
+notion of constraints and constraints handling rules, which makes it
+more expressive than the Teyjus implementation \cite{nadathur99cade}
+of \lP. Second, the plugin version of Elpi comes equipped with a
+quotation and anti-quotation syntax for mixing Coq expressions with
+\lP program elements and an API for accessing the Coq environment,
+type checking included.
 
@@
 
-Since relations (not functions) are central in \lP, Elpi is
-capable to providing direct support for the many relations that have a
-role in proof assistants implementation and usage.  Such relations include typing (e.g.,
-$\Gamma\vdash M\colon\sigma$), evaluation (e.g., natural semantics
-specifications \cite{kahn87stacs,hannan93jfp}), and interaction (e.g.,
-structured operational semantics \cite{plotkin81,milner89book}).
+Since relations (not functions) are central in \lP, Elpi is capable to
+providing direct support for the many relations that have a role in
+proof assistants implementation and usage.  Such relations include
+typing (e.g., $\Gamma\vdash M\colon\sigma$), evaluation (e.g., natural
+semantics specifications \cite{kahn87stacs,hannan93jfp}), and
+interaction (e.g., structured operational semantics
+\cite{milner89book,plotkin81}).
 
-Felty has also
-made the point that LCF-style tactics and tacticals can be given an
-elegant and natural specification using the higher-order relational
-specifications provided by \lP \cite{felty93jar}.  Some recent
-implementations built using Elpi
-support the usefulness of
-higher-order logic programming as a meta-programming language 
-for proof assistants in general~\cite{DunchevCT16,coen19mscs} and in particular
-for Coq via the \texttt{coq-elpi} plug in~\cite{tassi19itp,CohenST20} .
-%
+Felty has also made the point that LCF-style tactics and tacticals can
+be given an elegant and natural specification using the higher-order
+relational specifications provided by \lP \cite{felty93jar}.  Some
+recent implementations built using Elpi support the usefulness of
+higher-order logic programming as a meta-programming language for
+proof assistants in general~\cite{DunchevCT16,coen19mscs} and, in
+particular, for Coq via the \texttt{coq-elpi} plug
+in~\cite{CohenST20,tassi19itp} .
 
@@
 these examples, we shall illustrate that Elpi is useful not only
-because of its meta-programming features,
-% (e.g., a builtin treatment of bindings in syntax)
-but also because it soundly
-implements a higher-order intuitionistic logic.
-\todo{Explain better? Our kernels are deep embedding, aren't they? -am}The latter aspect of
-Elpi makes it possible to encode the proofs and proof theory of
-specific subsets of the Coq logic.
+because of its meta-programming features, but also because it soundly
+implements a higher-order intuitionistic logic: such implementations
+of higher-order logic have long been known to provide powerful and
+flexible approaches to implementing many different logics and their
+proof systems \cite{felty89phd,felty88cade,paulson89jar}.  Following
+that tradition, the Elpi system makes it possible to encode the proofs
+and proof theory of various subsets of the logic behind Coq (see also
+\cite{felty93lics}).
 
 While other meta-programming frameworks based on functional
-programming such as \emph{MetaCoq}~\cite{sozeau2020metacoq} can and have been used for related endeavors, we believe (together with~\cite{DunchevCT16}) that they would require much more boilerplate.
+programming such as \emph{MetaCoq}~\cite{sozeau2020metacoq} can and
+have been used for related endeavors, we believe (together
+with~\cite{DunchevCT16}) that they would require much more
+boilerplate code.
 

### 2020-11-05 papers/types2020/sec3.tex
proof system fixed

@@
   %
-  \renewcommand{\Gamma}{}
+%  \renewcommand{\Gamma}{}
     \newcommand{\bc}[5]{#1#2\vdash #3 : #4 \Downarrow #5}
@@
 % \qquad 
-\infer{\XXi\Gamma\vdash A : s}
+\infer{\XXi\vdash A : s}
 {%(A,s) \in \mathcal{A}
-  \mbox{call to Coq?}
+  {E[] \vdash_{CIC} A  : s}
 }
@@
 \qquad
-\infer{\XXi_1\Gamma\vdash k\ l : A}
-      {\bc{\XXi_2}{\Gamma} l D A \quad k : D \in \Pscr_A\qquad \unfoldExpert{\XXi_1}{\XXi_2}} 
+\infer{\XXi_1\vdash k\ l : A}
+      {\mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad (\mathit{head} \ A : T)\in \Gamma_I\quad  k : D \in \Gamma_C\quad \bc{\XXi_2}{} l D A \quad\unfoldExpert{\XXi_1}{\XXi_2}} 
 \]
@@
 \[
-  \infer{\bc {\XXi}{\Gamma} {[]} A A}{\trueExpert{\XXi}}{}\qquad
+  \infer{\bc {\XXi}{} {[]} A A}{\trueExpert{\XXi}}{}\qquad
 %  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
-  \infer{\bc{\XXi}{\Gamma}{(t\ ::\ l)} {\forall x:B.\, D} A}
-  {\XXi_1;\Gamma\vdash t : A\qquad \bc{\XXi_2}{\Gamma}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
+  \infer{\bc{\XXi}{}{(t\ ::\ l)} {\forall x:B.\, D} A}
+  {\XXi_1\vdash t : A\qquad \bc{\XXi_2}{}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
 \]
@@
 judgments for the two categories. In our case, where proof search steps are
-viewed as computation, the term judgment corresponds to a goal-reduction
-step while the list judgment corresponds to a backchaining step. This coincides
+viewed as computation, the term judgment $\XXi\vdash t \colon A$ corresponds to a goal-reduction
+step while the judgment $\XXi\vdash l \colon D \Downarrow A$ corresponds to a backchaining step. This coincides
 with the idea behind the \emph{spine calculus}~\cite{Cervesato97tr}.
 
-It may be at first surprising that there are no introduction rules for
-propositional connectives, nor equality for that matter.
-However, one of the beauty of the Calculus of Inductive Construction is that they are in fact defined inductively and therefore the \emph{unfold} rule will handle those; this also simplifies proof-terms.
+The proof system is parameterized by Coq' global environment $E$, which
+here can be thought to consist of a set of inductive definitions,
+which following Coq's reference manual we denote as
+$\mathrm{Ind}[p]  (\Gamma_I := \Gamma_C)$.  The first rule delegates
+to Coq's type checking enforcing the well-sortedness of such types. We
+do not have a \emph{local context} since we are only dealing with types
+that correspond to Horn clauses, and atomic types are inductively
+defined. In fact, we do not have a $\forall$ rule on the right,
+although the proof theory would gladly allow it. This means, at the
+term level, that we do not have variables in our grammar of
+terms. Terms are always applied to a (possibly empty) list of
+arguments.
 
-Since we are only dealing with types that correspond to Horn clauses, and
-atomic types are inductively defined, we do not need a $\forall$ rule on the right, although the proof theory would gladly allow it. This means, at the term level, that we
-do not have variables in our grammar of terms. Terms are always
-applied to a (possibly empty) list of arguments% , and as is usual in Pure Type
+The \emph{unfold} rule is the equivalent in our simplified setting of
+a \emph{decide} rule in a explicitly focused system. From a logic
+programming viewpoint given an atom, it selects a clause on which to
+back-chain on: we look-up the costructors from an \lsti{Inductive}
+definition from the global enviroment, one that matches the atom we
+atom we wish to bachchain on and then call the latter judgment. The rules for backchaining are standard.
+
+It may be at first surprising that there are no
+introduction rules for propositional connectives, nor equality for
+that matter. However, one of the beauty of the Calculus of Inductive
+Construction is that they are in fact defined inductively and
+therefore the \emph{unfold} rule will handle those; this also
+simplifies proof-terms.
+
+  % , and as is usual in Pure Type
 % Systems we have the same grammar for terms and types
-.
+%
 %, as shown in figure~\ref{fig:terms}.
@@
 
-\todo{Lookup Coq rules for unfold rule}
-The proof system is parameterized by  external set of inductive definitions available in
-the (implicit) global context: we denote by $\Pscr_A$ the set of type declarations for the
-constructors of type $A$, of the form $k : D$ where $D$ is of the form
-$\forall x_1 : t_1.\, \dots \forall x_n : t_n.\, D\ u_1 \dots u_m$. We denote
-by $\mathcal{A}$ the set of arity declaration for the inductive types.
+% we denote by $\Pscr_A$ the
+% set of type declarations for the constructors of type $A$, of the form
+% $k : D$ where $D$ is of the form
+% $\forall x_1 : t_1.\, \dots \forall x_n : t_n.\, D\ u_1 \dots u_m$. We
+% denote by $\mathcal{A}$ the set of arity declaration for the inductive
+% types.
 
@@
 % LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE
-% LocalWords:  Elpi modularity Herbelin bc PTSC's dprolog
+% LocalWords:  Elpi modularity Herbelin bc PTSC's dprolog CIC

### 2020-11-05 papers/types2020/cultures.tex

@@
 illustrated by considering how they are used in the specification of
-tactics.  The origin story for the ML functional programming
-language was that it was the meta-language for implementing the LCF
-suite of tactics and tacticals \cite{gordon79}.  To implement tactics,
-ML adopted not only novel features such as polymorphically typed
+tactics.  The origin story for the ML functional programming language
+was that it was the meta-language for implementing the LCF suite of
+tactics and tacticals \cite{gordon79}.  To implement tactics, ML
+adopted not only novel features such as polymorphically typed
 higher-order functional programming but also the non-functional
 mechanisms of failure and exception handling.  While \lP is also based
-on polymorphically typed higher-order relational programming, it also
-comes with a completely declarative version of failure and
-backtracking search.  Combining those features along with its support
-of unification (even in the presence of term-level bindings), \lP
-provides a rather different approach to the specification of tactics
-\cite{felty93jar}.
+on ML-style-polymorphically typed higher-order relational
+programming\footnote{It uses types to indicate syntactic categories, while
+  any dependency information needs to be captured using predicates and
+  logical connectives and quantifiers.}, it also comes with a
+completely declarative version of failure and backtracking search.
+Combining those features along with its support of unification (even
+in the presence of term-level bindings), \lP provides a rather
+different approach to the specification of tactics \cite{felty93jar}.
 
-While \lP is a typed language, it only uses ML-style polymorphism
-types, and these are more often used to indicate syntactic categories.
-Any dependency information needs to be captured using predicates and
-logical connectives and quantifiers.
+% While \lP is a typed language, it only uses ML-style polymorphic
+% types, and these are more often
 
 % LocalWords:  Elpi polymorphically
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:

### 2020-11-05 papers/types2020/app.tex
first pass to sec 3
Fixed the proof rules in Fig 1 again.  Changed the type setting of some code (might need more adjustments).

@@
 % formulas.
- \lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
-\lstinputlisting[linerange={interp-end}]{code/dep-kernel-v2.mod}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={interp-end}]{code/dep-kernel.sig}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={interp-end}]{code/dep-kernel-v2.mod}
 \caption{Encoding in coq-elpi of the vanilla meta-interpreter}

### 2020-11-05 papers/types2020/sec3.tex
Minor adjustments.
Fixed the proof rules in Fig 1 again.  Changed the type setting of some code (might need more adjustments).

@@
-%\newpage
+\newpage
 \section{Proof theory and proof certificates}
@@
 % \todo{Figure now uses Dale kernel-v2 --- needs further abstraction and explanation, see appendix}
-\begin{footnotesize}
+
+%\begin{footnotesize} % No effect on typesetting
 \begin{figure}[t]
@@
 %  \renewcommand{\Gamma}{}
-    \newcommand{\bc}[5]{#1#2\vdash #3 : #4 \Downarrow #5}
+    \newcommand{\bc}[5]{#1#2\Downarrow #3 : #4 \vdash #5}
 \[
@@
 % \qquad 
-\infer{\XXi\vdash A : s}
-{%(A,s) \in \mathcal{A}
-  {E[] \vdash_{CIC} A  : s}
-}
 % \qquad
@@
 % }
-\qquad
+%\qquad
 \infer{\XXi_1\vdash k\ l : A}
@@
 \[
+\infer{\XXi\vdash  s : A}
+{%(A,s) \in \mathcal{A}
+  {E[] \vdash_{CIC} s : A}% The order should be s : A, not A : s ?
+}
+\qquad
   \infer{\bc {\XXi}{} {[]} A A}{\trueExpert{\XXi}}{}\qquad
@@
   \infer{\bc{\XXi}{}{(t\ ::\ l)} {\forall x:B.\, D} A}
-  {\XXi_1\vdash t : A\qquad \bc{\XXi_2}{}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
+  {\XXi_1\vdash t : B\qquad \bc{\XXi_2}{}l {D[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
 \]
 %\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
-\caption{dep kernel.}
+\caption{Specification of the checking of a proof certificate and the
+  synthesis of a dependently typed $\lambda$-term.}
 \label{fig:augmented}
@@
 \begin{figure}
-\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
-\lstinputlisting[linerange={check-end}]{code/dep-kernel-v2.mod}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={sigs-end}]{code/dep-kernel.sig}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={check-end}]{code/dep-kernel-v2.mod}
 \caption{Implementation of a  dependent kernel for Coq terms.}
@@
 \end{figure}
-\end{footnotesize}
+%\end{footnotesize}
 
@@
 \begin{figure}[t]
-\lstinputlisting[linerange={resources-end}]{code/fpcs.sig}
-\lstinputlisting[linerange={resources-end}]{code/fpcs.mod}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={resources-end}]{code/fpcs.sig}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={resources-end}]{code/fpcs.mod}
 \caption{Sample FPCs}

### 2020-11-05 papers/types2020/sec3.tex
some typos fixed in 1 2 and 3

@@
 
-\todo{Anything here?}
+\todo{Right now there is no intro to FPC}
 % \begin{metanote}
@@
 \todo{Change title? }
-\todo{Use same names in Fig and 3: A for Atom, L for ListArgs etc.}
 
@@
 viewed as computation, the term judgment $\XXi\vdash t \colon A$ corresponds to a goal-reduction
-step while the judgment $\XXi\vdash l \colon D \Downarrow A$ corresponds to a backchaining step. This coincides
+step while the judgment $\XXi \Downarrow  l \colon D\vdash A$ corresponds to a backchaining step. This coincides
 with the idea behind the \emph{spine calculus}~\cite{Cervesato97tr}.
@@
 here can be thought to consist of a set of inductive definitions,
-which following Coq's reference manual we denote as
-$\mathrm{Ind}[p]  (\Gamma_I := \Gamma_C)$.  The first rule delegates
+which following Coq's reference manual we denote with
+$\mathrm{Ind}[p]  (\Gamma_I := \Gamma_C)$.  One  rule delegates
 to Coq's type checking enforcing the well-sortedness of such types. We
@@
 The \emph{unfold} rule is the equivalent in our simplified setting of
-a \emph{decide} rule in a explicitly focused system. From a logic
-programming viewpoint given an atom, it selects a clause on which to
+a \emph{decide} rule in an explicitly focused system. From a logic
+programming viewpoint, given an atom, the rule selects a clause on which to
 back-chain on: we look-up the costructors from an \lsti{Inductive}
-definition from the global enviroment, one that matches the atom we
-atom we wish to bachchain on and then call the latter judgment. The rules for backchaining are standard.
+definition from the global enviroment, one that matches the atom we wish to bachchain on and then call the latter judgment.
+The rules for backchaining are standard, corresponding to the \emph{init} and $\forall$-left rules.
 
@@
 therefore the \emph{unfold} rule will handle those; this also
-simplifies proof-terms.
+simplifies the syntax of proof-terms.
 
@@
 inference rules in Figure~\ref{fig:augmented}: $\XXi\vdash t \colon A$ corresponds to
-\lsti{check $\XXi$ (go A T)} and $\XXi\vdash l \colon D \Downarrow A$ corresponds to
+\lsti{check $\XXi$ (go A T)} and $\XXi \Downarrow  l \colon D\vdash A$ corresponds to
 \lsti{check $\XXi$ (bc D A L)}.
@@
   kind gref    type                                 % reification of refernces to global terms
-  type global gref -> term.
+  type global gref -> term.                      % coercion to term
   type indt   inductive -> gref.                 % reification of inductive types
@@
   type app    list term -> term.               % reification of nary application
-  type prod   name -> term -> (term -> term) -> term.
+  type prod   name -> term -> (term -> term) -> term. % reification of dependent product
 \end{lstlisting}
@@
 Note that \lsti{prod} encodes dependent products taking a name for pretty printing,  a term and a \lP function from
-terms to terms, namely $\forall x : A, B$ is encoded by \texttt{prod ``x'' A (x\ B x)};
+terms to terms: namely $\forall x : A, B$ is encoded by \texttt{prod ``x'' A (x\ B x)};
 when, in the implementation of the product-left rule, we
@@
 
-\smallskip
-In figure~\ref{fig:augmented}  each
+In Figure~\ref{fig:augmented}  each
 inference rule is augmented with an additional premise involving an
@@
 In our case here, an FPC is a collection of \lP clauses that
-provide the remaining details not supplied in Figure~\ref{fig:kernel}:
+provide the remaining details not supplied in Figure~\ref{fig:augmented} and~\ref{fig:kernel}:
 that is, the exact set of constructors for the  type of certificates as
@@
 %
-The top of Figure~\ref{fig:resources} displays two such FPCs,
+The top of Figure~\ref{fig:resources} displays two FPCs,
 both of which can be used to describe proofs where we bound
@@
 restricted to depth-first search, since it follows the dictate of the
-given FPC. For example the FPC can provide a \emph{trace} that may be
-more customable that the one offered by the \lsti{auto}'s hard-wired
-\lsti{Debug} facility.
+given FPC, for example iterative-deepening. Furthermore, the FPC can
+provide a \emph{trace} that may be more customable that the one
+offered by the \lsti{auto}'s hard-wired \lsti{Debug} facility.
 %

### 2020-11-05 papers/types2020/conc.tex
proof system fixed

@@
 The
-use of constraints is of course for data generation is well
+use of constraints for data generation is well
 developed~\cite{FioravantiPS15} and we could try to leverage it to
-improve PBT using partially instantiated terms: for example generating
+improve our PBT tactic so that it can use partially instantiated terms: for example generating
 say lists with at least two elements as \texttt{[1,2|Tail]}, without
-recurring to needed narrowing as in in
+recurring to needed narrowing as in 
 LazySmallCheck~\cite{smallcheck}. On the more practical side, it would be worthwhile to investigate \emph{random} generation, following the ideas in~\cite{pltredexconstraintlogic,blanco19ppdp}.
 
-Another obvious limitation of our PBT tactic is that is does not 
- handle any computation within an inductive relation.
+Another obvious limitation of our PBT tactic is that it does not 
+ handle any computation (conversion) within an inductive relation.
 We plan to tackle this either at the Ltac level by co-routining it with simplification tactics and/or by adding rewriting features to our FPC system~\cite{ChihaniM16}.
@@
  Finally it makes sense to tie together the two threads of this paper
- and provide a way checking and elaborating proof evidence for
+ and provide a way of checking and elaborating proof evidence for
  intutionistic logic \emph{with} (inductively) defined atoms. This
@@
 
- \noindent
+ \smallskip
  The source files mentioned in this paper are available at \url{https://github.com/proofcert/fpc-elpi}.

### 2020-11-05 papers/types2020/sec3.tex
Some thinking about a good introduction to Section 3... Commiting to resolve a merge.
Minor updates.  Will return tomorrow morning.
Fix typos and a bit more text on implementation

--- 2020-11-05 papers/types2020/sec3.tex
@@
 % \emph{backchaining} phase~\cite{miller91apal}.
-
-%
 
@@
 in \lP takes advantage of the native \lP constructs such as lists and binders.
-The following is a list of some of signatures of the notions that we use:
+The following is a list of some of the signatures of the notions that we use:
 \begin{lstlisting}
@@
 
-Note that \lsti{prod} encodes dependent products taking a name for pretty printing,  a term and a \lP function from
-terms to terms, namely $\forall x : A, B$ is encoded by \texttt{prod ``x'' A (x\ B x)};
-when, in the implementation of the product-left rule, we
-apply $B$ to the variable \lsti{Tm},
-we get a term that can be used to continue backchaining. This
-application is obtained via meta-level substitution, in the style of HOAS.
-In this sense, our calculus uses \emph{implicit} substitutions, rather than explicit ones as in LJT and PTSC's tradition, the more since proof search in our application is cut-free.
+Note that \lsti{prod} encodes dependent products taking a name for pretty
+printing, a term and a \lP function from terms to terms, namely $\forall x :
+A, B$ is encoded by \texttt{prod ``x'' A (x\textbackslash{} B x)}; when, in the
+implementation of the product-left rule, we apply $B$ to the variable
+\lsti{Tm}, we get a term that can be used to continue backchaining. This
+application is obtained via meta-level substitution, in the style of HOAS. In
+this sense, our calculus uses \emph{implicit} substitutions, rather than
+explicit ones as in LJT and PTSC's tradition, the more since proof search in
+our application is cut-free.
 % Syntax: one can think as if something of type \lsti{term -> term} were a
 % \lsti{term} with a hole in it for another \lsti{term}.
-\todo{Reput safe-dest-app in code, a bit more details about the unfold case}
 The unfolding rule makes use of the Coq-Elpi primitives
@@
 environment of inductive definitions and query for information about them.
+The \lsti{unfold_expert} predicate, then, has the role of selecting which
+constructor to introduce for the inductive type. The kernel will successively
+obtain the type of the selected constructor, and initiate the backchaining
+phase on it.
 
@@
 %
+\todo{MM: Talk about randomized FPC?}
 In our case here, an FPC is a collection of \lP clauses that

### 2020-11-05 papers/types2020/cultures.tex

@@
 would allow for \emph{exotic terms} \cite{despeyroux95tlca} and for
-structures on which induction was not possible \cite{roeckl01fossacs}.
-Later approaches yielded non-canonical and complex encodings, although
-these could support inductive and coinductive reasoning
-\cite{chlipala08icfp,honsell01tcs,pientka10ijcar}. 
+structures on which induction was not immediately possible \cite{roeckl01fossacs}.
+Later approaches yielded non-canonical and complex encodings~\cite{chlipala08icfp,honsell01tcs}, as well as sophisticated type theories~\cite{pientka10ijcar}. All of
+these could support inductive and coinductive reasoning. 
 % DM I'm hoping that the additional sentence and references are enough

### 2020-11-05 papers/types2020/intro.tex

@@
 implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
-more recently, Tassi has made Elpi available as the \texttt{coq-elpi}
-plugin to the Coq proof assistant \cite{tassi18coqpl}
-(\url{https://github.com/LPCIC/coq-elpi}).  This implementation of \lP
-extends earlier ones in primarily two directions: First, Elpi adds a
-notion of constraints and constraints handling rules, which makes it
-more expressive than the Teyjus implementation \cite{nadathur99cade}
-of \lP. Second, the plugin version of Elpi comes equipped with a
+more recently, Tassi has made Elpi available as the Coq-Elpi
+plugin~\cite{tassi18coqpl} (\url{https://github.com/LPCIC/coq-elpi})
+to the Coq proof assistant.  This implementation of \lP extends
+earlier ones in primarily two directions: First, Elpi adds a notion of
+constraints and constraints handling rules, which makes it more
+expressive than the Teyjus implementation \cite{nadathur99cade} of
+\lP. Second, the plugin version of Elpi comes equipped with a
 quotation and anti-quotation syntax for mixing Coq expressions with
@@
 proof assistants in general~\cite{DunchevCT16,coen19mscs} and, in
-particular, for Coq via the \texttt{coq-elpi} plug
-in~\cite{CohenST20,tassi19itp} .
+particular, for Coq via the Coq-Elpi plug~\cite{CohenST20,tassi19itp} .
 

### 2020-11-05 papers/types2020/sec3.tex

--- 2020-11-05 papers/types2020/sec3.tex
@@
 \label{sec:three}
+
 \newcommand{\XXi}{{\color{blue}{\Xi}}}
+\newcommand{\bc}[5]{#1#2\Downarrow #3 : #4 \vdash #5}
 
@@
 
-%
-
 %\subsection{Certificate checking with expert predicates}
@@
 
-%\begin{footnotesize} % No effect on typesetting
 \begin{figure}[t]
-  %
-%  \renewcommand{\Gamma}{}
-    \newcommand{\bc}[5]{#1#2\Downarrow #3 : #4 \vdash #5}
 \[
-% \infer{\XXi\vdash t = t}
-% {%\eqExpert{\XXi}
-% }
-% \qquad
-% \infer{\XXi\vdash \true}
-% {%\trueExpert{\XXi}
-% }
-% \qquad
-% \infer{\XXi\vdash G_1\vee G_2}
-% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
-% }
-% \qquad
-% \infer{\XXi\vdash \exists x. G}
-% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
-% }
-% \qquad 
-% \qquad
-% \infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
-% {\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
-% }
-%\qquad
 \infer{\XXi_1\vdash k\ l : A}
-      {\mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad (\mathit{head} \ A : T)\in \Gamma_I\quad  k : D \in \Gamma_C\quad \bc{\XXi_2}{} l D A \quad\unfoldExpert{\XXi_1}{\XXi_2}} 
+      { \mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad 
+       (\mathit{head} \ A : T)\in \Gamma_I\quad  
+       k : D \in \Gamma_C\quad 
+       \bc{\XXi_2}{} l D A \quad\unfoldExpert{\XXi_1}{\XXi_2}} 
 \]
@@
 \[
-\infer{\XXi\vdash  s : A}
-{%(A,s) \in \mathcal{A}
-  {E[] \vdash_{CIC} s : A}% The order should be s : A, not A : s ?
-}
+  \infer{\XXi\vdash  s : A}{{E[] \vdash_{CIC} s : A}}% The order should be s : A, not A : s ?
   \qquad
-  \infer{\bc {\XXi}{} {[]} A A}{\trueExpert{\XXi}}{}\qquad
-%  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
-  \infer{\bc{\XXi}{}{(t\ ::\ l)} {\forall x:B.\, D} A}
-  {\XXi_1\vdash t : B\qquad \bc{\XXi_2}{}l {D[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
+  \infer{\bc{\XXi}{}{[]} A A}{\trueExpert{\XXi}}
+  \qquad
+  \infer{\bc{\XXi}{}{(t\ ::\ l)}{\forall x:B.\, D} A}
+        {\XXi_1\vdash t : B\qquad
+         \bc{\XXi_2}{}l {D[t/x]} A\qquad
+         \prodExpert{\XXi}{\XXi_1}{\XXi_2}}
 \]
-%\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
 \caption{Specification of the checking of a proof certificate and the
@@
 \end{figure}
+
+\[
+  \infer{\Pscr\vdash A}{D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
+  \qquad
+  \infer{\Pscr\Downarrow A\vdash A}{}
+  \qquad
+  \infer{\Pscr\Downarrow\forall x. B(x)\supset D(x)\vdash A}
+        {\vdash B(t)\qquad \Pscr\Downarrow D(t)\vdash A}
+\]
+Here, $\Pscr$ is a collection of Horn clauses, i.e., formulas of the
+form $\forall \bar x_1. A_1\supset \forall \bar x_2. A_2\supset
+\forall \bar x_n. A_n\supset A_0$.
+
 \begin{figure}
@@
 \end{figure}
-%\end{footnotesize}
 
@@
 
-Note that \lsti{prod} encodes dependent products taking a name for pretty printing,  a term and a \lP function from
-terms to terms: namely $\forall x : A, B$ is encoded by \texttt{prod ``x'' A (x\ B x)};
-when, in the implementation of the product-left rule, we
-apply $B$ to the variable \lsti{Tm},
-we get a term that can be used to continue backchaining. This
-application is obtained via meta-level substitution, in the style of HOAS.
-In this sense, our calculus uses \emph{implicit} substitutions, rather than explicit ones as in LJT and PTSC's tradition, the more since proof search in our application is cut-free.
+Note that \lsti{prod} encodes dependent products taking a name for pretty
+printing, a term and a \lP function from terms to terms: namely $\forall x :
+A, B$ is encoded by \texttt{prod ``x'' A (x\textbackslash{} B x)}; when, in the
+implementation of the product-left rule, we apply $B$ to the variable
+\lsti{Tm}, we get a term that can be used to continue backchaining. This
+application is obtained via meta-level substitution, in the style of HOAS. In
+this sense, our calculus uses \emph{implicit} substitutions, rather than
+explicit ones as in LJT and PTSC's tradition, the more since proof search in
+our application is cut-free.
 % Syntax: one can think as if something of type \lsti{term -> term} were a
 % \lsti{term} with a hole in it for another \lsti{term}.
-\todo{Reput safe-dest-app in code, a bit more details about the unfold case}
 The unfolding rule makes use of the Coq-Elpi primitives
@@
 environment of inductive definitions and query for information about them.
+The \lsti{unfold_expert} predicate, then, has the role of selecting which
+constructor to introduce for the inductive type. The kernel will successively
+obtain the type of the selected constructor, and initiate the backchaining
+phase on it.
 
-In Figure~\ref{fig:augmented}  each
+In figure~\ref{fig:augmented}  each
 inference rule is augmented with an additional premise involving an

### 2020-11-05 papers/types2020/sec3.tex
Explain prolog tactic

@@
 
-\todo{explain Ev = Term as an Elpi thing}
-
 Thanks to the Coq-Elpi interface, in particular to the ``main''
@@
 \end{lstlisting}
-The following example show how we can use the above tactic to do
+
+The glue code between Coq-Elpi and the implementation of our calculus is
+straightforward: the goal is given to us as a quadruple of a context, an
+evar, a type and a list of extra information. In additon, the certificate is
+supplied by the user when calling the tactic (it simply is an integer in this
+case). We just need to call \lsti{check} on the goal's type, together with
+the certificate, in order to obtain a reconstructed proof term. We avoid
+calling the reconstruction directly on the evar because Coq-Elpi ensures that
+evars manipulated by \lP are well-typed at all times, and we cannot guarantee
+that since we work with partially reconstructed term; getting around this is
+simply a matter of assigning the reconstructed term to the evar right
+afterwards.
+The following example shows how we can use the above tactic to do
 FPC-driven logic programming in Coq and return a Coq proof-term:

### 2020-11-06 papers/types2020/pbt.tex
small cleanup of sec 5

@@
 We have presented in a previous paper~\cite{blanco19ppdp} a
-proof-theoretical reconstruction of property-based testing (PBT) of
+proof-theoretical reconstruction of property-based testing (PBT) \cite{fink97sen}of
 relational specifications, adopting techniques from foundational proof
@@
 
-Given the connection that \textsf{coq-elpi} offers between logic
+Given the connection that Coq-Elpi offers between logic
 programming and the internals of Coq, it is natural to extend the
@@
 
-While nothing prevents us from porting all the features we discussed
-in~\cite{blanco19ppdp}, for the sake of this paper we will implement
-only FPC corresponding to different flavors \emph{exhaustive}
-generation, as adopted, e.g., in SmallCheck~\cite{smallcheck} and
-$\alpha$Check~\cite{cheney_momigliano_2017}, and their combination.
-Note however that it would take additional two lines of code to the \lsti{unfold} expert to implement a form of random data generation in the sense of randomized backtracking~\cite{pltredexconstraintlogic}.
+While nothing prevents us from porting all the PBT features we have
+accounted for in~\cite{blanco19ppdp}, for the sake of this paper we
+will implement only FPC corresponding to different flavors
+\emph{exhaustive} generation, as adopted, e.g., in
+SmallCheck~\cite{smallcheck} and
+$\alpha$Check~\cite{cheney_momigliano_2017}, and their
+combination. Note however that it would take  no more than two lines of
+code in the \lsti{unfoldE} expert to implement a form of random data
+generation in the sense of randomized
+backtracking~\cite{pltredexconstraintlogic}.
 
@@
 increase automation, it is fair to say that testing with
-~\cite{QChick}, in particular of relational specifications, is still
-very labor intensive. We do not intend to compete with
-\textsf{QuickChick}, but we shall see that we can test immediately
+\textsf{QuickChick}, in particular of relational specifications, is
+still very labor intensive. We do not intend to compete with
+\textsf{QuickChick} at this stage, but we shall see that we can test immediately
 \lsti{Inductive} definitions that corresponds to pure Horn
@@
 proof for those definitions. Furthermore, we are not committed to a
-fixed random generation strategy, which in gnereal requires additional
+fixed random generation strategy, which in general requires additional
 work in the configuration of generators and shrinkers.
 
-\todo{Drop the subsections?}
 \subsection{PBT as proof reconstruction}
@@
 using Horn clause specifications, a counter-example to this conjecture
-consists of a witness to the negated formula
+consists of a witness of the negated formula
 \[
@@
 
-In our Coq current setting $A_i$ and $B$ will be \lsti{Inductive}
+In our Coq  setting $A_i$ and $B$ will be \lsti{Inductive}
 propositions, while the $\tau_j$ are honest-to-goodness datatypes. We
 will treat the two quite differently, in so far as elements of those
-types will be \emph{generated}, while predicates will only be
+datatypes will be \emph{generated}, while predicates will only be
 \emph{checked}. This distinction is reminiscent of bi-directional
@@
 points~\cite{baelde12tocl}, negation corresponds to the usual
-intutionistic interpretation, which is  what Coq supports.  % proofs of formulas such as
+intuitionistic interpretation, which is  what Coq supports.  % proofs of formulas such as
 % \[
@@
 % %
-
+%
 % Instead of giving a full focused proof system of a logic including
@@
 % Within the tactic
-the user therefore
+the user 
 specifies which variables of the environment should be used for
@@
 against data \lsti{x,xs} generated exhaustively up to height
-at most $5$  out of the library \lsti{Inductive} definitions of
+at most $5$  form the library \lsti{Inductive} definitions of
 \lsti{nat} and \lsti{list}. We do not generate values for \lsti{rs},
@@
 successfully~\cite{Klein12}. Here we will consider a far simpler example,
-a typed arithmetic language featuring numerals with predecessor
-and Booleans with if-then-else and test-for-zero, which comes from the \emph{Software Foundations} book series (\url{https://softwarefoundations.cis.upenn.edu/plf-current/Types.html}):
+a typed arithmetic language featuring numerals with predecessor  and test-for-zero, 
+and Booleans with if-then-else, which comes from the \emph{Software Foundations} book series (\url{https://softwarefoundations.cis.upenn.edu/plf-current/Types.html}):
 \begin{lstlisting}
@@
  elpi dep_pbt height 2 (HS /\ HT) (e).
+ Abort.
 \end{lstlisting}
@@
 Another way to asses the fault detection capability of a PBT setup is via \emph{mutation analysis}~\cite{CavadaCM20}, whereby localized bugs are voluntarily inserted, with the view that they should caught (``killed'') by a ``good enough'' testing suite.
-Following on  an exercise if the aforementioned chapter of SF, we modify the typing relation by adding the following (nonsensical)  clause:
+Following on  an exercise in the aforementioned chapter of SF, we modify the typing relation by adding the following (nonsensical)  clause:
 
@@
  elpi dep_pbt height 2 (Ht) (e) .
+ Abort.
 \end{lstlisting}
@@
 % LocalWords:  tfalse tpred tiszero TBool TNat subexp SuccBool
-% LocalWords:  notstuck
+% LocalWords:  notstuck unfoldE

### 2020-11-06 papers/types2020/elab.tex
Minor adjustments.
Refactored sections a bit.  There is still a lot of inconsistency in the names of experts, etc.

--- 2020-11-03 papers/types2020/elab.tex
@@
-%\newpage
 \section{Elaboration of external proof certificates for the Coq
   kernel}
-
-% Follow the text in \cite{blanco20coq}.
 
@@
 that type is, in fact, accepted by Coq users as a theorem of
-intuitionistic logic.
-\begin{metanote}
-  Small point: the Coq kernel can be soundly extended with axioms and will also validate classical tautologies
-\end{metanote}
-The rest of the Coq system, especially its
-tactic language, is designed to help a human user build
-proofs-as-$\lambda$-terms that can be checked by the kernel.
+intuitionistic logic along with any axioms that have been adopted.
+The rest of the Coq system, especially its tactic language, is
+designed to help a human user build proofs-as-$\lambda$-terms that can
+be checked by the kernel.
 
@@
 Following the general outline that has been described in
-\cite{blanco17cade}, we assume that the following step are taken.
+\cite{blanco17cade,blanco20coq}, we assume that the following step are
+taken.
 %
@@
 synthesis of such proof structures directly into the FPC proof
-checker.  If one is interested in checking only one kind of external
-proof structure, then the FPC for that structure could also be built
-into the checker (via, say, partial evaluation of logic programs
-\cite{lloyd91jlp}).
+checker.  (We took exactly this step in Figure~\ref{fig:augmented}
+where proof checking involved both proof certificates as well as
+Coq-term structures.)  If one is interested in checking only one kind
+of external proof structure, then the FPC for that structure could
+also be built into the checker (via, say, partial evaluation of logic
+programs \cite{lloyd91jlp}).
 
-To illustrate this approach to elaborating externally provided proof
-structures into terms the Coq kernel can check, we consider here
-examples take from the file \verb+coq_fpc.v+, which is part of the
-repository \url{https://github.com/proofcert/fpc-elpi/}.
-Figure~\ref{fig:debruijn} presents the specific FPC that defines the
-proof semantics of using the De Bruijn notions for untyped
-$\lambda$-terms as proof evidence of a proof in intuitionistic
-propositional logic.  That FPC definition, along with a proof checker
-capable of synthesizing Coq proof terms during proof checker, makes it
-possible to use the term denoting the untyped $\lambda$-term
-$\lambda x (x (\lambda y (y (\lambda z (x (\lambda u\,z))))))$.
+As an example, consider an external theorem prover for propositional
+intuitionistic logic which return proof structures that are De Bruijn
+expressions for the underlying untype $\lambda$-proof that encodes
+(part of) the proofs it finds.  Consider the De Buijn expression
+example at the end of Section~\ref{ssec:beyond}.  This can be used as
+evidence that the following is a theorem of Coq.
 
-\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
-(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 
-                                                    [lambda (apply 1 [])])])]))
-\end{lstlisting}
-as a proof of following theorem of Coq.
 \begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
@@
 
-\begin{figure}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.sig}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.mod}
-\caption{The FPC definition of De Bruijn notation as proof evidence.}
-\label{fig:debruijn}
-\end{figure}
+Of course, the proof information in the untyped term must be
+elaborated to be a fully typed and functional expression.  Proof
+checking using the FPC in Figure~\ref{fig:debruijn} and (an extension
+of) the proof Figure~\ref{fig:kernel}.
 

### 2020-11-06 papers/types2020/sec3.tex
Refactored sections a bit.  There is still a lot of inconsistency in the names of experts, etc.

@@
-\newpage
 \section{Proof theory and proof certificates}
@@
 
-\newcommand{\XXi}{{\color{blue}{\Xi}}}
-\newcommand{\bc}[5]{#1#2\Downarrow #3 : #4 \vdash #5}
-
-\todo{Right now there is no intro to FPC}
+%\todo{Right now there is no intro to FPC}
 % \begin{metanote}
@@
 
-\subsection{Proof systems with certificates}
+\subsection{Proof for the Horn fragment}
 \label{ssec:focused}
-\todo{Change title? }
 
@@
 % \todo{Figure now uses Dale kernel-v2 --- needs further abstraction and explanation, see appendix}
+
+A \emph{Horn clause} is a formula of the form $\forall \bar
+x_1. A_1\supset \forall \bar x_2. A_2\supset \forall \bar
+x_n. A_n\supset A_0$ where $\forall\bar x_i$ denote a list of
+universal quantifiers ($i\in\{1,\ldots,n\}$) and $A_0,\ldots,A_n$ are
+atomic formulas.  It is well known that the following set of sequent
+calculus proof rules are complete for both classical and
+intuitionistic logic when one is attempting to prove that a given
+atomic formula $A$ is provable from a give set $\Pscr$ of Horn
+clauses. 
+\[
+  \infer[decide]{\Pscr\vdash A}{D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
+  \qquad
+  \infer[init]{\Pscr\Downarrow A\vdash A}{}
+\]
+\[
+  \infer[\forall L]{\Pscr\Downarrow\forall x. D\vdash A}
+        {\Pscr\Downarrow D[t/x]\vdash A}
+  \qquad
+  \infer[\supset L]{\Pscr\Downarrow B\supset D\vdash A}
+        {\Pscr\vdash B\qquad \Pscr\Downarrow D\vdash A}
+\]
+Here, we use two different styles of sequents.  The sequent
+$\Pscr\vdash A$ is the usual sequent which we generally use as the end
+sequent (the conclusion) of a proof.  The sequent $\Pscr\Downarrow
+D\vdash A$ is a \emph{focused} sequent in which the formula $D$ is the
+\emph{focus} of the sequent.  The two left introduction rules and the
+initial rule can only be applied to the focused formula.  This latter
+point is in constrast to Gentzen's sequent calculus where these rules
+can involve \emph{any} formula on the left of the $\vdash$.  The fact
+that this proof system is complete for both classical and
+intuitionistic logic (when restricted to the Horn clause fragment)
+follows from rather simple considerations of Horn clauses
+\cite{nadathur90jacm} and from the completeness of LJT proofs
+\cite{Herbelin94}.  The use of the term \emph{focus} comes from
+Andreoli's proof system for linear logic \cite{andreoli92jlc}.
 
@@
   \qquad
-  \infer{\bc{\XXi}{}{[]} A A}{\trueExpert{\XXi}}
+  \infer{\bc{\XXi}{}{[]} A A}{\initLExpert{\XXi}}
   \qquad
-  \infer{\bc{\XXi}{}{(t\ ::\ l)}{\forall x:B.\, D} A}
+  \infer{\bc{\XXi}{}{(t\ ::\ l)}{\depprod{x}{B}{D}}{A}}
         {\XXi_1\vdash t : B\qquad
          \bc{\XXi_2}{}l {D[t/x]} A\qquad
-         \prodExpert{\XXi}{\XXi_1}{\XXi_2}}
+         \prodE{\XXi}{\XXi_1}{\XXi_2}{t}}
 \]
@@
 \end{figure}
-
-\[
-  \infer{\Pscr\vdash A}{D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
-  \qquad
-  \infer{\Pscr\Downarrow A\vdash A}{}
-  \qquad
-  \infer{\Pscr\Downarrow\forall x. B(x)\supset D(x)\vdash A}
-        {\vdash B(t)\qquad \Pscr\Downarrow D(t)\vdash A}
-\]
-Here, $\Pscr$ is a collection of Horn clauses, i.e., formulas of the
-form $\forall \bar x_1. A_1\supset \forall \bar x_2. A_2\supset
-\forall \bar x_n. A_n\supset A_0$.
 
@@
 
-Figure~\ref{fig:augmented} contains an implicitly focused proof system  for a small fragment of the Calculus of Inductive Constructions that corresponds to Horn logic programming over pure \lsti{Inductive} definitions.
-%
-It is a sequent calculus inspired by the calculus for
-proof search in Pure Type Systems introduced in \cite{LengrandDM06}, based in
- turn on ideas stemming from focusing (in particular, the LJT calculus of
-Herbelin \cite{Herbelin94}). Similarly to that calculus, we have a term
-language that includes terms and lists of terms, and two distinct typing
-judgments for the two categories. In our case, where proof search steps are
-viewed as computation, the term judgment $\XXi\vdash t \colon A$ corresponds to a goal-reduction
-step while the judgment $\XXi \Downarrow  l \colon D\vdash A$ corresponds to a backchaining step. This coincides
-with the idea behind the \emph{spine calculus}~\cite{Cervesato97tr}.
+Figure~\ref{fig:augmented} contains these proof rules and extends it
+in the following fours ways so that it more appropriately fits into
+the Coq proof system.
+\begin{enumerate}
+  \item Instead of having separate connectives for $\forall$ and
+    $\supset$, we have the dependent product connective $(x:A)D$. 
+  \item We have incorporated both \emph{proof certificates} (using the
+    schematic variable $\XXi$) along with expert predicates
+    (predicates with the $e$ subscript).
+  \item The inference rules are also anotated by terms structures that
+    can be give directly to the Coq kernel for checking.
+  \item We have added various premises which are responsible to
+    pulling information across the Elpi API with Coq.
+\end{enumerate}
+The proof system described in Figure~\ref{fig:kernel} corresponds to
+the Calculus of Inductive Constructions restricted to the Horn clause
+fragment of pure \lsti{Inductive} definitions.
+This proof system is inspired by the calculus for proof search in
+Pure Type Systems introduced in \cite{LengrandDM06}, based in turn on
+ideas stemming from focusing (in particular, the LJT calculus of
+Herbelin \cite{Herbelin94}). Similarly to that calculus, we have a
+term language that includes terms and lists of terms, and two distinct
+typing judgments for the two categories. In our case, where proof
+search steps are viewed as computation, the term judgment $\XXi\vdash
+t \colon A$ corresponds to a goal-reduction step while the judgment
+$\XXi \Downarrow l \colon D\vdash A$ corresponds to a backchaining
+step. This coincides with the idea behind the \emph{spine
+  calculus}~\cite{Cervesato97tr}.  The main novelity of our proof
+system here is that proof terms and proof certificates are used
+simultaneously in all inference rules.
 
-The proof system is parameterized by Coq' global environment $E$, which
+The proof system is parameterized by Coq's global environment $E$, which
 here can be thought to consist of a set of inductive definitions,
@@
 a \emph{decide} rule in an explicitly focused system. From a logic
-programming viewpoint, given an atom, the rule selects a clause on which to
-back-chain on: we look-up the costructors from an \lsti{Inductive}
-definition from the global enviroment, one that matches the atom we wish to bachchain on and then call the latter judgment.
-The rules for backchaining are standard, corresponding to the \emph{init} and $\forall$-left rules.
+programming viewpoint, given an atom, the rule selects a clause on
+which to back-chain on: we look-up the costructors from an
+\lsti{Inductive} definition from the global enviroment, one that
+matches the atom we wish to bachchain on and then call the latter
+judgment. The rules for backchaining are standard, corresponding to
+the \emph{init} and $\forall$-left rules.
 
@@
 that matter. However, one of the beauty of the Calculus of Inductive
-Construction is that they are in fact defined inductively and
+Construction is that they are, in fact,  defined inductively and
 therefore the \emph{unfold} rule will handle those; this also
@@
 
-  % , and as is usual in Pure Type
-% Systems we have the same grammar for terms and types
-%
-%, as shown in figure~\ref{fig:terms}.
-% \todo{Let's think about that: we need A at least, but A is also a metavariable for atoms}
-% %\begin{figure}
-% \begin{align*}
-% t,u ::= & k\ l \;|\; \forall x : t.\, u \\
-% l ::= & [] \;|\; (t\ ::\ l)
-% \end{align*}
-% % \caption{Grammar of terms}
-% % \label{fig:terms}
-% % \end{figure}
-
-% we denote by $\Pscr_A$ the
-% set of type declarations for the constructors of type $A$, of the form
-% $k : D$ where $D$ is of the form
-% $\forall x_1 : t_1.\, \dots \forall x_n : t_n.\, D\ u_1 \dots u_m$. We
-% denote by $\mathcal{A}$ the set of arity declaration for the inductive
-% types.
-
-%
-%
-%done Explain relations with other term assignments. Note that we do not use
-% explicit substitutions (we are cut free)
-% Explain the choice of not having certificates represent proof terms.
-% \todo{  Explain that the unfold clause handles
-% defined atoms and defined connectives (conjunction, disjunction, existential
-% and also equality): this may be less efficient and proof-theoretically induce
-% pointless phase changes, but it is arguably very elegant e simplifies proof
-% terms that are inherited by Coq }
+\subsection{Proof certificate checking}
 
@@
 kind term   type.                                   % reification of Coq terms
-  kind gref    type                                 % reification of refernces to global terms
+kind gref   type                                    % reification of references to global terms
 type global gref -> term.                            % coercion to term
@@
 offered by the \lsti{auto}'s hard-wired \lsti{Debug} facility.
-%
+
+\subsection{Moving beyond Horn clauses}
+\label{ssec:beyond}
+
+If we do not restrict our attention to only Horn clauses then there is
+no problem will presenting a focused proof system for all of
+first-order intuitionistic logic using the connectives $\forall$ and
+$\supset$.  The following rules extend those given earlier.
+\[
+  \infer[decide]{\Pscr\vdash A}{D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
+  \qquad
+  \infer[init]{\Pscr\Downarrow A\vdash A}{}
+\]
+\[
+  \infer[\forall L]{\Pscr\Downarrow\forall x. D\vdash A}
+        {\Pscr\Downarrow D[t/x]\vdash A}
+  \qquad
+  \infer[\supset L]{\Pscr\Downarrow B\supset D\vdash A}
+        {\Pscr\vdash B\quad \Pscr\Downarrow D\vdash A}
+  \qquad
+  \infer[\supset R]{\Pscr\vdash B\supset D}{\Pscr, B\vdash D}
+  \qquad
+  \infer[\forall R]{\Pscr\vdash \forall x.D}{\Pscr\vdash D[y/x]}
+\]
+As before, the syntactic variable $A$ denotes atomic formulas and, as
+usual, the $\forall R$ rule has the restriction that $y$ is not free in
+its conclusion.
+
+As has been detailed in earlier work on foundational proof
+certificates, this richer notion of proof system can provide for
+richer proof certificates.  The main differences with what we have
+seen before is that the left-hand sides of sequents can now grow
+during the proof checking process.  As a result, an indexing mechanism
+is invoked so that it is possible for the decide expert to properly
+select locally added assumptions instead of just those assumptions
+accepted globally.
+
+*** To illustrate this approach to elaborating externally provided proof
+structures into terms the Coq kernel can check, we consider here
+examples take from the file \verb+coq_fpc.v+, which is part of the
+repository \url{https://github.com/proofcert/fpc-elpi/}.
+Figure~\ref{fig:debruijn} presents the specific FPC that defines the
+proof semantics of using the De Bruijn notions for untyped
+$\lambda$-terms as proof evidence of a proof in intuitionistic
+propositional logic.  That FPC definition, along with a proof checker
+capable of synthesizing Coq proof terms during proof checker, makes it
+possible to use the term denoting the untyped $\lambda$-term
+$\lambda x (x (\lambda y (y (\lambda z (x (\lambda u\,z))))))$.
+
+\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
+(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 
+                                                    [lambda (apply 1 [])])])]))
+\end{lstlisting}
+
+\begin{figure}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.sig}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.mod}
+\caption{The FPC definition of De Bruijn notation as proof evidence.}
+\label{fig:debruijn}
+\end{figure}
+
 %%% Local Variables:

### 2020-11-06 papers/types2020/conc.tex
conclusion finished

@@
 
-\todo{sum up contribution}
+This paper follows on a line of research  starting in the late 80 and gaining more steam in the last five years that advocates the usefulness of proof theory and higher-order logic programming  for the many tasks around the development, enrichment and even formal verification of proof assistants.
+The development of the Elpi-Coq plug-in  has made this connection tighter.
+
+We have presented two application of this synergy: one geared towards
+providing a flexible approach to connecting external provers of
+first-order intuitionistic logic to Coq; the other supporting an out-of-the-box way to do property-based testing for pure inductive relations.
+%
+
+%\todo{sum up contribution}
 
@@
  and provide a way of checking and elaborating proof evidence for
- intutionistic logic \emph{with} (inductively) defined atoms. This
+ intuitionistic logic \emph{with} (inductively) defined atoms. This
  can go as far as FPC for inductive proofs~\cite{blanco15wof}.
@@
 
-% LocalWords:  Elpi simp LJF
+% LocalWords:  Elpi simp LJF PBT LazySmallCheck Ltac

### 2020-11-06 papers/types2020/intro.tex
conclusion finished

@@
 
-Recently, C. Sacerdoti Coen and E. Tassi developed the Elpi
+Recently, Enrico Tassi et al.\ developed the Elpi
 implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and

### 2020-11-06 papers/types2020/sec3.tex
Staging things to do a merge.

@@
 
-%\todo{Right now there is no intro to FPC}
-% \begin{metanote}
-%   (AM): Start with the logical reconstruction of \texttt{check}, to and then reverse engineer what we want/need to say about focusing and FPC.
-We \textbf{do} not talk about PBT yet, just logic programming (with proof terms).
-% In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
-% \end{metanote}
+In this section, we introduce the main ideas from \emph{focused proof
+  system}, \emph{foundational proof certificates}, and the
+\emph{Coq-Elpi} plugin.
 
@@
 
-% The proof search approach to encoding Horn clause computation
-% results in the structuring of proofs with repeated
-% switching between a \emph{goal-reduction} phase and a
-% \emph{backchaining} phase~\cite{miller91apal}.
-
-%\subsection{Certificate checking with expert predicates}
-%\subsection{Adding certificates and experts to a proof system}
-%\label{ssec:fpc}
-
-% \todo{Figure now uses Dale kernel-v2 --- needs further abstraction and explanation, see appendix}
+% \todo{Figure now uses Dale kernel-v2 --- needs further abstraction
+% and explanation, see appendix}
 
@@
 \[
-  \infer[decide]{\Pscr\vdash A}{D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
+  \infer[decide]{\Pscr\vdash A}
+                {D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
   \qquad
@@
 initial rule can only be applied to the focused formula.  This latter
-point is in constrast to Gentzen's sequent calculus where these rules
+point is in contrast to Gentzen's sequent calculus where these rules
 can involve \emph{any} formula on the left of the $\vdash$.  The fact
@@
 
-Figure~\ref{fig:augmented} contains these proof rules and extends it
+Figure~\ref{fig:augmented} contains these proof rules and extends them
 in the following fours ways so that it more appropriately fits into
@@
     (predicates with the $e$ subscript).
-  \item The inference rules are also anotated by terms structures that
+  \item The inference rules are also annotated by terms structures that
     can be give directly to the Coq kernel for checking.
@@
 The proof system described in Figure~\ref{fig:kernel} corresponds to
-the Calculus of Inductive Constructions restricted to the Horn clause
-fragment of pure \lsti{Inductive} definitions.
+the Calculus of Inductive Constructions in which all inductive
+definitions are limited to the Horn clause.
+%
 This proof system is inspired by the calculus for proof search in
@@
 ideas stemming from focusing (in particular, the LJT calculus of
-Herbelin \cite{Herbelin94}). Similarly to that calculus, we have a
+Herbelin \cite{Herbelin94}). Similar to that calculus, we have a
 term language that includes terms and lists of terms, and two distinct
-typing judgments for the two categories. In our case, where proof
-search steps are viewed as computation, the term judgment $\XXi\vdash
-t \colon A$ corresponds to a goal-reduction step while the judgment
-$\XXi \Downarrow l \colon D\vdash A$ corresponds to a backchaining
-step. This coincides with the idea behind the \emph{spine
-  calculus}~\cite{Cervesato97tr}.  The main novelity of our proof
+typing judgments for the two categories.
+%
+In our case, the \lP goal formula \lsti{(check Xi (go B T))}
+corresponds to $\XXi\vdash B \colon T$ and the formula
+%
+\lsti{(check Xi (bc D B L))} corresponds to $\XXi\Downarrow L\colon
+D\vdash B$.
+%
+This style of encode coincides with the idea behind the \emph{spine
+  calculus}~\cite{Cervesato97tr}.  The main novelty of our proof
 system here is that proof terms and proof certificates are used
@@
 The \emph{unfold} rule is the equivalent in our simplified setting of
-a \emph{decide} rule in an explicitly focused system. From a logic
-programming viewpoint, given an atom, the rule selects a clause on
-which to back-chain on: we look-up the costructors from an
-\lsti{Inductive} definition from the global enviroment, one that
-matches the atom we wish to bachchain on and then call the latter
-judgment. The rules for backchaining are standard, corresponding to
-the \emph{init} and $\forall$-left rules.
+a \emph{decide} rule in an explicitly focused system
+\cite{liang09tcs}.  From a logic programming viewpoint, given an atom,
+the rule selects a clause on which to back-chain on: we look-up the
+constructors from an \lsti{Inductive} definition from the global
+environment, one that matches the atom we wish to backchain on and then
+call the latter judgment. The rules for backchaining are standard,
+corresponding to the \emph{init} and the left introduction rules for
+$\forall$ and $\supset$.
 
@@
 
-In figure~\ref{fig:augmented}  each
+In Figure~\ref{fig:augmented}  each
 inference rule is augmented with an additional premise involving an
@@
 Thanks to the Coq-Elpi interface, in particular to the ``main''
-proocedure \lsti{solve}, we can package the \lP's code for the checker
+procedure \lsti{solve}, we can package the \lP's code for the checker
 as a tactic that can be called as any other tactic in a Coq script.
@@
 straightforward: the goal is given to us as a quadruple of a context, an
-evar, a type and a list of extra information. In additon, the certificate is
+evar, a type and a list of extra information. In addition, the certificate is
 supplied by the user when calling the tactic (it simply is an integer in this
@@
 given FPC, for example iterative-deepening. Furthermore, the FPC can
-provide a \emph{trace} that may be more customable that the one
+provide a \emph{trace} that may be more customized than the one
 offered by the \lsti{auto}'s hard-wired \lsti{Debug} facility.
@@
 during the proof checking process.  As a result, an indexing mechanism
-is invoked so that it is possible for the decide expert to properly
+is used so that it is possible for the decide expert to properly
 select locally added assumptions instead of just those assumptions
@@
 
-*** To illustrate this approach to elaborating externally provided proof
-structures into terms the Coq kernel can check, we consider here
-examples take from the file \verb+coq_fpc.v+, which is part of the
-repository \url{https://github.com/proofcert/fpc-elpi/}.
-Figure~\ref{fig:debruijn} presents the specific FPC that defines the
-proof semantics of using the De Bruijn notions for untyped
-$\lambda$-terms as proof evidence of a proof in intuitionistic
-propositional logic.  That FPC definition, along with a proof checker
-capable of synthesizing Coq proof terms during proof checker, makes it
-possible to use the term denoting the untyped $\lambda$-term
-$\lambda x (x (\lambda y (y (\lambda z (x (\lambda u\,z))))))$.
+\newpage
 
+To illustrate a proof certificate format that uses this richer logic
+and proof system, consider the simple \lP code in
+Figure~\ref{fig:debruijn}.  Using the constants provided in that
+figure, the untyped $\lambda$-term
+$\lambda x (x (\lambda y (y (\lambda z (x (\lambda u\,z))))))$ can be
+encoded as the following Elpi term of type \lsti{deb}.
+%
 \begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
@@
 \end{lstlisting}
+
+ presents the specific FPC that defines the
+proof semantics of using the De Bruijn notions for untyped
+$\lambda$-terms as proof evidence of a proof in intuitionistic
+propositional logic.
+
+This FPC can be interpreted by the proof
+certificate checker in the file \verb+ljf-dep.mod+ of the repository
+\url{https://github.com/proofcert/fpc-elpi/}.
+We note that this proof certificate checker is also able to
+simulataneously synthesize a Coq proof term while checking (see
+Section~\ref{}) all the while remaining less than 90 lines of Elpi
+code.
+
+  That FPC definition, along with a proof checker
+capable of synthesizing Coq proof terms during proof checker, makes it
+possible to use the term denoting the
 
@@
 
+\newpage
+
 %%% Local Variables:
@@
 % LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE
-% LocalWords:  Elpi modularity Herbelin bc PTSC's dprolog CIC
+% LocalWords:  Elpi modularity Herbelin bc PTSC's dprolog CIC init lp
+% LocalWords:  Andreoli's basicstyle lprolog sigs Coq's sortedness Ev
+% LocalWords:  backchain gref indt indc coq dest unfoldings str fpc
+% LocalWords:  evar nat forall ys elpi Qed

### 2020-11-06 papers/types2020/app.tex

@@
 
-In this appendix we list some definitions and pieces of code that may be helpful
+In this appendix we list some definitions and pieces of code that have been mentioned in the main paper.
 
-\subsection{Interp}
+\subsection{The vanilla meta-interpreter}
 
@@
 the testing phase of the \lsti{dep_pbt} tactic.
-Differently from Fig.~\ref{fig:augmented} we appeal, via quotations, to Coq's defined connectives.
+Differently from Fig.~\ref{fig:kernel} we appeal, via quotations, to Coq's defined connectives.
 An \lsti{atomic} proposition is one defined \lsti{Inductive}ly.
-The predicate \lsti{definition} hides the API call to the constructors of such a type.
+The predicate \lsti{atomic} makes sure that \lsti{Atom} is \emph{not} a defined logical connective.
 
@@
    | T_Zro : has_type tzero TNat
-   | T_Scc : forall t1,
-         has_type t1 TNat -> has_type (tsucc t1) TNat
-   | T_Prd : forall t1,
-        has_type t1 TNat -> has_type (tpred t1 ) TNat
-   | T_Iszro : forall t1,
-        has_type t1 TNat ->  has_type (tiszero t1) TBool.
+   | T_Scc : forall t1,  has_type t1 TNat -> has_type (tsucc t1) TNat
+   | T_Prd : forall t1,  has_type t1 TNat -> has_type (tpred t1 ) TNat
+   | T_Iszro : forall t1,  has_type t1 TNat ->  has_type (tiszero t1) TBool.
+\end{lstlisting}
+
+\begin{lstlisting}
 Inductive nvalue : tm -> Prop :=
-| nv_zero : nvalue tzero   | nv_succ : forall t, nvalue t -> nvalue (tsucc t).
+ | nv_zero : nvalue tzero
+ | nv_succ : forall t, nvalue t -> nvalue (tsucc t).
 Inductive bvalue : tm -> Prop :=
-| bv_t : bvalue ttrue   | bv_f : bvalue tfalse. 
+ | bv_t : bvalue ttrue
+ | bv_f : bvalue tfalse. 
 Reserved Notation "t1 '===>' t2" (at level 40).
 Inductive step : tm -> tm -> Prop :=
-   | ST_IfTrue : forall t1 t2,
-       (tif ttrue t1 t2) ===> t1
-   | ST_IfFalse : forall t1 t2,
-       (tif tfalse t1 t2) ===> t2
+   | ST_IfTrue : forall t1 t2,  (tif ttrue t1 t2) ===> t1
+   | ST_IfFalse : forall t1 t2, (tif tfalse t1 t2) ===> t2
    | ST_If : forall t1 t1' t2 t3,

### 2020-11-06 papers/types2020/sec3.tex
Finished a first pass over Sections 3 and 4.  I give up the lock on section 3.  I wish to continue with Section 4 a bit more.

@@
 
-\newpage
-
 To illustrate a proof certificate format that uses this richer logic
@@
 \end{lstlisting}
-
- presents the specific FPC that defines the
-proof semantics of using the De Bruijn notions for untyped
-$\lambda$-terms as proof evidence of a proof in intuitionistic
-propositional logic.
-
-This FPC can be interpreted by the proof
-certificate checker in the file \verb+ljf-dep.mod+ of the repository
-\url{https://github.com/proofcert/fpc-elpi/}.
-We note that this proof certificate checker is also able to
-simulataneously synthesize a Coq proof term while checking (see
-Section~\ref{}) all the while remaining less than 90 lines of Elpi
-code.
-
-  That FPC definition, along with a proof checker
-capable of synthesizing Coq proof terms during proof checker, makes it
-possible to use the term denoting the
+Using the constructor \lsti{lc} and \lsti{args}, terms in De Bruijn
+syntax (terms of type \lsti{deb}) are incorporated into proof
+certificates (terms of type \lsti{cert}) along with other integer
+arguments that are needed to compute offsets to address bound
+variables.  
+%
+This FPC can be interpreted by the proof certificate checker in the
+file \verb+ljf-dep.mod+ of the repository
+\url{https://github.com/proofcert/fpc-elpi/}.  We note that this proof
+certificate checker is also able to simultaneously synthesize a Coq
+proof term while checking (see Section~\ref{sec:elab}) all the while
+remaining less than 90 lines of Elpi code.
 
@@
 \end{figure}
-
-\newpage
 
@@
 % LocalWords:  backchain gref indt indc coq dest unfoldings str fpc
-% LocalWords:  evar nat forall ys elpi Qed
+% LocalWords:  evar nat forall ys elpi Qed lc args ljf

### 2020-11-06 papers/types2020/sec3.tex
Improve listings

--- 2020-11-06 papers/types2020/sec3.tex
@@
 The following is a list of some of the signatures of the notions that we use:
-\begin{lstlisting}
+\begin{lstlisting}[language=lprolog]
 kind term   type.                                   % reification of Coq terms
-kind gref   type                                    % reification of references to global terms
+kind gref   type.                                   % reif. of references to global terms
 type global gref -> term.                            % coercion to term
@@
 printing, a term and a \lP function from terms to terms: namely $\forall x :
-A, B$ is encoded by \texttt{prod ``x'' A (x\textbackslash{} B x)}; when, in the
+A, B$ is encoded by \lsti{prod ``x'' A (x\ B x)}; when, in the
 implementation of the product-left rule, we apply $B$ to the variable
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[deletekeywords={goal}]
 Elpi Tactic dprolog.

### 2020-11-06 papers/types2020/sec3.tex
some more metanotes

--- 2020-11-06 papers/types2020/sec3.tex
@@
 
-\subsection{Proof for the Horn fragment}
+\subsection{Proofs for the Horn fragment}
 \label{ssec:focused}
@@
 \[
-  \infer{\XXi\vdash  s : A}{{E[] \vdash_{CIC} s : A}}% The order should be s : A, not A : s ?
+  \infer{\XXi\vdash  A : s}{{E[] \vdash_{CIC} A : s}}% The order should be s : A, not A : s ?
+  %% we are checking nat : Set, for example, so A : s --AM
   \qquad
@@
     $\supset$, we have the dependent product connective $(x:A)D$. 
-  \item We have incorporated both \emph{proof certificates} (using the
+  \item We have incorporated both \emph{proof certificates}~\cite{chihani17jar} (using the
     schematic variable $\XXi$) along with expert predicates
-    (predicates with the $e$ subscript).
+    (predicates with the $e$ subscript). We explain this in Section~\ref{ssec:cert}.
   \item The inference rules are also annotated by terms structures that
@@
 \end{enumerate}
-The proof system described in Figure~\ref{fig:kernel} corresponds to
-the Calculus of Inductive Constructions in which all inductive
-definitions are limited to the Horn clause.
+The proof system described in Figure~\ref{fig:augmented} (and
+implemented in Figure~\ref{fig:kernel}) corresponds to the Calculus
+of Inductive Constructions in which all inductive definitions are
+limited to the Horn clause.
 %
@@
 typing judgments for the two categories.
+\begin{metanote}
+  This is repeated in 3.2 and isn't too early to talk about code here? -am
+\end{metanote}
 %
@@
 %
-This style of encode coincides with the idea behind the \emph{spine
+This style of  proof terms coincides with the idea behind the \emph{spine
   calculus}~\cite{Cervesato97tr}.  The main novelty of our proof
@@
 The proof system is parameterized by Coq's global environment $E$, which
-here can be thought to consist of a set of inductive definitions,
-which following Coq's reference manual we denote with
+here we take as a set of inductive definitions,
+which, following Coq's reference manual, we denote with
 $\mathrm{Ind}[p]  (\Gamma_I := \Gamma_C)$.  One  rule delegates
-to Coq's type checking enforcing the well-sortedness of such types. We
+to Coq's type checking the enforcement of the well-sortedness of such types. We
 do not have a \emph{local context} since we are only dealing with types
@@
 defined. In fact, we do not have a $\forall$ rule on the right,
-although the proof theory would gladly allow it. This means, at the
-term level, that we do not have variables in our grammar of
+although the proof theory would gladly allow it. This means at the
+term level that we do not have variables in our grammar of
 terms. Terms are always applied to a (possibly empty) list of
 arguments.
-
+\begin{metanote}
+  Should we call ``unfold'' \emph{decide} (left) and be done with it? AM
+\end{metanote}
 The \emph{unfold} rule is the equivalent in our simplified setting of
 a \emph{decide} rule in an explicitly focused system
-\cite{liang09tcs}.  From a logic programming viewpoint, given an atom,
-the rule selects a clause on which to back-chain on: we look-up the
+\cite{liang09tcs}.  As in the previous proof system for Horn logic,  given an atom,
+the rule selects a clause on which to back-chain on: we lookup the
 constructors from an \lsti{Inductive} definition from the global
@@
 call the latter judgment. The rules for backchaining are standard,
-corresponding to the \emph{init} and the left introduction rules for
+corresponding to the \emph{init} and the (conflation of the) left introduction rules for
 $\forall$ and $\supset$.
@@
 \subsection{Proof certificate checking}
-
+\label{ssec:cert}
 Figure~\ref{fig:kernel} contains the Elpi implementation of the
@@
 Note that \lsti{prod} encodes dependent products taking a name for pretty
-printing, a term and a \lP function from terms to terms: namely $\forall x :
-A, B$ is encoded by \texttt{prod ``x'' A (x\textbackslash{} B x)}; when, in the
-implementation of the product-left rule, we apply $B$ to the variable
+printing, a term and a \lP function from terms to terms: namely $(x :
+B) D$ is encoded by \texttt{prod ``x'' B (x\textbackslash{} D x)}; when, in the
+implementation of the product-left rule, we apply $D$ to the logic variable
 \lsti{Tm}, we get a term that can be used to continue backchaining. This
@@
 environment of inductive definitions and query for information about them.
-The \lsti{unfold_expert} predicate, then, has the role of selecting which
+The \lsti{unfoldE} predicate,  has, among others, the role of selecting which
 constructor to introduce for the inductive type. The kernel will successively
@@
 
+\begin{metanote}
+  I still don't like this transition -am
+\end{metanote}
 In Figure~\ref{fig:augmented}  each
@@
 %
-\todo{MM: Talk about randomized FPC?}
+%%\todo{MM: Talk about randomized FPC? They mentioned afterwards -am}
 In our case here, an FPC is a collection of \lP clauses that
@@
 %
-For example, the first FPC dictates that the query \mbox{\lsti{(check (qheight 5) B)}} is
+For example, the first FPC dictates that the query \mbox{\lsti{(check (qheight 5) A)}} is
 provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
@@
 %
-Thus, the query \lsti{check ((qheight 4) <c> (qsize 10)) B}
-will succeed if there is a proof of \lsti{B} that has a height less
+Thus, the query \lsti{check ((qheight 4) <c> (qsize 10)) A}
+will succeed if there is a proof of \lsti{A} that has a height less
 than or equal to 4 while also being of size less than or equal to 10.
@@
 
-\subsection{A prolog-like tactic}
+\subsection{A Prolog-like tactic}
 
@@
 
-The glue code between Coq-Elpi and the implementation of our calculus is
-straightforward: the goal is given to us as a quadruple of a context, an
-evar, a type and a list of extra information. In addition, the certificate is
-supplied by the user when calling the tactic (it simply is an integer in this
-case). We just need to call \lsti{check} on the goal's type, together with
-the certificate, in order to obtain a reconstructed proof term. We avoid
-calling the reconstruction directly on the evar because Coq-Elpi ensures that
-evars manipulated by \lP are well-typed at all times, and we cannot guarantee
-that since we work with partially reconstructed term; getting around this is
-simply a matter of assigning the reconstructed term to the evar right
-afterwards.
-The following example shows how we can use the above tactic to do
-FPC-driven logic programming in Coq and return a Coq proof-term:
+The glue code between Coq-Elpi and the implementation of our calculus
+is straightforward: the goal is given to us as a quadruple of a (here
+inactive) context, an \emph{evar}, a type (goal) and a list of extra
+information. In addition, we supply the certificate  when calling the tactic (it simply is an integer and a string to
+identify the ``resource'' FPC we will use in this case). We just need
+to call \lsti{check} on the goal, together with the
+certificate, in order to obtain a reconstructed proof term. We do not
+call the reconstruction directly on the evar because Coq-Elpi
+ensures that evars manipulated by \lP are well-typed at all times; since
+we cannot guarantee that, as we work with partially reconstructed
+term, we  get  around this by an explicit unification.
+
+The following
+example shows how we can use the above tactic to do FPC-driven logic
+programming in Coq and return a Coq proof-term:
 \begin{lstlisting}
@@
 If we do not restrict our attention to only Horn clauses then there is
-no problem will presenting a focused proof system for all of
+no problem  presenting a focused proof system for all of
 first-order intuitionistic logic using the connectives $\forall$ and
 $\supset$.  The following rules extend those given earlier.
+\begin{metanote}
+  Why repeat the previous rules? am
+\end{metanote}
 \[
@@
 % LocalWords:  backchain gref indt indc coq dest unfoldings str fpc
-% LocalWords:  evar nat forall ys elpi Qed lc args ljf
+% LocalWords:  evar nat forall ys elpi Qed lc args ljf HOAS unfoldE
+% LocalWords:  FPCs Prolog evars untyped

### 2020-11-07 papers/types2020/sec3.tex
merge
Fix merge conflict

@@
 printing, a term and a \lP function from terms to terms: namely $(x :
-B) D$ is encoded by \texttt{prod ``x'' B (x\textbackslash{} D x)}; when, in the
-implementation of the product-left rule, we apply $D$ to the logic variable
+B) D$ is encoded by \lsti{prod ``x'' B (x\ D x)}; when, in the
+implementation of the product-left rule, we apply $D$ to the variable
 \lsti{Tm}, we get a term that can be used to continue backchaining. This

### 2020-11-07 papers/types2020/elab.tex
Finished a first pass over Sections 3 and 4.  I give up the lock on section 3.  I wish to continue with Section 4 a bit more.
Moved around large pieces of text as discussed this morning.

@@
-\section{Elaboration of external proof certificates for the Coq
-  kernel}
+\section{Elaboration of external proof certificates for the Coq kernel}
+\label{sec:elab}
 
@@
 that type is, in fact, accepted by Coq users as a theorem of
-intuitionistic logic along with any axioms that have been adopted.
+intuitionistic logic (along with any axioms that have been asserted).
 The rest of the Coq system, especially its tactic language, is
@@
 Certificate (FPC) framework has been designed for formally defining a
-rich collection of proof certificates that do not necessarily contain
+broad collection of proof certificates that do not necessarily contain
 full details.  Proof checkers for FPC certificates can be written
@@
 proofs \cite{blanco17phd,blanco17cade}.
+
+\subsection{Moving beyond Horn clauses}
+\label{ssec:beyond}
+
+If we do not restrict our attention to only Horn clauses then there is
+no problem presenting a focused proof system for all of first-order
+intuitionistic logic using the connectives $\forall$ and $\supset$.
+The following rules extend those given earlier in
+Section~\ref{ssec:focused}. 
+\[
+  \infer[\supset R]{\Pscr\vdash B\supset D}{\Pscr, B\vdash D}
+  \qquad
+  \infer[\forall R]{\Pscr\vdash \forall x.D}{\Pscr\vdash D[y/x]}
+\]
+As usual, the $\forall R$ rule has the restriction that $y$ is not
+free in its conclusion.
+
+As has been detailed in earlier work on foundational proof
+certificates, this richer notion of proof system can provide for
+richer proof certificates.  The main differences with what we have
+seen before is that the left-hand sides of sequents can now grow
+during the proof checking process.  As a result, an indexing mechanism
+is used so that it is possible for the decide expert to properly
+select locally added assumptions instead of just those assumptions
+accepted globally.
+
+To illustrate a proof certificate format that uses this richer logic
+and proof system, consider the simple \lP code in
+Figure~\ref{fig:debruijn}.  Using the constants provided in that
+figure, the untyped $\lambda$-term
+$\lambda x (x (\lambda y (y (\lambda z (x (\lambda u\,z))))))$ can be
+encoded as the following Elpi term of type \lsti{deb}.
+%
+\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
+(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 
+                                                    [lambda (apply 1 [])])])]))
+\end{lstlisting}
+Using the constructor \lsti{lc} and \lsti{args}, terms in De Bruijn
+syntax (terms of type \lsti{deb}) are incorporated into proof
+certificates (terms of type \lsti{cert}) along with other integer
+arguments that are needed to compute offsets to address bound
+variables.  
+%
+This FPC can be interpreted by the proof certificate checker in the
+file \verb+ljf-dep.mod+ of the repository
+\url{https://github.com/proofcert/fpc-elpi/}.  We note that this proof
+certificate checker is also able to simultaneously synthesize a Coq
+proof term while checking (see Section~\ref{sec:elab}) all the while
+remaining less than 90 lines of Elpi code.
+
+\begin{figure}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.sig}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.mod}
+\caption{The FPC definition of De Bruijn notation as proof evidence.}
+\label{fig:debruijn}
+\end{figure}
+
+****************
 
@@
 \begin{enumerate}
-\item Have an external prover output some form of proof evidence
+\item Modify an external prover to output some form of proof evidence
   (proof certificate) for formulas it claims are theorems.
@@
 %
-It is also possible to pair two different proof certificates, defined
-by two different FPC definitions, and do the proof checking in
+It is also possible to pair together two different proof certificates,
+defined by two different FPC definitions, and do the proof checking in
 parallel.  The logic programming setting allows such parallel checking
@@
 As an example, consider an external theorem prover for propositional
-intuitionistic logic which return proof structures that are De Bruijn
-expressions for the underlying untype $\lambda$-proof that encodes
-(part of) the proofs it finds.  Consider the De Buijn expression
-example at the end of Section~\ref{ssec:beyond}.  This can be used as
-evidence that the following is a theorem of Coq.
+intuitionistic logic which returns proof structures as untyped
+$\lambda$-terms using De Bruijn's notation \cite{}.  For example, the
+De Bruijn expression at the end of Section~\ref{ssec:beyond} can be
+used as evidence that the following is a theorem of Coq.
 
@@
 checking using the FPC in Figure~\ref{fig:debruijn} and (an extension
-of) the proof Figure~\ref{fig:kernel}.
-
-\begin{metanote}
-  AM: I think we need more: the FPC may go to the appendix, we may want to say that there is a tactic that proves these theorems etc. 
-\end{metanote}
-
-%\newpage
+of) the proof checker in Figure~\ref{fig:kernel}.
 
@@
 
-% LocalWords:  intuitionistic
+% LocalWords:  intuitionistic dneg peirce forall

### 2020-11-07 papers/types2020/sec3.tex
Moved around large pieces of text as discussed this morning.

@@
 
-\subsection{Moving beyond Horn clauses}
-\label{ssec:beyond}
-
-If we do not restrict our attention to only Horn clauses then there is
-no problem  presenting a focused proof system for all of
-first-order intuitionistic logic using the connectives $\forall$ and
-$\supset$.  The following rules extend those given earlier.
-\begin{metanote}
-  Why repeat the previous rules? am
-\end{metanote}
-\[
-  \infer[decide]{\Pscr\vdash A}{D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
-  \qquad
-  \infer[init]{\Pscr\Downarrow A\vdash A}{}
-\]
-\[
-  \infer[\forall L]{\Pscr\Downarrow\forall x. D\vdash A}
-        {\Pscr\Downarrow D[t/x]\vdash A}
-  \qquad
-  \infer[\supset L]{\Pscr\Downarrow B\supset D\vdash A}
-        {\Pscr\vdash B\quad \Pscr\Downarrow D\vdash A}
-  \qquad
-  \infer[\supset R]{\Pscr\vdash B\supset D}{\Pscr, B\vdash D}
-  \qquad
-  \infer[\forall R]{\Pscr\vdash \forall x.D}{\Pscr\vdash D[y/x]}
-\]
-As before, the syntactic variable $A$ denotes atomic formulas and, as
-usual, the $\forall R$ rule has the restriction that $y$ is not free in
-its conclusion.
-
-As has been detailed in earlier work on foundational proof
-certificates, this richer notion of proof system can provide for
-richer proof certificates.  The main differences with what we have
-seen before is that the left-hand sides of sequents can now grow
-during the proof checking process.  As a result, an indexing mechanism
-is used so that it is possible for the decide expert to properly
-select locally added assumptions instead of just those assumptions
-accepted globally.
-
-To illustrate a proof certificate format that uses this richer logic
-and proof system, consider the simple \lP code in
-Figure~\ref{fig:debruijn}.  Using the constants provided in that
-figure, the untyped $\lambda$-term
-$\lambda x (x (\lambda y (y (\lambda z (x (\lambda u\,z))))))$ can be
-encoded as the following Elpi term of type \lsti{deb}.
-%
-\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
-(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 
-                                                    [lambda (apply 1 [])])])]))
-\end{lstlisting}
-Using the constructor \lsti{lc} and \lsti{args}, terms in De Bruijn
-syntax (terms of type \lsti{deb}) are incorporated into proof
-certificates (terms of type \lsti{cert}) along with other integer
-arguments that are needed to compute offsets to address bound
-variables.  
-%
-This FPC can be interpreted by the proof certificate checker in the
-file \verb+ljf-dep.mod+ of the repository
-\url{https://github.com/proofcert/fpc-elpi/}.  We note that this proof
-certificate checker is also able to simultaneously synthesize a Coq
-proof term while checking (see Section~\ref{sec:elab}) all the while
-remaining less than 90 lines of Elpi code.
-
-\begin{figure}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.sig}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.mod}
-\caption{The FPC definition of De Bruijn notation as proof evidence.}
-\label{fig:debruijn}
-\end{figure}
-
 %%% Local Variables:

### 2020-11-07 papers/types2020/elab.tex
More updates to sec3 and elab.

@@
+%\newpage\quad\newpage
 \section{Elaboration of external proof certificates for the Coq kernel}
@@
 
-\subsection{Moving beyond Horn clauses}
+\subsection{Moving beyond Horn clauses - because most provers are
+  without induction}
 \label{ssec:beyond}
@@
 %
-It is also possible to pair together two different proof certificates,
-defined by two different FPC definitions, and do the proof checking in
-parallel.  The logic programming setting allows such parallel checking
+***
+The logic programming setting allows such parallel checking
 to also serve as checking-and-synthesizing.  That is, during the
@@
 
+%\newpage
+
 %%% Local Variables:

### 2020-11-07 papers/types2020/sec3.tex
More updates to sec3 and elab.

@@
 \lsti{check $\XXi$ (bc D A L)}.
+%
+The code in that figure mixes both Coq-specific items with
+FPC-specific items.  We describe both of these separately.
 
-% here the infix
-% turnstile $\vdash$ symbol is replaced by the \lsti{check} predicate, and every rule
-% is directly mapped to a clause in the code. The argument of the clause is
-% either given by a \lsti{go} or a \lsti{bc} constructor, signaling whether the clause
-% is a goal-reduction or a backchaining one.
-
+\subsubsection{Coq-specific code}
 Coq terms are accessed through the Coq-Elpi API, and their representation
 in \lP takes advantage of the native \lP constructs such as lists and binders.
-The following is a list of some of the signatures of the notions that we use:
+The following is part of the Coq-Elpi API signature of constants we use:
 \begin{lstlisting}[language=lprolog]
@@
 \end{lstlisting}
-
+%
 Note that \lsti{prod} encodes dependent products taking a name for pretty
-printing, a term and a \lP function from terms to terms: namely $(x :
+printing, a term and a \lP abstraction from terms to terms: namely $(x :
 B) D$ is encoded by \lsti{prod ``x'' B (x\ D x)}; when, in the
@@
 
-\begin{metanote}
-  I still don't like this transition -am
-\end{metanote}
-In Figure~\ref{fig:augmented}  each
-inference rule is augmented with an additional premise involving an
-\emph{expert predicate}, a certificate $\Xi$, and possibly continuations of
-certificates ($\Xi'$, $\Xi_1$, $\Xi_2$) with extracted information from
+\subsubsection{FPC-specific code}
+
+Each inference rule in Figure~\ref{fig:augmented} is augmented with an
+additional premise involving an \emph{expert predicate} and is
+annotated with variables $\Xi$, $\Xi'$, $\Xi_1$, $\Xi_2$ that range
+over proof certificates.  These expert predicates are used to extract
+information from a certificate.  For example, the premise
+$\prodE{\XXi}{\XXi_1}{\XXi_2}{t}$ can be understood as a procedure for
+extracting from the certificate $\XXi$ a substitution term $t$ and two
+continuation certificates $\XXi_1$ and $\XXi_2$.
+%
 These notions are taken from the general setting of \emph{foundational
@@
 %%\todo{MM: Talk about randomized FPC? They mentioned afterwards -am}
-In our case here, an FPC is a collection of \lP clauses that
-provide the remaining details not supplied in Figure~\ref{fig:augmented} and~\ref{fig:kernel}:
-that is, the exact set of constructors for the  type of certificates as
-well as the specification of the expert predicates listed \emph{ibidem}.
+In our case here, an FPC is a collection of \lP clauses that provide
+the remaining details not supplied in Figures~\ref{fig:augmented}
+and~\ref{fig:kernel}: that is, the exact set of constructors for the
+type of certificates \lsti{cert} as well as the specification of the expert
+predicates listed \emph{ibidem}.
 %
@@
 
-\todo{We could move here from elab the paragraph ``It is also possible to pair two different proof certificates, defined
-by two different FPC definitions,'' and/or reduce redundancy}
-Further, if we view a particular FPC as a means of restricting proofs, it is
-possible to build an FPC that \emph{restricts} proofs satisfying two FPCs
-simultaneously.
-%
-% In particular, Figure~\ref{fig:pairing} defines an FPC based on the
-% (infix) constructor \lsti{<c>}, which \emph{pairs} two terms of type
-% \lsti{cert}.
-%
-The pairing experts for the certificate \lsti{Cert1 <c> Cert2} simply
-request that the corresponding experts also succeed for both
-\lsti{Cert1} and \lsti{Cert2} 
-%
-Thus, the query \lsti{check ((qheight 4) <c> (qsize 10)) A}
-will succeed if there is a proof of \lsti{A} that has a height less
-than or equal to 4 while also being of size less than or equal to 10.
-
-% Various  additional examples and experiments using the pairing of FPCs
-% can be found in~\cite{blanco17cade}. Using similar techniques, it is possible to define FPCs that target
-% specific types for special treatment: for example, when generating
-% integers, only (user-defined) small integers could be produced.
+As it has been described in \cite{blanco17cade}, it is also possible
+to pair together two different proof certificates, defined by two
+different FPC definitions, and do the proof checking in parallel.
+This means that we can build an FPC that \emph{restricts} proofs
+satisfying two FPCs simultaneously.  In particular, the infix
+constructor \lsti{<c>} in Figure~\ref{fig:resources} forms the pair of
+two proof certificates and the pairing experts for the certificate
+\lsti{Cert1 <c> Cert2} simply request that the corresponding experts
+also succeed for both \lsti{Cert1} and \lsti{Cert2} Thus, the query
+\lsti{check ((qheight 4) <c> (qsize 10)) A} will succeed if there is a
+proof of \lsti{A} that has a height less than or equal to 4 while also
+being of size less than or equal to 10.
 

### 2020-11-07 papers/types2020/pbt.tex
two metnotes taken care of; I'd like to keep the appendix

@@
 combination. Note however that it would take  no more than two lines of
-code in the \lsti{unfoldE} expert to implement a form of random data
+code in the \lsti{decideE} expert to implement a form of random data
 generation in the sense of randomized
@@
 
-% \begin{metanote}
-%   I'm just saying the bare minimum -am
-% \end{metanote}
-
-If we view  a property as a many-sorted logical formula:
+If we view  a Horn property (in uncurried form) as a many-sorted logical formula:
 \[

### 2020-11-07 papers/types2020/sec3.tex
two metnotes taken care of; I'd like to keep the appendix

@@
 typing judgments for the two categories.
-\begin{metanote}
-  This is repeated in 3.2 and isn't too early to talk about code here? -am
-\end{metanote}
-%
-In our case, the \lP goal formula \lsti{(check Xi (go B T))}
-corresponds to $\XXi\vdash B \colon T$ and the formula
-%
-\lsti{(check Xi (bc D B L))} corresponds to $\XXi\Downarrow L\colon
-D\vdash B$.
+% %
+% In our case, the \lP goal formula \lsti{(check Xi (go B T))}
+% corresponds to $\XXi\vdash B \colon T$ and the formula
+% %
+% \lsti{(check Xi (bc D B L))} corresponds to $\XXi\Downarrow L\colon
+% D\vdash B$.
 %
@@
 arguments.
-\begin{metanote}
-  Should we call ``unfold'' \emph{decide} (left) and be done with it? AM
-\end{metanote}
-The \emph{unfold} rule is the equivalent in our simplified setting of
-a \emph{decide} rule in an explicitly focused system
-\cite{liang09tcs}.  As in the previous proof system for Horn logic,  given an atom,
+The \emph{decide} rule, %  is the equivalent in our 
+% a \emph{decide} rule in an explicitly focused system
+% \cite{liang09tcs}.  
+as in the previous proof system for Horn logic,  given an atom,
 the rule selects a clause on which to back-chain on: we lookup the
@@
 Construction is that they are, in fact,  defined inductively and
-therefore the \emph{unfold} rule will handle those; this also
+therefore the \emph{decide} rule will handle those; this also
 simplifies the syntax of proof-terms.
@@
 % \lsti{term} with a hole in it for another \lsti{term}.
-The unfolding rule makes use of the Coq-Elpi primitives
+The decide rule makes use of the Coq-Elpi primitives
 \lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
@@
 environment of inductive definitions and query for information about them.
-The \lsti{unfoldE} predicate,  has, among others, the role of selecting which
+The \lsti{decideE} predicate,  has, among others, the role of selecting which
 constructor to introduce for the inductive type. The kernel will successively
@@
 %
-Similarly, the second FPC can be used to bound the total number of instances of
-unfoldings in a proof.
+Similarly, the second FPC can be used to bound the total number of instances of uses of the decide rule in a proof.
 %

### 2020-11-07 papers/types2020/conc.tex
happy about sec 3 and 4

@@
 
-This paper follows on a line of research  starting in the late 80 and gaining more steam in the last five years that advocates the usefulness of proof theory and higher-order logic programming  for the many tasks around the development, enrichment and even formal verification of proof assistants.
-The development of the Elpi-Coq plug-in  has made this connection tighter.
+This paper follows on a line of research starting in the late 80's and
+gaining more steam in the last five years, which advocates the
+usefulness of proof theory and higher-order logic programming for the
+many tasks concerning the development, enrichment and even formal
+verification of proof assistants. The development of the Elpi-Coq
+plug-in has made this connection tighter.
 

### 2020-11-07 papers/types2020/pbt.tex
happy about sec 3 and 4

@@
 We have presented in a previous paper~\cite{blanco19ppdp} a
-proof-theoretical reconstruction of property-based testing (PBT) \cite{fink97sen}of
+proof-theoretical reconstruction of property-based testing (PBT)~\cite{fink97sen}
+of
 relational specifications, adopting techniques from foundational proof
@@
 accounted for in~\cite{blanco19ppdp}, for the sake of this paper we
-will implement only FPC corresponding to different flavors
+will implement only FPC corresponding to different flavors of
 \emph{exhaustive} generation, as adopted, e.g., in
@@
 % LocalWords:  tfalse tpred tiszero TBool TNat subexp SuccBool
-% LocalWords:  notstuck unfoldE
+% LocalWords:  notstuck unfoldE decideE

### 2020-11-07 papers/types2020/sec3.tex
happy about sec 3 and 4

@@
   \item We have added various premises which are responsible to
-    pulling information across the Elpi API with Coq.
+    pulling information across  Coq's global environment.
 \end{enumerate}
 The proof system described in Figure~\ref{fig:augmented} (and
-implemented in Figure~\ref{fig:kernel}) corresponds to the Calculus
+implemented in Figure~\ref{fig:kernel}) corresponds to a subset of the Calculus
 of Inductive Constructions in which all inductive definitions are
-limited to the Horn clause.
+limited to pure Horn clauses.
 %
-This proof system is inspired by the calculus for proof search in
+This  system is inspired by the calculus for proof search in
 Pure Type Systems introduced in \cite{LengrandDM06}, based in turn on
@@
 
-The proof system is parameterized by Coq's global environment $E$, which
-here we take as a set of inductive definitions,
-which, following Coq's reference manual, we denote with
-$\mathrm{Ind}[p]  (\Gamma_I := \Gamma_C)$.  One  rule delegates
-to Coq's type checking the enforcement of the well-sortedness of such types. We
-do not have a \emph{local context} since we are only dealing with types
-that correspond to Horn clauses, and atomic types are inductively
-defined. In fact, we do not have a $\forall$ rule on the right,
-although the proof theory would gladly allow it. This means at the
-term level that we do not have variables in our grammar of
+The proof system is parameterized by Coq's global environment $E$,
+which here we take as a set of inductive definitions; following Coq's
+reference manual, we denote such a definition with
+$\mathrm{Ind}[p] (\Gamma_I := \Gamma_C)$, where $\Gamma_I$ determines
+the names and types of the inductive type and $\Gamma_C$ the names
+and types of its constructors. A rule delegates to Coq's
+type checking the enforcement of the well-sortedness of such types. We
+do not have a \emph{local context} since we are only dealing with
+types that correspond to Horn clauses, and atomic types are
+inductively defined. In fact, we do not have a $\forall$ rule on the
+right, although the proof theory would gladly allow it. This means at
+the term level that we do not have variables in our grammar of
 terms. Terms are always applied to a (possibly empty) list of
-arguments.
-The \emph{decide} rule, %  is the equivalent in our 
+arguments.  The \emph{decide} rule, % is the equivalent in our
 % a \emph{decide} rule in an explicitly focused system
@@
 as in the previous proof system for Horn logic,  given an atom,
-the rule selects a clause on which to back-chain on: we lookup the
+ selects a clause on which to back-chain on: we lookup the
 constructors from an \lsti{Inductive} definition from the global
-environment, one that matches the atom we wish to backchain on and then
+environment, one that matches the atom we wish to backchain on, and then
 call the latter judgment. The rules for backchaining are standard,
@@
 Note that \lsti{prod} encodes dependent products taking a name for pretty
-printing, a term and a \lP abstraction from terms to terms: namely $(x :
-B) D$ is encoded by \lsti{prod ``x'' B (x\ D x)}; when, in the
+printing, a term and a \lP abstraction from terms to terms: i.e., $(x :
+B) D$ is encoded by \verb|prod ``x'' B (x\ D x)|; when, in the
 implementation of the product-left rule, we apply $D$ to the variable
-\lsti{Tm}, we get a term that can be used to continue backchaining. This
+\lsti{Tm}, we get a new formula that can be used to continue backchaining. This
 application is obtained via meta-level substitution, in the style of HOAS. In
@@
 The \lsti{decideE} predicate,  has, among others, the role of selecting which
-constructor to introduce for the inductive type. The kernel will successively
+constructor to focus on from the inductive type. The kernel will successively
 obtain the type of the selected constructor, and initiate the backchaining
-phase on it.
+phase.
 
@@
 also succeed for both \lsti{Cert1} and \lsti{Cert2} Thus, the query
-\lsti{check ((qheight 4) <c> (qsize 10)) A} will succeed if there is a
+\verb+check ((qheight 4) <c> (qsize 10)) A+ will succeed if there is a
 proof of \lsti{A} that has a height less than or equal to 4 while also
@@
 inactive) context, an \emph{evar}, a type (goal) and a list of extra
-information. In addition, we supply the certificate  when calling the tactic (it simply is an integer and a string to
+information. In addition, we supply the certificate  when calling the tactic (it is simply  an integer and a string to
 identify the ``resource'' FPC we will use in this case). We just need
@@
 % LocalWords:  evar nat forall ys elpi Qed lc args ljf HOAS unfoldE
-% LocalWords:  FPCs Prolog evars untyped
+% LocalWords:  FPCs Prolog evars untyped decideE

### 2020-11-07 papers/types2020/elab.tex
Finished a draft of elab.tex.  I'll send email about this section in a few minutes.

@@
-%\newpage\quad\newpage
 \section{Elaboration of external proof certificates for the Coq kernel}
@@
 \end{enumerate}
-
-As we have describe in Section~\ref{sec:three}, the Foundational Proof
-Certificate (FPC) framework has been designed for formally defining a
-broad collection of proof certificates that do not necessarily contain
-full details.  Proof checkers for FPC certificates can be written
-using simple logic programs: the standard implementation mechanisms of
-unification and backtracking search can then be used to reconstruct
-missing proof details \cite{miller17fac}.  Such proof checking kernels
-can also be used to elaborate such certificates into fully detailed
-proofs \cite{blanco17phd,blanco17cade}.
-
-\subsection{Moving beyond Horn clauses - because most provers are
-  without induction}
-\label{ssec:beyond}
-
-If we do not restrict our attention to only Horn clauses then there is
-no problem presenting a focused proof system for all of first-order
-intuitionistic logic using the connectives $\forall$ and $\supset$.
-The following rules extend those given earlier in
-Section~\ref{ssec:focused}. 
+A majority of the external and automatic theorem provers for
+intuitionistic logic do not involve induction.  Instead, they go
+beyond Horn clause by permitting formulas with no restriction on
+occurrences of $\forall$ and $\supset$.  In that case, we need to
+modify the focused proof rules that we have seen in
+Section~\ref{ssec:focused} by adding the following rules.
 \[
@@
 \]
-As usual, the $\forall R$ rule has the restriction that $y$ is not
-free in its conclusion.
+As usual, the $\forall R$ rule has the restriction that the
+eigenvariable $y$ is not free in its conclusion.
 
@@
 seen before is that the left-hand sides of sequents can now grow
-during the proof checking process.  As a result, an indexing mechanism
-is used so that it is possible for the decide expert to properly
-select locally added assumptions instead of just those assumptions
-accepted globally.
+during the proof checking process.  When reading the right
+introduction rule for $\supset$ from conclusion to premise, we shall
+say that the antecedent of the implication is \emph{stored} in the
+left side of the context.  When this store action occurs, an
+\emph{index} is used by the store command to name that new, left-hand
+formula occurrence.  In this extended situation, the decide expert
+uses the index of an assumption in order to enter a focus phase of
+inference.  A full proof checking kernel for first-order
+intuitionistic logic has been given in \cite{chihani17jar} so we do
+not reproduce it here.
 
-To illustrate a proof certificate format that uses this richer logic
-and proof system, consider the simple \lP code in
-Figure~\ref{fig:debruijn}.  Using the constants provided in that
-figure, the untyped $\lambda$-term
-$\lambda x (x (\lambda y (y (\lambda z (x (\lambda u\,z))))))$ can be
-encoded as the following Elpi term of type \lsti{deb}.
+\begin{figure}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.sig}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.mod}
+\caption{The FPC definition of De Bruijn notation as proof evidence.}
+\label{fig:debruijn}
+\end{figure}
+
+To give an example, consider using untyped $\lambda$-terms encoded
+using De Bruijn's notation as proof certificates for the propositional
+intuitionistic logic over just $\supset$.  The fact that such terms
+can be used as proof certificates for such formulas (denoting simple
+types) can be formally defined using the FPC description in
+Figure~\ref{fig:debruijn}.
+%
+Using the constants provided in that figure, the untyped
+$\lambda$-term $\lambda x (x (\lambda y (y (\lambda z (x (\lambda
+u\,z))))))$ can be encoded as the following Elpi term of type
+\lsti{deb}.
 %
@@
 variables.
-%
-This FPC can be interpreted by the proof certificate checker in the
-file \verb+ljf-dep.mod+ of the repository
-\url{https://github.com/proofcert/fpc-elpi/}.  We note that this proof
-certificate checker is also able to simultaneously synthesize a Coq
-proof term while checking (see Section~\ref{sec:elab}) all the while
-remaining less than 90 lines of Elpi code.
-
-\begin{figure}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.sig}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.mod}
-\caption{The FPC definition of De Bruijn notation as proof evidence.}
-\label{fig:debruijn}
-\end{figure}
-
-****************
 
@@
 connecting external provers of first-order intuitionistic logic to
-Coq.
-%
-Following the general outline that has been described in
-\cite{blanco17cade,blanco20coq}, we assume that the following step are
+Coq.  Following the general outline that has been described in
+\cite{blanco17cade,blanco20coq}, we assume that the following steps are
 taken.
-%
 \begin{enumerate}
@@
 \item Develop a formal definition of the semantics behind such proof
-  certificates using the FPC framework.
+  certificates using the FPC framework.  The FPC for De Bruijn
+  expressions given in Figure~\ref{fig:debruijn} is an example of this
+  step.
 
@@
 \end{enumerate}
-%
-***
-The logic programming setting allows such parallel checking
-to also serve as checking-and-synthesizing.  That is, during the
-checking of one certificate, it is possible to synthesize, for
-example, a fully detailed term that is appropriate for handing to the
-Coq kernel.
-
+As we have mentioned in Section~\ref{ssec:fpc}, the logic programming
+setting allows parallel checking and synthesizing of a pair of
+certificates.  That is, during the checking of one certificate, it is
+possible to synthesize, for example, a fully detailed term that is
+appropriate for handing to the Coq kernel.
 If one is interested only in building Coq kernel proof structures, we
@@
 intuitionistic logic which returns proof structures as untyped
-$\lambda$-terms using De Bruijn's notation \cite{}.  For example, the
-De Bruijn expression at the end of Section~\ref{ssec:beyond} can be
-used as evidence that the following is a theorem of Coq.
+$\lambda$-terms using De Bruijn's notation.   Using the FPC provided
+in Figure~\ref{fig:debruijn} and the proof certificate checker in the
+file \verb+ljf-dep.mod+ of the repository
+\url{https://github.com/proofcert/fpc-elpi/},  the De Bruijn term
+display above can be elaborated into a proper proof for the following
+Coq theorem.
 
@@
 
-Of course, the proof information in the untyped term must be
-elaborated to be a fully typed and functional expression.  Proof
-checking using the FPC in Figure~\ref{fig:debruijn} and (an extension
-of) the proof checker in Figure~\ref{fig:kernel}.
-
-%\newpage
+We note that this proof certificate checker and Coq proof synthesizer
+is rather compact, comprising less than 90 lines of Elpi code.
 
@@
 
-% LocalWords:  intuitionistic dneg peirce forall
+% LocalWords:  intuitionistic dneg peirce forall basicstyle lprolog
+% LocalWords:  Elpi lc args ljf dep

### 2020-11-07 papers/types2020/cultures.tex

@@
 mechanisms of failure and exception handling.  While \lP is also based
-on ML-style-polymorphically typed higher-order relational
+on ML-style polymorphically typed higher-order relational
 programming\footnote{It uses types to indicate syntactic categories, while
-  any dependency information needs to be captured using predicates and
-  logical connectives and quantifiers.}, it also comes with a
+  any dependency information needs to be captured using predicates,
+  logical connectives, and quantifiers.}, it also comes with a
 completely declarative version of failure and backtracking search.
@@
 
-% LocalWords:  Elpi polymorphically
+% LocalWords:  Elpi polymorphically Gentzen disjunction Herbrand HOAS
 
@@
 %%% End:
+% LocalWords:  coinductive Abella LCF tacticals

### 2020-11-07 papers/types2020/intro.tex
Various minor updates, mostly to normalize the name of some FPC predicates.
Uniform spelling of Coq-Elpi

@@
 striking differences in notions of computing and reasoning that arise
-on each side of the \texttt{coq-elpi} API.   We will also present a quick
+on each side of the Coq-Elpi API.   We will also present a quick
 summary of the key proof theory concepts that are used by our example
@@
 % LocalWords:  Sacerdoti Coen Tassi Elpi Teyjus Felty LCF tacticals
-% LocalWords:  intuitionistic elpi coq MetaCoq
+% LocalWords:  intuitionistic elpi coq MetaCoq et al

### 2020-11-07 papers/types2020/elab.tex
Various minor updates, mostly to normalize the name of some FPC predicates.
small fixes before last pass to elab

@@
 %
+\begin{small}
 \begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
-(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 
-                                                    [lambda (apply 1 [])])])]))
+(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 [lambda (apply 1 [])])])]))
 \end{lstlisting}
+\end{small}
 Using the constructor \lsti{lc} and \lsti{args}, terms in De Bruijn

### 2020-11-07 papers/types2020/intro.tex
small fixes before last pass to elab

--- 2020-11-06 papers/types2020/intro.tex
@@
 Recently, Enrico Tassi et al.\ developed the Elpi
-implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
+implementation~\cite{dunchev15lpar} of \lP~\cite{miller12proghol} and
 more recently, Tassi has made Elpi available as the Coq-Elpi
@@
 constraints and constraints handling rules, which makes it more
-expressive than the Teyjus implementation \cite{nadathur99cade} of
+expressive than the Teyjus implementation~\cite{nadathur99cade} of
 \lP. Second, the plugin version of Elpi comes equipped with a
@@
 typing (e.g., $\Gamma\vdash M\colon\sigma$), evaluation (e.g., natural
-semantics specifications \cite{kahn87stacs,hannan93jfp}), and
-interaction (e.g., structured operational semantics
-\cite{milner89book,plotkin81}).
+semantics specifications~\cite{kahn87stacs,hannan93jfp}), and
+interaction (e.g., structured operational semantics~\cite{milner89book,plotkin81}).
 
@@
 be given an elegant and natural specification using the higher-order
-relational specifications provided by \lP \cite{felty93jar}.  Some
+relational specifications provided by \lP~\cite{felty93jar}.  Some
 recent implementations built using Elpi support the usefulness of
@@
 proof assistants in general~\cite{DunchevCT16,coen19mscs} and, in
-particular, for Coq via the Coq-Elpi plug~\cite{CohenST20,tassi19itp} .
+particular, for Coq via the Coq-Elpi plug~\cite{CohenST20,tassi19itp}.
 
@@
 flexible approaches to implementing many different logics and their
-proof systems \cite{felty89phd,felty88cade,paulson89jar}.  Following
+proof systems~\cite{felty89phd,felty88cade,paulson89jar}.  Following
 that tradition, the Elpi system makes it possible to encode the proofs
-and proof theory of various subsets of the logic behind Coq (see also
-\cite{felty93lics}).
+and proof theory of various subsets of the logic behind Coq (see also~\cite{felty93lics}).
 

### 2020-11-07 papers/types2020/pbt.tex
small fixes before last pass to elab

@@
 
-While nothing prevents us from porting all the PBT features we have
+While nothing prevents us from porting all the PBT features that we have
 accounted for in~\cite{blanco19ppdp}, for the sake of this paper we
@@
 perspective: being a clone of Haskell's QuickCheck, it emphasizes
-testing executable (read \emph{decidable)} specifications with random
+testing executable (read \emph{decidable}) specifications with random
 generators. While current research~\cite{LampropoulosPP18} aims to
@@
 \lsti{Inductive} definitions that corresponds to pure Horn
-programs. We can do that without having to provide a decidability
+programs, without having to provide a decidability
 proof for those definitions. Furthermore, we are not committed to a
@@
 
-If we view  a Horn property (in uncurried form) as a many-sorted logical formula:
+If we view  a Horn property (in uncurried form) as a many-sorted logical formula
 \[
@@
 \[
-  \exists x_1 \colon \tau_1\dots x_n \colon \tau_n [A_1\land\dots\land A_m\wedge \neg B]
+  \exists x_1 \colon \tau_1\dots x_n \colon \tau_n , A_1\land\dots\land A_m\wedge \neg B
   \tag{**}\label{eq:full}
@@
 
-In our Coq  setting $A_i$ and $B$ will be \lsti{Inductive}
+In our Coq  setting $A_i$ and $B$ will be 
 propositions, while the $\tau_j$ are honest-to-goodness datatypes. We
@@
 %
-When using a focused proof system for logic extended with fixed
+In a proof system for intuitionistic logic extended with fixed
 points~\cite{baelde12tocl}, negation corresponds to the usual
@@
 $
- \exists x_1 \colon \tau_1\dots x_n \colon \tau_n [A_1\land\dots\land A_m]
+ \exists x_1 \colon \tau_1\dots x_n \colon \tau_n , A_1\land\dots\land A_m
 $
@@
 The tactic calls \lsti{check} with the given FPC on the dependent
-variables and will simply call a vanilla meta-interpreter (see
-Fig.~\ref{fig:interp} in the appendix) to test the hypotheses and the negation of the
+variables and delegates to  a vanilla meta-interpreter (see
+Fig.~\ref{fig:interp} in the appendix) the  test of the hypotheses and of the negation of the
 goal:
@@
 against data \lsti{x,xs} generated exhaustively up to height
-at most $5$  form the library \lsti{Inductive} definitions of
+at most $5$  from the library \lsti{Inductive} definitions of
 \lsti{nat} and \lsti{list}. We do not generate values for \lsti{rs},
@@
 In order to generate the PBT query, some
-pre-processing is needed. In particular, we need to turn the data
+pre-processing is needed. In particular, we turn the data
 eigenvariables into \lP logic variables when they appear inside a
-specification, and to generate queries %to \lsti|check|
+specification, and  generate queries %to \lsti|check|
 for each of
@@
  | tpred : tm -> tm | tiszero : tm -> tm.
- Inductive typ : Type := |TBool : typ |TNat : typ.
+ Inductive typ : Type := | TBool : typ | TNat : typ.
 \end{lstlisting}
-The completely standard static and small step dynamic semantics are
+The completely standard static and small step dynamic semantics rules are
 reported in appendix~\ref{ssec:types}.

### 2020-11-07 papers/types2020/sec3.tex
Finished a draft of elab.tex.  I'll send email about this section in a few minutes.
small fixes before last pass to elab

@@
     (predicates with the $e$ subscript). We explain this in Section~\ref{ssec:cert}.
-  \item The inference rules are also annotated by terms structures that
-    can be give directly to the Coq kernel for checking.
-  \item We have added various premises which are responsible to
+  \item The inference rules are  annotated by terms structures that
+    can be given directly to the Coq kernel for checking.
+  \item We have added various premises which are responsible for
     pulling information across  Coq's global environment.
@@
 implemented in Figure~\ref{fig:kernel}) corresponds to a subset of the Calculus
-of Inductive Constructions in which all inductive definitions are
+of Inductive Constructions in which inductive definitions are
 limited to pure Horn clauses.
@@
  selects a clause on which to back-chain on: we lookup the
-constructors from an \lsti{Inductive} definition from the global
+constructors of an \lsti{Inductive} definition from the global
 environment, one that matches the atom we wish to backchain on, and then
@@
 %
-The code in that figure mixes both Coq-specific items with
+The code in that figure mixes both Coq-specific and
 FPC-specific items.  We describe both of these separately.
@@
 %
-Note that \lsti{prod} encodes dependent products taking a name for pretty
+Note that \lsti{prod} encodes dependent products by taking a name for pretty
 printing, a term and a \lP abstraction from terms to terms: i.e., $(x :
 B) D$ is encoded by \verb|prod ``x'' B (x\ D x)|; when, in the
-implementation of the product-left rule, we apply $D$ to the variable
+implementation of the product-left rule, we apply \lsti{D} to the variable
 \lsti{Tm}, we get a new formula that can be used to continue backchaining. This
@@
 this sense, our calculus uses \emph{implicit} substitutions, rather than
-explicit ones as in LJT and PTSC's tradition, the more since proof search in
+explicit ones as in the  LJT and PTSC's tradition, the more since proof search in
 our application is cut-free.
@@
 \subsubsection{FPC-specific code}
+\label{ssec:fpc}
 
@@
 %
-Similarly, the second FPC can be used to bound the total number of instances of uses of the decide rule in a proof.
+Similarly, the second FPC can be used to bound the total number of
+instances of the decide rule in a proof.
 %
@@
 \lsti{Cert1 <c> Cert2} simply request that the corresponding experts
-also succeed for both \lsti{Cert1} and \lsti{Cert2} Thus, the query
+also succeed for both \lsti{Cert1} and \lsti{Cert2}. Thus, the query
 \verb+check ((qheight 4) <c> (qsize 10)) A+ will succeed if there is a
@@
 The glue code between Coq-Elpi and the implementation of our calculus
-is straightforward: the goal is given to us as a quadruple of a (here
+is straightforward: the goal consists of a quadruple of a (here
 inactive) context, an \emph{evar}, a type (goal) and a list of extra
-information. In addition, we supply the certificate  when calling the tactic (it is simply  an integer and a string to
-identify the ``resource'' FPC we will use in this case). We just need
-to call \lsti{check} on the goal, together with the
-certificate, in order to obtain a reconstructed proof term. We do not
-call the reconstruction directly on the evar because Coq-Elpi
-ensures that evars manipulated by \lP are well-typed at all times; since
-we cannot guarantee that, as we work with partially reconstructed
-term, we  get  around this by an explicit unification.
+information, also inactive. In addition, we supply the certificate: it
+is simply an integer and a string to identify the ``resource'' FPC that we
+will use in this case. We just need to call \lsti{check} on the goal,
+together with the certificate, in order to obtain a reconstructed
+proof term. We do not call the reconstruction directly on the evar
+because Coq-Elpi ensures that evars manipulated by \lP are well-typed
+at all times; since we cannot guarantee that, as we work with
+partially reconstructed term, we get around this by an explicit
+unification.
 
@@
 restricted to depth-first search, since it follows the dictate of the
-given FPC, for example iterative-deepening. Furthermore, the FPC can
+given FPC, for example adding iterative-deepening search. Furthermore, the FPC can
 provide a \emph{trace} that may be more customized than the one

### 2020-11-07 papers/types2020/elab.tex
done

@@
 \item Detailed typing information might not need to be stored within a
-  proof since types can often be reconstructed during proof checking.
+  proof since types can often be reconstructed during proof
+  checking~\cite{miller17fac}.
 
@@
 To give an example, consider using untyped $\lambda$-terms encoded
-using De Bruijn's notation as proof certificates for the propositional
+using De Bruijn's notation as proof certificates for propositional
 intuitionistic logic over just $\supset$.  The fact that such terms
@@
 synthesis of such proof structures directly into the FPC proof
-checker.  (We took exactly this step in Figure~\ref{fig:augmented}
+checker.  We took exactly this step in Figure~\ref{fig:augmented}
 where proof checking involved both proof certificates as well as
-Coq-term structures.)  If one is interested in checking only one kind
+Coq terms.  If one is interested in checking only one kind
 of external proof structure, then the FPC for that structure could
@@
 
-As an example, consider an external theorem prover for propositional
-intuitionistic logic which returns proof structures as untyped
-$\lambda$-terms using De Bruijn's notation.   Using the FPC provided
-in Figure~\ref{fig:debruijn} and the proof certificate checker in the
-file \verb+ljf-dep.mod+ of the repository
-\url{https://github.com/proofcert/fpc-elpi/},  the De Bruijn term
-display above can be elaborated into a proper proof for the following
-Coq theorem.
+Continuing with the previous example, consider an external theorem
+prover for propositional intuitionistic logic which returns proof
+structures as untyped $\lambda$-terms using De Bruijn's notation.
+Using the FPC provided in Figure~\ref{fig:debruijn} and the proof
+certificate checker in the file \verb+ljf-dep.mod+ of the repository
+\url{https://github.com/proofcert/fpc-elpi}, the De Bruijn term
+displayed above can be elaborated into a proper proof for the
+following Coq theorem.
 
-\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
+\begin{lstlisting}[basicstyle=\ttfamily,language=coq]
 Theorem dneg_peirce_mid : forall P Q: Prop, (((((P -> Q) -> P) -> P) -> Q) -> Q).

### 2020-11-09 papers/types2020/preamble.tex
progress on 4
sec 4 ready some work on conclusion
related work bits spear thru the paper
ported the appendix and sec3 to kernel-v2

sec3 is still a mess
ACM Category
commenti di Enrico

@@
 
-\Copyright{M. Manighetti, D. Miller, and A. Momigliano}
+\Copyright{M. Manighetti, D. Miller and A. Momigliano}
 %TODO mandatory, please use full first names. LIPIcs license is
@@
 
-\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} 
+\ccsdesc[500]{Theory of computation~Logic and verification}
 %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
 
-\keywords{Proof assistant, logic programming, Coq, \lP}
+\keywords{Proof assistants, logic programming, Coq, \lP, property-based testing}
 %TODO mandatory; please add comma-separated list of keywords
@@
 
+ \acknowledgements{Thanks to Enrico Tassi for his help with Coq-Elpi and his comments on the paper.}
 \nolinenumbers %uncomment to disable line numbering
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% LocalWords:  Matteo Manighetti Inria

### 2020-11-09 papers/types2020/sec3.tex
commenti di Enrico

@@
 
-The \lsti{dprolog} tactic can be seen as a programmable version of
-Coq's \lsti{auto}: it currently lacks the latter's capability of
-solving goals by \lsti{reflexivity} as well as the modularity of using
-\lsti{Hints} as introduction rules; on the bright side, it is not
-restricted to depth-first search, since it follows the dictate of the
-given FPC, for example adding iterative-deepening search. Furthermore, the FPC can
-provide a \emph{trace} that may be more customized than the one
-offered by the \lsti{auto}'s hard-wired \lsti{Debug} facility.
+The \lsti{dprolog} tactic can be seen as a programmable (and possibly
+faster) version of Coq's \lsti{eauto}, although it currently lacks the
+latter's capability of solving goals by
+\lsti{reflexivity}; % as well as the modularity of using
+% \lsti{Hints} as introduction rules
+ \lsti{dprolog} is not restricted to depth-first search, since
+it follows the dictate of the given FPC, for example we could easily add
+iterative-deepening search. Furthermore, the tactic can still use
+\lsti{Hints} via Elpi's \lsti{Db} feature, and FPCs can provide a
+\emph{trace} that may be more customizable than the one offered by the
+\lsti{(e)auto}'s hard-wired \lsti{Debug} facility.
 
@@
 % LocalWords:  evar nat forall ys elpi Qed lc args ljf HOAS unfoldE
-% LocalWords:  FPCs Prolog evars untyped decideE
+% LocalWords:  FPCs Prolog evars untyped decideE eauto

### 2021-02-20 papers/types2020/revisions/app.tex
starting a revision dir for types20. Initial work on sec 3

--- 2020-11-06 papers/types2020/app.tex
@@
 
-In this appendix we list some definitions and pieces of code that have been mentioned in the main paper.
+In this appendix we list some definitions and pieces of code that we have  mentioned in the main paper.
 
@@
 An \lsti{atomic} proposition is one defined \lsti{Inductive}ly.
-The predicate \lsti{atomic} makes sure that \lsti{Atom} is \emph{not} a defined logical connective.
+% The predicate \lsti{atomic} makes sure that \lsti{Atom} is \emph{not} a defined logical connective.
 
@@
 % formulas.
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={interp-end}]{code/dep-kernel.sig}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={interp-end}]{code/dep-kernel-v2.mod}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={interp-end}]{../code/dep-kernel.sig}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={interp-end}]{../code/dep-kernel-v2.mod}
 \caption{Encoding in coq-elpi of the vanilla meta-interpreter}
@@
 
-% LocalWords:  LJF Interp dep pbt coq elpi
+% LocalWords:  LJF Interp dep pbt coq elpi ly

### 2021-02-20 papers/types2020/revisions/conc.tex
starting a revision dir for types20. Initial work on sec 3

--- 2020-11-07 papers/types2020/conc.tex
@@
 many tasks concerning the development, enrichment and even formal
-verification of proof assistants. The development of the Elpi-Coq
+verification of proof assistants. The development of the Coq-Elpi
 plug-in has made this connection tighter.
@@
 
-Another obvious limitation of our PBT tactic is that it does not 
- handle any computation (conversion) within an inductive relation.
-We plan to tackle this either at the Ltac level by co-routining it with simplification tactics and/or by adding rewriting features to our FPC system~\cite{ChihaniM16}.
+Another current limitation of our PBT tactic is that it does not
+handle conversion within an inductive relation. One way to address
+this is to have the \emph{init} rule use Coq's unification rather than
+\lP's, via Coq-Elpi's \lsti{coq.unif-eq} hook.
+%We plan to tackle this either at the Ltac level by co-routining it with simplification tactics and/or by adding rewriting features to our FPC system~\cite{ChihaniM16}.
 % Mention \emph{Harpoon}, Beluga's new tactic language?
@@
  The source files mentioned in this paper are available at \url{https://github.com/proofcert/fpc-elpi}.
+
  % \begin{metanote}

### 2021-02-20 papers/types2020/revisions/cultures.tex
starting a revision dir for types20. Initial work on sec 3

--- 2020-11-07 papers/types2020/cultures.tex
@@
 linear logic in type theory have difficulty treating linear negation
-and the multiplicative disjunction $\lpar$ and its unit $\bot$.
+and the multiplicative disjunction  and its unit.
 

### 2021-02-20 papers/types2020/revisions/elab.tex
starting a revision dir for types20. Initial work on sec 3

--- 2020-11-07 papers/types2020/elab.tex
@@
 \begin{figure}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.sig}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.mod}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{../code/deb-fpc.sig}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{../code/deb-fpc.mod}
 \caption{The FPC definition of De Bruijn notation as proof evidence.}

### 2021-02-20 papers/types2020/revisions/pbt.tex
starting a revision dir for types20. Initial work on sec 3

--- 2020-11-07 papers/types2020/pbt.tex
@@
 | o_s : forall x : nat, ordered [x]
-| o_c : forall (x y : nat), forall xs,  ordered xs  -> x <= y -> ordered (x:: y :: xs).     
+| o_c : forall (x y : nat), forall xs,  ordered xs  -> x <= y -> ordered (x :: y :: xs).     
 \end{lstlisting}
@@
 %
-Since we did slip in an error (where?), our tactic reports a
-counter-example, namely \verb|Proof Term: [0, [0; 1; 0]]|, and
-we therefore abort the proof.
+Since we did slip in an error, our tactic reports a counter-example,
+namely \verb|Proof Term: [0, [0; 1; 0]]|, which unpacks to \lsti{x = 0}
+and \lsti{xs = [0; 1; 0]}. As the latter is definitely not an
+ordered list, this points to a quite evident bug in the definition of
+\lsti{ordered}. We leave it to the reader and we abort the proof.
 
 In order to generate the PBT query, some
-pre-processing is needed. In particular, we turn the data
-eigenvariables into \lP logic variables when they appear inside a
+pre-processing is needed. In particular, we turn 
+variables inhabiting datatypes into \lP logic variables when they appear inside a
 specification, and  generate queries %to \lsti|check|

### 2021-02-20 papers/types2020/revisions/sec3.tex
starting a revision dir for types20. Initial work on sec 3

--- 2020-11-09 papers/types2020/sec3.tex
@@
 intuitionistic logic when one is attempting to prove that a given
-atomic formula $A$ is provable from a give set $\Pscr$ of Horn
+atomic formula $A$ is provable from a  set $\Pscr$ of Horn
 clauses. 
@@
       { \mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad 
-       (\mathit{head} \ A : T)\in \Gamma_I\quad  
+       (\mathit{head} \ A) : T \in \Gamma_I\quad  
        k : D \in \Gamma_C\quad 
@@
 \begin{figure}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={sigs-end}]{code/dep-kernel.sig}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={check-end}]{code/dep-kernel-v2.mod}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={sigs-end}]{../code/dep-kernel.sig}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={check-end}]{../code/dep-kernel-v2.mod}
 \caption{Implementation of a  dependent kernel for Coq terms.}
@@
 $\mathrm{Ind}[p] (\Gamma_I := \Gamma_C)$, where $\Gamma_I$ determines
-the names and types of the inductive type and $\Gamma_C$ the names
-and types of its constructors. A rule delegates to Coq's
-type checking the enforcement of the well-sortedness of such types. We
-do not have a \emph{local context} since we are only dealing with
-types that correspond to Horn clauses, and atomic types are
-inductively defined. In fact, we do not have a $\forall$ rule on the
-right, although the proof theory would gladly allow it. This means at
-the term level that we do not have variables in our grammar of
+the names and types of the (possibly mutually)inductive type and
+$\Gamma_C$ the names and types of its constructors; finally $p$
+denotes the number of parameters and playes no role. We delegate to
+Coq's type checking the enforcement of the well-sortedness of
+inductive types. The \emph{local context} is empty, since we are only
+dealing with types that correspond to Horn clauses, and atomic types
+are inductively defined. In fact, we do not have a $\forall$ rule on
+the right, although the proof theory would gladly allow it. This means
+at the term level that we do not have variables in our grammar of
 terms. Terms are always applied to a (possibly empty) list of
@@
 constructors of an \lsti{Inductive} definition from the global
-environment, one that matches the atom we wish to backchain on, and then
-call the latter judgment. The rules for backchaining are standard,
+environment, one that matches the head symbol of the atom we aim to backchain on, and then
+call the latter judgment that will find a correct instantiation, if any. The rules for backchaining are standard,
 corresponding to the \emph{init} and the (conflation of the) left introduction rules for
@@
 \begin{figure}[t]
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={resources-end}]{code/fpcs.sig}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={resources-end}]{code/fpcs.mod}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={resources-end}]{../code/fpcs.sig}
+\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={resources-end}]{../code/fpcs.mod}
 \caption{Sample FPCs}

### 2021-02-22 papers/types2020/revisions/pbt.tex
cleanup of the code directoty under types2020

@@
 increase automation, it is fair to say that testing with
-\textsf{QuickChick}, in particular of relational specifications, is
+\textsf{QuickChick}, in particular relational specifications, is
 still very labor intensive. We do not intend to compete with
@@
 \lsti{nat} and \lsti{list}. We do not generate values for \lsti{rs},
-since by mode information we know that it will be computed.
+since by (informal) mode information we know that it will be computed.
 %
@@
 ordered list, this points to a quite evident bug in the definition of
-\lsti{ordered}. We leave it to the reader and we abort the proof.
+\lsti{ordered}. We leave the fix to the reader. % and we abort the proof.
 
@@
 
-Another way to asses the fault detection capability of a PBT setup is via \emph{mutation analysis}~\cite{CavadaCM20}, whereby localized bugs are voluntarily inserted, with the view that they should caught (``killed'') by a ``good enough'' testing suite.
+Another way to asses the fault detection capability of a PBT setup is via \emph{mutation analysis}~\cite{CavadaCM20}, whereby localized bugs are purposely inserted, with the view that they should caught (``killed'') by a ``good enough'' testing suite.
 Following on  an exercise in the aforementioned chapter of SF, we modify the typing relation by adding the following (nonsensical)  clause:
@@
 \end{lstlisting}
-Some of the required properties for our SUT now fails: not
+Some of the desired properties for our SUT now fails: not
 only type uniqueness, but also progress with counterexample \lsti{e = tsucc ttrue}:
 \begin{lstlisting}
-Definition progress (e :tm) (Has_type : tm -> typ -> Prop) (Step : tm -> tm -> Prop):= 
+Definition progress (e : tm) (Has_type : tm -> typ -> Prop) (Step : tm -> tm -> Prop):= 
      forall t, Has_type e t -> notstuck e Step.
@@
  intros e t Ht.    
- elpi dep_pbt height 2 (Ht) (e) .
+ elpi dep_pbt height 2 (Ht) (e).
  Abort.

### 2021-02-22 papers/types2020/revisions/sec3.tex
cleanup of the code directoty under types2020

@@
 $\Gamma_C$ the names and types of its constructors; finally $p$
-denotes the number of parameters and playes no role. We delegate to
+denotes the number of parameters and plays no role. We delegate to
 Coq's type checking the enforcement of the well-sortedness of
@@
 Coq terms are accessed through the Coq-Elpi API, and their representation
-in \lP takes advantage of the native \lP constructs such as lists and binders.
-The following is part of the Coq-Elpi API signature of constants we use:
+in \lP takes advantage of native \lP constructs such as lists and binders.
+The following is part of the Coq-Elpi API signature of constants that we use:
 \begin{lstlisting}[language=lprolog]
@@
 implementation of the product-left rule, we apply \lsti{D} to the variable
-\lsti{Tm}, we get a new formula that can be used to continue backchaining. This
+\lsti{Tm}, we get a new term that can be used to continue backchaining. This
 application is obtained via meta-level substitution, in the style of HOAS. In
 this sense, our calculus uses \emph{implicit} substitutions, rather than
-explicit ones as in the  LJT and PTSC's tradition, the more since proof search in
-our application is cut-free.
+explicit ones as in the  LJT and PTSC's tradition; this is consistent with proof search in
+our application being cut-free, whereas explicit substitutions are linked to cuts.
 % Syntax: one can think as if something of type \lsti{term -> term} were a
@@
 The decide rule makes use of the Coq-Elpi primitives
-\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
-nested) application, and \lsti{coq.env.indt} in order to access the global
+\lsti{coq.safe-dest-app} to obtain the head term of a (possibly
+nested) application, and \lsti{coq.env.indt} to access the global
 environment of inductive definitions and query for information about them.
@@
 Thanks to the Coq-Elpi interface, in particular to the ``main''
-procedure \lsti{solve}, we can package the \lP's code for the checker
+procedure \lsti{solve}, we can package the \lP code for the checker
 as a tactic that can be called as any other tactic in a Coq script.
@@
     coq.say "Proof:" {coq.term->string Ev}.
-...
-(* Other clauses for different fpc omitted *)
+... (* Other clauses for different fpc omitted *)
 }}.
@@
 is simply an integer and a string to identify the ``resource'' FPC that we
-will use in this case. We just need to call \lsti{check} on the goal,
+will use in this case. We just need to call \lsti{check} on the goal  \lsti{G},
 together with the certificate, in order to obtain a reconstructed
@@
  \lsti{dprolog} is not restricted to depth-first search, since
-it follows the dictate of the given FPC, for example we could easily add
+it follows the dictate of the given FPC; for example we could easily add
 iterative-deepening search. Furthermore, the tactic can still use
 \lsti{Hints} via Elpi's \lsti{Db} feature, and FPCs can provide a
-\emph{trace} that may be more customizable than the one offered by the
+\emph{trace} that may be more customizable than the one offered by 
 \lsti{(e)auto}'s hard-wired \lsti{Debug} facility.

### 2021-03-04 papers/types2020/revisions/sec3.tex
3.3. talks abiyt conversion. More todos, a fix to fig 3, new file for dependent prolog tactic

@@
 programming in Coq and return a Coq proof-term:
-\begin{lstlisting}
+\begin{lstlisting}[]
 Inductive insert (x:nat) : list nat -> list nat -> Prop :=
-| i_n : is_nat x -> insert x [] [x]
+| i_n : insert x [] [x]
 | i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
 | i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
-Lemma i1:  exists R, insert 2 [0;1] R.
-eexists.
+Lemma i1:  exists R, insert 2 ([0] $++$ [1]) R.
 elpi dprolog height 10.
@@
 Print i1.
-ex_intro (fun R : list nat => insert 2 [0; 1] R) [0; 1; 2]
-  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2)) : exists R : list nat, insert 2 [0; 1] R
+ex_intro (fun R : list nat => insert 2 ([0] $++$ [1]) R)  [0; 1; 2]
+  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2))) : exists R : list nat, insert 2 ([0] $++$ [1]) R
 \end{lstlisting}
@@
 The \lsti{dprolog} tactic can be seen as a programmable (and possibly
-faster) version of Coq's \lsti{eauto}, although it currently lacks the
-latter's capability of solving goals by
-\lsti{reflexivity}; % as well as the modularity of using
+faster) version of Coq's \lsti{eauto}% , although it currently lacks the
+% latter's capability of solving goals by
+% \lsti{reflexivity}
+; % as well as the modularity of using
 % \lsti{Hints} as introduction rules

### 2021-03-04 papers/types2020/sec3.tex
3.3. talks abiyt conversion. More todos, a fix to fig 3, new file for dependent prolog tactic

@@
 Inductive insert (x:nat) : list nat -> list nat -> Prop :=
-| i_n : is_nat x -> insert x [] [x]
+| i_n : insert x [] [x]
 | i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
@@
 The \lsti{dprolog} tactic can be seen as a programmable (and possibly
-faster) version of Coq's \lsti{eauto}, although it currently lacks the
-latter's capability of solving goals by
-\lsti{reflexivity}; % as well as the modularity of using
+faster) version of Coq's \lsti{eauto}; % , although it currently lacks the
+% latter's capability of solving goals by
+% \lsti{reflexivity}; % as well as the modularity of using
 % \lsti{Hints} as introduction rules
  \lsti{dprolog} is not restricted to depth-first search, since
-it follows the dictate of the given FPC, for example we could easily add
+it follows the dictates of the given FPC --- for example we could easily add
 iterative-deepening search. Furthermore, the tactic can still use
 \lsti{Hints} via Elpi's \lsti{Db} feature, and FPCs can provide a
-\emph{trace} that may be more customizable than the one offered by the
+\emph{trace} that may be more customizable than the one offered by 
 \lsti{(e)auto}'s hard-wired \lsti{Debug} facility.

### 2021-03-04 papers/types2020/revisions/sec3.tex
added conversion to proof rules and (local) code, some DOI to bibs, start of cover letter

@@
 \begin{figure}[t]
+\centering
   \[
@@
 \[
-  \infer{\XXi\vdash  A : s}{{E[] \vdash_{CIC} A : s}}% The order should be s : A, not A : s ?
+  \infer{\XXi\vdash  A : s}{E[] \vdash_{CIC} A : s\quad \sortExpert{\XXi}}% The order should be s : A, not A : s ?
   %% we are checking nat : Set, for example, so A : s --AM
   \qquad
-  \infer{\bc{\XXi}{}{[]} A A}{\initLExpert{\XXi}}
-  \qquad
+  \infer{\bc{\XXi}{}{[]} A A'}{E[]\vdash_{CIC} A =_{\beta\delta\iota\xi}A'\quad\initLExpert{\XXi}}
+\]
+  \vskip -18pt
+\[
   \infer{\bc{\XXi}{}{(t\ ::\ l)}{\depprod{x}{B}{D}}{A}}
-        {\XXi_1\vdash t : B\qquad
-         \bc{\XXi_2}{}l {D[t/x]} A\qquad
+        {\XXi_1\vdash t : B\quad
+         \bc{\XXi_2}{}l {D[t/x]} A\quad
          \prodE{\XXi}{\XXi_1}{\XXi_2}{t}}
@@
 \begin{figure}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={sigs-end}]{../code/dep-kernel.sig}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={check-end}]{../code/dep-kernel-v2.mod}
+\lstinputlisting[language=lprolog,linerange={sigs-end}]{../code/dep-kernel.sig}
+\lstinputlisting[language=lprolog,linerange={check-end}]{../code/dep-kernel-v2.mod}
 \caption{Implementation of a  dependent kernel for Coq terms.}
@@
 Figure~\ref{fig:augmented} contains these proof rules and extends them
-in the following fours ways so that it more appropriately fits into
+in the following ways so that it more appropriately fits into
 the Coq proof system.
@@
     $\supset$, we have the dependent product connective $(x:A)D$. 
+  \item We account for \emph{computation} inside atoms by generalizing
+    to \emph{init} rule to work modulo conversion.
   \item We have incorporated both \emph{proof certificates}~\cite{chihani17jar} (using the
@@
 of Inductive Constructions in which inductive definitions are
-limited to pure Horn clauses.
+limited to first-order Horn clauses.
 %
@@
 The proof system is parameterized by Coq's global environment $E$,
-which here we take as a set of inductive definitions; following Coq's
+here a set of constant and inductive definitions; following Coq's
 reference manual, we denote such a definition with
 $\mathrm{Ind}[p] (\Gamma_I := \Gamma_C)$, where $\Gamma_I$ determines
-the names and types of the (possibly mutually)inductive type and
+the names and types of the (possibly mutually) inductive type and
 $\Gamma_C$ the names and types of its constructors; finally $p$
-denotes the number of parameters and plays no role. We delegate to
+denotes the number of parameters and plays here no role. We delegate to
 Coq's type checking the enforcement of the well-sortedness of
@@
 the right, although the proof theory would gladly allow it. This means
-at the term level that we do not have variables in our grammar of
+at the term level that there are no variables in our grammar of
 terms. Terms are always applied to a (possibly empty) list of
@@
 environment, one that matches the head symbol of the atom we aim to backchain on, and then
-call the latter judgment that will find a correct instantiation, if any. The rules for backchaining are standard,
-corresponding to the \emph{init} and the (conflation of the) left introduction rules for
-$\forall$ and $\supset$.
+call the latter judgment that will find a correct instantiation, if any. The rules for backchaining include
+ the (conflation of the) left introduction rules for
+$\forall$ and $\supset$, as well as the \emph{init} rule, which incorporates Coq's conversion.
 
@@
 \begin{figure}[t]
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={resources-end}]{../code/fpcs.sig}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={resources-end}]{../code/fpcs.mod}
+\lstinputlisting[language=lprolog,linerange={resources-end}]{../code/fpcs.sig}
+\lstinputlisting[language=lprolog,linerange={resources-end}]{../code/fpcs.mod}
 \caption{Sample FPCs}
@@
 ex_intro (fun R : list nat => insert 2 ([0] $++$ [1]) R)   [0; 1; 2]
-  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2))) : exists R : list nat, insert 2 ([0] $++$ [1]) R
+  (i_c 2 0 [1] [1; 2] (le_S 0 1 (le_S 0 0 (le_n 0)))
+	 (i_c 2 1 [] [2] (le_S 1 1 (le_n 1)) (i_n 2)))
 \end{lstlisting}

### 2021-03-05 papers/types2020/revisions/main.tex
starting a revision dir for types20. Initial work on sec 3
cleanup of the code directoty under types2020
Improved the abstract.

--- 2020-10-26 papers/types2020/main.tex
@@
 \documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
-\input head
+
+\input{../head}
 \input preamble
 
-\title{Interfacing logic programming and Coq} % Other suggestions?
-\titlerunning{Interfacing logic programming and Coq}
+\title{Two applications of logic programming to Coq} 
+\titlerunning{Two applications of logic programming to Coq}
 
@@
 \begin{abstract}
-We do stuff
+The logic programming paradigm provides a flexible setting for
+representing, manipulating, checking, and elaborating proof
+structures.  This is particularly true when the logic programming
+language allows for bindings in terms and proofs.  In this paper,
+we make use of two recent innovations at the intersection of logic
+programming and proof checking.  One of these is the
+\emph{foundational proof certificate} (FPC) framework which provides a
+flexible means of defining the semantics of a range of proof
+structures for classical and intuitionistic logic.  A second
+innovation is the recently released Coq-Elpi plugin for Coq in which
+the Elpi implementation of \lP can send and retrieve information to
+and from the Coq kernel.  We illustrate the use of both this Coq
+plugin and FPCs with two example applications.  First, we implement an
+FPC-driven sequent calculus for a fragment of the Calculus of
+Inductive Constructions and we package it into a tactic to perform
+property-based testing of inductive types corresponding to pure
+(higher-order) Horn programs.  Second, we implement in Elpi a proof
+checker for first-order intuitionistic logic and demonstrate how proof
+certificates can be supplied by external (to Coq) provers and then
+elaborated into the fully detailed proof terms that can be checked by
+the Coq kernel.
 \end{abstract}
@@
 
-\section{Conclusion}
-
-\bibliography{l}
+\input conc
+\bibliography{../l}
+%\newpage
+\appendix
+\input app
 \end{document}
@@
 %%% End:
+
+% LocalWords:  pbt elab conc Elpi LJF FPC Coq sequent provers

### 2021-03-05 papers/types2020/revisions/preamble.tex
starting a revision dir for types20. Initial work on sec 3
cleanup of the code directoty under types2020
Updated acknowledgments

--- 2020-11-09 papers/types2020/preamble.tex
@@
 
-\Copyright{M. Manighetti, D. Miller and A. Momigliano}
+\Copyright{Matteo Manighetti, Dale Miller and Alberto Momigliano}
 %TODO mandatory, please use full first names. LIPIcs license is
@@
 
- \acknowledgements{Thanks to Enrico Tassi for his help with Coq-Elpi and his comments on the paper.}
+ \acknowledgements{Thanks to Enrico Tassi for his help with Coq-Elpi
+   and his comments on the paper.  The comments of the anonymous
+   reviewers of an early draft of this paper have also been very helpful.} 
 \nolinenumbers %uncomment to disable line numbering
@@
 
-% LocalWords:  Matteo Manighetti Inria
+% LocalWords:  Matteo Manighetti Inria Palaiseau Ecole Polytechnique
+% LocalWords:  Momigliano Dipartimento di Informatica Universit degli
+% LocalWords:  studi Milano Coq Tassi Elpi
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:

### 2021-03-05 papers/types2020/revisions/intro.tex
Updated intro with mild changes.

@@
 \lP program elements and an API for accessing the Coq environment,
-type checking included.
+including its type checker.
 
@@
 In this paper, we present two applications of Elpi within Coq. With
-these examples, we shall illustrate that Elpi is useful not only
-because of its meta-programming features, but also because it soundly
+these applications, we shall illustrate that Elpi is useful not only
+because of its meta-programming features but also because it soundly
 implements a higher-order intuitionistic logic: such implementations
@@
 that tradition, the Elpi system makes it possible to encode the proofs
-and proof theory of various subsets of the logic behind Coq (see also~\cite{felty93lics}).
+and proof theory of various subsets of the logic behind Coq 
+(see also~\cite{felty93lics}).
 

### 2021-03-05 papers/types2020/revisions/cultures.tex
Updates to section 2.

@@
 Gentzen's discovered that the key to treating classical and
-intuitionistic logics in the same proof system was the weakening and
-contraction structural rules on the right-side of
-sequents~\cite{gentzen35}.  This discovery lead him away from natural
-deduction to multiple-conclusion sequent calculus, a very important
-kind of proof system that has not been given a satisfactory
-representation in type theory.  This same innovation of Gentzen also
-opened the way to another key proof-theoretic discovery, that of
-linear logic \cite{girard87tcs}.  Similarly, sequent calculus provides
-an elegant presentation of linear logic while most treatments of
-linear logic in type theory have difficulty treating linear negation
-and the multiplicative disjunction  and its unit.
+intuitionistic logics in the same proof system was identifying the
+structural rules of the weakening and contraction and
+placing them on the right-side of sequents~\cite{gentzen35}.
+This discovery lead him away from natural deduction to
+multiple-conclusion sequent calculus, a very important kind of proof
+system that has not been given a satisfactory representation in type
+theory.  This same innovation of Gentzen's also opened the way to
+another key proof-theoretic discovery, that of linear logic
+\cite{girard87tcs}.  Similarly, sequent calculus provides an elegant
+presentation of linear logic while most treatments of linear logic in
+type theory have difficulty treating linear negation and the
+multiplicative disjunction and its unit.
 
@@
 Curry-Howard correspondence, models computation abstractly as
-$\beta$-reduction of natural deduction proof: that is, computation is
-modeled as \emph{proof normalization}.  On the other hand, the logic
-programming paradigm, following the notion of goal-directed proof
-search, models computation abstractly as a regimented search for
-cut-free sequent calculus proofs: that is, computation is modeled as
-\emph{proof search}.
+$\beta$-reduction within natural deduction proofs: that is,
+computation is modeled as \emph{proof normalization}.  On the other
+hand, the logic programming paradigm, following the notion of
+goal-directed proof search, models computation abstractly as a
+regimented search for cut-free sequent calculus proofs: that is,
+computation is modeled as \emph{proof search}.
 
@@
 structures more generally.  Such eigenvariables can be used by the
-logic programming paradigm to provide a natural, elegant, and powerful
-computing approach to bindings within syntax.
+logic programming paradigm to provide a natural and powerful
+approach to computing with bindings within syntax.
 

### 2021-03-05 papers/types2020/revisions/main.tex
cleanup of addition of conversion

@@
   structures.  This is particularly true when the logic programming
-language allows for bindings in terms and proofs.  In this paper,
-we make use of two recent innovations at the intersection of logic
+  language allows for bindings in terms and proofs.  In this paper, we
+  make use of two recent innovations at the intersection of logic
   programming and proof checking.  One of these is the
-\emph{foundational proof certificate} (FPC) framework which provides a
-flexible means of defining the semantics of a range of proof
+  \emph{foundational proof certificate} (FPC) framework which provides
+  a flexible means of defining the semantics of a range of proof
   structures for classical and intuitionistic logic.  A second
@@
   and from the Coq kernel.  We illustrate the use of both this Coq
-plugin and FPCs with two example applications.  First, we implement an
-FPC-driven sequent calculus for a fragment of the Calculus of
+  plugin and FPCs with two example applications.  First, we implement
+  an FPC-driven sequent calculus for a fragment of the Calculus of
   Inductive Constructions and we package it into a tactic to perform
-property-based testing of inductive types corresponding to pure
-(higher-order) Horn programs.  Second, we implement in Elpi a proof
-checker for first-order intuitionistic logic and demonstrate how proof
+  property-based testing of inductive types corresponding to Horn
+  clauses.  Second, we implement in Elpi a proof checker for
+  first-order intuitionistic logic and demonstrate how proof
   certificates can be supplied by external (to Coq) provers and then
-elaborated into the fully detailed proof terms that can be checked by
-the Coq kernel.
+  elaborated into the fully detailed proof terms that can be checked
+  by the Coq kernel.
 \end{abstract}

### 2021-03-05 papers/types2020/revisions/sec3.tex
cleanup of addition of conversion

@@
 \]
-\caption{Specification of the checking of a proof certificate and the
-  synthesis of a dependently typed $\lambda$-term.}
+\caption{Specification of a core calculus% of the checking of a proof certificate and the
+  % synthesis of a dependently typed $\lambda$-term
+  .}
 \label{fig:augmented}
@@
 \lstinputlisting[language=lprolog,linerange={sigs-end}]{../code/dep-kernel.sig}
-\lstinputlisting[language=lprolog,linerange={check-end}]{../code/dep-kernel-v2.mod}
-\caption{Implementation of a  dependent kernel for Coq terms.}
+\lstinputlisting[language=lprolog,linerange={check-end}]{../code/dep-kernel.mod}
+\caption{Implementation of the core calculus.}
 \label{fig:kernel}
@@
   \item We account for \emph{computation} inside atoms by generalizing
-    to \emph{init} rule to work modulo conversion.
+    the \emph{init} rule to work modulo conversion.
   \item We have incorporated both \emph{proof certificates}~\cite{chihani17jar} (using the
@@
 of Inductive Constructions in which inductive definitions are
-limited to first-order Horn clauses.
+limited to % first-order
+Horn clauses.
 %
@@
 here a set of constant and inductive definitions; following Coq's
-reference manual, we denote such a definition with
+reference manual, inductive definition are denoted by
 $\mathrm{Ind}[p] (\Gamma_I := \Gamma_C)$, where $\Gamma_I$ determines
@@
  selects a clause on which to back-chain on: we lookup the
-constructors of an \lsti{Inductive} definition from the global
+constructors of an inductive definition from the global
 environment, one that matches the head symbol of the atom we aim to backchain on, and then
@@
 obtain the type of the selected constructor, and initiate the backchaining
-phase.
+phase. The latter is terminated when the focused atom unifies with the current goal, via
+Elpi's primitive \lsti{coq.unify-eq}.
 
@@
 example shows how we can use the above tactic to do FPC-driven logic
-programming in Coq and return a Coq proof-term:
+programming modulo conversion in Coq and return a Coq proof-term:
 \begin{lstlisting}[]
@@
 it follows the dictate of the given FPC; for example we could easily add
-iterative-deepening search. Furthermore, the tactic can still use
-\lsti{Hints} via Elpi's \lsti{Db} feature, and FPCs can provide a
+iterative-deepening search. Furthermore,
+%% commented out till we know for sure:
+%%the tactic can still use\lsti{Hints} via Elpi's \lsti{Db} feature, and
+ FPCs can provide a
 \emph{trace} that may be more customizable than the one offered by 
-\lsti{(e)auto}'s hard-wired \lsti{Debug} facility.
+\texttt{(e)auto}'s hard-wired \texttt{Debug} facility.
 

### 2021-03-05 papers/types2020/revisions/cultures.tex
Switching machine

@@
 on ML-style polymorphically typed higher-order relational
-programming\footnote{It uses types to indicate syntactic categories, while
-  any dependency information needs to be captured using predicates,
-  logical connectives, and quantifiers.}, it also comes with a
+programming,
+%\footnote{It uses types to indicate syntactic categories, while
+%  any dependency information needs to be captured using predicates,
+%  logical connectives, and quantifiers.}
+it also comes with a
 completely declarative version of failure and backtracking search.

### 2021-03-05 papers/types2020/revisions/sec3.tex
Switching machine

--- 2021-03-04 papers/types2020/revisions/sec3.tex
@@
   system}, \emph{foundational proof certificates}, and the
-\emph{Coq-Elpi} plugin.
+\emph{Coq-Elpi} plugin that we need for this paper.
 
@@
 
-Figure~\ref{fig:augmented} contains these proof rules and extends them
-in the following ways so that it more appropriately fits into
-the Coq proof system.
+Figure~\ref{fig:augmented} contains an annotated version of these
+proof rules: the annotations help us connect to elements of the Coq
+proof system.
 \begin{enumerate}
@@
   \item We account for \emph{computation} inside atoms by generalizing
-    to \emph{init} rule to work modulo conversion.
-  \item We have incorporated both \emph{proof certificates}~\cite{chihani17jar} (using the
-    schematic variable $\XXi$) along with expert predicates
-    (predicates with the $e$ subscript). We explain this in Section~\ref{ssec:cert}.
+    the \emph{init} rule to allow for term-level conversion.
+  \item We have incorporated both \emph{proof
+  certificates}~\cite{chihani17jar} (using the schematic variable
+    $\XXi$) along with expert predicates (predicates with the $e$
+    subscript). We explain these in Section~\ref{ssec:cert}.
   \item The inference rules are  annotated by terms structures that
@@
   \item We have added various premises which are responsible for
-    pulling information across  Coq's global environment.
+    pulling information from Coq's global environment.
 \end{enumerate}

### 2021-03-05 papers/types2020/revisions/sec3.tex

--- 2021-03-05 papers/types2020/revisions/sec3.tex
@@
   \item We account for \emph{computation} inside atoms by generalizing
-    the \emph{init} rule to work modulo conversion.
-  \item We have incorporated both \emph{proof certificates}~\cite{chihani17jar} (using the
-    schematic variable $\XXi$) along with expert predicates
-    (predicates with the $e$ subscript). We explain this in Section~\ref{ssec:cert}.
+    the \emph{init} rule to allow for term-level conversion.
+  \item We have incorporated both \emph{proof
+  certificates}~\cite{chihani17jar} (using the schematic variable
+    $\XXi$) along with expert predicates (predicates with the $e$
+    subscript). We explain these in Section~\ref{ssec:cert}.
   \item The inference rules are  annotated by terms structures that

### 2021-03-05 papers/types2020/revisions/cover.tex

\documentclass{article}

\begin{document}
\title{Cover letter for ``Two applications\dots''}
\maketitle{}

\section*{Main Changes}

\begin{itemize}
\item ``auto'' with conversion
\item ``auto'' with Hints (?)
\item other \dots
\end{itemize}
\section*{Reviewer 1}
\begin{verbatim}
An introduction to the topic, or at
least a section explaining the application in the paper, would be
necessary to open the paper to a larger audience.
\end{verbatim}
comment
\begin{verbatim}
It is misterious why you choosed to only use constructors: you never
justify this restriction. There is also an hint that you can use the Hints
that eauto uses, but this is not reflected in your code. How can it be?
\end{verbatim}
comment
\section*{Reviewer 2}

\end{document}

### 2021-03-05 papers/types2020/revisions/elab.tex

--- 2020-11-07 papers/types2020/elab.tex
@@
 \begin{figure}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.sig}
-\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.mod}
+\lstinputlisting[language=lprolog,linerange={start-end}]{../code/deb-fpc.sig}
+\lstinputlisting[language=lprolog,linerange={start-end}]{../code/deb-fpc.mod}
 \caption{The FPC definition of De Bruijn notation as proof evidence.}
@@
 % LocalWords:  intuitionistic dneg peirce forall basicstyle lprolog
-% LocalWords:  Elpi lc args ljf dep
+% LocalWords:  Elpi lc args ljf dep Coq

### 2021-03-06 papers/types2020/revisions/intro.tex
cleanup of pbt dir, small changes to 3.2

@@
 flexible approaches to implementing many different logics and their
-proof systems~\cite{felty89phd,felty88cade,paulson89jar}.  Following
+proof systems~\cite{felty88cade,paulson89jar}.  Following
 that tradition, the Elpi system makes it possible to encode the proofs

### 2021-03-06 papers/types2020/revisions/sec3.tex
cleanup of pbt dir, small changes to 3.2

@@
   \item We account for \emph{computation} inside atoms by generalizing
-    the \emph{init} rule to allow for term-level conversion.
-  \item We have incorporated both \emph{proof
+    the \emph{init} rule to allow type-level conversion.
+  \item We have incorporated  \emph{proof
   certificates}~\cite{chihani17jar} (using the schematic variable
-    $\XXi$) along with expert predicates (predicates with the $e$
+    $\XXi$) along with \emph{expert predicates} (predicates with the $e$
     subscript). We explain these in Section~\ref{ssec:cert}.
@@
   \item We have added various premises which are responsible for
-    pulling information from Coq's global environment.
+    interacting with Coq.
 \end{enumerate}
@@
 Herbelin \cite{Herbelin94}). Similar to that calculus, we have a
-term language that includes terms and lists of terms, and two distinct
+term language that includes terms and lists of terms, and two %distinct
 typing judgments for the two categories.
@@
 $\Gamma_C$ the names and types of its constructors; finally $p$
-denotes the number of parameters and plays here no role. We delegate to
-Coq's type checking the enforcement of the well-sortedness of
-inductive types. The \emph{local context} is empty, since we are only
+denotes the number of parameters and plays here no role. The {local context} is empty, since we are only
 dealing with types that correspond to Horn clauses, and atomic types
@@
 the right, although the proof theory would gladly allow it. This means
-at the term level that there are no variables in our grammar of
+that there are no variables in our grammar of
 terms. Terms are always applied to a (possibly empty) list of
-arguments.  The \emph{decide} rule, % is the equivalent in our
+arguments.   We delegate to
+Coq's type checking the enforcement of the well-sortedness of
+inductive types.  The {decide} rule, % is the equivalent in our
 % a \emph{decide} rule in an explicitly focused system
@@
 inference rules in Figure~\ref{fig:augmented}: $\XXi\vdash t \colon A$ corresponds to
-\lsti{check $\XXi$ (go A T)} and $\XXi \Downarrow  l \colon D\vdash A$ corresponds to
-\lsti{check $\XXi$ (bc D A L)}.
+\lsti{check Cert (go A T)} and $\XXi \Downarrow  l \colon D\vdash A$ corresponds to
+\lsti{check Cert (bc D A L)}.
 %
@@
 information, also inactive. In addition, we supply the certificate: it
-is simply an integer and a string to identify the ``resource'' FPC that we
-will use in this case. We just need to call \lsti{check} on the goal  \lsti{G},
-together with the certificate, in order to obtain a reconstructed
-proof term. We do not call the reconstruction directly on the evar
-because Coq-Elpi ensures that evars manipulated by \lP are well-typed
-at all times; since we cannot guarantee that, as we work with
-partially reconstructed term, we get around this by an explicit
-unification.
+consists of an integer (or two in the case of pairing) and a string to
+identify the ``resource'' FPC that we will use in this case. We just
+need to call \lsti{check} on the goal \lsti{G}, together with the
+certificate, in order to obtain a reconstructed proof term. We do not
+call the reconstruction directly on the evar because Coq-Elpi ensures
+that evars manipulated by \lP are well-typed at all times; since we
+cannot guarantee that, as we work with partially reconstructed term,
+we get around this by an explicit unification.
 
@@
  \lsti{dprolog} is not restricted to depth-first search, since
-it follows the dictate of the given FPC; for example we could easily add
+it follows the dictates of the given FPC; for example we could easily add
 iterative-deepening search. Furthermore,

### 2021-03-06 papers/types2020/revisions/sec3.tex
Working on the fpc section.

--- 2021-03-05 papers/types2020/revisions/sec3.tex
@@
   \[
-\infer{\XXi_1\vdash k\ l : A}
+\infer{\XXi\vdash k\ l : A}
       { \mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad 
@@
        k : D \in \Gamma_C\quad 
-       \bc{\XXi_2}{} l D A \quad\unfoldExpert{\XXi_1}{\XXi_2}} 
+       \bc{\XXi_1}{} l D A \quad\unfoldExpert{\XXi}{\XXi_1}} 
 \]
@@
 as in the previous proof system for Horn logic,  given an atom,
- selects a clause on which to back-chain on: we lookup the
+ selects a clause on which to backchain on: we lookup the
 constructors of an inductive definition from the global
@@
 introduction rules for propositional connectives, nor equality for
-that matter. However, one of the beauty of the Calculus of Inductive
+that matter. However, one of the beauties of the Calculus of Inductive
 Construction is that they are, in fact, defined inductively and
-therefore the \emph{decide} rule will handle those; this also
-simplifies the syntax of proof-terms.
+therefore the \emph{decide} rule will handle those.  Thus, the syntax
+of proof-terms is rather simple.
 
@@
 in \lP takes advantage of native \lP constructs such as lists and binders.
-The following is part of the Coq-Elpi API signature of constants that we use:
+The following is part of the Coq-Elpi API signature of constants that we use.
 \begin{lstlisting}[language=lprolog]
 kind term   type.                                   % reification of Coq terms
-kind gref   type.                                 % reif. of references to global terms
+kind gref   type.                                   % reification of refs to globals
 type global gref -> term.                            % coercion to term
@@
 
+\newpage
 \subsubsection{FPC-specific code}
@@
 
-Each inference rule in Figure~\ref{fig:augmented} is augmented with an
-additional premise involving an \emph{expert predicate} and is
-annotated with variables $\Xi$, $\Xi'$, $\Xi_1$, $\Xi_2$ that range
-over proof certificates.  These expert predicates are used to extract
-information from a certificate.  For example, the premise
-$\prodE{\XXi}{\XXi_1}{\XXi_2}{t}$ can be understood as a procedure for
-extracting from the certificate $\XXi$ a substitution term $t$ and two
-continuation certificates $\XXi_1$ and $\XXi_2$.
+The \emph{foundational proof certificates} (FPC) framework was
+proposed in \cite{chihani17jar} as a flexible approach to specifying a
+range of proof structures in first-order classical and intuitionistic
+logics.
 %
-These notions are taken from the general setting of \emph{foundational
-  proof certificates}~\cite{chihani17jar}.
+Such specifications are also executable using a simple logic
+programming interpreter.
+%
+As a result of using logic programming, proof certificates in this
+framework are allowed to drop details that could be reconstructed during
+the checking phase.
+%
+For example, substitution instances of quantifiers do not need to be
+explicitly described within a certificate since unification within the
+logic programming checker is often capable of reconstructing such
+substitutions.
+
+In this and the next section, we shall only use a much reduced subset
+of the FPC framework.
+%
+In particular, a proof certificate, denoted by the schematic variable
+$\XXi$, is a particular term that is threaded throughout a logic
+programming interperter.
+%
+For example, the inference rules in Figure~\ref{fig:augmented} are
+augmented with an additional premise called an \emph{expert predicate}
+with are responsible for extracting relevant information from a proof
+certificate $\XXi$ as well as constructing continuation certificates,
+such as, $\XXi_1$ and $\XXi_2$.
+%
+For example, the premise $\prodE{\XXi}{\XXi_1}{\XXi_2}{t}$ can be
+understood as a procedure for extracting from the certificate $\XXi$ a
+substitution term $t$ and two continuation certificates $\XXi_1$ and
+$\XXi_2$ for the two premises of this rule.
+%
+If the certificate $\XXi$ does not contain an explicit substitution
+term, the expert predicate can simply return a logic variable which would
+denote any term that satisfies subsequent unification problems arising
+in completing the check of this certificate.
+
+***
+
+%
 %
@@
 being of size less than or equal to 10.
+
+\newpage
 

### 2021-03-06 papers/types2020/revisions/sec3.tex
Fixing merge.

@@
 
-\newpage
+%\newpage
 \subsubsection{FPC-specific code}
@@
 
-***
+*** DM Continue here
 
@@
 
-\newpage
+%\newpage
 

### 2021-03-06 papers/types2020/revisions/sec3.tex

@@
 
-%\newpage
 \subsubsection{FPC-specific code}
@@
 range of proof structures in first-order classical and intuitionistic
-logics.
-%
-Such specifications are also executable using a simple logic
-programming interpreter.
-%
-As a result of using logic programming, proof certificates in this
-framework are allowed to drop details that could be reconstructed during
-the checking phase.
-%
-For example, substitution instances of quantifiers do not need to be
-explicitly described within a certificate since unification within the
-logic programming checker is often capable of reconstructing such
+logics.  Such specifications are also executable using a simple logic
+programming interpreter.  As a result of using logic programming,
+proof certificates in this framework are allowed to lack details that
+can be reconstructed during the checking phase.  For example,
+substitution instances of quantifiers do not need to be explicitly
+described within a certificate since unification within the logic
+programming checker is often capable of reconstructing such
 substitutions.
@@
 In this and the next section, we shall only use a much reduced subset
-of the FPC framework.
-%
-In particular, a proof certificate, denoted by the schematic variable
-$\XXi$, is a particular term that is threaded throughout a logic
-programming interperter.
-%
-For example, the inference rules in Figure~\ref{fig:augmented} are
-augmented with an additional premise called an \emph{expert predicate}
-with are responsible for extracting relevant information from a proof
+of the FPC framework.  In particular, a proof certificate, denoted by
+the schematic variable $\XXi$, is a particular term that is threaded
+throughout a logic programming interperter.  For example, the
+inference rules in Figure~\ref{fig:augmented} are augmented with an
+additional premise that invokes an \emph{expert predicate} with is
+responsible for extracting relevant information from a proof
 certificate $\XXi$ as well as constructing continuation certificates,
-such as, $\XXi_1$ and $\XXi_2$.
-%
-For example, the premise $\prodE{\XXi}{\XXi_1}{\XXi_2}{t}$ can be
-understood as a procedure for extracting from the certificate $\XXi$ a
-substitution term $t$ and two continuation certificates $\XXi_1$ and
-$\XXi_2$ for the two premises of this rule.
-%
-If the certificate $\XXi$ does not contain an explicit substitution
-term, the expert predicate can simply return a logic variable which would
-denote any term that satisfies subsequent unification problems arising
-in completing the check of this certificate.
+such as, $\XXi_1$ and $\XXi_2$.  For example, the premise
+$\prodE{\XXi}{\XXi_1}{\XXi_2}{t}$ calls the expert for products which
+should extract from the certificate $\XXi$ a substitution term $t$ and
+two continuation certificates $\XXi_1$ and $\XXi_2$ for the two
+premises of this rule.  If the certificate $\XXi$ does not contain an
+explicit substitution term, the expert predicate can simply return a
+logic variable which would denote any term that satisfies subsequent
+unification problems arising in completing the check of this
+certificate.
 
-*** DM Continue here
-
-%
-%
-%%\todo{MM: Talk about randomized FPC? They mentioned afterwards -am}
 In our case here, an FPC is a collection of \lP clauses that provide
@@
 and~\ref{fig:kernel}: that is, the exact set of constructors for the
-type of certificates \lsti{cert} as well as the specification of the expert
-predicates listed \emph{ibidem}.
-%
-The top of Figure~\ref{fig:resources} displays two FPCs,
-both of which can be used to describe proofs where we bound
-the dimension of  a proof.
-%
-For example, the first FPC dictates that the query \mbox{\lsti{(check (qheight 5) A)}} is
-provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
-and~\ref{fig:resources} if and only if the height of that proof is 5
-or less.
-%
-Similarly, the second FPC can be used to bound the total number of
-instances of the decide rule in a proof.
+type of certificates \lsti{cert} as well as the specification of the
+expert predicates listed \emph{ibidem}.  The top of
+Figure~\ref{fig:resources} displays two FPCs, both of which can be
+used to describe proofs where we bound the dimension of a proof.  For
+example, the first FPC dictates that the query \mbox{\lsti{(check
+    (qheight 5) A)}} is provable in the kernel using the clauses in
+Figures~\ref{fig:kernel} and~\ref{fig:resources} if and only if the
+height of that proof is 5 or less.  Similarly, the second FPC can be
+used to bound the total number of instances of the decide rule in a
+proof.  (Obviously, such proof certificates do not contain, for
+example, substitution terms.)
 %
@@
 
-%\newpage
+%%\todo{MM: Talk about randomized FPC? They mentioned afterwards -am}
 

### 2021-03-07 papers/types2020/revisions/cultures.tex
Revisions to sections 2 and 3.  Added two references.

@@
 
+%% DM Reviewer 1 want more references to claims made here.  I
+%% lightened the claims and added two references.
+
 Gentzen's discovered that the key to treating classical and
 intuitionistic logics in the same proof system was identifying the
-structural rules of the weakening and contraction and
-placing them on the right-side of sequents~\cite{gentzen35}.
-This discovery lead him away from natural deduction to
-multiple-conclusion sequent calculus, a very important kind of proof
-system that has not been given a satisfactory representation in type
-theory.  This same innovation of Gentzen's also opened the way to
+structural rules of the weakening and contraction and placing them on
+the right-side of sequents~\cite{gentzen35}.  This discovery lead him
+away from natural deduction to multiple-conclusion sequent calculus.
+%
+This same innovation of Gentzen's also opened the way to
 another key proof-theoretic discovery, that of linear logic
-\cite{girard87tcs}.  Similarly, sequent calculus provides an elegant
-presentation of linear logic while most treatments of linear logic in
-type theory have difficulty treating linear negation and the
-multiplicative disjunction and its unit.
+and the linear negation~\cite{girard87tcs}.
+%
+While sequent calculus provides an elegant presentation of full linear
+logic, most treatments of linear logic in type theory have
+been limited to single-conclusion sequents~\cite{bierman95phd}
+or to restricted, multiple-conclusion variants~\cite{hyland93apal}.
 

### 2021-03-07 papers/types2020/revisions/sec3.tex
Revisions to sections 2 and 3.  Added two references.

@@
 
+% DM I tried to down play FPC here so we would not need to say too
+% much about the general picture.  Hopefully, the examples we provide
+% and the few motivating sentences will make the reviewers happier.
+
 The \emph{foundational proof certificates} (FPC) framework was
@@
 In this and the next section, we shall only use a much reduced subset
-of the FPC framework.  In particular, a proof certificate, denoted by
-the schematic variable $\XXi$, is a particular term that is threaded
-throughout a logic programming interperter.  For example, the
-inference rules in Figure~\ref{fig:augmented} are augmented with an
-additional premise that invokes an \emph{expert predicate} with is
-responsible for extracting relevant information from a proof
-certificate $\XXi$ as well as constructing continuation certificates,
-such as, $\XXi_1$ and $\XXi_2$.  For example, the premise
-$\prodE{\XXi}{\XXi_1}{\XXi_2}{t}$ calls the expert for products which
-should extract from the certificate $\XXi$ a substitution term $t$ and
-two continuation certificates $\XXi_1$ and $\XXi_2$ for the two
-premises of this rule.  If the certificate $\XXi$ does not contain an
-explicit substitution term, the expert predicate can simply return a
-logic variable which would denote any term that satisfies subsequent
-unification problems arising in completing the check of this
-certificate.
+of the FPC framework: in essence, an FPC will be used as a simple
+mechanism for bounding the search for proofs.  In our examples, a
+proof certificate, denoted by the schematic variable $\XXi$, is a
+particular term that is threaded throughout a logic programming
+interperter.  For example, the inference rules in
+Figure~\ref{fig:augmented} are augmented with an additional premise
+that invokes an \emph{expert predicate} with is responsible for
+extracting relevant information from a proof certificate $\XXi$ as
+well as constructing continuation certificates, such as, $\XXi_1$ and
+$\XXi_2$.  For example, the premise $\prodE{\XXi}{\XXi_1}{\XXi_2}{t}$
+calls the expert for products which should extract from the
+certificate $\XXi$ a substitution term $t$ and two continuation
+certificates $\XXi_1$ and $\XXi_2$ for the two premises of this rule.
+If the certificate $\XXi$ does not contain an explicit substitution
+term, the expert predicate can simply return a logic variable which
+would denote any term that satisfies subsequent unification problems
+arising in completing the check of this certificate.
 

### 2021-03-08 papers/types2020/revisions/conc.tex
some ednote in conc.tex examples reveisted with new kernel todo updated

@@
 to handle directly existential goals) and more importantly one that uses
- Elpi's \emph{delay} mechanism. % This a distinguished feature of Elpi, whose
+ Elpi's \emph{delay} mechanism.\ednote{This may not be accurate anymore -am} % This a distinguished feature of Elpi, whose
 % usef has been advocated in ~\cite{DunchevCT16,coen19mscs}.
@@
 
-Another current limitation of our PBT tactic is that it does not
-handle conversion within an inductive relation. One way to address
-this is to have the \emph{init} rule use Coq's unification rather than
-\lP's, via Coq-Elpi's \lsti{coq.unif-eq} hook.
+% Another current limitation of our PBT tactic is that it does not
+% handle conversion within an inductive relation. One way to address
+% this is to have the \emph{init} rule use Coq's unification rather than
+% \lP's, via Coq-Elpi's \lsti{coq.unif-eq} hook.
 %We plan to tackle this either at the Ltac level by co-routining it with simplification tactics and/or by adding rewriting features to our FPC system~\cite{ChihaniM16}.
@@
  \smallskip
- The source files mentioned in this paper are available at \url{https://github.com/proofcert/fpc-elpi}.
+ The source files mentioned in this paper are available at \url{https://github.com/proofcert/fpc-elpi}.\ednote{Or not. Fix this -am}
 

### 2021-03-08 papers/types2020/revisions/elab.tex
some ednote in conc.tex examples reveisted with new kernel todo updated

@@
 Using the FPC provided in Figure~\ref{fig:debruijn} and the proof
-certificate checker in the file \verb+ljf-dep.mod+ of the repository
-\url{https://github.com/proofcert/fpc-elpi}, the De Bruijn term
+certificate checker in the file \verb+ljf-dep.mod+,
+%% may nobe this repo -am
+%%of the repository \url{https://github.com/proofcert/fpc-elpi},
+the De Bruijn term
 displayed above can be elaborated into a proper proof for the

### 2021-03-08 papers/types2020/revisions/cover.tex
Filled in cover.tex more with reviewer requests.  I filled in a few responses already.  A few more remain.

@@
 \begin{document}
-\title{Cover letter for ``Two applications\dots''}
+\title{Revisions made to the paper \\
+  ``Two applications of logic programming to Coq''}
 \maketitle{}
@@
 \end{itemize}
+
 \section*{Reviewer 1}
 \begin{verbatim}
-An introduction to the topic, or at
-least a section explaining the application in the paper, would be
-necessary to open the paper to a larger audience.
+An introduction to the topic, or at least a section explaining
+ the application in the paper, would be necessary to open the
+ paper to a larger audience.
+\end{verbatim}
+
+The reviewer is asking for more of an introduction about FPC.  We
+have done two things to the paper.  First, we have added a
+paragraph providing more of an explanation about FPC in Section
+3.2.2.  Second, since the FPCs in the ``property-based testing''
+part of his paper are rather simple, we illustrate their
+operation on a few examples.  Those examples can be understood
+fully from the little bit of Prolog code used to describe them.
+The role of FPCs in Section 5 is more substantial but we are able
+to treat them more as a blackbox: the elaboration aspect of FPCs
+come from using parallel interpretation within logic programming:
+these concepts have now been illustrated in Section 4.
+
+\begin{verbatim}
+page 5, Figure 1: the first inference rule (the decide inference
+ rule) seems wrong. Indeed you are adopting Coq's syntx
+ Ind[p](\Gamma_I := \Gamma_C) where p is a list of abstractions
+ over so called uniform parameters.  You seem to never
+ instantiate them in the rule. Moverover, the premise "(head A :
+ T) \in \Gamma_I" is also suspicious. I would expect that premise
+ to compute a substitution \sigma for p, to be applied to D in
+ both premises (e.g. D\sigma).
+\end{verbatim}
+
+comment
+
+\begin{verbatim}
+It is mysterious why you choose to only use constructors: you
+ never justify this restriction. There is also an hint that you
+ can use the Hints that eauto uses, but this is not reflected in
+ your code. How can it be?
 \end{verbatim}
 comment
+
 \begin{verbatim}
-It is misterious why you choosed to only use constructors: you never
-justify this restriction. There is also an hint that you can use the Hints
-that eauto uses, but this is not reflected in your code. How can it be?
+page 1: put some references for "that has not been given a
+ satisfactory representation in type theory" and for "have
+ difficulty treating linear negation and the multiplicative
+ disjunction"
+\end{verbatim}
+
+We modified the claims in that paragraph and added two references
+to help support those modified claims.
+
+\section*{Reviewer 2}
+
+\begin{verbatim}
+1. The paper is centered on the idea that it is possible to use
+ Coq-elpi to construct tactics following the ideas from FPC, but
+ it lacks a proper evaluation of the tactics to show its
+ benefits. It would make for a stronger story to have a thorough
+ evaluation with its competitors (eauto for the first tactic and
+ QuickChick for the second), even in restricted scenarios.
 \end{verbatim}
 comment
-\section*{Reviewer 2}
+
+\begin{verbatim}
+2. The code is not available. The cited
+ https://github.com/proofcert/fpc-elpi returns 404, and the
+ search for fpc-elpi failed, as well inspecting the repos from
+ the proofcert organization.
+\end{verbatim}
+
+We are sorry for not having converted our repository from private
+to public.  We have fixed this problem now.
+
+\begin{verbatim}
+1. The proof term shown as the result of calling the tactic to
+ prove i1 is missing some terms, namely, the inequalities
+ proofs. Are these also automatically obtained by the tactic?
+\end{verbatim}
+
+comment
 

### 2021-03-08 papers/types2020/revisions/cover.tex
some comments to rev's questions

@@
 
-\begin{itemize}
-\item ``auto'' with conversion
-\item ``auto'' with Hints (?)
-\item other \dots
-\end{itemize}
+From the implementation point of view, we have extended the kernel to
+account for conversion: this allows us to do prolog search and
+therefore PBT on Inductive definitions that mix deduction and
+computation. We have also implemented some optimizations to the kernel
+itself, as briefly mentioned in the conclusion. We have experimented
+with adding Hints to the kernel via Coq-Elpi's \texttt{Accumulate DB}
+and while this is doable, it is not particularly useful for the
+intended application, namely PBT. What we're working on is modifying
+the elaboration tactic of Sec 5, which already sees the local context,
+to handle Hints and Inductive. However,  given the deadline, this
+will be future work.
 
@@
  Ind[p](\Gamma_I := \Gamma_C) where p is a list of abstractions
- over so called uniform parameters.  You seem to never
- instantiate them in the rule. Moverover, the premise "(head A :
- T) \in \Gamma_I" is also suspicious. I would expect that premise
- to compute a substitution \sigma for p, to be applied to D in
- both premises (e.g. D\sigma).
+ over so called uniform parameters ...
+The alternative is that the parameters are explicitly quantified both in
+\Gamma_i and \Gamma_C (the way the inductive type packet is represented
+internally into Coq) and p becomes just a number that says how many of the
+abstractions are parameters.
 \end{verbatim}
 
-comment
+We meant the latter alternative. This is now made explicit in the text.ù
 
@@
 \end{verbatim}
-comment
+As anticipated before, it is immediate to add another ``decide'' rule
+that picks up previously proven lemmas. That would be close to the
+functionality of \texttt{Hint Resolve}, but does not literally uses
+Coq's Hints facility. It is conceivable that we could write some glue
+code to connect the two, but handling all the fairly sophistcated features of the Hint
+mechanism is out of the scope of this paper.
 
@@
 
-comment
+Well spotted: we had forgot the guard in the third clause of \texttt{insert}. The roof terms now reflects those evidence.
 
 \end{document}
+
+% LocalWords:  FPC FPCs blackbox

### 2021-03-09 papers/types2020/revisions/cover.tex
almost done

@@
 the elaboration tactic of Sec 5, which already sees the local context,
-to handle Hints and Inductive. However,  given the deadline, this
-will be future work.
+to handle Hints and Inductive. All the peices are in place, however,  given the deadline, this
+will not be part of this submission.
 
@@
 
-We meant the latter alternative. This is now made explicit in the text.ù
+We meant the latter alternative. This is now made explicit in the text.
 
@@
 functionality of \texttt{Hint Resolve}, but does not literally uses
-Coq's Hints facility. It is conceivable that we could write some glue
+Coq's Hints facility. It is feasable that to write some glue
 code to connect the two, but handling all the fairly sophistcated features of the Hint
@@
 \end{verbatim}
-comment
+
+We meant this paper to be a proof-of-concept rather than a system
+paper. We are not in the position to carry out a ``thorough evaluation'' with
+Coq's automation not with QuickChick, if not for the high-level, qualitative remarks we have made in the paper. [AM: add more if I do more about dependent terms]
 
@@
 
-Well spotted: we had forgot the guard in the third clause of \texttt{insert}. The roof terms now reflects those evidence.
+Well spotted: we had forgot the inequality guard in the third clause of
+\texttt{insert}. The proof terms now reflects those evidence.
 
@@
 % LocalWords:  FPC FPCs blackbox
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2021-03-09 papers/types2020/revisions/conc.tex
more points in cover letter
conclusion fixed

toned down relation with eauto

@@
 
-We have presented two application of this synergy: one geared towards
-providing a flexible approach to connecting external provers of
-first-order intuitionistic logic to Coq; the other supporting an out-of-the-box way to do property-based testing for pure inductive relations.
+We have presented two application of this synergy: one supporting an
+out-of-the-box way to do property-based testing for inductive
+relations; the other geared towards providing a flexible approach to
+connecting external provers of first-order intuitionistic logic to Coq
 %
@@
 
+ The
+code reported in Fig.~\ref{fig:augmented} is a simplification for
+exposition purposes of the real implementation of the
+kernel. Following ideas from bidirectional type checking, we have
+factored out the product left rule in $\forall-L$ and $\supset-L$,
+where the former delegates to Coq the \emph{check} that the
+instantiation term $t$ is well-typed w.r.t.~$B$, while in the latter
+proof search will \emph{generate} such a term, given the type
+$B$. There are also other minor tweaks, such as
+a rule performing weak-head reduction (allowing
+us to handle directly existential goals).
+
 There are many avenues of developments for this line of research. 
-The code reported in Fig.~\ref{fig:augmented} is somewhat idealized in
-so far that we omitted a rule performing weak-head reduction (allowing
-to handle directly existential goals) and more importantly one that uses
- Elpi's \emph{delay} mechanism.\ednote{This may not be accurate anymore -am} % This a distinguished feature of Elpi, whose
-% usef has been advocated in ~\cite{DunchevCT16,coen19mscs}.
-The
-use of constraints for data generation is well
-developed~\cite{FioravantiPS15} and we could try to leverage it to
-improve our PBT tactic so that it can use partially instantiated terms: for example generating
-say lists with at least two elements as \texttt{[1,2|Tail]}, without
-recurring to needed narrowing as in 
-LazySmallCheck~\cite{smallcheck}. On the more practical side, it would be worthwhile to investigate \emph{random} generation, following the ideas in~\cite{pltredexconstraintlogic,blanco19ppdp}.
-
-% Another current limitation of our PBT tactic is that it does not
-% handle conversion within an inductive relation. One way to address
-% this is to have the \emph{init} rule use Coq's unification rather than
-% \lP's, via Coq-Elpi's \lsti{coq.unif-eq} hook.
-%We plan to tackle this either at the Ltac level by co-routining it with simplification tactics and/or by adding rewriting features to our FPC system~\cite{ChihaniM16}.
-% Mention \emph{Harpoon}, Beluga's new tactic language?
+We would like to exploit one of distinguished features of Elpi: the
+\emph{delay} mechanism.  The use of constraints for data generation is
+well developed~\cite{FioravantiPS15} and we could try to leverage it
+to improve our PBT tactic so as to generate partially instantiated
+terms, % : for example generating say lists with at least two elements as
+% \texttt{[1,2|Tail]}
+without recurring to needed narrowing as in
+LazySmallCheck~\cite{smallcheck}. On the more practical side, it would
+be worthwhile to investigate \emph{random} generation, following the
+ideas in~\cite{pltredexconstraintlogic,blanco19ppdp}.
 
@@
 and provide a way of checking and elaborating proof evidence for
- intuitionistic logic \emph{with} (inductively) defined atoms. This
- can go as far as FPC for inductive proofs~\cite{blanco15wof}.
+intuitionistic logic \emph{over} (inductively) defined atoms and
+previously proven lemmata. This can go as far as FPC for inductive
+proofs~\cite{blanco15wof}.
 
@@
 %\todo{finish future work}
+% Another current limitation of our PBT tactic is that it does not
+% handle conversion within an inductive relation. One way to address
+% this is to have the \emph{init} rule use Coq's unification rather than
+% \lP's, via Coq-Elpi's \lsti{coq.unif-eq} hook.
+%We plan to tackle this either at the Ltac level by co-routining it with simplification tactics and/or by adding rewriting features to our FPC system~\cite{ChihaniM16}.
+% Mention \emph{Harpoon}, Beluga's new tactic language?
 
@@
 
-% LocalWords:  Elpi simp LJF PBT LazySmallCheck Ltac
+% LocalWords:  Elpi simp LJF PBT LazySmallCheck Ltac Coq provers FPC
+% LocalWords:  lemmata

### 2021-03-09 papers/types2020/revisions/cover.tex
more points in cover letter
conclusion fixed

toned down relation with eauto

@@
 From the implementation point of view, we have extended the kernel to
-account for conversion: this allows us to do prolog search and
+account for conversion: this allows us to do Prolog search and
 therefore PBT on Inductive definitions that mix deduction and
@@
 the elaboration tactic of Sec 5, which already sees the local context,
-to handle Hints and Inductive. All the peices are in place, however,  given the deadline, this
+to handle Hints and Inductive. All the pieces are in place, however,  given the deadline, this
 will not be part of this submission.
@@
 functionality of \texttt{Hint Resolve}, but does not literally uses
-Coq's Hints facility. It is feasable that to write some glue
-code to connect the two, but handling all the fairly sophistcated features of the Hint
+Coq's Hints facility. It is feasible that to write some glue
+code to connect the two, but handling all the fairly sophisticated features of the Hint
 mechanism is out of the scope of this paper.
+\begin{verbatim}
+you are re-implementing a better eauto. However, your
+implementation only applies as lemmas constructors of an inductive type.
+Note: I understand, because I know it in advance, that a reason could be
+that you plan to use LP negation as failure in the later section on testing.
+Therefore you need a close world specification. If that's the main reason,
+it should be made explicit in the paper.
+\end{verbatim}
+If we have an inductive predicate $p$ and then prove a lemma about it,
+this does not violate the CWA, since the lemma is true in the same
+least model of $p$.  Operationally though, it could alter NF's
+capability of finding \emph{finite} failure, since the latter is
+implemented by the vanilla meta-interpreter, which is not FPC-driven. At this stage
+we do not see a reason to have lemmata in PBT. We have mentioned this in the paper.
 
@@
 We meant this paper to be a proof-of-concept rather than a system
-paper. We are not in the position to carry out a ``thorough evaluation'' with
-Coq's automation not with QuickChick, if not for the high-level, qualitative remarks we have made in the paper. [AM: add more if I do more about dependent terms]
+paper. We are not in the position to carry out a ``thorough
+evaluation'' with Coq's automation not with QuickChick, if not for the
+high-level, qualitative remarks we have made in the paper.  We have
+carried out in QuickChick the example in Section 4, both with a
+functional and a relational encoding. The code is present in our
+examples repo. We have added a few remarks in the paper but space
+prevents us form explaining the features of QuickChick and their
+peculiarity. [AM: add more if I do more about dependent
+terms]. Finally, we may have misstated the ambitions of our Prolog
+tactic, which does not mean to compete with eauto, but merely is the
+building block of FPC-driven PBT of relational specifications. We have rephrased its characterization in the paper.
 
@@
 
-Well spotted: we had forgot the inequality guard in the third clause of
+Well spotted: we had simply forgot the inequality guard in the third clause of
 \texttt{insert}. The proof terms now reflects those evidence.
@@
 
-% LocalWords:  FPC FPCs blackbox
+% LocalWords:  FPC FPCs blackbox Coq PBT QuickChick repo CWA NF's
 
@@
 %%% End:
+% LocalWords:  lemmata

### 2021-03-09 papers/types2020/revisions/sec3.tex
more points in cover letter
conclusion fixed

toned down relation with eauto

@@
 particular term that is threaded throughout a logic programming
-interperter.  For example, the inference rules in
+interpreter.  For example, the inference rules in
 Figure~\ref{fig:augmented} are augmented with an additional premise
@@
 
-The \lsti{dprolog} tactic can be seen as a programmable (and possibly
-faster) version of Coq's \lsti{eauto}% , although it currently lacks the
+The \lsti{dprolog} tactic implements some of the features of Coq's \lsti{eauto}% , although it currently lacks the
 % latter's capability of solving goals by
@@
 % \lsti{Hints} as introduction rules
- \lsti{dprolog} is not restricted to depth-first search, since
+ it is programmable and as such  not restricted to depth-first search, since
 it follows the dictates of the given FPC; for example we could easily add

### 2021-03-11 papers/coLP/code/haskell/l.hs
Move papers

@@ -1,26 +1,0 @@
-import Test.QuickCheck
-
-eqi [] [] = True
-eqi (x : xs) (y : ys) = (x == y) && eqi xs ys
-eqi _ _ = False
-
-zeros = 0 : zeros
-ones = 1 : ones
-ones' = map succ zeros
-
-
-prop_eq :: [Integer] -> [Integer] -> Property
-prop_eq xs ys =
-  eqi xs ys ==> (xs == ys)
-
-
-prop_d :: [Int] -> Int -> Property
-prop_d xs n =
-  not (null xs) && n >= 0 ==>
-  take n (cycle xs) == take n (cycle $ xs ++ xs)
-
-main = quickCheck prop_d
-
-prop_o :: Int -> Property
-prop_o n =
-  n >= 0 ==> take n ones == take n zeros

### 2021-03-11 papers/coLP/code/prolog/aut.pl

@@ -1,54 +1,0 @@
-% based on history meta-interpreter
-:- ensure_loaded('meta-co.pl').
-
-:- discontiguous coinductive/2.
-
-:- dynamic automata/3.
-:- dynamic accepts/2.
-:- dynamic drop/3.
-:- dynamic trans/3.
-:- dynamic comember/2.
-:- dynamic final/1.
-
-% true is X occurs in Xs infinitely oftern. Fail if finte list, loops
-% if it does not.
-coinductive(comember,2).
-comember(X,L) :-
-    drop(X,L,L1), comember(X,L1).
-drop(H,[H|T],T).
-drop(H,[_|T],T1) :- drop(H,T,T1).
-
-% eg. Xs = [1,2,3|Xs], query(comember(2,Xs)).
-% Xs = [1, 2, 3|Xs]
-
-% Xs = [1,2,3], query(comember(2,Xs)).
-% false.
-
-
-coinductive(automata,3).
-automata([X|T], St, Sts) :-
-    trans(St, X, NewSt), automata(T, NewSt,[St|Sts]).
-trans(s0, a, s1).
-trans(s1, b, s2).
-trans(s2, c, s3).
-trans(s3, d, s0).
-trans(s2, e, s0).
-final(s2).
-
-
-% String accepted by fixed automata  starting with InitSt and with Final
-% occurring inf often
-
-% (abcd)* accepted and final s2 occurs infintely often
-
-accepts(String,InitSt) :-
-    automata(String,InitSt,Sts),
-    final(Final),
-    comember(Final,Sts).
-
-%  query((automata(X,s0, States ), comember(s2,States))).
-%  X = [a, b, c, d|X],
-% States = [s3, s2, s1, s0|States] ;
-
-%query(accepts(String,s0)).
-%String = [a, b, c, d|String]

### 2021-03-11 papers/coLP/code/prolog/auts.pl

@@ -1,62 +1,0 @@
-% automata over different LTS
-
-:- ensure_loaded('meta-co.pl').
-:- discontiguous final/2.
-:- discontiguous trans/4.
-:- discontiguous coinductive/2.
-
-:- dynamic automata/4.
-:- dynamic accepts/3.
-:- dynamic drop/3.
-:- dynamic trans/3.
-:- dynamic comember/2.
-:- dynamic final/2.
-:- dynamic is_aut/2.
-:- dynamic not_sim/2.
-
-coinductive(comember,2).
-comember(X,L) :-
-    drop(X,L,L1), comember(X,L1).
-drop(H,[H|T],T).
-drop(H,[_|T],T1) :- drop(H,T,T1).
-
-
-coinductive(automata,4).
-
-accepts(String,InitSt,Trans) :-
-    automata(String,InitSt,Trans,Sts),
-    final(Trans,Final),
-    comember(Final,Sts).
-
-
-automata([X|T], St, Trans, Sts) :-
-    trans(Trans, St, X, NewSt),
-    automata(T, NewSt,Trans,[St|Sts]).
-
-% graph g1
-trans(g1, s0, a, s1).
-trans(g1, s1, b, s2).
-trans(g1, s2, c, s3).
-trans(g1, s3, d, s0).
-trans(g1, s2, e, s0).
-final(g1, s2).
-
-% graph g2, 
-trans(g2, s0, a, s1).
-trans(g2, s1, b, s2).
-%trans(g2, s2, c, s3).
-%trans(g2, s3, d, s0).
-trans(g2, s2, e, s0).
-final(g2, s2).
-
-is_aut(g1,s0).
-is_aut(g2,s0).
-
-not_sim(A1,A2) :-
-    is_aut(g1,InitSt1),
-    is_aut(g2,InitSt2),
-    accepts(String1,InitSt1,A1),
-%    not(accepts(String,InitSt2,A2)). % not not-handled by the meta
-    accepts(String2,InitSt2,A2),
-    not(String1 =  String2).				
-    

### 2021-03-11 papers/coLP/code/prolog/co.pl

@@ -1,73 +1,0 @@
-:- use_module(library(coinduction)).
-
-:- coinductive p/1.
-
-p([1|T]) :- p(T).
-
-
-:- coinductive from/2.
-from(N, [N|T]) :- from(s(N), T).
-
-% :- from(0,_). %loops, why?
-:- coinductive stream/1.
-stream([H | T]) :-
-    num(H), stream(T).
-
-num(0).
-num(s(N)) :- num(N).
-
-:- coinductive cmem/2.
-cmem( H, [ H | _ ] ).
-cmem( H, [ _ | T ] ) :- cmem( H, T ).
-
-%  cmem(0,[s(z),0]). % true
-%  X = [0|X],cmem(a,X). %true: anything member of a inf list
-
-% called comember/2, which is true if and only if the desired element
-% occurs an infnite number of times in the list.
-
-drop( H, [ H | T ], T ).
-drop( H, [ _ | T ], T1 ) :- drop( H, T, T1 ).
-
-:- coinductive comember/2.
-comember( X, L ) :-
-    drop( X, L, L1 ), comember( X, L1 ).
-
-% X = [ 1, 2, 3 | X ], comember( 2, X ). yes
-% X = [ 1, 2, 3 | X ], comember( 4, X ). loops
-%   X = [ 1, 2, 3], comember( E, X ). fails
-
-%--------------------------------
-
-
-%:-  stream([0, s(0), s(s(0)) | T ]).
-% TODO: some of the bewdyr examples
-
-% from Luke's thesis p.77
-% does not seem to work in SWI
-:- coinductive automata/2.
-automata([X|T], St) :- trans(St, X, NewSt), automata(T, NewSt).
-trans(s0, a, s1).
-trans(s1, b, s2).
-trans(s2, c, s3).
-trans(s3, d, s0).
-trans(s2, e, s0).
-final(s2).
-
-% :- automata(X, s0), comember(s2, X).
-:- coinductive bin/1.
-bin([0|T]) :- bin(T).
-bin([1|T]) :- bin(T).
-
-:-  X = [1|X] , bin(X).
-% contrary to paper yap bin(X) does not retun  X = [1|X]. (df)
-
-				% maxlist, example from Ancona should prove - Os = [1|Os], maxList(Os,2). but it does not
-
-max(A,B,Max) :-
-	(A < B -> Max = B; Max = A).
-mmax(1,2,2).
[…]

### 2021-03-11 papers/coLP/code/prolog/meta-co.pl

@@ -1,44 +1,0 @@
-% metainterpreter with depth-first history mechanism for "coinduction"
-:- discontiguous coinductive/2.
-query( Goal ) :-
-    solve( [], Goal ).
-solve( Hypothesis, ( Goal1, Goal2 ) ) :-
-    !,
-    solve( Hypothesis, Goal1 ),
-    solve( Hypothesis, Goal2 ).
-solve( _ , Atom ) :-
-    builtin( Atom ),
-    !,
-    Atom.
-solve( Hypothesis, Atom ) :-
-    coinductive( Atom ),
-    member( Atom, Hypothesis ).
-solve( Hypothesis, Atom ) :-
-    coinductive( Atom ),
-    !,
-    clause( Atom, Atoms ),
-    solve( [ Atom | Hypothesis ], Atoms ).
-
-solve( Hypothesis, Atom ) :-
-    %inductive( Atom ),
-    clause( Atom, Atoms ),
-    solve( Hypothesis, Atoms ).
-
-inductive( Atom ) :-
-    Atom \= ( _, _ ),
-    \+ builtin( Atom ),
-    functor( Atom, Predicate, Arity ),
-    inductive( Predicate, Arity ).
-coinductive( Atom ) :-
-    Atom \= ( _, _ ),
-    \+ builtin( Atom ),
-    functor( Atom, Predicate, Arity ),
-    coinductive( Predicate, Arity ).
-
-builtin( _ = _ ).
-builtin( true ).
-builtin( _ is _ ).
-builtin( _ > _ ).
-builtin( _ < _ ).
-builtin(\+ _).
-builtin(not(_G)).
